### 2013-03-12 paper/paper.tex
moved the paper into tex

\documentclass[twocolumn,floatfix,nofootinbib,aps]{revtex4-1}
\usepackage[utf8]{inputenc}

\usepackage{amsmath}    % need for subequations
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
\usepackage{hyperref}   % use for hypertext links, including those to external documents and URLs

\begin{document}

\title{Opt-K}
\author{Robert T. McGibbon}
\author{Christian R. Schwantes}
\author{Vijay S. Pande}
\maketitle

Likelihood framework for choosing the optimal number of states in a
Markov State Model

\section{Overview}

A significant challenge in the automated construction of markov state
models is choosing the number of microstates. The number of microstates,
$k$ governs a bias-variance tradeoff -- when $k$ is large, the models
have a lot of parameters enabling it to fit the data with low bias, but
the simultaneous estimation of these parameters from finite data can
lead to high variance. On the other hand, when $k$ is small, we have
fewer parameters and better statistics, but at the expense of model
flexibility.

Currently, most MSM construction pipelines involve manually selecting
the number of states. This is done primarily by human intuition, and is
a significant bottleneck.

Here, we introduce a likelihood framework enabling the automatic
selection of the number of states.

Within the discrete state space, the likelihood of an ensemble of
trajectories given the MSM is straightforward. We simply take the
product of the state to state transition probabilities along the path.
These state to state transition probabilities are our main central
parameters, the entries of the transition matrix. However, as we vary
the number of states, it is not permissible to simply compare these
likelihoods to select an optimal number of states. Doing so, we would
always chose the trivial model, with only one state, as the transition
probability in that model would be 1 and thus the likelihood of any
trajectory within that state space would be 1.

The proper likelihood is not of the trajectory within the discrete state
space; instead it's the likelihood of the trajectory within the
continuous phase space, on which the discrete states are merely an
indicator function basis.

$$
P[x(t)] dx^N = O_{s=0}(x_0) \prod_{i=0}^{N-1} T(x_i \rightarrow x_{i+1}) \cdot O_{s=i}(x_i)
$$

With a discrete, non-overlapping state space, the likelihood of the
trajectory can be decomposed into a product over the trajectory of two
types of terms: the state to state transition probabilities and the
``emission'' probabilities of each state, the probability of observing a
conformation at a given location in phase space given that the
conformation is within a certain state.

\section{Emission Distributions}

Up until this point, the emission probability distributions have not
been a model parameter for MSMs. Nevertheless, they are a critical
parameter from the point of view of evaluating trajectory likelihoods.
For example, consider two MSMs with the same transition probabilities.
However, in one case, the emission distributions for each state are
highly peaked at specific locations in phase space, whereas in the other
model the emission distributions are uniform over the volume of the
clusters. If the trajectory actually does go through the regions of high
likelihood in the first model, we would say that the first model more
accurately fits the data.

However, the long timescale behavior of the MSM, the quantity of most
interest scientifically, is independent of the choice of the emission
distributions. It's only determined by the eigenspectrum of the
transition matrix elements. (The emission distributions determine the
estimated eigenfunctions, but not the eigenvectors.)

Therefore, the most appropriate emmision distribution for discrete state
MSMs is that of the uniform distribution over the phase-space volume of
the state. That is, the likelihood of observing a conformation in phase
space given that the conformation is assigned to state \texttt{i} is 0
if the conformation is outside of the bounding volume of the state and
constant if the conformation is within the volume. The constant is set
so that the distribution integrates to 1, and is thus the reciprocal
volume of the microstate.

\section{Algorithm}

To use this uniform distribution emission model, computationally, we
need to compute the volume of our MSM states, which are high-dimensional
Voronoi cells. While trivial in two or three dimensions, this
computational geometry task\\becomes challenging in large dimensional
settings.

One challenge is how to model the volume of states which are at the
``edge'' of the model. Is the volume of these states infinite, extending
all the way out to infinity in some direction? This seems problematic.
Instead, it seems appropriate to assert that the volume is bounded by
the convex hull of the data. That is, the volume of a state is defined
as the volume of the intersection of its Voronoi cell and the convex
hell of the dataset.

Computing both N-dimensional Voronoi volumes and convex hulls is
difficult too. We adopt the following approximate algorithm. First, we
compute the axis-aligned bounding box of all of the conformations. Next,
we fit a multi-dimensional gaussian to the coordinates of each state.
(We should probably use a diagonal covariance matrix, or even a
spherical gaussian?).

Now, we sample random points from the uniform distribution over the
bounding box. We check these points against all of the gaussian PDFs. If
the point is farther than \texttt{3-sigma} away from every cluster
center, we reject it, reasoning that it is probably outside the convex
hull. The remaining points are now a representation of the uniform
distribution over the convex hull of phase space. For each point, we
compute its distance to each of the cluster centers, ``assigning'' it to
the center it is closest to. As the number of randomly sampled points
goes to infinity, the fraction of the points assigned to each state
converges to being proportional to the state's volume.

This algorithm is highly amenable to parallel computation.

\section{Choosing the Optimal Number of States}

Choose both the clustering algorithm (k-centers, k-means, etc) and the
number of states by maximizing the BIC score of the model, using this
likelihood.

\bibliography{bibliography}
\end{document}

### 2013-03-12 paper/postprocess_markdown.py

@@ -1,0 +1,13 @@
+import sys
+import string
+
+msg = "This file was automatically generated from latex using pandoc\n\n"
+
+sys.stdout.write(msg)
+
+f = open(sys.argv[1])
+for line in f.readlines():
+    # replace with backticks
+    line = string.replace(line, '$$', '`')
+    
+    sys.stdout.write(line)

### 2013-03-12 paper/paper.tex
math mode

@@
 always chose the trivial model, with only one state, as the transition
-probability in that model would be 1 and thus the likelihood of any
-trajectory within that state space would be 1.
+probability in that model would be $1$ and thus the likelihood of any
+trajectory within that state space would be $1$.
 
@@
 the state. That is, the likelihood of observing a conformation in phase
-space given that the conformation is assigned to state \texttt{i} is 0
+space given that the conformation is assigned to state $i$ is $0$
 if the conformation is outside of the bounding volume of the state and
 constant if the conformation is within the volume. The constant is set
-so that the distribution integrates to 1, and is thus the reciprocal
+so that the distribution integrates to $1$, and is thus the reciprocal
 volume of the microstate.
@@
 Voronoi cells. While trivial in two or three dimensions, this
-computational geometry task\\becomes challenging in large dimensional
+computational geometry task becomes challenging in large dimensional
 settings.

### 2013-03-12 paper/postprocess_markdown.py
latex in readme

@@ -1,13 +1,29 @@
 import sys
 import string
+import re
+import urllib
 
 msg = "This file was automatically generated from latex using pandoc\n\n"
 
 sys.stdout.write(msg)
+inline_math = re.compile('[$]+([^$]*)[$]+')
+
+base_url = 'http://latex.codecogs.com/gif.latex?'
 
 f = open(sys.argv[1])
 for line in f.readlines():
     # replace with backticks
-    line = string.replace(line, '$$', '`')
+    m1 = inline_math.search(line)
+    if m1:
+        #print >> sys.stderr, m1.group(1)
+        query = urllib.quote(m1.group(1))
+        line = '![equation](' + inline_math.sub(base_url + query, line) + ')'
+        
+    #line = string.replace(line, '$$', '`')
+    #line = string.replace(line, '$', '`')
+    
+    #![equation](http://latex.codecogs.com/gif.latex?1%2Bsin%28mc%5E2%29%0D%0A)
+
+    
     
     sys.stdout.write(line)

### 2013-03-12 paper/postprocess_markdown.py
fix

@@ -17,8 +17,10 @@ for line in f.readlines():
     if m1:
         #print >> sys.stderr, m1.group(1)
         query = urllib.quote(m1.group(1))
-        line = '![equation](' + inline_math.sub(base_url + query, line) + ')'
+        repl = '![equation](%s)' % (base_url + query)
+        line =  inline_math.sub(repl, line)
         
+        print >> sys.stderr, line
     #line = string.replace(line, '$$', '`')
     #line = string.replace(line, '$', '`')
     

### 2013-03-12 paper/postprocess_markdown.py
math better

@@ -6,26 +6,19 @@ import urllib
 msg = "This file was automatically generated from latex using pandoc\n\n"
 
 sys.stdout.write(msg)
-inline_math = re.compile('[$]+([^$]*)[$]+')
 
-base_url = 'http://latex.codecogs.com/gif.latex?'
+# match inline math or displayed math
+math_re = re.compile('[$]+([^$]*)[$]+')
+
+base_url = 'http://latex.codecogs.com/gif.latex'
 
 f = open(sys.argv[1])
 for line in f.readlines():
-    # replace with backticks
-    m1 = inline_math.search(line)
+    m1 = math_re.search(line)
     if m1:
-        #print >> sys.stderr, m1.group(1)
+        # replace latex with a url that renders the math
         query = urllib.quote(m1.group(1))
-        repl = '![equation](%s)' % (base_url + query)
-        line =  inline_math.sub(repl, line)
-        
-        print >> sys.stderr, line
-    #line = string.replace(line, '$$', '`')
-    #line = string.replace(line, '$', '`')
-    
-    #![equation](http://latex.codecogs.com/gif.latex?1%2Bsin%28mc%5E2%29%0D%0A)
-
-    
+        repl = '![equation](%s)' % (base_url + '?' + query)
+        line =  math_re.sub(repl, line)
     
     sys.stdout.write(line)

### 2013-03-12 paper/paper.tex
math better
3sigma
two eqs on same line

@@
 
-Therefore, the most appropriate emmision distribution for discrete state
+Therefore, the most appropriate emission distribution for discrete state
 MSMs is that of the uniform distribution over the phase-space volume of
@@
 bounding box. We check these points against all of the gaussian PDFs. If
-the point is farther than \texttt{3-sigma} away from every cluster
+the point is farther than $3\sigma$ away from every cluster
 center, we reject it, reasoning that it is probably outside the convex

### 2013-03-12 paper/postprocess_markdown.py

@@ -14,11 +14,15 @@ base_url = 'http://latex.codecogs.com/gif.latex'
 
 f = open(sys.argv[1])
 for line in f.readlines():
-    m1 = math_re.search(line)
-    if m1:
-        # replace latex with a url that renders the math
-        query = urllib.quote(m1.group(1))
-        repl = '![equation](%s)' % (base_url + '?' + query)
-        line =  math_re.sub(repl, line)
+    
+    while True:
+        m1 = math_re.search(line)
+        if m1:
+            # replace latex with a url that renders the math
+            query = urllib.quote(m1.group(1))
+            repl = '![equation](%s)' % (base_url + '?' + query)
+            line, _ =  math_re.subn(repl, line, 1)
+        else:
+            break
     
     sys.stdout.write(line)

### 2013-03-12 paper/paper.tex
added sentence about RMSD
Changed the likelihood equation to make more sense to me. If you don't like it, then switch it back, but more importantly define what you mean by O_{s=0}(x_0).

@@
 $$ 
-P[x(t)] dx^N = O_{s=0}(x_0) \prod_{i=0}^{N-1} T(x_i \rightarrow x_{i+1}) \cdot O_{s=i}(x_i)
+P[x_t] dx^N = \prod_{i=0}^{N-1} T(x_i \rightarrow x_{i+1}) \cdot \prod_{i=0}^{N} p(x_{i} | \sigma(x_{i}))
 $$
+%% CRS: I think this is a better way of writing the likelihood
+%% This is essentially what Liz Kellogg wrote in her likelihood.
 
@@
 conformation at a given location in phase space given that the
-conformation is within a certain state.
+conformation ($x_t$) is within a certain state ($\sigma(x_t)$).
 
@@
 
+Unfortunately, this doesn't really help with picking the projection operator
+to vectorize the conformations (e.g. dihedrals, contact maps, etc). Also,
+it's not going to work with RMSD.
+
 \bibliography{bibliography}

### 2013-03-12 pre-commit.sh
test commit

@@ -1,0 +1,11 @@
+#!/bin/sh
+files=`git diff --cached --name-status`
+re="<files of importance>"
+if [[ $files =~ $re ]]
+then
+  echo "Creating files"
+  bundle exec create_my_files
+  git add my_files
+fi
+
+exit 1

### 2013-03-12 post-commit.sh
changes

@@ -1,0 +1,10 @@
+#!/bin/sh
+#http://stackoverflow.com/questions/3284292/can-a-git-hook-automatically-add-files-to-the-commit
+if [ -a .commit ]
+    then
+    cd "$(git rev-parse --show-toplevel)/paper"
+    del .commit
+    git add README.md
+    git commit --ammend -C HEAD --no-verify
+fi
+exit

### 2013-03-12 pre-commit.sh

@@ -1,11 +1,6 @@
 #!/bin/sh
-files=`git diff --cached --name-status`
-re="<files of importance>"
-if [[ $files =~ $re ]]
-then
-  echo "Creating files"
-  bundle exec create_my_files
-  git add my_files
-fi
-
-exit 1
+echo "Rebuilding README.md from LaTeX"
+cd "$(git rev-parse --show-toplevel)/paper"
+make
+git add README.md
+touch .commit

### 2013-03-12 post-commit.sh
test

@@ -7,4 +7,3 @@ if [ -a .commit ]
     git add README.md
     git commit --ammend -C HEAD --no-verify
 fi
-exit

### 2013-03-12 post-commit.sh
stuff

@@ -1,9 +1,9 @@
 #!/bin/sh
 #http://stackoverflow.com/questions/3284292/can-a-git-hook-automatically-add-files-to-the-commit
-if [ -a .commit ]
-    then
-    cd "$(git rev-parse --show-toplevel)/paper"
-    del .commit
-    git add README.md
+echo "post commit hook"
+if [ -a "$(git rev-parse --show-toplevel)/.commit" ] then
+    rm "$(git rev-parse --show-toplevel)/.commit"
+    git add paper/README.md
     git commit --ammend -C HEAD --no-verify
+    echo "post commit hook"
 fi

### 2013-03-12 pre-commit.sh

@@ -1,6 +1,5 @@
 #!/bin/sh
 echo "Rebuilding README.md from LaTeX"
-cd "$(git rev-parse --show-toplevel)/paper"
-make
-git add README.md
+(cd "$(git rev-parse --show-toplevel)/paper" && make)
+#cd "$(git rev-parse --show-toplevel)" || exit
 touch .commit

### 2013-03-12 post-commit.sh
stuff

@@ -1,9 +1,9 @@
 #!/bin/sh
 #http://stackoverflow.com/questions/3284292/can-a-git-hook-automatically-add-files-to-the-commit
-echo "post commit hook"
-if [ -a "$(git rev-parse --show-toplevel)/.commit" ] then
+if [ -a "$(git rev-parse --show-toplevel)/.commit" ]
+then
     rm "$(git rev-parse --show-toplevel)/.commit"
     git add paper/README.md
-    git commit --ammend -C HEAD --no-verify
+    git commit --amend -C HEAD --no-verify
     echo "post commit hook"
 fi

### 2013-03-12 post-commit.sh
stuff

@@ -5,5 +5,5 @@ then
     rm "$(git rev-parse --show-toplevel)/.commit"
     git add paper/README.md
     git commit --amend -C HEAD --no-verify
-    echo "post commit hook"
+    echo "Adding new README.md"
 fi

### 2013-03-12 pre-commit.sh

@@ -1,5 +1,4 @@
 #!/bin/sh
 echo "Rebuilding README.md from LaTeX"
 (cd "$(git rev-parse --show-toplevel)/paper" && make)
-#cd "$(git rev-parse --show-toplevel)" || exit
 touch .commit

### 2013-03-12 post-commit.sh
make the hooks fail early

@@ -1,5 +1,6 @@
 #!/bin/sh
 #http://stackoverflow.com/questions/3284292/can-a-git-hook-automatically-add-files-to-the-commit
+set -e
 if [ -a "$(git rev-parse --show-toplevel)/.commit" ]
 then
     rm "$(git rev-parse --show-toplevel)/.commit"

### 2013-03-12 pre-commit.sh

@@ -1,4 +1,5 @@
 #!/bin/sh
+set -e
 echo "Rebuilding README.md from LaTeX"
 (cd "$(git rev-parse --show-toplevel)/paper" && make)
 touch .commit

### 2013-03-12 paper/paper.tex
changes
test
stuff
remove comment
changed notation in likelihood slightly
added ball tree

@@
 $$ 
-P[x_t] dx^N = \prod_{i=0}^{N-1} T(x_i \rightarrow x_{i+1}) \cdot \prod_{i=0}^{N} p(x_{i} | \sigma(x_{i}))
+P[x_{0...T-1}] dx^N = \prod_{i=0}^{T-1} T(x_i \rightarrow x_{i+1}) \cdot \prod_{i=0}^{N} p(x_{i} | \sigma(x_{i}))
 $$
-%% CRS: I think this is a better way of writing the likelihood
-%% This is essentially what Liz Kellogg wrote in her likelihood.
 
@@
 
-This algorithm is highly amenable to parallel computation.
+This algorithm is highly amenable to parallel computation. For a euclidean
+distance metric, the assignment can be sped up by using a BallTree data
+structure for fast neighbor search. (We can probably actually use this
+data structure within msmbuilder's assign step...)
+
+The documentation for the sklearn cython BallTree is given here \url{http://scikit-learn.org/dev/modules/generated/sklearn.neighbors.BallTree.html#sklearn.neighbors.BallTree}
 

### 2013-03-12 src/likelihood.py
initial code to calculate state volumes. This is by no means clean and polished but I just want to get something that works so that we can test these ideas.

@@ -1,0 +1,4 @@
+"""
+This library is used to calculate the likelihood of a particular MSM.
+
+The likelihood takes as input the 

### 2013-03-12 src/montecarlo.py

@@ -1,0 +1,81 @@
+
+from volumes import VolumeEstimator
+
+class MonteCarlo(VolumeEstimator):
+    """
+    This class uses a brute force Monte Carlo algorithm to compute the volume
+    of a state by randomly drawing samples from R^d and assigning them to a 
+    state.
+    """
+    
+    def __init__(self, *args, num_points=1E6, cushion=1.):
+        """
+        Compute the volume of states by using a simple monte carlo algorithm.
+
+        Parameters
+        ----------
+        metric : msmbuilder.metrics.Vectorized
+            metric to compute distances. Must be vectorized because we assume
+            that phase space is euclidean
+        generators : msmbuilder.Trajectory
+            generators that define the voronoi cells
+        num_points : int, optional
+            number of points to simulate to calculate the volumes
+        cushion : float, optional
+            cushion to use to define the bounding box for the data. The box
+            will be defined based on a hyberrectangle with corners in each 
+            dimension equal to 
+                [min_d(generators) - cushion, max_d(generators) + cushion]
+        """
+        super(self, MonteCarlo).__init__(*args)
+
+        self.num_points = num_points
+
+        self.mins = self.prep_generators.min(axis=0) - cushion
+        self.maxs = self.prep_generators.max(axis=0) + cushion
+        self.lengths = self.maxs - self.mins
+        
+        self.total_volume = np.prod(self.lengths)
+
+    def get_state_volumes(self, which_states=None):
+        """
+        Compute the volume of each state.
+        
+        Parameters
+        ----------
+        which_states : Placeholder. This is not used.
+
+        Returns
+        -------
+        volumes : np.ndarray
+            array of volumes for each state
+
+        """
+        # This algorithm is more efficient if we do all states at the same time
+        # so we will not use the same framework as the base class VolumeEstimator
+    
+        state_counts = np.zeros(len(self.prep_generators))
+
+        for i in xrange(self.num_points):
+        # It might be more efficient to write this vectorially, but I don't 
+        # want to run out of memory
+            random_sample = np.random.random(self.dimension)
+            random_sample = random_sample * self.lengths + self.mins
+            random_sample = random_sample.reshape((1, -1))
+            
+            state_dists = self.metric.one_to_all(random_sample, self.prep_generators, 0)
+
+            state_counts[np.argmin(state_dists)] += 1
+
+        points_per_state = state_counts.astype(float) / self.num_points
+
+        volumes = points_per_state * self.total_volume
+
+        return volumes
+        
[…]

### 2013-03-12 src/volumes.py

@@ -1,0 +1,62 @@
+"""
+This library is used to compute the volume of states.
+"""
+
+class VolumeEstimator(object):
+    """
+    This base class can be subclassed by implementations that use an algorithm
+    to compute the volume of a state.
+
+    Each implementation should use the distance metric and generators to return
+    a list of volumes
+    """
+    def __init__(self, metric, generators):
+    """
+    Parameters
+    ----------
+    metric : msmbuilder.metric.Vectorized instance or subclass instance
+        distance metric to use to calculate distances between points. This
+        needs to be a Vectorized type because we assume that phase space is
+        euclidean
+    generators : msmbuilder.Trajectory instance
+        each state is defined by the voronoi cell defined by the generators
+    """
+
+        if not isinstance(metric, msmbuilder.metrics.Vectorized):
+            raise Exception("Metric must be a subclass of msmbuilder.metrics.Vectorized")
+
+        self.metric = metric
+        self.generators = generators
+
+        self.prep_generators = self.metric.prepare_trajectory(self.generators)
+        self.dimension = self.prep_generators.shape[1]
+    
+    def get_state_volumes(self, which_states=None):
+    """
+    Parameters
+    ----------
+    which_states : np.ndarray, optional
+        Calculate volumes only for requested states. If None given, then 
+        get volume for each state
+    
+    Returns
+    -------
+    volumes : np.ndarray
+        volume for each state in same units as metric.prepare_trajectory 
+    """
+
+        if which_states is None:
+            which_states = np.arange(len(self.generators))
+
+        volumes = np.ones(len(which_states)) * -1
+        for i, s in enumerate(which_states):
+            print "Working on state %d" % i
+            
+            volumes[i] = self._get_volume(s)
+            
+        return volumes
+    
+    def _get_volume(state):
+    """Internal function to get the volume of a particular state.
+    """
+        raise NotImplementedError

### 2013-03-12 scripts/get_vols.py
Added scripts/get_vols.py which will get state volumes with a specified distance metric and generators. Also added test/Gens.lh5 which is the generators file from msmbuillder/reference

@@ -1,0 +1,10 @@
+
+
+from msmbuilder import Trajectory
+from msmbuilder import arglib
+import numpy as np
+from montecarlo import MonteCarlo
+
+parser = arglib.ArgumentParser(get_metric=True)
+
+parser.add_argument('generators', help='Generators filename')

### 2013-03-13 lib/montecarlo.py
Added testALA which calculates the volumes associated with the generators of an Alanine dipeptide state decomposition. For these systems, the volumes at least appear to converge in 1E6 samples. They are in 2 dimensions however, so the number of samples will be very very large for higher dimensional cases.

@@ -1,4 +1,6 @@
 
+from msmbuilder import metrics
+import numpy as np
 from volumes import VolumeEstimator
 
 class MonteCarlo(VolumeEstimator):
@@ -8,7 +10,7 @@ class MonteCarlo(VolumeEstimator):
     state.
     """
     
-    def __init__(self, *args, num_points=1E6, cushion=1.):
+    def __init__(self, metric, generators, num_points=1E6, cushion=1.):
         """
         Compute the volume of states by using a simple monte carlo algorithm.
 
@@ -27,9 +29,9 @@ class MonteCarlo(VolumeEstimator):
             dimension equal to 
                 [min_d(generators) - cushion, max_d(generators) + cushion]
         """
-        super(self, MonteCarlo).__init__(*args)
+        super(MonteCarlo, self).__init__(metric, generators)
 
-        self.num_points = num_points
+        self.num_points = int(num_points)
 
         self.mins = self.prep_generators.min(axis=0) - cushion
         self.maxs = self.prep_generators.max(axis=0) + cushion

### 2013-03-13 lib/montecarlo.py
Added directory to hold the muller test cases. Also changed MonteCarlo to reject points that are not 'close enough' to the data. The way I defined that was whether that point is closer than some cutoff (cushion) to a set of space_generators, that can be the same as the clustering generators or different (e.g., the whole dataset, or a different clustering's generators).

@@ -10,7 +10,8 @@ class MonteCarlo(VolumeEstimator):
     state.
     """
     
-    def __init__(self, metric, generators, num_points=1E6, cushion=1.):
+    def __init__(self, metric, generators, space_generators=None,
+         num_points=1E6, cushion=1.):
         """
         Compute the volume of states by using a simple monte carlo algorithm.
 
@@ -21,23 +22,22 @@ class MonteCarlo(VolumeEstimator):
             that phase space is euclidean
         generators : msmbuilder.Trajectory
             generators that define the voronoi cells
+        space_generators : msmbuilder.Trajectory, optional
+            conformations that define the full space. If None, then generators
+            are used
         num_points : int, optional
             number of points to simulate to calculate the volumes
         cushion : float, optional
-            cushion to use to define the bounding box for the data. The box
-            will be defined based on a hyberrectangle with corners in each 
-            dimension equal to 
-                [min_d(generators) - cushion, max_d(generators) + cushion]
+            cushion to use to define full space of the data. A point is in
+            this space if it is within <cushion> of any of the space_generators
         """
         super(MonteCarlo, self).__init__(metric, generators)
 
         self.num_points = int(num_points)
+        self.cushion = float(cushion)
 
-        self.mins = self.prep_generators.min(axis=0) - cushion
-        self.maxs = self.prep_generators.max(axis=0) + cushion
-        self.lengths = self.maxs - self.mins
-        
-        self.total_volume = np.prod(self.lengths)
+        self.space_generators = space_metrics
+        self.prep_space_generators = self.metric.prepare_trajectory(self.space_generators)
 
     def get_state_volumes(self, which_states=None):
         """
@@ -58,12 +58,17 @@ class MonteCarlo(VolumeEstimator):
     
         state_counts = np.zeros(len(self.prep_generators))
 
-        for i in xrange(self.num_points):
+        while state_counts.sum() < num_points:
         # It might be more efficient to write this vectorially, but I don't 
         # want to run out of memory
             random_sample = np.random.random(self.dimension)
             random_sample = random_sample * self.lengths + self.mins
             random_sample = random_sample.reshape((1, -1))
+            
+            space_dists = self.metric.one_to_all(random_sample, self.prep_space_generators, 0)
+
+            if np.min(space_dists) > self.cushion:
+                continue
             
             state_dists = self.metric.one_to_all(random_sample, self.prep_generators, 0)
 

### 2013-03-13 muller/simulate.py

@@ -1,0 +1,23 @@
+import random
+from mullermsm import muller
+mullerforce = muller.muller_force()
+from sklearn import mixture
+import scipy.linalg
+from matplotlib.pyplot import *
+from argparse import ArgumentParser
+
+parser = ArgumentParser()
+parser.add_argument('--dt', dest='dt', type=float, default=0.1)
+parser.add_argument('-n', dest='num_frames', type=int, default=100000)
+parser.add_argument('-o', dest='output', default='pos.npy')
+
+args = parser.parse_args()
+
+kT = 15.0
+dt = args.dt
+mGamma = 1000.0
+traj_length = args.num_frames 
+initial_x = [random.uniform(-1.5, 1.2), random.uniform(-0.2, 2)]
+positions = muller.propagate(traj_length, initial_x, kT, dt, mGamma, mullerforce)
+
+np.save(args.output, positions)

### 2013-03-13 paper/paper.tex
initial code to calculate state volumes. This is by no means clean and polished but I just want to get something that works so that we can test these ideas.
Modifications to MonteCarlo to work how we want: i.e. calculate the volumes relative to the volume of the data. We are approximating the volume of the data, by some buffer around all of the data points or some subset of those datapoints.

@@
 \usepackage{amsmath}    % need for subequations
+\usepackage{amssymb}    % for symbols
 \usepackage{graphicx}   % need for figures
@@
 $$ 
-P[x_{0...T-1}] dx^N = \prod_{i=0}^{T-1} T(x_i \rightarrow x_{i+1}) \cdot \prod_{i=0}^{N} p(x_{i} | \sigma(x_{i}))
+P[x_{0...T-1}] dx^N = \prod_{i=0}^{T-1} T(x_i \rightarrow x_{i+1}) \cdot \prod_{i=0}^{T} p(x_{i} | \sigma(x_{i}))
 $$
@@
 
+% \begin{align*}
+% p(x | s) &= C \\
+% \int p(x | s) dx &= 1 \\
+% \int C \mathbf{1}_s dx &= 1 \\
+% C &= \frac{1}{\int \mathbf{1}_s dx}
+% \end{align*}
+
+$$ 
+P[x_{0...T-1}] dx^N = \prod_{i=0}^{T-1} T(x_i \rightarrow x_{i+1}) \cdot \prod_{i=0}^T \frac{1}{V_{\sigma(x_{i})}}
+$$
+
 \section{Algorithm}
@@
 computational geometry task becomes challenging in large dimensional
-settings.
+settings. The computation of high dimensional volumes has occupied significant
+attention in recent years in the computational geometry literature, especially
+via randomized algorithms. [See issue \#1 on the github]
 
@@
 Instead, it seems appropriate to assert that the volume is bounded by
-the convex hull of the data. That is, the volume of a state is defined
+the convex hull of the data. That is, the volume of a state might be defined
 as the volume of the intersection of its Voronoi cell and the convex
-hell of the dataset.
+hull of the dataset.
 
-Computing both N-dimensional Voronoi volumes and convex hulls is
-difficult too. We adopt the following approximate algorithm. First, we
-compute the axis-aligned bounding box of all of the conformations. Next,
-we fit a multi-dimensional gaussian to the coordinates of each state.
-(We should probably use a diagonal covariance matrix, or even a
-spherical gaussian?).
+Instead, we use a slightly modified version of this definition that adopts
+the same spirit. Instead of taking the outer bounding region to be the convex
+hull of the data, we take it to be the set of all trail points such that the nearest neighbor of the trial point within the dataset is less than a certain
+cutoff, $R$. This can be computed relatively efficiently using a BallTree data
+structure. For further efficiency, we might use only a random subsample of the
+dataset for this nearest neighbor computation.
+
+Instead of computing the volume of the Voronoi cells explicitly, we instead
+compute the ratio of volume of each Voronoi cell to the volume of the entire
+bounding region.
+Because the bounding region's volume is independent of the clustering parameters, its inclusion changes all of the calculated likelihoods by a constant multiplicative factor, and can thus be discarded from the perspective of model comparison.
+
+To compute the fractional volumes of the Voronoi cells, we use the following randomized algorithm.
+First, we generate points inside of the bounding region using the lazy walker markov chain monte carlo algorithm.
+
+% Computing both N-dimensional Voronoi volumes and convex hulls is
+% difficult too. We adopt the following approximate algorithm. First, we
+% compute the axis-aligned bounding box of all of the conformations. Next,
+% we fit a multi-dimensional gaussian to the coordinates of each state.
+% (We should probably use a diagonal covariance matrix, or even a
+% spherical gaussian?).
 

### 2013-03-13 lib/likelihood.py

@@ -1,4 +1,47 @@
 """
 This library is used to calculate the likelihood of a particular MSM.
 
-The likelihood takes as input the 
+The likelihood takes as input the transition matrix as well as the volume
+of each state.
+"""
+
+
+def get_log_likelihood(tProb, trajs, volumes, lagtime=1):
+    """
+    Function to calculate the log likelihood according to:
+
+    .. math:: \log L = \sum_{i,j} C_{ij} \log T_{ij} - 
+        \sum_{i,j} c_i \log V_i
+
+    where :math:`C_{ij}` is the number of transitions from state i to state
+    j, :math:`T_{ij}` is the probability of transistioning from state i to
+    state j, :math:`c_i` is the number of samples of state i, and :math:`V_i` 
+    is the volume of state i in phase space.
+
+    Parameters
+    ----------
+    tProb : scipy.sparse matrix or np.ndarray
+        transition probability matrix from msmbuilder
+    trajs : np.ndarray
+        trajectory of states
+    volumes : np.ndarray
+        volume for each state in the msm
+    lagtime : int
+        lag time used to build the msm. Will subsample the trajs accordingly
+
+    Returns
+    -------
+    log_like : float
+        log-likelihood as defined above
+    """
+
+    counts = msm_analysis.get_counts_from_assignments(trajs, lag_time=lagtime)
+
+    state_counts = counts.sum(axis=1)
+
+    ij = counts.nonzero()
+
+    log_like = np.sum(counts.data * np.log(tProb[ij])) - \
+        np.sum(state_counts * np.log(volumes))
+    
+    return log_like

### 2013-03-13 lib/montecarlo.py

@@ -36,8 +36,12 @@ class MonteCarlo(VolumeEstimator):
         self.num_points = int(num_points)
         self.cushion = float(cushion)
 
-        self.space_generators = space_metrics
+        self.space_generators = space_generators
         self.prep_space_generators = self.metric.prepare_trajectory(self.space_generators)
+
+        self.mins = np.min(self.prep_space_generators, axis=0) - self.cushion
+        self.maxs = np.max(self.prep_space_generators, axis=0) + self.cushion
+        self.lengths = self.maxs - self.mins
 
     def get_state_volumes(self, which_states=None):
         """
@@ -58,9 +62,10 @@ class MonteCarlo(VolumeEstimator):
     
         state_counts = np.zeros(len(self.prep_generators))
 
-        while state_counts.sum() < num_points:
+        while state_counts.sum() < self.num_points:
         # It might be more efficient to write this vectorially, but I don't 
         # want to run out of memory
+            print state_counts.sum()
             random_sample = np.random.random(self.dimension)
             random_sample = random_sample * self.lengths + self.mins
             random_sample = random_sample.reshape((1, -1))
@@ -74,9 +79,8 @@ class MonteCarlo(VolumeEstimator):
 
             state_counts[np.argmin(state_dists)] += 1
 
-        points_per_state = state_counts.astype(float) / self.num_points
-
-        volumes = points_per_state * self.total_volume
+        volumes = state_counts.astype(float) / self.num_points
+        # volumes are just percentages of the total space
 
         return volumes
         

### 2013-03-13 muller/simulate.py

@@ -1,7 +1,6 @@
 import random
 from mullermsm import muller
 mullerforce = muller.muller_force()
-from sklearn import mixture
 import scipy.linalg
 from matplotlib.pyplot import *
 from argparse import ArgumentParser

### 2013-03-13 scripts/get_vols.py

@@ -27,21 +27,39 @@ class DihedralWrap(metrics.Vectorized):
         print 'hi'
         return arcC
 
+class Euclidean(metrics.Vectorized):
+    def prepare_trajectory(self, traj):
+        return traj
+
 parser = arglib.ArgumentParser(get_metric=True)
 
 parser.add_argument('generators', help='Generators filename')
+parser.add_argument('space_gens', help='Generators to use to define the whole space')
+parser.add_argument('cushion', type=float, help='Cushion to require points be within a generator to include it as a random sample.')
 parser.add_argument('output', help='output filename to save volumes.')
 parser.add_argument('num_points', help='Number of points to sample.')
 parser.add_argument('use_dih', action='store_true', default=False)
+parser.add_argument('use_euc', action='store_true', default=False)
 
 args, metric = parser.parse_args()
 
-if args.use_dih:
+if args.use_euc:
+    metric = Euclidean()
+elif args.use_dih:
     metric = DihedralWrap(dih_metric=copy.deepcopy(metric))
 
-gens = Trajectory.load_from_lhdf(args.generators)
+try: 
+    gens = np.load(args.generators)
+except: 
+    gens = Trajectory.load_from_lhdf(args.generators)
 
-MC = MonteCarlo(metric, gens, num_points=args.num_points)
+try:
+    space_gens = np.load(args.space_gens)
+except:
+    space_gens = Trajectory.load_from_lhdf(args.space_gens)
+
+MC = MonteCarlo(metric, gens, space_gens, num_points=args.num_points, 
+        cushion=args.cushion)
 
 volumes = MC.get_state_volumes()
 

### 2013-03-13 paper/paper.tex
updated paper

@@
 \usepackage{subfigure}  % use for side-by-side figures
-\usepackage{hyperref}   % use for hypertext links, including those to external documents and URLs
+%\usepackage{hyperref}   % use for hypertext links, including those to external documents and URLs
 
@@
 To compute the fractional volumes of the Voronoi cells, we use the following randomized algorithm.
-First, we generate points inside of the bounding region using the lazy walker markov chain monte carlo algorithm.
-
-% Computing both N-dimensional Voronoi volumes and convex hulls is
-% difficult too. We adopt the following approximate algorithm. First, we
-% compute the axis-aligned bounding box of all of the conformations. Next,
-% we fit a multi-dimensional gaussian to the coordinates of each state.
-% (We should probably use a diagonal covariance matrix, or even a
-% spherical gaussian?).
-
-Now, we sample random points from the uniform distribution over the
-bounding box. We check these points against all of the gaussian PDFs. If
-the point is farther than $3\sigma$ away from every cluster
-center, we reject it, reasoning that it is probably outside the convex
-hull. The remaining points are now a representation of the uniform
-distribution over the convex hull of phase space. For each point, we
-compute its distance to each of the cluster centers, ``assigning'' it to
-the center it is closest to. As the number of randomly sampled points
-goes to infinity, the fraction of the points assigned to each state
-converges to being proportional to the state's volume.
+First, we generate points inside of the bounding region using the lazy random walk Markov chain Monte Carlo algorithm.\cite{Kannan97} This random walk converges to sampling from the uniform distribution over the interior.
+After a sufficient burn in period, for each sampled point we compute its nearest MSM state center, assigning it to that state.
+As the number of randomly sampled points goes to infinity, the fraction of the points assigned to each state converges to being proportional to the state's volume.
 
 This algorithm is highly amenable to parallel computation. For a euclidean
-distance metric, the assignment can be sped up by using a BallTree data
-structure for fast neighbor search. (We can probably actually use this
-data structure within msmbuilder's assign step...)
-
-The documentation for the sklearn cython BallTree is given here \url{http://scikit-learn.org/dev/modules/generated/sklearn.neighbors.BallTree.html#sklearn.neighbors.BallTree}
+distance metric, the assignment can be performed efficiently by using a BallTree 
+data structure for fast neighbor search.
 
@@
 
-Choose both the clustering algorithm (k-centers, k-means, etc) and the
-number of states by maximizing the BIC score of the model, using this
+We choose both the clustering algorithm (k-centers, k-means, etc) and the
+number of states by maximizing the BIC/AIC scores of the model, using this
 likelihood.
@@
 to vectorize the conformations (e.g. dihedrals, contact maps, etc). Also,
-it's not going to work with RMSD.
+it's not going to work rigorously with RMSD.
+
+\section{Validation}
+We being by using the procedure described above on the M\"{u}ller potential.
 

### 2013-03-13 lib/likelihood.py
bug fixes in lib/likelihood.py

@@ -4,7 +4,8 @@ This library is used to calculate the likelihood of a particular MSM.
 The likelihood takes as input the transition matrix as well as the volume
 of each state.
 """
-
+from msmbuilder import MSMLib
+import numpy as np
 
 def get_log_likelihood(tProb, trajs, volumes, lagtime=1):
     """
@@ -35,13 +36,16 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1):
         log-likelihood as defined above
     """
 
-    counts = msm_analysis.get_counts_from_assignments(trajs, lag_time=lagtime)
+    counts = MSMLib.get_count_matrix_from_assignments(trajs, lag_time=lagtime)
 
+
+    tProb = tProb.tocsr()
+    counts.eliminate_zeros()
     state_counts = counts.sum(axis=1)
 
     ij = counts.nonzero()
 
-    log_like = np.sum(counts.data * np.log(tProb[ij])) - \
+    log_like = np.sum(np.array(counts.data) * np.array(np.log(tProb[ij]))) - \
         np.sum(state_counts * np.log(volumes))
     
     return log_like

### 2013-03-13 paper/paper.tex

@@
 One challenge is how to model the volume of states which are at the
-``edge'' of the model. Is the volume of these states infinite, extending
-all the way out to infinity in some direction? This seems problematic.
-Instead, it seems appropriate to assert that the volume is bounded by
-the convex hull of the data. That is, the volume of a state might be defined
-as the volume of the intersection of its Voronoi cell and the convex
-hull of the dataset.
+``edge'' of the model. Is the volume of these states unbounded, given that they extend all the way out to infinity in some direction? This would be problematic.
+It seems appropriate to assert that the volume of these edge states is bounded in some way by the extent of our dataset.
+For example, the volume of a state might be defined as the volume of the intersection of its Voronoi cell and the convex hull of the whole dataset.
 
-Instead, we use a slightly modified version of this definition that adopts
-the same spirit. Instead of taking the outer bounding region to be the convex
-hull of the data, we take it to be the set of all trail points such that the nearest neighbor of the trial point within the dataset is less than a certain
-cutoff, $R$. This can be computed relatively efficiently using a BallTree data
-structure. For further efficiency, we might use only a random subsample of the
-dataset for this nearest neighbor computation.
+We use a slightly modified version of this definition that adopts
+the same spirit.
+Instead of taking the outer bounding region to be the convex hull of the data, we take it to be the set of all trial points such that the nearest data point to the trial point is closer than a certain cutoff, $R$.
+This can be computed relatively efficiently using a BallTree data structure.
+For further efficiency, we might use only a random subsample of the dataset for this nearest neighbor computation.
 

### 2013-03-13 lib/likelihood.py
Bugfix in likelihood.py The problem was that scipy.sparse matrices use np.matrices as opposed to np.arrays, so when I was using the '*' operator it was doing matrix multiplication when I really wanted element-wise multiplication. The solution was to cast everything to arrays so that it is obvious what is going on.

@@ -12,7 +12,7 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1):
     Function to calculate the log likelihood according to:
 
     .. math:: \log L = \sum_{i,j} C_{ij} \log T_{ij} - 
-        \sum_{i,j} c_i \log V_i
+        \sum_{i} c_i \log V_i
 
     where :math:`C_{ij}` is the number of transitions from state i to state
     j, :math:`T_{ij}` is the probability of transistioning from state i to
@@ -38,14 +38,21 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1):
 
     counts = MSMLib.get_count_matrix_from_assignments(trajs, lag_time=lagtime)
 
-
+    counts = counts.tocsr()
     tProb = tProb.tocsr()
-    counts.eliminate_zeros()
-    state_counts = counts.sum(axis=1)
 
     ij = counts.nonzero()
 
-    log_like = np.sum(np.array(counts.data) * np.array(np.log(tProb[ij]))) - \
-        np.sum(state_counts * np.log(volumes))
+    state_counts = np.array(counts.sum(axis=1)).flatten()
+    nonzero_counts = np.array(counts.data).flatten()
+    nonzero_tprobs = np.array(tProb[ij]).flatten()
+    volumes = np.array(volumes).flatten()
+    # scipy.sparse uses np.matrices which behave differently than arrays
+    # so we cast everything to arrays and do the math with them
     
+    #log_like = np.sum(np.array(counts.data) * np.array(np.log(tProb[ij]))) - \
+    #    np.sum(state_counts * np.log(volumes))
+    log_like = np.sum(nonzero_counts * np.log(nonzero_tprobs)) - \
+        np.sum(state_counts * np.log(volumes))
+
     return log_like

### 2013-03-13 muller/MSMs/get_like.py
Added likelihoods for 1-100 + 1000 state models

@@ -1,0 +1,18 @@
+import likelihood
+import numpy as np
+from msmbuilder import io
+from scipy.io import mmread
+
+
+Ls = []
+Ls.append(0) # for a one state model the likelihood is 0
+
+for k in range(2, 101) + [ 1000 ]:
+    print k
+    t = mmread('k%d/Lag30/tProb.mtx' % k)
+    v = np.load('k%d/vols.npy' % k)
+    a = io.loadh('k%d/Assignments.h5' % k)
+    a = a['arr_0']
+    Ls.append(likelihood.get_log_likelihood(t, a, v, lagtime=30))
+
+np.save('likes.npy', np.array(Ls))

### 2013-03-14 muller/MSMs/get_like.py
Added additional data

@@ -7,7 +7,7 @@ from scipy.io import mmread
 Ls = []
 Ls.append(0) # for a one state model the likelihood is 0
 
-for k in range(2, 101) + [ 1000 ]:
+for k in range(50,1050,50):
     print k
     t = mmread('k%d/Lag30/tProb.mtx' % k)
     v = np.load('k%d/vols.npy' % k)
@@ -15,4 +15,4 @@ for k in range(2, 101) + [ 1000 ]:
     a = a['arr_0']
     Ls.append(likelihood.get_log_likelihood(t, a, v, lagtime=30))
 
-np.save('likes.npy', np.array(Ls))
+np.save('likes1000_s50.npy', np.array(Ls))

### 2013-03-14 muller/MSMs/get_like.py
New data generated by volumes calculated with 100,000 MC samples.

@@ -10,9 +10,9 @@ Ls.append(0) # for a one state model the likelihood is 0
 for k in range(50,1050,50):
     print k
     t = mmread('k%d/Lag30/tProb.mtx' % k)
-    v = np.load('k%d/vols.npy' % k)
+    v = np.load('k%d/vols100k.npy' % k)
     a = io.loadh('k%d/Assignments.h5' % k)
     a = a['arr_0']
     Ls.append(likelihood.get_log_likelihood(t, a, v, lagtime=30))
 
-np.save('likes1000_s50.npy', np.array(Ls))
+np.save('likes1000_s50_n100k.npy', np.array(Ls))

### 2013-03-14 implementation/setup.py
removed the implemenation/ folder that wasn't being used

@@ -1,9 +1,0 @@
-from setuptools import setup
-
-setup(name='opt-k',
-      version='0.1',
-      description='Likelihood framework for choosing the optimal number of states in a markov state model',
-      url='https://github.com/rmcgibbo/opt-k',
-      packages=['opt-k'],
-      install_requires=['scikit-learn>=0.13.1']
-      )

### 2013-03-14 lib/likelihood.py
added just transition probability based likelihood with it's BIC to show that we cannot compare different state spaces using the transition likelihood alone.

@@ -52,7 +52,10 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1):
     
     #log_like = np.sum(np.array(counts.data) * np.array(np.log(tProb[ij]))) - \
     #    np.sum(state_counts * np.log(volumes))
-    log_like = np.sum(nonzero_counts * np.log(nonzero_tprobs)) - \
-        np.sum(state_counts * np.log(volumes))
+    #log_like = np.sum(nonzero_counts * np.log(nonzero_tprobs)) - \
+    #    np.sum(state_counts * np.log(volumes))
 
-    return log_like
+    t_like = np.sum(nonzero_counts * np.log(nonzero_tprobs))
+    v_like = - np.sum(state_counts * np.log(volumes))
+
+    return t_like, v_like

### 2013-03-14 lib/likelihood.py
Added option to separate the likelihood contributions vs. adding them together

@@ -7,7 +7,7 @@ of each state.
 from msmbuilder import MSMLib
 import numpy as np
 
-def get_log_likelihood(tProb, trajs, volumes, lagtime=1):
+def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False):
     """
     Function to calculate the log likelihood according to:
 
@@ -58,4 +58,7 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1):
     t_like = np.sum(nonzero_counts * np.log(nonzero_tprobs))
     v_like = - np.sum(state_counts * np.log(volumes))
 
-    return t_like, v_like
+    if separate:
+        return t_like, v_like
+    
+    return t_like + v_like

### 2013-03-14 muller/get_like.py
renamed locations of clustering data.

@@ -9,9 +9,9 @@ Ls.append(0) # for a one state model the likelihood is 0
 
 for k in range(50,1050,50):
     print k
-    t = mmread('k%d/Lag30/tProb.mtx' % k)
-    v = np.load('k%d/vols100k.npy' % k)
-    a = io.loadh('k%d/Assignments.h5' % k)
+    t = mmread('KCenters/k%d/Lag30/tProb.mtx' % k)
+    v = np.load('KCenters/k%d/vols100k.npy' % k)
+    a = io.loadh('KCenters/k%d/Assignments.h5' % k)
     a = a['arr_0']
     Ls.append(likelihood.get_log_likelihood(t, a, v, lagtime=30))
 

### 2013-03-14 paper/paper.tex
test hooks
Started adding validation results to the paper. This is definitely messy, but it's a start.

@@
 %\usepackage{hyperref}   % use for hypertext links, including those to external documents and URLs
-
+\usepackage[capitalise]{cleveref}   % use for referencing figures/equations
 \begin{document}
@@
 \section{Validation}
-We being by using the procedure described above on the M\"{u}ller potential.
+
+\subsection{Langevin Dynamics on the M\"uller Potential}
+We being by using the procedure described above on data generated by simulating Langevin dynamics on the M\"{u}ller potential (\cref{fig:muller_pot}).
+This data was clustered into models of different sizes using one of two algorithms.
+The first, K-Centers, tends to make equal sized states as it works to minimize the maximum distance between a conformation and its generator.
+The second, Hybrid, will make varying size states because it works to minimize the average distance between conformations and their generators.
+In both cases, we built models ranging from 50 to 1,000 total states.
+Since this space is only two-dimensional, we used the simple, brute-force MC algorithm for calculating volumes.
+
+Briefly, we generated uniform random vectors within a bounding box whose size was slightly larger than the data.
+This random sample was then accepted if it was 'close enough' to a generator in the 100 state model, where close enough meant it was within 0.2 units from any generator in that model.
+This was used to approximate the convex hull of the data.
+Then the accepted points were assigned to a state.
+Then the volume of a state is proportional to the number of samples assigned to that state divided by the total number of samples.
+We note, that these volumes are actually in units of the total acceptable volume.
+This affects the absolute value of the likelihood, however, since the total volume was constant in all cases, the relative likelihoods are the same.
+
+\begin{figure}[h!]
+\includegraphics[width=3in]{figs/muller_pot_traj.pdf}
+\caption{The M\"{u}ller potential consists of three metastable wells. We simulated Langevin dynamics on this potential as validation of the new likelihood function.}
+\label{fig:muller_pot}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=2in]{figs/like_comp.pdf}
+\end{figure}
 

### 2013-03-14 muller/cluster.py

@@ -1,39 +1,0 @@
-
-from msmbuilder import clustering, io
-import numpy as np
-from mullermsm.metric import EuclideanMetric
-import os
-
-
-from argparse import ArgumentParser
-
-parser = ArgumentParser()
-parser.add_argument('-t', dest='data')
-#parser.add_argument('-d', dest='dist_cut')
-parser.add_argument('-k', dest='num_states', type=int)
-parser.add_argument('-o', dest='output_dir')
-
-args = parser.parse_args()
-
-traj = np.load(args.data)
-
-metric = EuclideanMetric()
-
-gen_ids, ass_gen_ids, distances = clustering._kcenters(metric, traj, k=args.num_states, seed=np.random.randint(len(traj)))
-
-ass_gen_ids = np.array([ass_gen_ids])
-
-if not os.path.exists(args.output_dir):
-    os.mkdir(args.output_dir)
-
-ass_contig = np.ones(ass_gen_ids.shape) * -1
-ass_contig = ass_contig.astype(int)
-
-for i, j in enumerate(gen_ids):
-    ass_contig[np.where(ass_gen_ids == j)] = i
-
-np.savetxt(os.path.join(args.output_dir, 'gen_ids.dat'), gen_ids)
-np.save(os.path.join(args.output_dir, 'gens.npy'), traj[gen_ids])
-io.saveh(os.path.join(args.output_dir, 'Assignments.h5'), ass_contig)
-io.saveh(os.path.join(args.output_dir, 'Assignments.h5.distances'), np.array([distances]))
-print "Saved output to %s" % args.output_dir

### 2013-03-15 paper/paper.tex
Added change to likelihood.py to allow you to compute Liz Kellogg's (Baker Lab) likelihood function. Also added stuff to the paper.tex as well as likelihoods.ipynb

@@
 
-$$ 
+\begin{equation}
+\label{eq:like}
 P[x_{0...T-1}] dx^N = \prod_{i=0}^{T-1} T(x_i \rightarrow x_{i+1}) \cdot \prod_{i=0}^{T} p(x_{i} | \sigma(x_{i}))
-$$
+\end{equation}
 
@@
 
-$$ 
+\begin{equation}
+\label{eq:like_vol}
 P[x_{0...T-1}] dx^N = \prod_{i=0}^{T-1} T(x_i \rightarrow x_{i+1}) \cdot \prod_{i=0}^T \frac{1}{V_{\sigma(x_{i})}}
-$$
+\end{equation}
 
@@
 This affects the absolute value of the likelihood, however, since the total volume was constant in all cases, the relative likelihoods are the same.
-
 \begin{figure}[h!]
 \includegraphics[width=3in]{figs/muller_pot_traj.pdf}
-\caption{The M\"{u}ller potential consists of three metastable wells. We simulated Langevin dynamics on this potential as validation of the new likelihood function.}
+\caption{The M\"{u}ller potential consists of three metastable wells. We simulated Langevin dynamics on this potential as a validation of the new likelihood function.}
 \label{fig:muller_pot}
@@
 
+We built an MSM using a fixed lag time for each state decomposition and then calculated the likelihood according to \cref{eq:like_vol}.
+These likelihoods increased rapidly with $k$, but then plateaued at a value of around 300 states \cref{fig:like_kcenters}.
+At this plateau, we can conclude that we do not gain any new information by adding new states to our model and so should pick $k$ to be this point.
+This is a heuristic argument that can be made more mathematically rigorous with the introduction of functions like the Bayesian Information Content (BIC), which is defined in \cref{eq:bic}.
+
+\begin{equation}
+BIC = -2 \log L + m \log(n) 
+\label{eq:bic}
+\end{equation} Here, $L$ denotes the likelihood of a model, while $m$ is the number of parameters used in the model, and $n$ is the number of data points.
+Essentially this is a mathematical way of saying we would like to use a model with the maximum likelihood, but without introducing too many parameters.
+% don't know how to say this yet:
+%We note that we cannot take this too seriously, and there are many other such criteria, such as the Aikaike Information Content (AIC). 
+We found that the BIC of the models built using K-Centers clustering has a minimum at $k=250$.
+This minimum is the ``optimal'' model according to the BIC of our likelihood.
+As a check of the usefulness of this minimum, we plotted the three slowest implied timescales in the models as a function of $k$.
+The optimal model was close to the point at which the implied timescales became constant.
+
 \begin{figure}
 \includegraphics[width=2in]{figs/like_comp.pdf}
+\caption{For models built between 50 and 1,000 states using the K-Centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states.
+Heuristically, at this point, adding another state does not add anything to the model.
+As a result, by using a criterion, like the BIC, that penalizes using additional parameters to fit the data, we can find an ``optimal'' model.
+Interestingly, the BIC optimal model occurs at 250 states, which also corresponds to where the three slowest timescales plateau.}
+\label{fig:like_kcenters}
 \end{figure}

### 2013-03-15 lib/likelihood.py

@@ -7,7 +7,8 @@ of each state.
 from msmbuilder import MSMLib
 import numpy as np
 
-def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False):
+def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False, 
+    kellogg=False):
     """
     Function to calculate the log likelihood according to:
 
@@ -29,6 +30,10 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False):
         volume for each state in the msm
     lagtime : int
         lag time used to build the msm. Will subsample the trajs accordingly
+    separate : bool, optional
+        separate transition probability component from the volume component
+    kellogg : bool, optional
+        if true then use Kellogg's approximation
 
     Returns
     -------
@@ -56,7 +61,11 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False):
     #    np.sum(state_counts * np.log(volumes))
 
     t_like = np.sum(nonzero_counts * np.log(nonzero_tprobs))
-    v_like = - np.sum(state_counts * np.log(volumes))
+
+    if not kellogg:
+        v_like = - np.sum(state_counts * np.log(volumes))
+    else:
+        v_like = - np.sum(state_counts * np.log(state_counts))
 
     if separate:
         return t_like, v_like

### 2013-03-15 muller/get_like.py

@@ -7,12 +7,12 @@ from scipy.io import mmread
 Ls = []
 Ls.append(0) # for a one state model the likelihood is 0
 
-for k in range(50,1050,50):
+for k in range(50,600,50):
     print k
-    t = mmread('KCenters/k%d/Lag30/tProb.mtx' % k)
-    v = np.load('KCenters/k%d/vols100k.npy' % k)
-    a = io.loadh('KCenters/k%d/Assignments.h5' % k)
+    t = mmread('Hybrid10/k%d/Lag30/tProb.mtx' % k)
+    v = np.load('Hybrid10/k%d/vols100k.npy' % k)
+    a = io.loadh('Hybrid10/k%d/Assignments.h5' % k)
     a = a['arr_0']
     Ls.append(likelihood.get_log_likelihood(t, a, v, lagtime=30))
 
-np.save('likes1000_s50_n100k.npy', np.array(Ls))
+np.save('hybrid10_600.npy', np.array(Ls))

### 2013-03-15 paper/postprocess_markdown.py
maybe png figures...

@@ -9,12 +9,18 @@ sys.stdout.write(msg)
 
 # match inline math or displayed math
 math_re = re.compile('[$]+([^$]*)[$]+')
-
+fig_re = re.compile('\((figs/\S*\.png)')
 base_url = 'http://latex.codecogs.com/gif.latex'
+
+github_url = "https://raw.github.com/rmcgibbo/opt-k/master/paper/"
 
 f = open(sys.argv[1])
 for line in f.readlines():
-    
+    m1 = fig_re.search(line)    
+    if m1:
+        print >> sys.stderr, 'inside m1'
+        line = fig_re.sub(github_url + m1.group(1), line)
+
     while True:
         m1 = math_re.search(line)
         if m1:

### 2013-03-15 paper/postprocess_markdown.py
maybe png figures...

@@ -9,7 +9,7 @@ sys.stdout.write(msg)
 
 # match inline math or displayed math
 math_re = re.compile('[$]+([^$]*)[$]+')
-fig_re = re.compile('\((figs/\S*\.png)')
+fig_re = re.compile('(figs/\S*\.png)')
 base_url = 'http://latex.codecogs.com/gif.latex'
 
 github_url = "https://raw.github.com/rmcgibbo/opt-k/master/paper/"

### 2013-03-15 paper/postprocess_markdown.py
maybe png figures...

@@ -10,6 +10,8 @@ sys.stdout.write(msg)
 # match inline math or displayed math
 math_re = re.compile('[$]+([^$]*)[$]+')
 fig_re = re.compile('(figs/\S*\.png)')
+label_re = re.compile('\\label{\S*}')
+
 base_url = 'http://latex.codecogs.com/gif.latex'
 
 github_url = "https://raw.github.com/rmcgibbo/opt-k/master/paper/"
@@ -20,6 +22,11 @@ for line in f.readlines():
     if m1:
         print >> sys.stderr, 'inside m1'
         line = fig_re.sub(github_url + m1.group(1), line)
+
+    m2 = label_re.search(line)
+    if m2:
+        line = label_re.sub('', line)
+        print >> sys.stderr, line
 
     while True:
         m1 = math_re.search(line)

### 2013-03-15 paper/postprocess_markdown.py
maybe png figures...

@@ -10,7 +10,7 @@ sys.stdout.write(msg)
 # match inline math or displayed math
 math_re = re.compile('[$]+([^$]*)[$]+')
 fig_re = re.compile('(figs/\S*\.png)')
-label_re = re.compile('\\label{\S*}')
+label_re = re.compile('[\]label{\S*}')
 
 base_url = 'http://latex.codecogs.com/gif.latex'
 

### 2013-03-15 paper/postprocess_markdown.py
maybe png figures...

@@ -10,7 +10,7 @@ sys.stdout.write(msg)
 # match inline math or displayed math
 math_re = re.compile('[$]+([^$]*)[$]+')
 fig_re = re.compile('(figs/\S*\.png)')
-label_re = re.compile('[\]label{\S*}')
+label_re = re.compile(r'\\label{\S*}')
 
 base_url = 'http://latex.codecogs.com/gif.latex'
 

### 2013-03-15 paper/postprocess_markdown.py
working better...

@@ -2,6 +2,7 @@ import sys
 import string
 import re
 import urllib
+import itertools
 
 msg = "This file was automatically generated from latex using pandoc\n\n"
 
@@ -9,33 +10,51 @@ sys.stdout.write(msg)
 
 # match inline math or displayed math
 math_re = re.compile('[$]+([^$]*)[$]+')
-fig_re = re.compile('(figs/\S*\.png)')
+fig_re = re.compile('\((figs/\S*\.png)')
 label_re = re.compile(r'\\label{\S*}')
+fig_location_re = re.compile('\[[ch!]+\]')
+wierd_re = re.compile('\[fig:\S+\]')
 
 base_url = 'http://latex.codecogs.com/gif.latex'
 
 github_url = "https://raw.github.com/rmcgibbo/opt-k/master/paper/"
 
 f = open(sys.argv[1])
-for line in f.readlines():
-    m1 = fig_re.search(line)    
+text = f.read()
+counter = itertools.count()
+while True:
+    print >> sys.stderr, counter.next()
+    matched = False
+
+    m0 = fig_location_re.search(text)
+    if m0:
+        matched = True
+        text, _ = fig_location_re.subn('', text)
+
+    m0 = wierd_re.search(text)
+    if m0:
+        matched = True
+        text, _ = wierd_re.subn('', text)
+
+    m1 = fig_re.search(text)
     if m1:
-        print >> sys.stderr, 'inside m1'
-        line = fig_re.sub(github_url + m1.group(1), line)
+        matched = True
+        text, _ = fig_re.subn('(' + github_url + m1.group(1), text, 1)
 
-    m2 = label_re.search(line)
+    m2 = label_re.search(text)
     if m2:
-        line = label_re.sub('', line)
-        print >> sys.stderr, line
-
-    while True:
-        m1 = math_re.search(line)
-        if m1:
-            # replace latex with a url that renders the math
-            query = urllib.quote(m1.group(1))
-            repl = '![equation](%s)' % (base_url + '?' + query)
-            line, _ =  math_re.subn(repl, line, 1)
-        else:
-            break
+        matched = True
+        text, _ = label_re.subn('', text)
     
-    sys.stdout.write(line)
+    m3 = math_re.search(text)
+    if m3:
+        matched = True
+        # replace latex with a url that renders the math
+        query = urllib.quote(m3.group(1))
+        repl = '![equation](%s)' % (base_url + '?' + query)
[…]

### 2013-03-15 paper/postprocess_markdown.py
working better...

@@ -51,7 +51,7 @@ while True:
         matched = True
         # replace latex with a url that renders the math
         query = urllib.quote(m3.group(1))
-        repl = '![equation](%s)' % (base_url + '?' + query)
+        repl = '![](%s)' % (base_url + '?' + query)
         text, _ =  math_re.subn(repl, text, 1)
 
     if not matched:

### 2013-03-15 paper/postprocess_markdown.py
figure captions

@@ -8,16 +8,16 @@ msg = "This file was automatically generated from latex using pandoc\n\n"
 
 sys.stdout.write(msg)
 
+base_url = 'http://latex.codecogs.com/gif.latex'
+github_url = "https://raw.github.com/rmcgibbo/opt-k/master/paper/"
+
 # match inline math or displayed math
 math_re = re.compile('[$]+([^$]*)[$]+')
 fig_re = re.compile('\((figs/\S*\.png)')
 label_re = re.compile(r'\\label{\S*}')
 fig_location_re = re.compile('\[[ch!]+\]')
 wierd_re = re.compile('\[fig:\S+\]')
-
-base_url = 'http://latex.codecogs.com/gif.latex'
-
-github_url = "https://raw.github.com/rmcgibbo/opt-k/master/paper/"
+caption_re = re.compile('!\[(.+)\]\(' + github_url)
 
 f = open(sys.argv[1])
 text = f.read()
@@ -25,6 +25,11 @@ counter = itertools.count()
 while True:
     print >> sys.stderr, counter.next()
     matched = False
+
+    m0 = caption_re.search(text)
+    if m0:
+        matched = True
+        text, _ = caption_re.subn('_' + m0.group(1) + '_\n' + '![](' + github_url, text)
 
     m0 = fig_location_re.search(text)
     if m0:

### 2013-03-15 paper/postprocess_markdown.py
added comments so i dont forget

@@ -29,25 +29,35 @@ while True:
     m0 = caption_re.search(text)
     if m0:
         matched = True
+        # caption re matches the first part of the ![caption](img_link)
+        # bit, and moves the caption out front, displaying it in italics
+        # (the underlines)
         text, _ = caption_re.subn('_' + m0.group(1) + '_\n' + '![](' + github_url, text)
 
     m0 = fig_location_re.search(text)
     if m0:
+        # just matches [h!] and removes it
         matched = True
         text, _ = fig_location_re.subn('', text)
 
     m0 = wierd_re.search(text)
     if m0:
+        # removes "[fig:<random_junk>]" that gets inseted by pandoc?
         matched = True
         text, _ = wierd_re.subn('', text)
 
     m1 = fig_re.search(text)
     if m1:
+        # remap the url in figures so that they point to the github
+        # raw url for that img. Right now we're only detecting the img
+        # links because they're in the fig/ directory and end in .png
         matched = True
         text, _ = fig_re.subn('(' + github_url + m1.group(1), text, 1)
 
     m2 = label_re.search(text)
     if m2:
+        # remove the "\label{eq:<whatever>}" statement from inside
+        # math equations
         matched = True
         text, _ = label_re.subn('', text)
     
@@ -55,6 +65,7 @@ while True:
     if m3:
         matched = True
         # replace latex with a url that renders the math
+        # looks for $$<stuff>$$ or $<stuff>$
         query = urllib.quote(m3.group(1))
         repl = '![](%s)' % (base_url + '?' + query)
         text, _ =  math_re.subn(repl, text, 1)

### 2013-03-15 muller/get_like.py
added Kmeans and Hybrid likelihoods. NOTE: I used the K-Centers generators from the 100 state model to define the 'whole volume' for the hybrid and kmeans cases. this means we can compare the clustering algorithms.

@@ -7,12 +7,13 @@ from scipy.io import mmread
 Ls = []
 Ls.append(0) # for a one state model the likelihood is 0
 
-for k in range(50,600,50):
+folder='KMeans'
+for k in range(50, 1050, 50):
     print k
-    t = mmread('Hybrid10/k%d/Lag30/tProb.mtx' % k)
-    v = np.load('Hybrid10/k%d/vols100k.npy' % k)
-    a = io.loadh('Hybrid10/k%d/Assignments.h5' % k)
+    t = mmread('%s/k%d/Lag30/tProb.mtx' % (folder, k))
+    v = np.load('%s/k%d/vols1M.npy' % (folder, k))
+    a = io.loadh('%s/k%d/Assignments.h5' % (folder, k))
     a = a['arr_0']
     Ls.append(likelihood.get_log_likelihood(t, a, v, lagtime=30))
 
-np.save('hybrid10_600.npy', np.array(Ls))
+np.save('%s.npy' % folder, np.array(Ls))

### 2013-03-15 paper/paper.tex
stuff
change
working better...
changed notation
notation shift
power
Added figures to the paper. I think I have too many in there so we'll have to decide what to keep.

@@
 \begin{equation}
+P[X_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(X_i \rightarrow X_{i+1}) \cdot \prod_{i=0}^{T} p(X_{i} | \sigma(X_{i}))
 \label{eq:like}
-P[x_{0...T-1}] dx^N = \prod_{i=0}^{T-1} T(x_i \rightarrow x_{i+1}) \cdot \prod_{i=0}^{T} p(x_{i} | \sigma(x_{i}))
 \end{equation}
@@
 \begin{figure}[h!]
-\includegraphics[width=3in]{figs/muller_pot_traj.pdf}
+\includegraphics[width=3in]{figs/muller_pot.png}
 \caption{The M\"{u}ller potential consists of three metastable wells. We simulated Langevin dynamics on this potential as a validation of the new likelihood function.}
@@
 % don't know how to say this yet:
-%We note that we cannot take this too seriously, and there are many other such criteria, such as the Aikaike Information Content (AIC). 
+%We note that we cannot take this too seriously, and there are many other such criteria, such as the Aikaike Information Content (AIC). %%%
 We found that the BIC of the models built using K-Centers clustering has a minimum at $k=250$.
@@
 
+\begin{figure*}
+\includegraphics[width=5in]{figs/kcent_vors.png}
+\caption{For K-Centers models, we observed a rapid increase in likelihood followed by a plateau.
+For models built with too few states, the states within the metastable wells are too large.
+By increasing the number of states, the resolution in increased, but using too many states leads to a model like the bottom right which is clearly over-fit and so has a non-optimal BIC despite having a higher likelihood.}
+\end{figure*}
+
+\begin{figure}
+\includegraphics[width=3in]{figs/cluster_comp.png}
+\caption{We also compared three clustering algorithms and used the likelihood function to determine which is best.
+The likelihood of the K-Means models was always larger than the other two.
+Additionally, the K-Means algorithm plateaued at a lower value of $k$, which corresponded to a minimum in the BIC curve at 200 states vs.
+the optimal models of 250 states in the K-Centers and Hybrid K-Medoids models.} \end{figure}
+
+\begin{figure}
+\includegraphics[width=2in]{figs/clust_vors.png}
+\caption{The likelihoods (and BIC's) of the three clustering algorithms predicted that the K-Means algorithm is the better clustering algorithm.
+The result of K-Means is many states clustered into the highest density areas of the potential.
+The Hybrid K-Medoids ends up somewhere in between the K-Centers model and the K-Means model.
+This is likely due to the implementation being an approximate algorithm that requires more iterations to converge to the best state decomposition.}
+\end{figure}
 \bibliography{bibliography}

### 2013-03-15 paper/postprocess_markdown.py
Commented out print counter in postprocess script. We should be able to know if it hangs because it will take more than a second or two to run.

@@ -23,7 +23,7 @@ f = open(sys.argv[1])
 text = f.read()
 counter = itertools.count()
 while True:
-    print >> sys.stderr, counter.next()
+    #print >> sys.stderr, counter.next()
     matched = False
 
     m0 = caption_re.search(text)

### 2013-03-22 lib/montecarlo.py
Added volume comparison from fip35

@@ -11,7 +11,7 @@ class MonteCarlo(VolumeEstimator):
     """
     
     def __init__(self, metric, generators, space_generators=None,
-         num_points=1E6, cushion=1.):
+         num_points=1E6, cushion=1., space_metric=None):
         """
         Compute the volume of states by using a simple monte carlo algorithm.
 
@@ -30,6 +30,9 @@ class MonteCarlo(VolumeEstimator):
         cushion : float, optional
             cushion to use to define full space of the data. A point is in
             this space if it is within <cushion> of any of the space_generators
+        space_metric : msmbuilder.metrics.Vectorized, optional
+            use this metric to compute distances to space_generators
+            if None, then use metric.
         """
         super(MonteCarlo, self).__init__(metric, generators)
 
@@ -37,7 +40,9 @@ class MonteCarlo(VolumeEstimator):
         self.cushion = float(cushion)
 
         self.space_generators = space_generators
-        self.prep_space_generators = self.metric.prepare_trajectory(self.space_generators)
+        if space_metric is None:
+            self.space_metric = self.metric
+        self.prep_space_generators = self.space_metric.prepare_trajectory(self.space_generators)
 
         self.mins = np.min(self.prep_space_generators, axis=0) - self.cushion
         self.maxs = np.max(self.prep_space_generators, axis=0) + self.cushion
@@ -70,7 +75,7 @@ class MonteCarlo(VolumeEstimator):
             random_sample = random_sample * self.lengths + self.mins
             random_sample = random_sample.reshape((1, -1))
             
-            space_dists = self.metric.one_to_all(random_sample, self.prep_space_generators, 0)
+            space_dists = self.space_metric.one_to_all(random_sample, self.prep_space_generators, 0)
 
             if np.min(space_dists) > self.cushion:
                 continue

### 2013-03-22 scripts/get_vols.py

@@ -40,6 +40,7 @@ parser.add_argument('output', help='output filename to save volumes.')
 parser.add_argument('num_points', help='Number of points to sample.')
 parser.add_argument('use_dih', action='store_true', default=False)
 parser.add_argument('use_euc', action='store_true', default=False)
+#parser.add_argument('space_metric', default=None, help='metric to use to define which points are in the whole space. None or pickled distance metric')
 
 args, metric = parser.parse_args()
 
@@ -57,9 +58,15 @@ try:
     space_gens = np.load(args.space_gens)
 except:
     space_gens = Trajectory.load_from_lhdf(args.space_gens)
+#
+#if not args.space_metric is None:
+#    space_metric = pickle.unpickle(args.space_metric)
+#else:
+#    space_metric = None
 
 MC = MonteCarlo(metric, gens, space_gens, num_points=args.num_points, 
         cushion=args.cushion)
+#, space_metric=space_metric)
 
 volumes = MC.get_state_volumes()
 

### 2013-03-22 paper/paper.tex
Change like_comp.pdf to like_comp.png and cropped the whitespace
updated muller_pot figure to png and the new colormap

@@
 \author{Vijay S. Pande}
-\maketitle
 
+\begin{abstract}
 Likelihood framework for choosing the optimal number of states in a
 Markov State Model
+\end{abstract}
+
+\maketitle
 
@@
 
-A significant challenge in the automated construction of markov state
-models is choosing the number of microstates. The number of microstates,
-$k$ governs a bias-variance tradeoff -- when $k$ is large, the models
-have a lot of parameters enabling it to fit the data with low bias, but
-the simultaneous estimation of these parameters from finite data can
-lead to high variance. On the other hand, when $k$ is small, we have
-fewer parameters and better statistics, but at the expense of model
-flexibility.
+A significant challenge in the automated construction of Markov state
+models is the choice of the number of microstates.
+Although classical Hamiltonian dynamics form a continuous-time Markov chain in $\mathbb{R}^{6N}$, the Markov property does not hold after the projecting the dynamics onto a basis of discrete indication functions -- the MSM microstates.
+In particular, when microstates contain within them free energy barriers of substantial magnitude, the validity of the Markov assumption begins to suffer considerably.
+While this source of modeling error can be addressed by increasing the number of microstates, the reduction in one error comes at the expense of the increase in another.
+This second source of error is statistical in origin.
+As the number of states in the model grows, so does the number of parameters required to specify the transition matrix.
+Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per
+parameter, which makes the model liable for over-fitting.
 
-Currently, most MSM construction pipelines involve manually selecting
-the number of states. This is done primarily by human intuition, and is
-a significant bottleneck.
+Currently, most MSM construction pipelines involve manually selecting the number of microstates.
+This is done primarily by human intuition, guided by the Chapman–Kolmogorov test, and is a significant bottleneck.
 
-Here, we introduce a likelihood framework enabling the automatic
-selection of the number of states.
+Here, we introduce a new procedure for scoring the likelihood of an MSM, which, together with the Bayesian information criterion (BIC), enables the optimal selection of both the number of states and clustering algorithm.
 
 Within the discrete state space, the likelihood of an ensemble of
-trajectories given the MSM is straightforward. We simply take the
-product of the state to state transition probabilities along the path.
-These state to state transition probabilities are our main central
-parameters, the entries of the transition matrix. However, as we vary
-the number of states, it is not permissible to simply compare these
-likelihoods to select an optimal number of states. Doing so, we would
-always chose the trivial model, with only one state, as the transition
-probability in that model would be $1$ and thus the likelihood of any
-trajectory within that state space would be $1$.
+trajectories given the MSM is straightforward. It is computed simply by taking the product of the state to state transition probabilities along the path.
+These state-to-state transition probabilities are entries in the MSM's so-called transition matrix, and are the central parameters of the model.
+However, as we vary the number of states, it is not permissible to simply compare these likelihoods to select an optimal number of states.
+Doing so, the optimal model would always be the trivial model, with only one state, as the transition probability in that model would be $1$ and thus the likelihood of any trajectory within that state space would be $1$.
 
 The proper likelihood is not of the trajectory within the discrete state
-space; instead it's the likelihood of the trajectory within the
+space; instead it is the likelihood of the trajectory within the
 continuous phase space, on which the discrete states are merely an
@@
 conformation at a given location in phase space given that the
-conformation ($x_t$) is within a certain state ($\sigma(x_t)$).
+conformation, $x_t$, is within a certain state, $\sigma(x_t)$.
 
@@
 been a model parameter for MSMs. Nevertheless, they are a critical
-parameter from the point of view of evaluating trajectory likelihoods.
+parameter for evaluating trajectory likelihoods.
 For example, consider two MSMs with the same transition probabilities.
-However, in one case, the emission distributions for each state are
-highly peaked at specific locations in phase space, whereas in the other
-model the emission distributions are uniform over the volume of the
-clusters. If the trajectory actually does go through the regions of high
-likelihood in the first model, we would say that the first model more
-accurately fits the data.
+However, in one case, the emission distributions for each state are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the clusters.
+If the trajectory actually does go through the first models' regions of high likelihood, it would be clear that the first model more accurately fits the data.
 
@@
 interest scientifically, is independent of the choice of the emission
-distributions. It's only determined by the eigenspectrum of the
-transition matrix elements. (The emission distributions determine the
-estimated eigenfunctions, but not the eigenvectors.)
+distributions. It is only determined by the eigenspectrum of the
+transition matrix elements. The emission distributions determine the
+estimated eigenfunctions, but not the eigenvectors.
 
@@
 if the conformation is outside of the bounding volume of the state and
-constant if the conformation is within the volume. The constant is set
+constant if the conformation is within the volume. This constant is set
 so that the distribution integrates to $1$, and is thus the reciprocal
@@
 \label{eq:like_vol}
-P[x_{0...T-1}] dx^N = \prod_{i=0}^{T-1} T(x_i \rightarrow x_{i+1}) \cdot \prod_{i=0}^T \frac{1}{V_{\sigma(x_{i})}}
+P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(x_i \rightarrow x_{i+1}) \cdot \prod_{i=0}^T \frac{1}{V_{\sigma(x_{i})}}
 \end{equation}
@@
 attention in recent years in the computational geometry literature, especially
-via randomized algorithms. [See issue \#1 on the github]
+via randomized algorithms.\cite{Simonovits03, Lovasz03}
 
-One challenge is how to model the volume of states which are at the
-``edge'' of the model. Is the volume of these states unbounded, given that they extend all the way out to infinity in some direction? This would be problematic.
-It seems appropriate to assert that the volume of these edge states is bounded in some way by the extent of our dataset.
-For example, the volume of a state might be defined as the volume of the intersection of its Voronoi cell and the convex hull of the whole dataset.
+A further challenge is the procedure by which to model the volume of states which are at the ``edge'' of the MSM, whose Voronoi cells extend to infinity in some direction.
+Is the volume of these states unbounded?
+It is appropriate to assert that the volume of these edge states is bounded in some way by the extent of our dataset.
+For example, the volume of a state might be defined as the volume of the intersection of its Voronoi cell and the convex hull of the whole dataset, which would encode the assumption that the likelihood of points
+outside the convex hull of the sampled data is vanishing.
 
@@
 the same spirit.
-Instead of taking the outer bounding region to be the convex hull of the data, we take it to be the set of all trial points such that the nearest data point to the trial point is closer than a certain cutoff, $R$.
-This can be computed relatively efficiently using a BallTree data structure.
-For further efficiency, we might use only a random subsample of the dataset for this nearest neighbor computation.
+Instead of taking the outer bounding envelope to be the convex hull of the data, we take it to be the set of all trial points such that the nearest sampled configuration to the trial point is closer than a certain cutoff, $R$.
+This can be computed relatively efficiently using a Balltree data structure.\cite{Omohundro1989} For further efficiency, it is feasible to use only a random subsample of the dataset for this nearest neighbor computation.
 
 Instead of computing the volume of the Voronoi cells explicitly, we instead
-compute the ratio of volume of each Voronoi cell to the volume of the entire
-bounding region.
-Because the bounding region's volume is independent of the clustering parameters, its inclusion changes all of the calculated likelihoods by a constant multiplicative factor, and can thus be discarded from the perspective of model comparison.
+compute the ratio of volume of each Voronoi cell to the volume of the
+bounding envelope.
+Because the bounding envelope's volume is independent of the clustering parameters, its inclusion changes all of the calculated likelihoods by a constant multiplicative factor, and can thus be discarded from the perspective of model comparison.
 
-To compute the fractional volumes of the Voronoi cells, we use the following randomized algorithm.
-First, we generate points inside of the bounding region using the lazy random walk Markov chain Monte Carlo algorithm.\cite{Kannan97} This random walk converges to sampling from the uniform distribution over the interior.
+To compute the fractional volumes of the Voronoi cells, we utilized a randomized algorithm, as exact algorithms have an impermissible exponential time complexity with respect to the dimensionality.
+First, we generate points inside of the bounding region using a lazy random walk Markov chain Monte Carlo algorithm.\cite{Kannan97} This random walk converges to sampling from the uniform distribution over the interior.
 After a sufficient burn in period, for each sampled point we compute its nearest MSM state center, assigning it to that state.
@@
 
-This algorithm is highly amenable to parallel computation. For a euclidean
-distance metric, the assignment can be performed efficiently by using a BallTree 
-data structure for fast neighbor search.
-
-\section{Choosing the Optimal Number of States}
-
-We choose both the clustering algorithm (k-centers, k-means, etc) and the
-number of states by maximizing the BIC/AIC scores of the model, using this
-likelihood.
-
-Unfortunately, this doesn't really help with picking the projection operator
-to vectorize the conformations (e.g. dihedrals, contact maps, etc). Also,
-it's not going to work rigorously with RMSD.
+This algorithm is highly amenable to parallel computation.
 
@@
 \subsection{Langevin Dynamics on the M\"uller Potential}
-We being by using the procedure described above on data generated by simulating Langevin dynamics on the M\"{u}ller potential (\cref{fig:muller_pot}).
+We being by using the procedure described above on simulated Langevin dynamics on the M\"{u}ller potential (\cref{fig:muller_pot}).
 This data was clustered into models of different sizes using one of two algorithms.
@@
 In both cases, we built models ranging from 50 to 1,000 total states.
-Since this space is only two-dimensional, we used the simple, brute-force MC algorithm for calculating volumes.
+Since this space is only two-dimensional, we used simple rejection sampling to compute the volumes.
 
@@
 Then the volume of a state is proportional to the number of samples assigned to that state divided by the total number of samples.
-We note, that these volumes are actually in units of the total acceptable volume.
+%We note, that these volumes are actually in units of the total acceptable volume.
 This affects the absolute value of the likelihood, however, since the total volume was constant in all cases, the relative likelihoods are the same.
 \begin{figure}[h!]
-\includegraphics[width=3in]{figs/muller_pot.png}
+\includegraphics[width=2in]{figs/muller_pot.png}
 \caption{The M\"{u}ller potential consists of three metastable wells. We simulated Langevin dynamics on this potential as a validation of the new likelihood function.}
@@
 These likelihoods increased rapidly with $k$, but then plateaued at a value of around 300 states \cref{fig:like_kcenters}.
-At this plateau, we can conclude that we do not gain any new information by adding new states to our model and so should pick $k$ to be this point.
-This is a heuristic argument that can be made more mathematically rigorous with the introduction of functions like the Bayesian Information Content (BIC), which is defined in \cref{eq:bic}.
+At this plateau, we can conclude that we do not gain any new information by adding new states to our model, to compensate for the additional parameters.
+This is a heuristic argument that can be made more rigorous with the introduction of model comparison metrics like the Bayesian information content (BIC), which is defined in \cref{eq:bic}.
 
@@
 \end{equation} Here, $L$ denotes the likelihood of a model, while $m$ is the number of parameters used in the model, and $n$ is the number of data points.
-Essentially this is a mathematical way of saying we would like to use a model with the maximum likelihood, but without introducing too many parameters.
-% don't know how to say this yet:
-%We note that we cannot take this too seriously, and there are many other such criteria, such as the Aikaike Information Content (AIC). %%%
+While somewhat heuristic, the BIC can be viewed as an approximation to the Bayes Factor, whose direct computation involves the infeasible task of integrating over the posterior distribution of Markov models.
+
 We found that the BIC of the models built using K-Centers clustering has a minimum at $k=250$.
-This minimum is the ``optimal'' model according to the BIC of our likelihood.
-As a check of the usefulness of this minimum, we plotted the three slowest implied timescales in the models as a function of $k$.
+This minimum is the ``optimal'' model according to the BIC.
+As an orthogonal check on the selection of this minimum, we plotted the three slowest implied timescales in the models as a function of $k$.
+Because the model's implied timescales are estimators of the eigenvalues of the underlying potential's transfer operator, their invariance with respect to the number of states is a standard check on model quality.
 The optimal model was close to the point at which the implied timescales became constant.
@@
 \begin{figure}
-\includegraphics[width=2in]{figs/like_comp.pdf}
+\includegraphics[width=2in]{figs/like_comp.png}
 \caption{For models built between 50 and 1,000 states using the K-Centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states.
@@
 \begin{figure}
-\includegraphics[width=3in]{figs/cluster_comp.png}
+\centering
+\includegraphics[width=3.4in]{figs/cluster_comp.png}
 \caption{We also compared three clustering algorithms and used the likelihood function to determine which is best.
@@
 \begin{figure}
+\centering
 \includegraphics[width=2in]{figs/clust_vors.png}

### 2013-04-17 fip35/get_like.py
added likelihoods as well as another notebook describing the penalizing coefficient in BIC and AIC.

@@ -1,0 +1,45 @@
+import likelihood
+import numpy as np
+from msmbuilder import io
+from scipy.io import mmread
+import os
+import argparse
+
+parser = argparse.ArgumentParser()
+parser.add_argument('-d', dest='msm_dirs', help='MSM directory containing, tProb.mtx, Mapping.dat, and Assignments.Fixed.h5', action='append')
+parser.add_argument('-v', dest='vols', help='Volumes to use for each MSM.', action='append')
+parser.add_argument('-o', dest='out_fn', default='likelihoods.npy', help='output filename (np.savetxt)')
+parser.add_argument('-l', type=int, dest='lagtime', help='lag time (in frames). Right now it only makes sense to compare MSMs built at the same lag time. Though I suppose it is possible to change this in the future.')
+args = parser.parse_args()
+
+if len(args.msm_dirs) != len(args.vols):
+    print "Need to input as many data directories as volume files."
+
+likelihoods = []
+states = []
+likelihoods.append(0) # for a one state model the likelihood is 0
+states.append(1)
+
+for i in xrange(len(args.msm_dirs)):
+
+    msm_folder = args.msm_dirs[i]
+    
+    print "Calculating likelihood for MSM in %s" % msm_folder
+
+    t = mmread(os.path.join(msm_folder, 'tProb.mtx'))
+    m = np.loadtxt(os.path.join(msm_folder, 'Mapping.dat'))
+    v = np.load(args.vols[i])
+    a = io.loadh(os.path.join(msm_folder, 'Assignments.Fixed.h5'))
+    try:
+        a = a['arr_0']
+    except:
+        a = a['Data']
+
+    v = v[np.where(m != -1)] # remove trimmed states
+    likelihoods.append(likelihood.get_log_likelihood(t, a, v, lagtime=args.lagtime, mapping=m))
+    states.append(np.where(m != -1)[0].shape[0])
+
+out_ary = np.array([states, likelihoods]).T
+
+np.save(args.out_fn, out_ary)
+

### 2013-04-17 paper/paper.tex
added another figure for WW which emphasizes that our method picks models in the 100s of states rather than 1000s. Furthermore, I am attempting to show that this is ok by monitoring the convergence of the RMSD ACF's but cannot come up with a reliable way of calculating a difference between each ACF. Maybe I can instead just look at the slowest implied timescales and make a plot like I did for the muller case.

@@
 \end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=2in]{figs/ww_acf_vs_aic.png}
+\caption{In the WW domain simulations, we projected the coordinates into reduced spaces from two to six dimensions, using tICA \cite{Schwantes:2013..}.
+The AIC optimal models were in the 200 to 400 state range.
+We compared the AIC with the autocorrelation convergence, and found that the AIC optimal model coincided generally with the point at which the autocorrelation functions converged.
+The BIC, penalizes the models more, and so gave rise to BIC optimal models between 100 and 300 states.}
+\end{figure}
+
 \bibliography{bibliography}

### 2013-06-13 paper/paper.tex
added some work on the analytic approximation

--- 2013-03-22 paper/paper.tex
@@
 In particular, when microstates contain within them free energy barriers of substantial magnitude, the validity of the Markov assumption begins to suffer considerably.
-While this source of modeling error can be addressed by increasing the number of microstates, the reduction in one error comes at the expense of the increase in another.
-This second source of error is statistical in origin.
+While this source of modeling error can be addressed by increasing the number of microstates, the reduction in one error comes at the expense of the increase in statistical error.
 As the number of states in the model grows, so does the number of parameters required to specify the transition matrix.
@@
 space; instead it is the likelihood of the trajectory within the
-continuous phase space, on which the discrete states are merely an
+continuous phase space on which the discrete states are merely an
 indicator function basis.
@@
 For example, consider two MSMs with the same transition probabilities.
-However, in one case, the emission distributions for each state are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the clusters.
+In one case, the emission distributions for each state are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the clusters.
 If the trajectory actually does go through the first models' regions of high likelihood, it would be clear that the first model more accurately fits the data.
@@
 However, the long timescale behavior of the MSM, the quantity of most
-interest scientifically, is independent of the choice of the emission
-distributions. It is only determined by the eigenspectrum of the
-transition matrix elements. The emission distributions determine the
-estimated eigenfunctions, but not the eigenvectors.
+interest scientifically, is largely independent of the choice of the
+emission distributions. %It is determined by the eigenspectrum of the
+%transition matrix elements. The emission distributions determine the
+%estimated eigenfunctions, but not the eigenvectors.
 
-Therefore, the most appropriate emission distribution for discrete state
-MSMs is that of the uniform distribution over the phase-space volume of
-the state. That is, the likelihood of observing a conformation in phase
-space given that the conformation is assigned to state $i$ is $0$
-if the conformation is outside of the bounding volume of the state and
-constant if the conformation is within the volume. This constant is set
-so that the distribution integrates to $1$, and is thus the reciprocal
-volume of the microstate.
+For this study, we consider the emission distribution of each state to be the 
+uniform distribution over the phase-space volume of the state. That is, the 
+likelihood of observing a conformation in phase space given that the 
+conformation is assigned to state $i$ is $0$ if the conformation is outside of 
+the bounding volume of the state and constant if the conformation is within 
+the volume. This constant is set so that the distribution integrates to $1$, 
+and is thus the reciprocal volume of the microstate.
 
@@
 Briefly, we generated uniform random vectors within a bounding box whose size was slightly larger than the data.
-This random sample was then accepted if it was 'close enough' to a generator in the 100 state model, where close enough meant it was within 0.2 units from any generator in that model.
+This random sample was then accepted if it was ``close enough'' to a generator in the 100 state model, where close enough meant it was within 0.2 units from any generator in that model.
 This was used to approximate the convex hull of the data.
@@
 \begin{figure}[h!]
-\includegraphics[width=2in]{figs/muller_pot.png}
+\includegraphics[width=2.5in]{figs/muller_pot.png}
 \caption{The M\"{u}ller potential consists of three metastable wells. We simulated Langevin dynamics on this potential as a validation of the new likelihood function.}

### 2013-06-13 analytical_approximation/analytic.tex
added some analytic work in progress

\documentclass{article}
\usepackage{amsmath, url, geometry}
\begin{document}

\section*{idea}    
If we have a point, $(x_1, \ldots, x_n)$, distributed within the $n$-ball of radius $R$, and we measure its ``radius'', $r = \sqrt{\sum_i^n x_i^2}$,
what is $P(R|r)$? If we know $P(R|r)$ and we measure $r$ for a lot of points,
can we come up with a good estimate for $R$? Intuitively, it seems like as $n\rightarrow\infty$, $P(R|r)$ should approach $\delta(R-r)$, since all of the
volume of the $n$-ball will be basically at the surface.

\section*{work}

According to \url{http://mathworld.wolfram.com/BallPointPicking.html}, a point
chosen from the uniform distribution over the unit $N$-ball is distributed like

$$
\frac{(X_1, \ldots X_n)}{\sqrt{Y + \sum_i^N X_i^2}}
$$

Where $X_i$ are drawn independently from a standard normal, and $Y$ is drawn independently from an exponential distribution with $\lambda=1$.

Thus, the radius of the point is distributed like

$$
R \sim \sqrt{\frac{\sum_i^N X_i^2}{Y  + \sum_i^N X_i^2}}
$$

or

$$
R \sim \frac{1}{\sqrt{\frac{Y}{\sum_i^N X_i^2} + 1}}
$$

This is basically an exponential distribution over a $\chi^2$ distribution.

Let's call the quotient part $Q$. That is, $Q \sim \frac{Y}{\sum_i^N X_i^2}$. What is the distribution of $Q$? According to \url{dx.doi.org/10.1214/aoms/1177731679}, it should be

$$
p_Q(q) = \int_{-\infty}^\infty |y| p_{exp}(qy)p_{\chi^2}(y)dy
$$

$$
p_Q(q) = \frac{\lambda}{2^{N/2} \Gamma(\frac{N}{2})} \int_{-\infty}^\infty |y| e^{-\lambda qy} y^{N/2-1}e^{-y/2}dy
$$

$$
p_Q(q) = \frac{\lambda}{2^{N/2} \Gamma(\frac{N}{2})} \int_{-\infty}^\infty |y|  y^{N/2-1}e^{-(\lambda q + 1/2) y}dy
$$

With some help from Wolfram alpha, I think that we can do this integral. 
Wolfram gives the identity
$$
\int x^n e^{-cx} \; dx = -\frac{x^n (cx)^{-n} \; \Gamma(n + 1, cx)}{c} + \text{constant}
$$
Where $\Gamma(a, x)$ is the incomplete gamma function.

I don't think we need the part of the integral below zero. Because the exponential random variable $p_\text{exp}(qy)$ is only nonzero for $qy > 0$. I think this implies that $y>0$, since obviously $q$ is going to be greater than zero too.

The indefinite integral given implies
$$
\int_{0}^{\infty} x^n e^{-cx} \; dx = c^{-n-1} \Gamma(n+1)
$$

Which would give

$$
p_Q(q) = \frac{\lambda}{2^{N/2} \Gamma(\frac{N}{2})} \left(\lambda q + \frac{1}{2}\right)^{-\frac{N}{2} - 1} \Gamma(N/2+1)
$$

$$
p_Q(q) = \frac{N\lambda}{2^{\frac{N}{2} + 1}} \left(\lambda q + \frac{1}{2}\right)^{-\frac{N}{2} - 1} 
$$

\end{document}

### 2013-06-13 analytical_approximation/analytic.tex
more

@@
 
+Since $R(q)$ is invertible, we get (?)
+
+$$
+p_R(r) = \frac{N\lambda}{2^{\frac{N}{2} + 1}} \left(\frac{\lambda}{r^2} -\lambda + \frac{1}{2} \right)^{-\frac{N}{2} - 1} 
+$$
+
+$\lambda$ was really assumed to be one from the beginning (it was the unit-ball), so
+
+$$
+p(r) = \frac{N}{2^{\frac{N}{2} + 1}} \left(\frac{1}{r^2} - \frac{1}{2} \right)^{-\frac{N}{2} - 1}
+$$
+
+Next step: What if it's \emph{not} a unit-ball?
+Then we'll get $p(r)$ as a function of the ball's radius.
+We dealt with the $\lambda$ separately, but there is also a term from the $\chi^2$ distributions (which should be $\lambda^2$ that I didn't carry through).
+
 \end{document}

### 2013-06-14 analytical_approximation/analytic.tex
stuff'

--- 2013-06-13 analytical_approximation/analytic.tex
@@
 \documentclass{article}
-\usepackage{amsmath, url, geometry}
+\usepackage{amsmath, url, geometry, graphicx}
 \begin{document}
@@
 $$
-p_Q(q) = \frac{\lambda}{2^{N/2} \Gamma(\frac{N}{2})} \int_{-\infty}^\infty |y| e^{-\lambda qy} y^{N/2-1}e^{-y/2}dy
+p_Q(q) = \frac{1}{2^{N/2} \Gamma(\frac{N}{2})} \int_{-\infty}^\infty |y| e^{-qy} y^{N/2-1}e^{-y/2}dy
 $$
@@
 $$
-p_Q(q) = \frac{\lambda}{2^{N/2} \Gamma(\frac{N}{2})} \int_{-\infty}^\infty |y|  y^{N/2-1}e^{-(\lambda q + 1/2) y}dy
+p_Q(q) = \frac{1}{2^{N/2} \Gamma(\frac{N}{2})} \int_{-\infty}^\infty |y|  y^{N/2-1}e^{-(q + 1/2) y}dy
 $$
@@
 $$
-p_Q(q) = \frac{\lambda}{2^{N/2} \Gamma(\frac{N}{2})} \left(\lambda q + \frac{1}{2}\right)^{-\frac{N}{2} - 1} \Gamma(N/2+1)
+p_Q(q) = \frac{1}{2^{N/2} \Gamma(\frac{N}{2})} \left(q + \frac{1}{2}\right)^{-\frac{N}{2} - 1} \Gamma(N/2+1)
 $$
@@
 $$
-p_Q(q) = \frac{N\lambda}{2^{\frac{N}{2} + 1}} \left(\lambda q + \frac{1}{2}\right)^{-\frac{N}{2} - 1} 
+p_Q(q) = \frac{N}{2^{\frac{N}{2} + 1}} \left(q + \frac{1}{2}\right)^{-\frac{N}{2} - 1} 
 $$
 
+Since $R(q)$ is invertible, we get (?)
+
+$$
+p(r) = \frac{N}{2^{\frac{N}{2} + 1}} \left(\frac{1}{r^2} - \frac{1}{2} \right)^{-\frac{N}{2} - 1}
+$$
+
+Next step: What if it's \emph{not} a unit-ball? If the ball of of radius $R$, then that just sets the unit system, so  we have something like: (need a new normalization condition?)
+
+$$
+p(\frac{r}{R}) \propto \frac{N}{2^{\frac{N}{2} + 1}} \left(\frac{R^2}{r^2} - \frac{1}{2} \right)^{-\frac{N}{2} - 1}
+$$
+
+So how do we go from here to the quantity we want, which is $p(r|R)$?
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=3in]{p_of_r.png}
+\caption{The $p(r)$ expression looks reasonable. In higher dimensions, the distribution is more peaked towards the surface at $r \approx 1$. But I think the normalization is off}
+\end{figure}
+
+We're going to have a lot of observations of $r$. If we assume that they're independent, then our log likelihood is
+
+$$
+\mathcal{L} = -\left( \frac{N}{2} - 1\right) \sum_i \log \left( \frac{R^2}{r_i^2} - \frac{1}{2} \right)
+$$
+
+Taking the derivative of the log likelihood with respect to $R$, we have
+$$
+\frac{\partial \mathcal{L}}{\partial R} = - \sum_i \frac{4R}{r_i^2 - 2R^2} = 0
+$$
+
+How do I solve this for $R$?
+
 \end{document}

### 2013-06-14 analytical_approximation/p_of_r.py

@@ -1,0 +1,16 @@
+from __future__ import division
+import numpy as np
+import matplotlib.pyplot as pp
+
+r = np.linspace(0,1)
+for N in range(3,10,2):
+    p = N / (2**(N/2 + 1)) * (1/r**2 - 1/2)**(-N/2 - 1)
+    np.log(1/r**2 - 0.5)
+    pp.plot(r,p, lw=2, label='N=%d' % N)
+
+pp.ylabel('p(r)')
+pp.xlabel('r')
+pp.legend(loc=2)
+#pp.show()
+pp.savefig('p_of_r.png')
+

### 2013-07-05 paper/paper.tex
added some intro notes

--- 2013-04-17 paper/paper.tex
@@
 \maketitle
+
+\section{Introduction}
+\begin{itemize}
+\item Many processes in proteins are fundamentally dynamical in nature. Folding, aggregation, cryptic binding sites. Governed by a delicate entropic and enthalpic balance.
+\item Set up a conflict between the detailed picture given by MD simulation and the simple picture seen by experiments.
+Multidimensional time series vs.
+phenomenological two-state behavior.
+But sometimes experiments see more complex dynamics too -- interpreted as alternate pathways, kinetic intermediates.
+\item Minimially-biased kenetic analysis of MD data has been a challenge for the field.
+Some procedures, in making strong assumptions about the form of the simplified dynamics -- e.g.
+rxn coord, G\={o} -- lack the capability to describe the complex systems in which those assumptions break down.
+\item Here, we seek to build models that are \emph{suitably} complex, given the data. Markov state models are able to capture both two-state and more complex multi-state kinetics.
+\item Like all statistical methods, Markov state models are chacterized by a bias-variance tradeoff.
+Picking the number of states in the MSM is a key tradeoff parameter, with statistical error balanced against modelling bias.
+\end{itemize}
+
+\subsection{Comparison to previous work}
+After we've introduced the idea:
+\begin{itemize}
+\item Contrast with lumping. Instead of going overboard on making too many states and then trying to correct it by merging some of them, we're going to try to pick the right number in the first place.
+\item Given the frame that lumping is an attempt to correct the statistical error comping from too many states (bias-variance), it makes sense that spectral methods like PCCA/PCCA+ fail: they're construction assumes that the transition matrix (its eigenspectrum) is known precisely.
+But the conditions where that is most likely to break down, when there are too many states  -- that's precisely the time that you're trying to apply these methods.
+So PCCA/PCCA+ is unsuitable here.
+\item BACE on the other hand, by explicitly modeling the statistical uncertainty when merging states, does not suffer from this drawback. (say something nice, get greg as reviewer?)
+\end{itemize}
 

### 2013-08-09 paper/paper2.tex
Wrote 1.5 pages of new paper. Crank this baby out.

\documentclass[twocolumn,floatfix,nofootinbib,aps]{revtex4-1}
\usepackage[utf8]{inputenc}

\usepackage{amsmath}    % need for subequations
\usepackage{amssymb}    % for symbols
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
%\usepackage{hyperref}   % use for hypertext links, including those to external documents and URLs
\usepackage[capitalise]{cleveref}   % use for referencing figures/equations
\begin{document}

\title{Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics}
\author{Robert T. McGibbon}
\author{Christian R. Schwantes}
\author{Vijay S. Pande}

\begin{abstract}
Likelihood framework for choosing the optimal number of states in a
Markov State Model
\end{abstract}

\maketitle

\section{Introduction}
Protein dynamics are an essential aspect of biomolecular function. These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, the conformational change between functional native substates, ligand binding, and allostery. Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex kinetics occurring via various pathways and with multiple intermediates. But atomic-resolution structural characterization of these dynamics is often a Herculean challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structural dynamics, generally reporting simultaneously on many degrees of freedom.

Computer simulations can complement experiments by providing atomic-resolution insight into the structural dynamics. With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation data sets. Nonetheless, minimally-biased kinetic analysis of such simulations present a major challenge to the field. The analysis employed often entails the construction of lower resolution models built from the high resolution simulation data set that captures its essential features in an interpretable framework. For example, by projecting the data down onto one or two degrees of freedom we create a simpler model for the system, such as one characterized by diffusion along a single reaction coordinate.

Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high-resolution models that maintain quantitative agreement with the underlying simulation data and low-resolution models which capture the salient features of the underlying potential energy landscape while sacrificing some degree of model complexity. In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states. The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which determine the kinetics.

A significant challenge in the automated construction of Markov state
models is the choice of the number of states. Although classical Hamiltonian dynamics form a continuous-time Markov chain in $\mathbb{R}^{6N}$, the Markov property does not hold after the projecting the dynamics onto a basis of discrete indicator functions. In particular, when states contain within them free energy barriers of substantial magnitude, the validity of the Markov assumption begins to suffer considerably. While this source of modeling error can be addressed by increasing the number of microstates, the reduction in one error comes at the expense of the increase in another. This second source of error is statistical in origin. As the number of states in the model grows, so does the number of parameters required to completely specify the kinetic model between all pairs of states. Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per model parameter, which makes the approach liable for over-fitting.

Here, we seek to build models that are \emph{suitably} complex, given the data, yielding complex descriptions of the system only to the extent that their additional parameters are implied by the observed dynamics. To that end, we introduce a new procedure for scoring the likelihood of an MSM, which, together with the Bayesian information criterion (BIC), enables the optimal selection of the state space, which we express both in terms of the number of states and the clustering algorithm employed to group sampled conformations into states. This approach complements validation procedures performed primarily based on human intuition, such as Chapman-Kolmogorov tests, and enables us to treat model selection as an optimization problem amenable to automated methods.

\section{Likelihood of a Markov State Model}
With the kinetic model expressed as a set of pairwise state to state transition probabilities at a given lag time, the likelihood of an ensemble of trajectories after projection into the indicator function basis is given simply by the product of the transition matrix elements along the observed trajectories. However, as we vary the number of states, it is not permissble to simply compare these likelihoods as part of an optimization of the state definitions. In doing so, the optimal model would always be the trivial one state model, whose computed likelihood is unity regardless of the data.

The appropriate likelihood is a path action in (continuous) phase space, on which the discrete states are merely an indicator function basis.

\begin{align}
P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(X_i) \rightarrow s(x_{i+1})) \cdot \prod_{i=0}^{T} p_{s(x_i)}(X_{i})
\label{eq:like}
\end{align}

With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state transition probabilities, $T(s_i \rightarrow s_j)$, and so-called emission distributions of each state, the conditional probablity of observing a conformation at a given location in phase space given that the conformation, $x_t$ is within a certain state, $s(x_t)$.

For example, consider two Markov state models sharing the same transition matrix, $T$. In one model, the state emission distributions are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the states. If an observed trajectory does go through the first models’ regions of high likelihood, it is appropatiely termed a more likely model given the data.

However, models' long timescale behavior -- the rates and fluxes between metastable basins are independent of the choice of the emission distributions. The emmission distributions characterize only the fine details of the equilibrium distribution within states, a quantity that the MSM approach does not seek to model; implicit in the decision to group conformations together into states is the idea that we decline to model the differences between conformations belonging to the same state. If such differences exist and are sufficiently large to warrant attention, than the states are too large.

Therefore, the most appropriate emission distribution for discrete state MSMs is that of the uniform distribution over the phase-space volume of the state. That is, the likelihood of observing a conformation in phase space given that the conformation is assigned to state i is 0 if the conformation is outside of the bounding volume of the state and constant if the conformation is within the volume. This constant is set so that the distribution integrates to 1, and is thus the reciprocal volume of themicrostate.

\begin{equation}
\label{eq:like_vol}
P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \cdot \prod_{i=0}^T \frac{1}{V_{s(x_{i})}}
\end{equation}

\section{Model Selection With the Bayes Information Criterion}

Likelihood maximization is insifficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to overfitting.

Full bayesian treatment. Intractable for problems of this size, but right in theory. Cross validation: can effective in practice but tricky with timeseries -- hard in data-poor regime. BIC: very useful. ``free'' computationally (no extra burden). Very nice heuristic.

\section{Methods}

\section{Results and Discussion}

\end{document}

### 2013-08-11 paper/paper2.tex
A little more work on paper2

@@
 With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation data sets.
-Nonetheless, minimally-biased kinetic analysis of such simulations present a major challenge to the field.
-The analysis employed often entails the construction of lower resolution models built from the high resolution simulation data set that captures its essential features in an interpretable framework.
+In fact, the minimally-biased kinetic analysis of such simulations is often a central bottleneck and presents a major challenge to the field.
+The analysis paradigms often entail the construction of lower resolution models parametrized from the high resolution simulation data set which capture the essential features in an interpretable framework.
 For example, by projecting the data down onto one or two degrees of freedom we create a simpler model for the system, such as one characterized by diffusion along a single reaction coordinate.
 
-Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high-resolution models that maintain quantitative agreement with the underlying simulation data and low-resolution models which capture the salient features of the underlying potential energy landscape while sacrificing some degree of model complexity.
+Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high-resolution models that maintain quantitative agreement with the underlying simulation data and low-resolution models which capture the salient features of the potential energy landscape while sacrificing some degree of model complexity.
 In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states.
@@
 
-A significant challenge in the automated construction of Markov state
-models is the choice of the number of states.
+A significant challenge in the automated construction of Markov state models is the choice of the number of states.
 Although classical Hamiltonian dynamics form a continuous-time Markov chain in $\mathbb{R}^{6N}$, the Markov property does not hold after the projecting the dynamics onto a basis of discrete indicator functions.
@@
 As the number of states in the model grows, so does the number of parameters required to completely specify the kinetic model between all pairs of states.
-Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per model parameter, which makes the approach liable for over-fitting.
+Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per model parameter, which makes the approach susceptible to over-fitting.
 
@@
 With the kinetic model expressed as a set of pairwise state to state transition probabilities at a given lag time, the likelihood of an ensemble of trajectories after projection into the indicator function basis is given simply by the product of the transition matrix elements along the observed trajectories.
-However, as we vary the number of states, it is not permissble to simply compare these likelihoods as part of an optimization of the state definitions.
+However, as we vary the number of states, it is not permissible to simply compare these likelihoods as part of an optimization of the state definitions.
 In doing so, the optimal model would always be the trivial one state model, whose computed likelihood is unity regardless of the data.
@@
 
-With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state transition probabilities, $T(s_i \rightarrow s_j)$, and so-called emission distributions of each state, the conditional probablity of observing a conformation at a given location in phase space given that the conformation, $x_t$ is within a certain state, $s(x_t)$.
+With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state transition probabilities, $T(s_i \rightarrow s_j)$, and so-called emission distributions of each state, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $x_t$ is within a certain state, $s(x_t)$.
 
@@
 In one model, the state emission distributions are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the states.
-If an observed trajectory does go through the first models’ regions of high likelihood, it is appropatiely termed a more likely model given the data.
+If an observed trajectory does go through the first models’ regions of high likelihood, it is appropriately termed a more likely model given the data.
 
 However, models' long timescale behavior -- the rates and fluxes between metastable basins are independent of the choice of the emission distributions.
-The emmission distributions characterize only the fine details of the equilibrium distribution within states, a quantity that the MSM approach does not seek to model; implicit in the decision to group conformations together into states is the idea that we decline to model the differences between conformations belonging to the same state.
+The emission distributions characterize only the fine details of the equilibrium distribution within states, a quantity that the MSM approach does not seek to model; implicit in the decision to group conformations together into states is the idea that we decline to model the differences between conformations belonging to the same state.
 If such differences exist and are sufficiently large to warrant attention, than the states are too large.
@@
 That is, the likelihood of observing a conformation in phase space given that the conformation is assigned to state i is 0 if the conformation is outside of the bounding volume of the state and constant if the conformation is within the volume.
-This constant is set so that the distribution integrates to 1, and is thus the reciprocal volume of themicrostate.
+This constant is set so that the distribution integrates to 1, and is thus the reciprocal volume of the microstate.
 
@@
 
-Likelihood maximization is insifficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to overfitting.
+Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to over-fitting\cite{Hastie01Elements}.
+Statistical learning theory provides a number of alternative approaches for this problem.
+Conceptually, the most straightfoward is a full Bayesian treatment in which all unknown model parameters are representated by probability distributions.
+The evidence for a model is computed by formally integrating over the model parameters.
+The evidence ratio, or Bayes factor\cite{Gelfand94Bayesian}, then provides a rigorous basis of model selection that appropriately punishes overly complex models as they become poorly constrained in parameter space.
+Unfortunately such approaches are intractable for problems of this size because of the need to integrate over all possible Markov models of a given size.
 
-Full bayesian treatment.
-Intractable for problems of this size, but right in theory.
-Cross validation: can effective in practice but tricky with timeseries -- hard in data-poor regime.
-BIC: very useful.
-``free'' computationally (no extra burden).
-Very nice heuristic.
+Instead, we find that Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} provides an efficient framework for choosing the number of states in a Markov state model.
 
-\section{Methods}
+What  about cross validation?
+
+\section{Computational Methods}
+
+\begin{figure}
+\centering
+\vspace{2em}
+\includegraphics[width=3in]{figs/fig1_rbow.png}
+\caption{Three systems studied in this work.
+(a) Langevin dynamics on the two dimensional M\"{u}ller potential.
+(b) 50 ns of dynamics of alanine dipeptide.
+(c) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{}, courtesy of D.E.
+Shaw research \cite{}.}
+\end{figure}
 
@@
 
+\bibliography{bibliography}
+
 \end{document}

### 2013-08-14 paper/paper2.tex
Added more text, and a new figure describing overfitting

@@
 
-Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high-resolution models that maintain quantitative agreement with the underlying simulation data and low-resolution models which capture the salient features of the potential energy landscape while sacrificing some degree of model complexity.
+Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models.
+Such detailed models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity.
 In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states.
@@
 Here, we seek to build models that are \emph{suitably} complex, given the data, yielding complex descriptions of the system only to the extent that their additional parameters are implied by the observed dynamics.
-To that end, we introduce a new procedure for scoring the likelihood of an MSM, which, together with the Bayesian information criterion (BIC), enables the optimal selection of the state space, which we express both in terms of the number of states and the clustering algorithm employed to group sampled conformations into states.
-This approach complements validation procedures performed primarily based on human intuition, such as Chapman-Kolmogorov tests, and enables us to treat model selection as an optimization problem amenable to automated methods.
+To that end, we introduce a new procedure for scoring the likelihood of an MSM, which, together with cross validation and the Bayesian information criterion (BIC), enables the optimal selection of the state space, which we express both in terms of the number of states and the clustering algorithm employed to group sampled conformations into states.
+This approach complements validation procedures performed primarily based on human intuition, such as Chapman-Kolmogorov tests, and enables the treatment of model selection as an optimization problem amenable to automated methods.
 
@@
 
-The appropriate likelihood is a path action in (continuous) phase space, on which the discrete states are merely an indicator function basis.
+The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis.
+With $s(X)$ as the function mapping conformations into the indicator function basis set, $s : \mathbb{R}^{3N} \rightarrow \{1, 2, \ldots, K\}$, the likelihood can be written as
 
 \begin{align}
-P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(X_i) \rightarrow s(x_{i+1})) \cdot \prod_{i=0}^{T} p_{s(x_i)}(X_{i})
+P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \cdot \prod_{i=0}^{T} p_{s(x_i)}(x_{i})
 \label{eq:like}
@@
 
-However, models' long timescale behavior -- the rates and fluxes between metastable basins are independent of the choice of the emission distributions.
+However, models' long timescale behavior -- the rates and fluxes between metastable basins -- are independent of the choice of the emission distributions.
 The emission distributions characterize only the fine details of the equilibrium distribution within states, a quantity that the MSM approach does not seek to model; implicit in the decision to group conformations together into states is the idea that we decline to model the differences between conformations belonging to the same state.
@@
 Therefore, the most appropriate emission distribution for discrete state MSMs is that of the uniform distribution over the phase-space volume of the state.
-That is, the likelihood of observing a conformation in phase space given that the conformation is assigned to state i is 0 if the conformation is outside of the bounding volume of the state and constant if the conformation is within the volume.
-This constant is set so that the distribution integrates to 1, and is thus the reciprocal volume of the microstate.
+That is, the likelihood of observing a conformation in phase space given that the conformation is assigned to state $i$ is $0$ if the conformation is outside of the bounding volume of the state and constant if the conformation is within the volume.
+This constant is set so that the distribution integrates to $1$, and is thus the reciprocal volume of the microstate.
 
@@
 
-\section{Model Selection With the Bayes Information Criterion}
+\section{Cross Validation and the Bayes Information Criterion}
 
-Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to over-fitting\cite{Hastie01Elements}.
+\begin{figure}
+\centering
+\includegraphics[width=3.1in]{figs/overfitting.png}
+\caption{An example of overfitting.
+Left: in the presence of noise, a more complex model is able to fit the observed training data better, with lower residues and a higher empirical likelihood, but fails to distinguish the signal from the noise.
+Right: more complex model classes exhibit lower training errors, but generalize poorly to unobserved data.}
+\end{figure}
+
+Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to overfitting\cite{Hastie01Elements}.
 Statistical learning theory provides a number of alternative approaches for this problem.
-Conceptually, the most straightfoward is a full Bayesian treatment in which all unknown model parameters are representated by probability distributions.
-The evidence for a model is computed by formally integrating over the model parameters.
-The evidence ratio, or Bayes factor\cite{Gelfand94Bayesian}, then provides a rigorous basis of model selection that appropriately punishes overly complex models as they become poorly constrained in parameter space.
+Conceptually, the most straightforward is a full Bayesian treatment in which all unknown model parameters are represented by probability distributions.
+The evidence for a model is computed by formally integrating over the model parameters and the evidence ratio, or Bayes factor\cite{Gelfand94Bayesian}, then provides a rigorous basis of model selection that appropriately punishes overly complex models as they become poorly constrained in parameter space.
 Unfortunately such approaches are intractable for problems of this size because of the need to integrate over all possible Markov models of a given size.
 
-Instead, we find that Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} provides an efficient framework for choosing the number of states in a Markov state model.
+Instead, we explore both cross validation and Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} for choosing the number of states in a Markov state model.
+For cross validation, we parameterize the model, building both the state space and transition matrix on a subset of the data, but evaluating the likelihood on the left-out portion in an attempt to directly measure the generalization error.
+The BIC on the other hand involves augmenting the likelihood with a penalty on the number of parameters, and is
+an asymptotic approximation Bayesian evidence.
 
-What  about cross validation?
+\begin{equation}
+\label{eq:bic}
+\mathrm{BIC} \equiv -2\cdot \ln L + k \ln (n)
+\end{equation}
+
+where $L$ is the likelihood, $k$ is the number of free parameters, and $n$ is the number of data points, assumed to be independent and identically distributed.
 
@@
 \centering
-\vspace{2em}
 \includegraphics[width=3in]{figs/fig1_rbow.png}
@@
 (b) 50 ns of dynamics of alanine dipeptide.
-(c) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{}, courtesy of D.E.
+(c) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E.
 Shaw research \cite{}.}
@@
 
+The uniform distribution emission model presents a computational challenge: its use requires the calculation of the (hyper)volume of the MSM's states, which, when defined by clustering are high-dimensional Voronoi cells.
+While trivial in two or three dimensions, this computational geometry task becomes challenging in high-dimensional settings.
+The computation of such volumes has occupied significant attention in recent years in the computational geometry literature, especially via randomized algorithms\cite{}.
+We opt to approximate the volumes using naive Monte Carlo rejection sampling, which we find tractable for large systems only when the molecular dynamics dataset is first projected into a suitable small vector space of up to perhaps ten dimensions.
+
+A further challenge is the procedure by which to model the volume of states which are at the ``edge'' of the MSM -- whose Voronoi cells extend to infinity in some direction.
+Is the volume of these states unbounded?
+It is appropriate to assert that the volume of these edge states is bounded in some way by the extent of our dataset.
+For example, the volume of a state might be defined as the volume of the intersection of its Voronoi cell and the convex hull of the whole dataset, which would encode the assumption that the likelihood of observing conformations outside the convex hull of the sampled data is vanishing.
+
+We use a slightly modified version of this definition that adopts the same spirit.
+Instead of taking the outer bounding envelope to be the convex hull of the data, we take it to be the set of all trial points such that the nearest sampled configuration to the trial point is closer than a certain cutoff, $R$.
+For further efficiency, it is feasible to use only a random subsample of the dataset for this nearest neighbor computation.
+
 \section{Results and Discussion}
+\subsection{M\"{u}ller Potential}
+We simulated [HOW MANY STEPS?] of Langevin dynamics on the M\"{u}ller potential, clustered the data using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with [HOW MANY ROUNDS] of Monte carlo rejection sampling, using a padding distance of $R=?$ around the state centers from the 500 state model to define the datasets envelope.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=2in]{figs/like_comp.png}
+\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states, very close to the region in which the BIC is minimized.
+The BIC optimal model is also where the models' slowest implied timescals begins to become invariant with respect to the number of states.}
+\label{fig:mullerlike}
+\end{figure}
+
+As shown in Fig.
+\ref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly.
+Here, the BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate.
+This model selection is consistent with tests based on the invariance of transition matrix eigenspectrum, but permits a direct optimization of the state space.
+
+\subsection{Alanine Dipeptide}
+
+\subsection{Fip35 WW Domain}
+
+\begin{figure}
+\centering
+\includegraphics[width=2in]{figs/ww_bic_vs_eval_slide_nscounts.png}
+\caption{}
+\label{fig:ww}
+\end{figure}
+
+To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E.
+Shaw research \cite{}.
+In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis, retaining between two and six uncorrelated(?) linear combination of pairwise distances(?).
 
 \bibliography{bibliography}
-
 \end{document}

### 2013-08-14 overfitting_figure/complexity.py

@@ -1,0 +1,26 @@
+import os
+import numpy as np
+import matplotlib.pyplot as pp
+from xkcdplot import XKCDify
+
+x = np.linspace(0.1, 5)
+training = 1/x + 1.0
+test = 1/x + 1.4 + 0.2*(x-1)**2
+
+pp.plot(x+0.1, test, label='test error')
+pp.plot(x, training, label='training error')
+
+#pp.legend()
+#pp.ylabel('Error')
+#pp.xlabel('Model Complexity')
+XKCDify(pp.gca())
+
+
+
+
+
+
+name = __file__ + '.png'
+pp.savefig(name)
+os.system('open %s' % name)
+

### 2013-08-14 overfitting_figure/overfittingfig.py

@@ -1,0 +1,50 @@
+import os
+import pandas as pd
+import numpy as np
+import statsmodels.formula.api as smf
+import matplotlib
+import matplotlib.pyplot as pp
+
+# Set pandas matplotlib style
+matplotlib.rcParams.update(pd.tools.plotting.mpl_stylesheet)
+matplotlib.rcParams.update({'font.family':'sans-serif'})
+
+# 3rd degree polynomial
+coeff = [0.4, -0.7, 0.3, 0]
+sigma = 0.01
+
+x = np.linspace(0, 1, 12)
+y = np.polyval(coeff, x) + sigma*np.random.randn(len(x))
+
+
+pp.figure(figsize=(8,4))#figsize=(3,3))
+pp.subplot(121)
+
+df = pd.DataFrame(data=dict(x=x, y=y))
+for i in range(2, 11):
+    df['x%d' % i] = df['x']**i
+df.plot(x='x', y='y', style='.', label='data', ms=15)
+
+m3 = smf.ols('y ~ x + x2 + x3', data=df).fit()
+m6 = smf.ols('y ~ x + x2 + x3 + x4 + x5 + x6', data=df).fit()
+
+df2 = pd.DataFrame(data=dict(x=np.linspace(0,1,1000)))
+for i in range(2, 11):
+    df2['x%d' % i] = df2['x']**i
+
+df2['y_pred_3'] = m3.predict(df2)
+df2['y_pred_6'] = m6.predict(df2)
+df2.plot(x='x', y='y_pred_3', label='3rd degree')
+df2.plot(x='x', y='y_pred_6', label='6th degree')
+pp.xlabel('Model Inputs', size=16)
+pp.ylabel('Predictions', size=16)
+pp.legend(loc=1)
+pp.gcf().subplots_adjust(bottom=0.15)
+
+pp.subplot(122)
+
+
+name = __file__ + '.png'
+pp.savefig(name)
+os.system('open %s' % name)
+

### 2013-08-14 overfitting_figure/text.py

@@ -1,0 +1,33 @@
+import os
+import pandas as pd
+import numpy as np
+import statsmodels.formula.api as smf
+import matplotlib
+import matplotlib.pyplot as pp
+
+# Set pandas matplotlib style
+matplotlib.rcParams.update(pd.tools.plotting.mpl_stylesheet)
+matplotlib.rcParams.update({'font.family':'sans-serif'})
+
+pp.figure(figsize=(8,4))
+
+pp.subplot(121)
+
+pd.DataFrame(data=dict(x=[0,1], y=[0,1.2])).plot(x='x', y='y', style='.', ms=15, label='Data')
+pp.plot([1], [1], label='Robust')
+pp.plot([1], [1], label='Overfit')
+
+pp.legend()
+
+pp.subplot(122)
+pp.plot([1], [1], label='Training Error')
+pp.plot([1], [1], label='Generalization Error')
+
+
+pp.xlabel('Model Complexity', size=16)
+pp.ylabel('Error', size=16)
+pp.legend()
+
+pp.gcf().subplots_adjust(bottom=0.15)
+pp.savefig('text.png')
+#os.system('open text.png')

### 2013-08-14 overfitting_figure/xkcdplot.py

@@ -1,0 +1,306 @@
+"""
+XKCD plot generator
+-------------------
+Author: Jake Vanderplas
+
+This is a script that will take any matplotlib line diagram, and convert it
+to an XKCD-style plot.  It will work for plots with line & text elements,
+including axes labels and titles (but not axes tick labels).
+
+The idea for this comes from work by Damon McDougall
+  http://www.mail-archive.com/matplotlib-users@lists.sourceforge.net/msg25499.html
+"""
+import numpy as np
+import pylab as pl
+from scipy import interpolate, signal
+import matplotlib.font_manager as fm
+
+
+# We need a special font for the code below.  It can be downloaded this way:
+import os
+import urllib2
+if not os.path.exists('Humor-Sans.ttf'):
+    fhandle = urllib2.urlopen('http://antiyawn.com/uploads/Humor-Sans.ttf')
+    open('Humor-Sans.ttf', 'w').write(fhandle.read())
+
+    
+def xkcd_line(x, y, xlim=None, ylim=None,
+              mag=1.0, f1=30, f2=0.05, f3=15):
+    """
+    Mimic a hand-drawn line from (x, y) data
+
+    Parameters
+    ----------
+    x, y : array_like
+        arrays to be modified
+    xlim, ylim : data range
+        the assumed plot range for the modification.  If not specified,
+        they will be guessed from the  data
+    mag : float
+        magnitude of distortions
+    f1, f2, f3 : int, float, int
+        filtering parameters.  f1 gives the size of the window, f2 gives
+        the high-frequency cutoff, f3 gives the size of the filter
+    
+    Returns
+    -------
+    x, y : ndarrays
+        The modified lines
+    """
+    x = np.asarray(x)
+    y = np.asarray(y)
+    
+    # get limits for rescaling
+    if xlim is None:
+        xlim = (x.min(), x.max())
+    if ylim is None:
+        ylim = (y.min(), y.max())
+
+    if xlim[1] == xlim[0]:
+        xlim = ylim
+        
+    if ylim[1] == ylim[0]:
+        ylim = xlim
+
+    # scale the data
+    x_scaled = (x - xlim[0]) * 1. / (xlim[1] - xlim[0])
[…]

### 2013-08-14 paper/paper2.tex
add some notes about the conclusions and stuff

@@
 
+\begin{figure}
+\centering
+\includegraphics[width=2in]{figs/ww_bic_vs_eval_slide_nscounts.png}
+\caption{Implied timescales and BIC scores for a series of Markov state models for the Fip35 WW domain.
+The panels differ by the dimensionality of the tICA projection.
+All models were built using Markovian lag time of [HOW MANY ns].}
+\label{fig:ww}
+\end{figure}
+
 \subsection{Alanine Dipeptide}
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=2in]{figs/ww_bic_vs_eval_slide_nscounts.png}
-\caption{}
-\label{fig:ww}
-\end{figure}
+To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E.
+Shaw Research \cite{}.
+In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis, retaining between two and six uncorrelated(?) linear combination of pairwise distances(?).
 
-To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E.
-Shaw research \cite{}.
-In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis, retaining between two and six uncorrelated(?) linear combination of pairwise distances(?).
+How many states are required for a Markov state model?
+Lane et.
+al., when analyzing this same dataset, proposed a 26,000 state Markov state model after clustering the data by RMSD, whereas Kellogg et.
+al.
+using an approach similar to that described herein, arrive at a [HOW MANY] state model after clustering on contact maps\cite{Lane2011Markov, Kellogg2012Evaluation}.
+
+\begin{itemize}
+\item Is the 26,000 state model ``overfit''?
+The obvious answer, based on the line of reasoning that this paper is taking, is yes.
+But I think we need to consider the question a little more subtly.
+Perhaps it depends on what questions we ask of the model.
+When you ask both a ``properly-fit'' model and an overfit model a question that basically falls within the training set -- and neither of them need to extrapolate or really infer out-of-sample information, they'll both do fine.
+Heck, the overfit model might do better at telling you exactly what was \emph{in} the dataset.
+\item The folding timescale is relatively insensitive to differences in MSM construction methodology.
+This has been observed in many datasets (McGibbon 2013, others).
+We hypothesize that this behavior is due to the large structural deviations at play -- even a two state model, given the same dataset, could basically get the right timescale just from counting the number of transitions.
+But finding these other timescales that are structurally subtle is different challenge.
+Also, how robust are other properties like experimental projections or the MFPT distribution to the number of states?
+\item Single trajectory datasets are ``easier'' for MSMs.
+One of the uses of MSMs is to ``stich together'' multiple trajectories -- this function only happens when the states are large enough that they \emph{connect} trajectories, so you can't have too many states in this case.
+But when you have a single trajectory, there's no possible way to create a disconnected model anyways.
+\item We need to state, elegantly if possible, the differences between our likelihood and Liz Kellogg's.
+This requires some careful language, since we think there are methodological problems with her approach.
+\end{itemize}
+
+\subsection{Limitations and Future Work}
+
+\begin{itemize}
+\item This model doesn't help us pick the lag time.
+\item It also doesn't help us pick the projected vector space, at least in a quantitative way. This is kind of a big deal, but it is what it is.
+\item Uniform distributions are extremely inconvenient to work with. In the future, we plan to extend this work to Markov state models without discrete states.
+\item \{I want to get this in print.\} Going forward, we anticipate that a combination of kinetic dimensionality reduction and clustering is going to be the key combination for robust Markov state model construction.
+The two approaches are characterized by complementary approximations and sources of error.
+In dimensionality reduction, the assumption is that the dynamics ``align'' with the vectors in some sense -- we talk about dynamics in the ``x direction'' as having some characteristic, regardless of whether they occur around $x=x_a$ or $x=x_b$.
+In clustering, if $x_a$ and $x_b$ are in different states, there's no crosstalk at all.
+A euclidean distance metric is invariant with respect to unitary transforms anyways.
+But clustering gets ``thrown off'' when you have kinetically irrelevant directions: to get the same resolution with the clusters in the kinetically \emph{relevant} directions, you need a number of states that is exponential in the number of included kinetically \emph{irrelevant} directions.
+Because the sources of error are complementary, the combination should be uniquely powerful.
+\end{itemize}
+
+\section{Conclusions}
+\begin{itemize}
+    \item MSMs are great
+    \item They have weaknesses. We are solving them.
+    \item The automated selection of the number of states is a critical part of procedures that build MSMs as part of a simulation protocol to drive adaptive sampling.
+If MD algorithms required user input every couple picoseconds, people would never get anything done.
+\end{itemize}
 

### 2013-08-14 paper/paper2.tex
Remade figure 3 to take up less space, grey background.

@@
 \label{eq:bic}
-\mathrm{BIC} \equiv -2\cdot \ln L + k \ln (n)
+\mathrm{BIC} \equiv -2\cdot \ln L + k \cdot \ln N
 \end{equation}
 
-where $L$ is the likelihood, $k$ is the number of free parameters, and $n$ is the number of data points, assumed to be independent and identically distributed.
+where $L$ is the likelihood, $k$ is the number of free parameters, and $N$ is the number of data points, assumed to be independent and identically distributed.
 
@@
 \centering
-\includegraphics[width=2in]{figs/like_comp.png}
+\includegraphics[width=3.3in]{figs/muller_bic_timescales.png}
 \caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states, very close to the region in which the BIC is minimized.

### 2013-08-15 paper/paper2.tex
added lots of citations. we're up to 30 and i've got most of the intro cited now

@@
 \usepackage{subfigure}  % use for side-by-side figures
-%\usepackage{hyperref}   % use for hypertext links, including those to external documents and URLs
+\usepackage{hyperref}   % use for hypertext links, including those to external documents and URLs
 \usepackage[capitalise]{cleveref}   % use for referencing figures/equations
@@
 Protein dynamics are an essential aspect of biomolecular function.
-These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, the conformational change between functional native substates, ligand binding, and allostery.
-Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex kinetics occurring via various pathways and with multiple intermediates.
-But atomic-resolution structural characterization of these dynamics is often a Herculean challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structural dynamics, generally reporting simultaneously on many degrees of freedom.
+These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, the conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}.
+Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex kinetics occurring via various pathways and with multiple intermediates \cite{Cosa2006Evidence, Zhang2011Direct}.
+But atomic-resolution structural characterization of these dynamics is often a Herculean challenge, particularly in dynamical settings -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.
 
 Computer simulations can complement experiments by providing atomic-resolution insight into the structural dynamics.
-With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation data sets.
+With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation data sets\cite{Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS}.
 In fact, the minimally-biased kinetic analysis of such simulations is often a central bottleneck and presents a major challenge to the field.
-The analysis paradigms often entail the construction of lower resolution models parametrized from the high resolution simulation data set which capture the essential features in an interpretable framework.
-For example, by projecting the data down onto one or two degrees of freedom we create a simpler model for the system, such as one characterized by diffusion along a single reaction coordinate.
+The analysis paradigms often entail the construction of lower resolution models parametrized from the high resolution simulation data set which capture the essential features in an interpretable framework\cite{Freddolino2010Challenges, Lane2013Milliseconds}.
+For example, by projecting the data down onto one or two degrees of freedom we create a simpler model for the system, such as one characterized by diffusion along a single reaction coordinate\cite{Best2010Coordinate}.
 
-Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models.
+Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models\cite{Pande2010Everything, Beauchamp2012Simple, Prinz2011Markov, Bowman2013Quantitative}.
 Such detailed models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity.
 In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states.
-The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which determine the kinetics.
+The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics.
 
-A significant challenge in the automated construction of Markov state models is the choice of the number of states.
+A significant challenge in the automated construction of Markov state models is the choice of the number of states\cite{McGibbon2013Learning}.
 Although classical Hamiltonian dynamics form a continuous-time Markov chain in $\mathbb{R}^{6N}$, the Markov property does not hold after the projecting the dynamics onto a basis of discrete indicator functions.
@@
 
-Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to overfitting\cite{Hastie01Elements}.
+Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to overfitting\cite{Liddle2007Information, Hastie01Elements}.
 Statistical learning theory provides a number of alternative approaches for this problem.
@@
 While trivial in two or three dimensions, this computational geometry task becomes challenging in high-dimensional settings.
-The computation of such volumes has occupied significant attention in recent years in the computational geometry literature, especially via randomized algorithms\cite{}.
+The computation of such volumes has occupied significant attention in recent years in the computational geometry literature, especially via randomized algorithms\cite{Kannan97, Simonovits03, Lovasz03}.
 We opt to approximate the volumes using naive Monte Carlo rejection sampling, which we find tractable for large systems only when the molecular dynamics dataset is first projected into a suitable small vector space of up to perhaps ten dimensions.

### 2013-08-15 paper/paper2.tex
Added the missing shaw citation for ww data
tica citation

@@
 (c) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E.
-Shaw research \cite{}.}
+Shaw research \cite{Shaw2010Atomic}.}
 \end{figure}
@@
 To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E.
-Shaw Research \cite{}.
-In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis, retaining between two and six uncorrelated(?) linear combination of pairwise distances(?).
+Shaw Research \cite{Shaw2010Atomic}.
+In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis\cite{Schwantes2013Improvements}, retaining between two and six uncorrelated(?) linear combination of pairwise distances(?).
 

### 2013-08-15 paper/paper2.tex
abstract

@@
 \begin{abstract}
-Likelihood framework for choosing the optimal number of states in a
-Markov State Model
+Markov state models (MSMs) are a powerful tool for the analysis of molecular dynamics simulations, but have been hampered by the need for manual selection of the number of states.
+We report a new method for the optimal selection of the number of states in an MSM based on the Bayesian information criterion.
+We demonstrate the approach on three systems of increasing complexity...
 \end{abstract}

### 2013-09-25 paper/paper2.tex
updated figures in paper2

@@
 \centering
-\includegraphics[width=3in]{figs/fig1_rbow.png}
+\includegraphics[width=3in]{figs/mull_ww.png}
 \caption{Three systems studied in this work.
@@
 \centering
-\includegraphics[width=3.3in]{figs/muller_bic_timescales.png}
+\includegraphics[width=3.3in]{figs/mull_cross.png}
 \caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states, very close to the region in which the BIC is minimized.
@@
 \centering
-\includegraphics[width=2in]{figs/ww_bic_vs_eval_slide_nscounts.png}
+\includegraphics[width=2in]{figs/ww_aic_vs_time.png}
 \caption{Implied timescales and BIC scores for a series of Markov state models for the Fip35 WW domain.

### 2013-10-01 paper/paper2.tex
paper changes. Wrote out most of RTM's notes into paragraph form. The most major thing we need is a discussion of why this is better than Liz's method. @rmcgibbo I think you said you had an idea of what to write here? Other than that we just need to make small changes to make it flow a bit better.

@@
 \usepackage[capitalise]{cleveref}   % use for referencing figures/equations
+\usepackage{soul}
+
 \begin{document}
@@
 These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, the conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}.
-Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex kinetics occurring via various pathways and with multiple intermediates \cite{Cosa2006Evidence, Zhang2011Direct}.
+Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with infcreasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex kinetics occurring via various pathways and with multiple intermediates \cite{Cosa2006Evidence, Zhang2011Direct}.
 But atomic-resolution structural characterization of these dynamics is often a Herculean challenge, particularly in dynamical settings -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.
@@
 (a) Langevin dynamics on the two dimensional M\"{u}ller potential.
-(b) 50 ns of dynamics of alanine dipeptide.
-(c) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E.
+(b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E.
 Shaw research \cite{Shaw2010Atomic}.}
@@
 \subsection{M\"{u}ller Potential}
-We simulated [HOW MANY STEPS?] of Langevin dynamics on the M\"{u}ller potential, clustered the data using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with [HOW MANY ROUNDS] of Monte carlo rejection sampling, using a padding distance of $R=?$ around the state centers from the 500 state model to define the datasets envelope.
+We simulated $1\times 10^7$ steps of Langevin dynamics on the M\"{u}ller potential, clustered the data using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with $1\times 10^5$ rounds of Monte carlo rejection sampling, using a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope.
+
+The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and furthermore, the likelihood remained almost exactly constant, which gives us some confidence that the volumes do not need to be perfect in order to have a good estimate of the likelihood.
+\hl{I can make a figure, and we can put it in the SI}.
 
@@
 \includegraphics[width=3.3in]{figs/mull_cross.png}
-\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states, very close to the region in which the BIC is minimized.
-The BIC optimal model is also where the models' slowest implied timescals begins to become invariant with respect to the number of states.}
+\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states.
+There are many methods that can be used to test the transferability of an MSM.
+We've plotted the AIC, BIC, and cross-validation scores for the WW domain.
+These results indicate that, at least for this dataset, the BIC and AIC may penalize the number of states too heavily.
+\hl{See SI section: ``AIC vs BIC" for a discussion of the various methods in detail.}}
 \label{fig:mullerlike}
@@
 \ref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly.
-Here, the BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate.
-This model selection is consistent with tests based on the invariance of transition matrix eigenspectrum, but permits a direct optimization of the state space.
+Here, the AIC and BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate.
+The optimal models according to the three methods are between 200 and 1000 states for this system.
+This observation is consistent with the convergence of the timescales of the MSMs, but allows for a direct optimization of the state space, as opposed to a heuristic way of comparing models.
+
+Interestingly, the AIC and BIC penalize the larger state models much more than the cross-validation approach.
+This could be due to the simplicity of this system.
+Since the M\"uller potential is only two-dimensional, it is difficult to actually produce a drastically over-fit model.
+We suspect that the cross-validation score on a real protein system would be closer to the BIC and AIC.
+However, we note, that cross-validation requires estimating a model without some subset of the data, in which case, it is not feasible for large systems that are not over-sampled.
 
@@
 \includegraphics[width=2in]{figs/ww_aic_vs_time.png}
-\caption{Implied timescales and BIC scores for a series of Markov state models for the Fip35 WW domain.
+\caption{Implied timescales and AIC scores for a series of Markov state models for the Fip35 WW domain.
 The panels differ by the dimensionality of the tICA projection.
-All models were built using Markovian lag time of [HOW MANY ns].}
+All models were built using a Markovian lag time of 50 ns.
+Qualitatively, the optimal model according to the AIC, corresponds roughly to where the timescales begin to flatten out.}
 \label{fig:ww}
 \end{figure}
-
-\subsection{Alanine Dipeptide}
 
@@
 Shaw Research \cite{Shaw2010Atomic}.
-In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis\cite{Schwantes2013Improvements}, retaining between two and six uncorrelated(?) linear combination of pairwise distances(?).
+In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis\cite{Schwantes2013Improvements}, retaining between two and six uncorrelated linear combinations of residue-residue distances.
 
@@
 al.
-using an approach similar to that described herein, arrive at a [HOW MANY] state model after clustering on contact maps\cite{Lane2011Markov, Kellogg2012Evaluation}.
+using an approach similar to that described herein, arrive at a \hl{1,000} state model after clustering on contact maps\cite{Lane2011Markov, Kellogg2012Evaluation}.
+\hl{That number is the micro state model from KCenters and CM's, but they also did KMeans and used 100 states}.
+
+According to this likelihood approach, the most likely models built on tICA projections are in the hundreds of states range.
+This does not mean, however, that the previous models are ``wrong." The value of a model can only be judged based on the questions one is asking.
+If the question is, what are the slowest dynamics in the data, then an over-fit model may actually give a better estimate than a ``perfectly-fit" model.
+However, if the question requires an inference of out-of-sample information, then we would trust the well-fit model over an over-fit one.
+
+For protein simulations, the slowest timescale (typically corresponding to folding) has been observed to be insensitive to differences in MSM construction methodology \cite{McGibbon2013Learning, others?}.
+We hypothesize that this behavior is due to the large structural changes associated with a folding event -- even a two state model, given the same dataset, could estimate the slowest timescale by simply counting the number of transitions.
+However, teasing out the slow timescales that are associated with subtler structural changes is a different challenge.
+
+\hl{Kellogg stuff.}
+
+\subsection{Limitations and Future Work}
+
+The likelihood described above allows for the comparison of MSMs with varying number of states, however it requires the definition of a vector space.
+As such, models built on different representations cannot be compared.
+This is because the likelihood requires the calculation of relative volumes, but the volumes will be different depending on what space you are working in.
+In addition, the lag time must be specified and cannot be selected using the likelihood.
+This is an improvement over the current state of constructing an MSM where the number of states is selected and then the lag time is tuned according to the eigenspectrum of the model.
+Selecting a lag time {\it a priori} is a much more intuitive parameter, that one typically knows based on the system.
+
+We can also extend these results to models that do not have discrete states.
+Since all we require is an emission probability into a vector space, it is easy to use this likelihood for evaluating models that no longer have discrete states.
+
+\section{Conclusions}
+
+Markov State Models are a promising analysis technique for automatically analyzing simulations generated from Markovian dynamics.
+There are a number of steps in the construction process, however, that require human intervention, which limits the use of MSMs to experts as well as introduces significant uncertainty into the model building process.
+These results take a large step toward fully automating the construction process by providing a quantitative way of selecting the number of states to use, as well as a way of selecting the clustering method.
+This will allow for non-experts to begin to use MSMs as well as allow adaptive sampling methods to be applied without human intervention.
+
+\bibliography{bibliography}
+\end{document}
 
@@
 \item Single trajectory datasets are ``easier'' for MSMs.
-One of the uses of MSMs is to ``stich together'' multiple trajectories -- this function only happens when the states are large enough that they \emph{connect} trajectories, so you can't have too many states in this case.
+One of the uses of MSMs is to ``stitch together'' multiple trajectories -- this function only happens when the states are large enough that they \emph{connect} trajectories, so you can't have too many states in this case.
 But when you have a single trajectory, there's no possible way to create a disconnected model anyways.
@@
 \end{itemize}
-
-\subsection{Limitations and Future Work}
 
@@
 
-\section{Conclusions}
 \begin{itemize}
@@
 \end{itemize}
-
-\bibliography{bibliography}
-\end{document}

### 2013-10-01 lib/likelihood.py

@@ -44,7 +44,8 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False,
     if not mapping is None:
         MSMLib.apply_mapping_to_assignments(trajs, mapping)
 
-    counts = MSMLib.get_count_matrix_from_assignments(trajs, lag_time=lagtime)
+    counts = MSMLib.get_count_matrix_from_assignments(trajs, lag_time=lagtime,
+        n_states=tProb.shape[0])
 
     counts = counts.tocsr()
     tProb = tProb.tocsr()

### 2013-10-01 lib/montecarlo.py

@@ -67,7 +67,8 @@ class MonteCarlo(VolumeEstimator):
     
         state_counts = np.zeros(len(self.prep_generators))
 
-        while state_counts.sum() < self.num_points:
+        count = 0
+        while count < self.num_points:
         # It might be more efficient to write this vectorially, but I don't 
         # want to run out of memory
             random_sample = np.random.random(self.dimension)
@@ -82,6 +83,8 @@ class MonteCarlo(VolumeEstimator):
             state_dists = self.metric.one_to_all(random_sample, self.prep_generators, 0)
 
             state_counts[np.argmin(state_dists)] += 1
+
+            count += 1
 
         volumes = state_counts.astype(float) / self.num_points
         # volumes are just percentages of the total space

### 2013-11-13 paper/paper2.tex
paper edits

@@
 Protein dynamics are an essential aspect of biomolecular function.
-These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, the conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}.
-Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with infcreasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex kinetics occurring via various pathways and with multiple intermediates \cite{Cosa2006Evidence, Zhang2011Direct}.
+These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}.
+Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex kinetics \cite{Cosa2006Evidence, Zhang2011Direct}.
 But atomic-resolution structural characterization of these dynamics is often a Herculean challenge, particularly in dynamical settings -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.
@@
 \includegraphics[width=3in]{figs/mull_ww.png}
-\caption{Three systems studied in this work.
-(a) Langevin dynamics on the two dimensional M\"{u}ller potential.
+\caption{Systems studied in this work.
+(A) Langevin dynamics on the two dimensional M\"{u}ller potential.
 (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E.
@@
 There are many methods that can be used to test the transferability of an MSM.
-We've plotted the AIC, BIC, and cross-validation scores for the WW domain.
+We've plotted the AIC, BIC, and cross-validation scores for the M\"uller potential simulations.
 These results indicate that, at least for this dataset, the BIC and AIC may penalize the number of states too heavily.

### 2013-11-13 lib/likelihood.py

@@ -75,3 +75,74 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False,
         return t_like, v_like
     
     return t_like + v_like
+
+
+def get_kmeans_log_likelihood(tprob, trajs, state_dists, num_dims, lagtime=1,
+    mapping=None, add_weights=False):
+
+    """
+    get the log likelihood of an MSM given that each state emits a conformation
+    according to a user-input probability model. 
+
+    Parameters
+    ----------
+    tprob : scipy.sparse matrix
+        transition probability matrix for your MSM
+    trajs : np.ndarray
+        two-dim. array corresponding to the assignments of each trajectory
+    state_dists : np.ndarray
+        distance of a conformation to its generator
+    num_dims : int
+        number of dimensions in your vector space
+    lagtime : int, optional
+        lag time of the MSM (default: 1)
+    mapping : mapping to map counts to a transition matrix
+    add_weights : bool, optional
+        add the per-state weights corresponding to R_n / R from the X-means 
+        paper (default: False)
+
+    Returns
+    -------
+    log_likelihood : float
+        log likelihood of the model
+
+    Notes
+    -----
+    These results come from Pelleg, D. and Moore A. X-means: Extending K-means
+        with Efficient Estimation of the Number of Clusters. 
+        (http://www.cs.cmu.edu/~dpelleg/download/xmeans.pdf)
+    """
+
+    if not mapping is None:
+        MSMLib.apply_mapping_to_assignments(trajs, mapping)
+
+    counts = MSMLib.get_count_matrix_from_assignments(trajs, lag_time=lagtime,
+        n_states=tprob.shape[0])
+
+    counts = counts.tocsr()
+    tprob = tprob.tocsr()
+
+    ij = counts.nonzero()
+
+    state_counts = np.array(counts.sum(axis=1)).flatten()
+    nonzero_counts = np.array(counts.data).flatten()
+    nonzero_tprobs = np.array(tprob[ij]).flatten()
+    # scipy.sparse uses np.matrices which behave differently than arrays
+    # so we cast everything to arrays and do the math with them
+   
+    #log_like = np.sum(np.array(counts.data) * np.array(np.log(tProb[ij]))) - \
+    #    np.sum(state_counts * np.log(volumes))
+    #log_like = np.sum(nonzero_counts * np.log(nonzero_tprobs)) - \
[…]

### 2013-11-15 lib/likelihood.py
added a figure

@@ -77,9 +77,8 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False,
     return t_like + v_like
 
 
-def get_kmeans_log_likelihood(tprob, trajs, state_dists, num_dims, lagtime=1,
-    mapping=None, add_weights=False):
-
+def get_kmeans_log_likelihood(tprob, trajs, state_dists, num_dims, 
+    lagtime=1, train_dists=None, train_ass=None, mapping=None, add_weights=False):
     """
     get the log likelihood of an MSM given that each state emits a conformation
     according to a user-input probability model. 
@@ -96,6 +95,11 @@ def get_kmeans_log_likelihood(tprob, trajs, state_dists, num_dims, lagtime=1,
         number of dimensions in your vector space
     lagtime : int, optional
         lag time of the MSM (default: 1)
+    train_dists : np.ndarray, optional
+        the distance to the cluster center for the training data. If this
+        is not None, then we assume you are doing cross-validation, 
+        which means the state variance is estimated from these distances
+        and the state_dists are used in the likelihood separately
     mapping : mapping to map counts to a transition matrix
     add_weights : bool, optional
         add the per-state weights corresponding to R_n / R from the X-means 
@@ -112,6 +116,9 @@ def get_kmeans_log_likelihood(tprob, trajs, state_dists, num_dims, lagtime=1,
         with Efficient Estimation of the Number of Clusters. 
         (http://www.cs.cmu.edu/~dpelleg/download/xmeans.pdf)
     """
+
+    if not train_dists is None and not train_ass is None:
+        cross = True
 
     if not mapping is None:
         MSMLib.apply_mapping_to_assignments(trajs, mapping)
@@ -139,10 +146,18 @@ def get_kmeans_log_likelihood(tprob, trajs, state_dists, num_dims, lagtime=1,
 
     #R = np.sum(state_counts) # total number of points
     R = np.where(trajs != -1)[0].shape[0]
-    state_var = np.square(state_dists[np.where(trajs != -1)]).sum() / (R - tprob.shape[0])
-    print state_var
+    if cross:
+        orig_R = np.where(train_ass != -1)[0].shape[0]
+        state_var = np.square(train_dists[np.where(train_ass != -1)]).sum() / (orig_R - tprob.shape[0])
+    else:
[…]

### 2013-11-15 paper/paper2.tex
a few paper edits plus added new figures for the MVN likelihood

@@
 In one model, the state emission distributions are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the states.
-If an observed trajectory does go through the first models’ regions of high likelihood, it is appropriately termed a more likely model given the data.
+If an observed trajectory does go through the first models' regions of high likelihood, it is appropriately termed a more likely model given the data.
 
-However, models' long timescale behavior -- the rates and fluxes between metastable basins -- are independent of the choice of the emission distributions.
-The emission distributions characterize only the fine details of the equilibrium distribution within states, a quantity that the MSM approach does not seek to model; implicit in the decision to group conformations together into states is the idea that we decline to model the differences between conformations belonging to the same state.
-If such differences exist and are sufficiently large to warrant attention, than the states are too large.
-
-Therefore, the most appropriate emission distribution for discrete state MSMs is that of the uniform distribution over the phase-space volume of the state.
-That is, the likelihood of observing a conformation in phase space given that the conformation is assigned to state $i$ is $0$ if the conformation is outside of the bounding volume of the state and constant if the conformation is within the volume.
-This constant is set so that the distribution integrates to $1$, and is thus the reciprocal volume of the microstate.
+This emission distribution can be defined in a number of ways.
+The simplest definition would be a uniform distribution since we are constructing an MSM that is agnostic to intra-state dynamics.
+When properly normalized, the uniform distribution takes a value of the reciprocal of the state's volume.
+This makes for an easy expression, however, volumes in high-dimensional states can be difficult to compute \ref{eq:like_vol}.
+For this reason, we can also choose other emission distributions that may be more tractable.
+In fact, \citet{Pelleg2000Xmeans} have used a gaussian emission distribution to define a likelihood to compare two k-means clusterings.
+The result of this emission distribution, is that the likelihood is proportional to the sum of square distances, which is precisely what is optimized in k-means clustering.
+If we think of k-centers as a poor man's version of k-means, then we can apply this same likelihood.
 
@@
 \end{equation}
+
+\begin{equation}
+\label{eq:like_mvn}
+\begin{split}
+P[x_{0...T-1}] & dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
+	&\cdot \prod_{i=0}^T \frac{1}{\left(2 \pi \sigma_{s(x_{i})}^2\right)^\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \sigma_{s(x_{i})}^2}\right)
+\end{split}
+\end{equation} where $\mu_{s(x_i)}$ is the cluster center for the state that $x_i$ is assigned to.
+For simplicity, we follow \citet{Pelleg2000Xmeans} for estimating the intrastate variances, and simply use a maximum likelihood estimator so that the likelihood becomes:
+\begin{equation}
+\label{eq:like_mvn}
+\begin{split}
+P[x_{0...T-1}] & dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
+	&\cdot \prod_{i=0}^T \frac{1}{\left(2 \pi \hat{\sigma}^2\right)^\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \hat{\sigma}^2}\right)
+\end{split}
+\end{equation} where
+\begin{equation}
+\label{eq:mle_sigma}
+\hat{\sigma}^2 = \frac{1}{T - k} \sum_{i=0}^T || x_i - \mu_{s(x_i)} ||^2
+\end{equation} where $k$ is the number of states.
 
@@
 
-Instead, we explore both cross validation and Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} for choosing the number of states in a Markov state model.
+Instead, we explore cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC} for choosing the number of states in a Markov state model.
 For cross validation, we parameterize the model, building both the state space and transition matrix on a subset of the data, but evaluating the likelihood on the left-out portion in an attempt to directly measure the generalization error.
@@
 \mathrm{BIC} \equiv -2\cdot \ln L + k \cdot \ln N
+\end{equation}
+\begin{equation}
+\label{eq:eic}
+\mathrm{AIC} \equiv -2\cdot \ln L + 2 \cdot k
 \end{equation}
@@
 \subsection{M\"{u}ller Potential}
-We simulated $1\times 10^7$ steps of Langevin dynamics on the M\"{u}ller potential, clustered the data using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with $1\times 10^5$ rounds of Monte carlo rejection sampling, using a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope.
+We simulated $1\times 10^7$ steps of Langevin dynamics on the M\"{u}ller potential, clustered the first $10^6$ steps using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with $1\times 10^5$ rounds of Monte carlo rejection sampling, using a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope.
 
 The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and furthermore, the likelihood remained almost exactly constant, which gives us some confidence that the volumes do not need to be perfect in order to have a good estimate of the likelihood.
-\hl{I can make a figure, and we can put it in the SI}.
+
+For cross-validation, we used the final $10^6$ frames as the test data.
 
@@
 \centering
-\includegraphics[width=3.3in]{figs/mull_cross.png}
+\includegraphics[width=2in]{figs/mull_cross_vert.png}
 \caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states.
@@
 \label{fig:mullerlike}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=3in]{figs/kmeans_vor_like.png}
+\caption{MSMs were built using K-Means clustering on the M\"uller potential simulations.
+Since K-Means optimizes an objective function that is related to the emission probability, our likelihood picks only a few states as the most likely model.
+Interestingly, the likelihood decreases as you introduce additional states.
+This behavior is due to the state-to-state likelihood decreased rapidly when adding additional states, which may be a result of the large number of states in the most stable well.}
 \end{figure}
@@
 \centering
-\includegraphics[width=2in]{figs/ww_aic_vs_time.png}
-\caption{Implied timescales and AIC scores for a series of Markov state models for the Fip35 WW domain.
-The panels differ by the dimensionality of the tICA projection.
-All models were built using a Markovian lag time of 50 ns.
-Qualitatively, the optimal model according to the AIC, corresponds roughly to where the timescales begin to flatten out.}
+\includegraphics[width=3in]{figs/ww_logLs.png}
+\caption{For models built using tICA, we were able to compute volumes as the dimensionality was low (A).
+These likelihoods were qualitatively similar to the MVN likelihoods (B).
+Because the MVN likelihood is easy to compute, we can use it to calculate the likelihood of a model clustered using the RMSD distance metric (C).
+Our likelihood indicates we should construct a model with 5000 states in the RMSD metric, whereas previous studies used a model with 26,000 states \cite{Lane2011Markov}.}
 \label{fig:ww}
@@
 \end{itemize}
+
+% old stuff about using uniform distributions
+However, models' long timescale behavior -- the rates and fluxes between metastable basins -- are independent of the choice of the emission distributions.
+The emission distributions characterize only the fine details of the equilibrium distribution within states, a quantity that the MSM approach does not seek to model; implicit in the decision to group conformations together into states is the idea that we decline to model the differences between conformations belonging to the same state.
+If such differences exist and are sufficiently large to warrant attention, than the states are too large.
+
+Therefore, the most appropriate emission distribution for discrete state MSMs is that of the uniform distribution over the phase-space volume of the state.
+That is, the likelihood of observing a conformation in phase space given that the conformation is assigned to state $i$ is $0$ if the conformation is outside of the bounding volume of the state and constant if the conformation is within the volume.
+This constant is set so that the distribution integrates to $1$, and is thus the reciprocal volume of the microstate.

### 2013-11-15 paper/paper2.tex
paper edits

@@
 Instead, we explore cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC} for choosing the number of states in a Markov state model.
-For cross validation, we parameterize the model, building both the state space and transition matrix on a subset of the data, but evaluating the likelihood on the left-out portion in an attempt to directly measure the generalization error.
+For cross validation, we parameterize the model, building both the state space and transition matrix on a subset of the data, but evaluate the likelihood on the left-out portion in an attempt to directly measure the generalization error.
 The BIC on the other hand involves augmenting the likelihood with a penalty on the number of parameters, and is
@@
 We've plotted the AIC, BIC, and cross-validation scores for the M\"uller potential simulations.
-These results indicate that, at least for this dataset, the BIC and AIC may penalize the number of states too heavily.
-\hl{See SI section: ``AIC vs BIC" for a discussion of the various methods in detail.}}
+These results indicate that, at least for this dataset, the BIC and AIC may penalize the number of states too heavily.}
 \label{fig:mullerlike}
 \end{figure}
+
+As shown in \cref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly.
+Here, the AIC and BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate.
+The optimal models according to the three methods are between 200 and 1000 states for this system.
+This observation is consistent with the convergence of the timescales of the MSMs, but allows for a direct optimization of the state space, as opposed to a heuristic way of comparing models.
+
+Interestingly, the AIC and BIC penalize the larger state models much more than the cross-validation approach.
+This could be due to the simplicity of this system.
+Since the M\"uller potential is only two-dimensional, it is difficult to actually produce a drastically over-fit model.
+We suspect that the cross-validation score on a real protein system would be closer to the BIC and AIC.
+However, we note, that cross-validation requires estimating a model without some subset of the data, which is not feasible for large systems that are not over-sampled.
+
+We also tested the MVN likelihood on MSMs built using k-means clustering on the M\"uller potential.
+The likelihood (\cref{fig:kmeans_mull}) peaks at a five-state model and actually decreases without the aid of the BIC or AIC.
+We can explain this maximum model by looking at where the cluster centers are located on the M\"uller potential.
+The five state model is the first model to have a cluster center located in the least stable well and as more states are added, the most stable well ends up being heavily split.
+We hypothesize that this over splitting in the most stable well gives rise to the decrease in likelihood.
 
@@
 This behavior is due to the state-to-state likelihood decreased rapidly when adding additional states, which may be a result of the large number of states in the most stable well.}
+\label{fig:kmeans_mull}
 \end{figure}
 
-As shown in Fig.
-\ref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly.
-Here, the AIC and BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate.
-The optimal models according to the three methods are between 200 and 1000 states for this system.
-This observation is consistent with the convergence of the timescales of the MSMs, but allows for a direct optimization of the state space, as opposed to a heuristic way of comparing models.
+\subsection{Fip35 WW Domain}
 
-Interestingly, the AIC and BIC penalize the larger state models much more than the cross-validation approach.
-This could be due to the simplicity of this system.
-Since the M\"uller potential is only two-dimensional, it is difficult to actually produce a drastically over-fit model.
-We suspect that the cross-validation score on a real protein system would be closer to the BIC and AIC.
-However, we note, that cross-validation requires estimating a model without some subset of the data, in which case, it is not feasible for large systems that are not over-sampled.
+To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E.
+Shaw Research \cite{Shaw2010Atomic}.
+In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements}, retaining between two and six uncorrelated linear combinations of residue-residue distances.
+
+How many states are required for a Markov state model?
+Lane et.
+al., when analyzing this same dataset, proposed a 26,000 state Markov state model after clustering the data by RMSD, whereas Kellogg et.
+al.
+using an approach similar to that described herein, arrive at a \hl{1,000} state model after clustering on contact maps\cite{Lane2011Markov, Kellogg2012Evaluation}.
+Both likelihoods, however, tell us that while using tICA, we should build models that have 200-500 states (\cref{fig:ww}).
+
+Since the MVN likelihood only requires the computation of distances, we can use it to analyze models built with in high dimensions as well as those built in spaces that are not vector spaces.
+We applied this methodology to models built using the minimum root mean square deviation in atom positions (RMSD) and k-centers clustering (\cref{fig:ww}).
+The MVN likelihood indicates that models built using RMSD should using on the order of 5,000 states, which is significantly higher than the tICA models.
+This is an illustration of how the tICA method can be used remarkably well to build Markov Models.
 
@@
 
-\subsection{Fip35 WW Domain}
-
-To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E.
-Shaw Research \cite{Shaw2010Atomic}.
-In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis\cite{Schwantes2013Improvements}, retaining between two and six uncorrelated linear combinations of residue-residue distances.
-
-How many states are required for a Markov state model?
-Lane et.
-al., when analyzing this same dataset, proposed a 26,000 state Markov state model after clustering the data by RMSD, whereas Kellogg et.
-al.
-using an approach similar to that described herein, arrive at a \hl{1,000} state model after clustering on contact maps\cite{Lane2011Markov, Kellogg2012Evaluation}.
-\hl{That number is the micro state model from KCenters and CM's, but they also did KMeans and used 100 states}.
-
 According to this likelihood approach, the most likely models built on tICA projections are in the hundreds of states range.
-This does not mean, however, that the previous models are ``wrong." The value of a model can only be judged based on the questions one is asking.
-If the question is, what are the slowest dynamics in the data, then an over-fit model may actually give a better estimate than a ``perfectly-fit" model.
-However, if the question requires an inference of out-of-sample information, then we would trust the well-fit model over an over-fit one.
 

### 2013-11-17 paper/paper2.tex
Many changes. The remaining thing is to write the prior work section

@@
 \usepackage{subfigure}  % use for side-by-side figures
-\usepackage{hyperref}   % use for hypertext links, including those to external documents and URLs
+\usepackage{hyperref}   % use for hypertext links, including those to external
+                        % documents and URLs
 \usepackage[capitalise]{cleveref}   % use for referencing figures/equations
@@
 \section{Introduction}
-Protein dynamics are an essential aspect of biomolecular function.
+Protein are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function.
 These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}.
-Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex kinetics \cite{Cosa2006Evidence, Zhang2011Direct}.
-But atomic-resolution structural characterization of these dynamics is often a Herculean challenge, particularly in dynamical settings -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.
+Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}.
+But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.
 
-Computer simulations can complement experiments by providing atomic-resolution insight into the structural dynamics.
-With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation data sets\cite{Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS}.
+Computer simulations can complement experiments by providing atomic insight into conformational dynamics.
+With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation data sets\cite{Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}.
 In fact, the minimally-biased kinetic analysis of such simulations is often a central bottleneck and presents a major challenge to the field.
@@
 The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics.
+The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors to identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{}.
 
@@
 Here, we seek to build models that are \emph{suitably} complex, given the data, yielding complex descriptions of the system only to the extent that their additional parameters are implied by the observed dynamics.
-To that end, we introduce a new procedure for scoring the likelihood of an MSM, which, together with cross validation and the Bayesian information criterion (BIC), enables the optimal selection of the state space, which we express both in terms of the number of states and the clustering algorithm employed to group sampled conformations into states.
+To that end, we introduce a procedure for scoring the likelihood of an MSM, which, together with standard statistical model selection techniques, enables the optimal selection of the state space, which we express both in terms of the number of states and the clustering algorithm employed to group sampled conformations into states.
 This approach complements validation procedures performed primarily based on human intuition, such as Chapman-Kolmogorov tests, and enables the treatment of model selection as an optimization problem amenable to automated methods.
+
+\section{Prior Work}
+In the context of MSMs, model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time.
+
+\begin{itemize}
+\item Swope-Pitera http://pubs.acs.org/doi/abs/10.1021/jp037421y
+\item Noe http://www.pnas.org/content/106/45/19011
+\item Park and Pande
+\item Sergio
+\item Kellogg
+\end{itemize}
+
+After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.
 
@@
 
-This emission distribution can be defined in a number of ways.
-The simplest definition would be a uniform distribution since we are constructing an MSM that is agnostic to intra-state dynamics.
+This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parameterizes a different statistical model.
+We propose two possible distributions: first, one may model the conditional distribution $P(x_t | s(x_t))$ as a uniform distribution over the (hyper)volume of $s(x_t)$, as shown in eq.
+\ref{eq:like_vol}.
+This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation.
+A more tractable alternative is a Gaussian emission model where $P(x_t | s(x_t))$ is modeled as multivariate normal, where the pairwise overlap between the emission distributions is treated as a truncation error, as shown in eq.
+\ref{eq:like_vol} and used in the popular X-means algorithm.\cite{Pelleg2000Xmeans} The truncation error steps from the fact that the emission probability distributions from every state assign nonzero probability to each conformation in a dataset, while the parameterization procedure for the Markov transition probabilities assumes each conformation to be uniquely assigned to a single state.
+Relaxing this approximation is possible with the forward-backward algorithm,\cite{Rabiner1989Tutorial} but is beyond the scope of this work.
+
+%The simplest definition would be a uniform distribution since we are constructing an MSM that is agnostic to intra-state dynamics.
 When properly normalized, the uniform distribution takes a value of the reciprocal of the state's volume.
@@
 \begin{split}
-P[x_{0...T-1}] & dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
-	&\cdot \prod_{i=0}^T \frac{1}{\left(2 \pi \sigma_{s(x_{i})}^2\right)^\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \sigma_{s(x_{i})}^2}\right)
+&P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
+    &\cdot \prod_{i=0}^T \left(2 \pi\right)^\frac{d}{2} |\Sigma_{s(x_i)}|^{-\frac{1}{2}}  \exp\left(-\frac{1}{2} (x_i - \mu_{s(x_i)})' \Sigma_{s(x_i)}^{-1} (x_i - \mu_{s(x_i)})\right)
 \end{split}
-\end{equation} where $\mu_{s(x_i)}$ is the cluster center for the state that $x_i$ is assigned to.
-For simplicity, we follow \citet{Pelleg2000Xmeans} for estimating the intrastate variances, and simply use a maximum likelihood estimator so that the likelihood becomes:
+\end{equation}
+
+where $\mu_{s(x_i)}$ is the cluster center assigned to $x_i$ and $\Sigma_{s(x_i)}$ is the corresponding covariance matrix.
+For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which conditions eq.
+\ref{eq:like_mvn} reduces to
 \begin{equation}
-\label{eq:like_mvn}
+\label{eq:like_mvn2}
 \begin{split}
 P[x_{0...T-1}] & dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
-	&\cdot \prod_{i=0}^T \frac{1}{\left(2 \pi \hat{\sigma}^2\right)^\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \hat{\sigma}^2}\right)
+	&\cdot \prod_{i=0}^T \left(2 \pi \hat{\sigma}^2\right)^{-\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \hat{\sigma}^2}\right)
 \end{split}
@@
 
-Instead, we explore cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC} for choosing the number of states in a Markov state model.
-For cross validation, we parameterize the model, building both the state space and transition matrix on a subset of the data, but evaluate the likelihood on the left-out portion in an attempt to directly measure the generalization error.
-The BIC on the other hand involves augmenting the likelihood with a penalty on the number of parameters, and is
-an asymptotic approximation Bayesian evidence.
+Instead, we explore three alternative procedures: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC} for choosing the number of states in a Markov state model.
+In cross validation, we attempt to directly measure the Markov model's generalization error.
+We parameterize the model, building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, but evaluate the likelihood on the left-out portion.
+The AIC and BIC are augmented likelihood function approaches which do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.
 
@@
 
-A further challenge is the procedure by which to model the volume of states which are at the ``edge'' of the MSM -- whose Voronoi cells extend to infinity in some direction.
-Is the volume of these states unbounded?
-It is appropriate to assert that the volume of these edge states is bounded in some way by the extent of our dataset.
-For example, the volume of a state might be defined as the volume of the intersection of its Voronoi cell and the convex hull of the whole dataset, which would encode the assumption that the likelihood of observing conformations outside the convex hull of the sampled data is vanishing.
+A further challenge is the procedure by which to model the volume of states which are at the edge of the MSM -- whose Voronoi cells extend to infinity in some direction.
+In these cases the Voronoi cells are of unbounded volume.
+Because the convex hull of our simulation datasets are computationally inaccessible, we define the outer extent of our data sets to be the set of all trial points such that the nearest sampled configuration to the trial point is closer than a certain cutoff, $R$.
 
-We use a slightly modified version of this definition that adopts the same spirit.
-Instead of taking the outer bounding envelope to be the convex hull of the data, we take it to be the set of all trial points such that the nearest sampled configuration to the trial point is closer than a certain cutoff, $R$.
-For further efficiency, it is feasible to use only a random subsample of the dataset for this nearest neighbor computation.
+For cross validation, we find it essential to use a dense prior during transition matrix estimation.
+Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes with maximum likelihood methods.
+For simplicity, we use the Jeffreys prior on each row of the transition matrix -- a Dirichlet prior with $\alpha=\frac{1}{2}$ -- which can be interpreted as a psuedocount of one half of an observed transition between each pair of states.
 
@@
 \subsection{M\"{u}ller Potential}
-We simulated $1\times 10^7$ steps of Langevin dynamics on the M\"{u}ller potential, clustered the first $10^6$ steps using the $k$-centers clustering algorithm with a euclidean distance metric, and computed the state volumes with $1\times 10^5$ rounds of Monte carlo rejection sampling, using a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope.
+We simulated two trajectories with $10^6$ steps of Langevin dynamics on the M\"{u}ller potential\cite{}.
+Simulations were performed with a timestep $t=0.1$, friction coefficient of $\gamma=10^3$, and $kT=15$.
+The first trajectory was clustered using the $k$-centers clustering algorithm, and state volumes were computed for the uniform emission model using $10^5$ rounds of Monte Carlo rejection sampling with a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope.
+The second trajectory was used as a test set.
 
-The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and furthermore, the likelihood remained almost exactly constant, which gives us some confidence that the volumes do not need to be perfect in order to have a good estimate of the likelihood.
-
-For cross-validation, we used the final $10^6$ frames as the test data.
+The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and the likelihood remained essentially constant, which indicates that the use of approximation volumes is sufficient to obtain robust estimates of the likelihood.
 
@@
 The optimal models according to the three methods are between 200 and 1000 states for this system.
-This observation is consistent with the convergence of the timescales of the MSMs, but allows for a direct optimization of the state space, as opposed to a heuristic way of comparing models.
+This observation is consistent with the convergence of the relaxation timescales of the models.
 
@@
 We suspect that the cross-validation score on a real protein system would be closer to the BIC and AIC.
-However, we note, that cross-validation requires estimating a model without some subset of the data, which is not feasible for large systems that are not over-sampled.
+However, we note, that cross-validation requires estimating a model without some subset of the data, which is not feasible for large systems in the data-sparse regime.
 
-We also tested the MVN likelihood on MSMs built using k-means clustering on the M\"uller potential.
-The likelihood (\cref{fig:kmeans_mull}) peaks at a five-state model and actually decreases without the aid of the BIC or AIC.
-We can explain this maximum model by looking at where the cluster centers are located on the M\"uller potential.
+We also tested the Gaussian likelihood on MSMs built using k-means clustering on the M\"uller potential.
+The likelihood (\cref{fig:kmeans_mull}) peaks at a five-state model and decreases with the addition of further states.
+This maximum can be interpreted by inspecting the k-means cluster centers on the M\"uller potential.
 The five state model is the first model to have a cluster center located in the least stable well and as more states are added, the most stable well ends up being heavily split.
-We hypothesize that this over splitting in the most stable well gives rise to the decrease in likelihood.
+We hypothesize that this over-splitting in the most stable minima gives rise to the decrease in MSM likelihood, and indicates a deficiency in clustering-based procedures which determine the state space without directly maximizing equations \ref{eq:like_vol} or \ref{eq:like_mvn}.
 
@@
 Shaw Research \cite{Shaw2010Atomic}.
-In order to reduce the dimensionality of the problem, especially critical for the computation of the state volumes, we first preprocess the trajectories using time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements}, retaining between two and six uncorrelated linear combinations of residue-residue distances.
+Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification}, extracting only the the four slowest uncorrelated linear combinations of residue-residue distances.
+Using the Gaussian emission model, we compare both models built in the 3 dimensional tICA subspace and those built using k-centers clustering with the cartesian root-mean squared deviation (RMSD) over the Fip35 WW domain's 252 non-redundant heavy atoms, computed using the quaternion characteristic polynomial method\cite{Theobald2005Rapid}.
 
-How many states are required for a Markov state model?
-Lane et.
-al., when analyzing this same dataset, proposed a 26,000 state Markov state model after clustering the data by RMSD, whereas Kellogg et.
-al.
-using an approach similar to that described herein, arrive at a \hl{1,000} state model after clustering on contact maps\cite{Lane2011Markov, Kellogg2012Evaluation}.
-Both likelihoods, however, tell us that while using tICA, we should build models that have 200-500 states (\cref{fig:ww}).
-
-Since the MVN likelihood only requires the computation of distances, we can use it to analyze models built with in high dimensions as well as those built in spaces that are not vector spaces.
-We applied this methodology to models built using the minimum root mean square deviation in atom positions (RMSD) and k-centers clustering (\cref{fig:ww}).
-The MVN likelihood indicates that models built using RMSD should using on the order of 5,000 states, which is significantly higher than the tICA models.
-This is an illustration of how the tICA method can be used remarkably well to build Markov Models.
+How many states are required in a Markov state model?
+The two 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain simulated by Shaw et.
+al.\cite{Shaw2010Atomic} have been used to parameterize multiple Markov state models.
+\citet{Lane2011Markov} proposed a 26,000 state MSM after clustering the data by RMSD, whereas \citet{Kellogg2012Evaluation} clustering on contact maps, present a 1000 state model.
+We find the optimal number of states to be highly dependent on space in which the data is clustered.
+With RMSD, the Gaussian likelihood indicates that Markov models built with on the order of 5,000 states optimally balance model quality and generalization error.
+On the other hand, the tICA method's identification of slow order parameters in the simulations has the effect of reducing the number of states necessary when constructing Markov models: here we find high-quality models can be constructed using only on the order of 500 states.
+Even with 200-600 states, Markov models constructed using tICA are able to accurately identify the simulations folding process on the 1-10 $\mu s$ timescale, as well as the faster near-native dynamics identified in this dataset by \citet{McGibbon2013Learning}.
 
@@
 
-According to this likelihood approach, the most likely models built on tICA projections are in the hundreds of states range.
-
-For protein simulations, the slowest timescale (typically corresponding to folding) has been observed to be insensitive to differences in MSM construction methodology \cite{McGibbon2013Learning, others?}.
-We hypothesize that this behavior is due to the large structural changes associated with a folding event -- even a two state model, given the same dataset, could estimate the slowest timescale by simply counting the number of transitions.
-However, teasing out the slow timescales that are associated with subtler structural changes is a different challenge.
-
-\hl{Kellogg stuff.}
-
 \subsection{Limitations and Future Work}
 
-The likelihood described above allows for the comparison of MSMs with varying number of states, however it requires the definition of a vector space.
-As such, models built on different representations cannot be compared.
-This is because the likelihood requires the calculation of relative volumes, but the volumes will be different depending on what space you are working in.
-In addition, the lag time must be specified and cannot be selected using the likelihood.
-This is an improvement over the current state of constructing an MSM where the number of states is selected and then the lag time is tuned according to the eigenspectrum of the model.
-Selecting a lag time {\it a priori} is a much more intuitive parameter, that one typically knows based on the system.
+The likelihood functions described herein permits the comparison of Markov state models with varying number of states, however it requires the definition of a vector space over which the emission model is defined.
+As such, models built on different projections of an underlying space cannot be directly compared.
+Furthermore, this approach is not amenable to optimizing the Markov lag time, the discretization interval for the Markov jump process.
+Nonetheless, we anticipate that automatic selection of the number of states in a Markov state model will remove much of the guess work in MSM construction, as the selection of the lag time is more amenable to existing Chapman-Kolmogorov tests.
 
-We can also extend these results to models that do not have discrete states.
-Since all we require is an emission probability into a vector space, it is easy to use this likelihood for evaluating models that no longer have discrete states.
+In the future, we plan to extend this work to the consideration of models without discrete states, where the requirement that states strictly partition phase space into a set of discrete indicator functions is relaxed.
 
@@
 
-Markov State Models are a promising analysis technique for automatically analyzing simulations generated from Markovian dynamics.
-There are a number of steps in the construction process, however, that require human intervention, which limits the use of MSMs to experts as well as introduces significant uncertainty into the model building process.
-These results take a large step toward fully automating the construction process by providing a quantitative way of selecting the number of states to use, as well as a way of selecting the clustering method.
-This will allow for non-experts to begin to use MSMs as well as allow adaptive sampling methods to be applied without human intervention.
+Markov State Models are powerful and popular tool for the analysis of molecular simulations, with a growing literature and two open source source software implementations\cite{Beauchamp2011Msmbuilder2, Senne2012EMMA}.
+There are, however a number of steps in the construction process that require human intervention, which limits the use of MSMs to experts and introduces significant biases into the model building process.
+Additionally, the ability to automatically construct Markov state models on the fly while simulations are in progress, an important point for so-called adaptive sampling procedures\cite{Bowman2010Enhanced}, is hampered when manual model selection is required.
+In the finite-data regime, additional complexity in the Markov model, in the form of increasing number of states, does come at a statistical cost.
+These results take a step toward fully automating the MSM construction process and controlling this statistical error by providing a quantitative method for by which the most suitable discretization procedure can be selected.
 

### 2013-11-25 paper/paper2.tex
final figs

@@
 
-\begin{figure}
-\centering
-\includegraphics[width=3.1in]{figs/overfitting.png}
-\caption{An example of overfitting.
-Left: in the presence of noise, a more complex model is able to fit the observed training data better, with lower residues and a higher empirical likelihood, but fails to distinguish the signal from the noise.
+%\begin{figure}
+%\centering
+%\includegraphics[width=3.1in]{figs/overfitting.png}
+%\caption{An example of overfitting.
+Left: in the presence of noise, a more complex model is able to fit the observed training data better, with lower residues and a higher empirical likelihood, %but fails to distinguish the signal from the noise.
 Right: more complex model classes exhibit lower training errors, but generalize poorly to unobserved data.}
-\end{figure}
+%\end{figure}
 
@@
 \centering
-\includegraphics[width=3in]{figs/mull_ww.png}
+\includegraphics[width=3in]{figs_final/mull_ww.png}
 \caption{Systems studied in this work.
@@
 \centering
-\includegraphics[width=2in]{figs/mull_cross_vert.png}
+\includegraphics[width=2in]{figs_final/mull_cross_vert.png}
 \caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states.

### 2013-11-25 lib/likelihood.py

@@ -117,6 +117,7 @@ def get_kmeans_log_likelihood(tprob, trajs, state_dists, num_dims,
         (http://www.cs.cmu.edu/~dpelleg/download/xmeans.pdf)
     """
 
+    cross = False
     if not train_dists is None and not train_ass is None:
         cross = True
 
@@ -152,6 +153,7 @@ def get_kmeans_log_likelihood(tprob, trajs, state_dists, num_dims,
     else:
         state_var = np.square(state_dists[np.where(trajs != -1)]).sum() / (R - tprob.shape[0])
 
+    print 'state variance: %.4e' % state_var
     s_like = - num_dims / 2. * R * np.log(2 * np.pi * state_var) 
     if cross:
         s_like -= np.square(state_dists[np.where(trajs != -1)]).sum() / (2. * state_var)

### 2013-11-26 fip35/get_like.py
added fig

@@ -1,45 +1,0 @@
-import likelihood
-import numpy as np
-from msmbuilder import io
-from scipy.io import mmread
-import os
-import argparse
-
-parser = argparse.ArgumentParser()
-parser.add_argument('-d', dest='msm_dirs', help='MSM directory containing, tProb.mtx, Mapping.dat, and Assignments.Fixed.h5', action='append')
-parser.add_argument('-v', dest='vols', help='Volumes to use for each MSM.', action='append')
-parser.add_argument('-o', dest='out_fn', default='likelihoods.npy', help='output filename (np.savetxt)')
-parser.add_argument('-l', type=int, dest='lagtime', help='lag time (in frames). Right now it only makes sense to compare MSMs built at the same lag time. Though I suppose it is possible to change this in the future.')
-args = parser.parse_args()
-
-if len(args.msm_dirs) != len(args.vols):
-    print "Need to input as many data directories as volume files."
-
-likelihoods = []
-states = []
-likelihoods.append(0) # for a one state model the likelihood is 0
-states.append(1)
-
-for i in xrange(len(args.msm_dirs)):
-
-    msm_folder = args.msm_dirs[i]
-    
-    print "Calculating likelihood for MSM in %s" % msm_folder
-
-    t = mmread(os.path.join(msm_folder, 'tProb.mtx'))
-    m = np.loadtxt(os.path.join(msm_folder, 'Mapping.dat'))
-    v = np.load(args.vols[i])
-    a = io.loadh(os.path.join(msm_folder, 'Assignments.Fixed.h5'))
-    try:
-        a = a['arr_0']
-    except:
-        a = a['Data']
-
-    v = v[np.where(m != -1)] # remove trimmed states
-    likelihoods.append(likelihood.get_log_likelihood(t, a, v, lagtime=args.lagtime, mapping=m))
-    states.append(np.where(m != -1)[0].shape[0])
-
-out_ary = np.array([states, likelihoods]).T
-
-np.save(args.out_fn, out_ary)
-

### 2013-11-26 paper/paper2.tex
added fig
changed style in the math stuff

@@
 \usepackage{soul}
+\usepackage{bm}
 
@@
 The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis.
-With $s(X)$ as the function mapping conformations into the indicator function basis set, $s : \mathbb{R}^{3N} \rightarrow \{1, 2, \ldots, K\}$, the likelihood can be written as
+With $s(X)$ as the function mapping conformations into the indicator function basis set, and $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory, the likelihood can be written as:
 
-\begin{align}
-P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \cdot \prod_{i=0}^{T} p_{s(x_i)}(x_{i})
+\begin{equation}
+\begin{split}
+P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
+\cdot \prod_{t=1}^{N} p_{s(\mathbf{x}_t)}(\mathbf{x}_{t}) \cdot d\mathbf{x}^N
 \label{eq:like}
-\end{align}
+\end{split}
+\end{equation}
 
-With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state transition probabilities, $T(s_i \rightarrow s_j)$, and so-called emission distributions of each state, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $x_t$ is within a certain state, $s(x_t)$.
+With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state transition probabilities, $T(s_i \rightarrow s_j)$, and so-called emission distributions of each state, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}_t$ is within a certain state, $s(\mathbf{x}_t)$.
 
@@
 This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parameterizes a different statistical model.
-We propose two possible distributions: first, one may model the conditional distribution $P(x_t | s(x_t))$ as a uniform distribution over the (hyper)volume of $s(x_t)$, as shown in eq.
+We propose two possible distributions: first, one may model the conditional distribution $P(\mathbf{x}_t | s(\mathbf{x}_t))$ as a uniform distribution over the (hyper)volume of $s(\mathbf{x}_t)$, as shown in eq.
 \ref{eq:like_vol}.
 This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation.
-A more tractable alternative is a Gaussian emission model where $P(x_t | s(x_t))$ is modeled as multivariate normal, where the pairwise overlap between the emission distributions is treated as a truncation error, as shown in eq.
+A more tractable alternative is a Gaussian emission model where $P(\mathbf{x}_t | s(\mathbf{x}_t))$ is modeled as multivariate normal, where the pairwise overlap between the emission distributions is treated as a truncation error, as shown in eq.
 \ref{eq:like_vol} and used in the popular X-means algorithm.\cite{Pelleg2000Xmeans} The truncation error steps from the fact that the emission probability distributions from every state assign nonzero probability to each conformation in a dataset, while the parameterization procedure for the Markov transition probabilities assumes each conformation to be uniquely assigned to a single state.
@@
 \label{eq:like_vol}
-P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \cdot \prod_{i=0}^T \frac{1}{V_{s(x_{i})}}
+\begin{split}
+P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
+	\cdot \prod_{t=1}^N \frac{1}{V_{s(\mathbf{x}_{t})}} \cdot d\mathbf{x}^N
+\end{split}
 \end{equation}
@@
 \begin{split}
-&P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
-    &\cdot \prod_{i=0}^T \left(2 \pi\right)^\frac{d}{2} |\Sigma_{s(x_i)}|^{-\frac{1}{2}}  \exp\left(-\frac{1}{2} (x_i - \mu_{s(x_i)})' \Sigma_{s(x_i)}^{-1} (x_i - \mu_{s(x_i)})\right)
+&P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
+	&\;\;\cdot \prod_{t=1}^N\Bigg[ \frac{1}{(2 \pi)^\frac{d}{2} |\Sigma_{s(\mathbf{x}_t)}|^\frac{1}{2}} \\
+    &\;\;\cdot \exp\left(-\frac{1}{2} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})^T \Sigma_{s(\mathbf{x}_t)}^{-1} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})\right)\Bigg] \cdot d\mathbf{x}^N
 \end{split}
@@
 
-where $\mu_{s(x_i)}$ is the cluster center assigned to $x_i$ and $\Sigma_{s(x_i)}$ is the corresponding covariance matrix.
+where $\bm{\mu}_{s(\mathbf{x}_t)}$ is the cluster center assigned to $\mathbf{x}_t$ and $\Sigma_{s(\mathbf{x}_t)}$ is the corresponding covariance matrix.
 For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which conditions eq.
@@
 \begin{split}
-P[x_{0...T-1}] & dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
-	&\cdot \prod_{i=0}^T \left(2 \pi \hat{\sigma}^2\right)^{-\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \hat{\sigma}^2}\right)
+P\left[\{\mathbf{x}_t\}_{t=1}^{N}\right] & d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
+	&\cdot \prod_{t=1}^N \left(2 \pi \hat{\sigma}^2\right)^{-\frac{d}{2}} \exp\left(-\frac{||\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)}||^2}{2 \hat{\sigma}^2}\right)
 \end{split}
@@
 Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes with maximum likelihood methods.
+We used a dirichlet prior on each entry of the transition matrix, which
+
 For simplicity, we use the Jeffreys prior on each row of the transition matrix -- a Dirichlet prior with $\alpha=\frac{1}{2}$ -- which can be interpreted as a psuedocount of one half of an observed transition between each pair of states.
@@
 \centering
-\includegraphics[width=2in]{figs_final/mull_cross_vert.png}
+\includegraphics[width=3in]{figs_final/mull_likes.png}
 \caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states.
@@
 \centering
-\includegraphics[width=3in]{figs/kmeans_vor_like.png}
+\includegraphics[width=2.5in]{figs_final/kmeans_like2_lbl.png}
 \caption{MSMs were built using K-Means clustering on the M\"uller potential simulations.
@@
 \centering
-\includegraphics[width=3in]{figs/ww_logLs.png}
+\includegraphics[width=3in]{figs_final/ww_likes.png}
 \caption{For models built using tICA, we were able to compute volumes as the dimensionality was low (A).

### 2013-11-28 paper/paper2.tex
Final updates to the paper including a new abstract, fixed citations, and reformatting a couple sentences

@@
-\documentclass[twocolumn,floatfix,nofootinbib,aps]{revtex4-1}
+%\documentclass[twocolumn,floatfix,nofootinbib,acs]{revtex4-1}
+\documentclass[journal=jpcbfk, layout=twocolumn, manuscript=article]{achemso}
+\setkeys{acs}{articletitle = true}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{bm}
+\usepackage{achemso}
+
+\title{Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics}
+\author{Christian R. Schwantes}
+\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
+\altaffiliation{Contributed equally to this work}
+\author{Robert T. McGibbon}
+\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
+\altaffiliation{Contributed equally to this work}
+\author{Vijay S. Pande}
+\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
+\alsoaffiliation{Biophysics Program, Stanford University, Stanford CA 94305, USA}
+\alsoaffiliation{Department of Computer Science, Stanford University, Stanford CA 94305, USA}
+\alsoaffiliation{Department of Structural Biology, Stanford University, Stanford CA 94305, USA}
+\email{pande@stanford.edu}
+
+\mciteErrorOnUnknownfalse
 
 \begin{document}
-
-\title{Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics}
-\author{Robert T. McGibbon}
-\author{Christian R. Schwantes}
-\author{Vijay S. Pande}
-
+\maketitle
 \begin{abstract}
-Markov state models (MSMs) are a powerful tool for the analysis of molecular dynamics simulations, but have been hampered by the need for manual selection of the number of states.
-We report a new method for the optimal selection of the number of states in an MSM based on the Bayesian information criterion.
-We demonstrate the approach on three systems of increasing complexity...
+Markov state models provide a powerful framework for the analysis of biomolecular conformation dynamics in terms of their metastable states and transition rates.
+These models provide both a quantitative and comprehensible description of the long-timescale dynamics of large molecular dynamics with a Master equation, and have been successfully used to study protein folding, protein conformational change, and protein-ligand binding.
+However, to achieve satisfactory performance, existing methodologies often require expert intervention when defining the model's discrete state space.
+Here, we show that by consideration of the states' conditional distribution over conformations, the state space can be evaluated and optimized using the Markov state model's likelihood function and standard statistical model selection tools.
+Application of these techniques on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those techniques encourage over-fitting.
+By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markvov state model construction.
 \end{abstract}
-
-\maketitle
 
@@
 Protein are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function.
-These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}.
+These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native sub-states, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}.
 Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}.
@@
 Computer simulations can complement experiments by providing atomic insight into conformational dynamics.
-With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation data sets\cite{Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}.
+With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation ensembles\cite{Gotz2012Routine, Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}.
 In fact, the minimally-biased kinetic analysis of such simulations is often a central bottleneck and presents a major challenge to the field.
@@
 
-Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models\cite{Pande2010Everything, Beauchamp2012Simple, Prinz2011Markov, Bowman2013Quantitative}.
+Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models\cite{Chodera2007Automatic, Prinz2011Markov, Beauchamp2012Simple, Bowman2013Quantitative}.
 Such detailed models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity.
@@
 \section{Prior Work}
+%After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.
 In the context of MSMs, model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time.
+As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
+$$ T(n\tau) = T(\tau)^n $$ As such, the eigenvalues, and correspondingly the relaxation timescales, should remain constant as the lag time increases.
+The resulting test of a model's quality consists of calculating the eigenspectra of models built at many lag times and choosing models whose eigenvalues have converged.
 
-\begin{itemize}
-\item Swope-Pitera http://pubs.acs.org/doi/abs/10.1021/jp037421y
-\item Noe http://www.pnas.org/content/106/45/19011
-\item Park and Pande
-\item Sergio
-\item Kellogg
-\end{itemize}
+In practice, while this test has been useful for many \cite{Noe2009Constructing}, it has some flaws.
+In particular, it is generally observed that most models' timescales do not converge, but steadily increase, making it is difficult to judge whether a particular model passes the test and encouraging the use of ever-more states.
+Additionally, it is difficult to compare models built on two state decompositions, as neither may have convergent timescales.
+Often, the choice is made to use the model whose timescales better match an experimental observable.
 
-After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.
+Other evaluation strategies include a test based on comparison to a second-order Markov model \cite{Park2006Validation}.
+Additionally, Bayesian methods that involve marginalizing over all possible MSMs have been developed \cite{Bacallado2009Bayesian, *Bacallado2011Bayesian}.
+However, these methods are often prohibitively expensive on larger data sets.
+In contrast, \citet{Kellogg2012Evaluation} proposed a tractable method for comparing MSMs built with different state decompositions, based on a likelihood function for an MD trajectory given an MSM, discussed below.
 
 \section{Likelihood of a Markov State Model}
-With the kinetic model expressed as a set of pairwise state to state transition probabilities at a given lag time, the likelihood of an ensemble of trajectories after projection into the indicator function basis is given simply by the product of the transition matrix elements along the observed trajectories.
+Bayes' rule provides a foundation for model selection, by establishing the proportionality of the probability of a model given the data to the probability of the data given the model.
+Without a strong prior, model selection is reduced to the search for the model that maximizes the likelihood of the data.
+
+Thus we seek to formulate this likelihood function for Markov state models.
+With the kinetic model expressed as a set of pairwise state to state transition probabilities at a given lag time, the likelihood of an ensemble of trajectories after projection into the MSM state space is given simply by the product of the transition matrix elements along the observed trajectories.
 However, as we vary the number of states, it is not permissible to simply compare these likelihoods as part of an optimization of the state definitions.
@@
 The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis.
-With $s(X)$ as the function mapping conformations into the indicator function basis set, and $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory, the likelihood can be written as:
+With $s(\mathbf{x})$ as the function mapping conformations into the indicator function basis set, and $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory, the likelihood can be written as:
 
@@
 P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
-\cdot \prod_{t=1}^{N} p_{s(\mathbf{x}_t)}(\mathbf{x}_{t}) \cdot d\mathbf{x}^N
+\cdot \prod_{t=1}^{N} P(\mathbf{x}_t | s(\mathbf{x}_t)) \cdot d\mathbf{x}^N
 \label{eq:like}
@@
 
-With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state transition probabilities, $T(s_i \rightarrow s_j)$, and so-called emission distributions of each state, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}_t$ is within a certain state, $s(\mathbf{x}_t)$.
+With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state Markov transition probabilities, $T(s_i \rightarrow s_j) \equiv P(S_{t+1}=s_j | S_{t}=s_i)$, and so-called emission distributions of each state, $P(\mathbf{x} | s(\mathbf{x}))$, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}$, is within a certain state, $s_i$.
+By convention, we define $P(\mathbf{x}| s(\mathbf{x}))$ to be zero for all $\mathbf{x}$ such that $s(\mathbf{x}) \neq s_i$.
 
-For example, consider two Markov state models sharing the same transition matrix, $T$.
+This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parametrizes a different statistical model.
+\citet{Kellogg2012Evaluation} used a discriminative model with:
+\begin{equation}
+P(\mathbf{x}_t | s(\mathbf{x}_t)) = \big| \{ \mathbf{x}_k : s(\mathbf{x}_k) = s(\mathbf{x}_t) \} \big|^{-1}
+\end{equation} This model has the training data set as its support rather than phase space, which makes it unable to generalize and assign probability to new data.
+In certain circumstances, this is an undesirable property.
+For example, protocols that involve fitting and validating models on separate data sets (e.g.
+cross-validation) are impossible when the statistical model lacks the capacity to describe new data.
+
+Instead, we seek a generative model whose emission distributions are supported on phase space (i.e.
+standard distributions over $\mathbb{R}^{3N_{atoms}}$).
+We propose two possible emission distributions, $P(\mathbf{x} | s(\mathbf{x}))$: first, a uniform distribution over the (hyper)volume of each state, as shown in \cref{eq:like_vol}.
+This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation.
+
+%For example, consider two Markov state models sharing the same transition matrix, $T$.
 In one model, the state emission distributions are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the states.
 If an observed trajectory does go through the first models' regions of high likelihood, it is appropriately termed a more likely model given the data.
-
-This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parameterizes a different statistical model.
-We propose two possible distributions: first, one may model the conditional distribution $P(\mathbf{x}_t | s(\mathbf{x}_t))$ as a uniform distribution over the (hyper)volume of $s(\mathbf{x}_t)$, as shown in eq.
-\ref{eq:like_vol}.
-This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation.
-A more tractable alternative is a Gaussian emission model where $P(\mathbf{x}_t | s(\mathbf{x}_t))$ is modeled as multivariate normal, where the pairwise overlap between the emission distributions is treated as a truncation error, as shown in eq.
-\ref{eq:like_vol} and used in the popular X-means algorithm.\cite{Pelleg2000Xmeans} The truncation error steps from the fact that the emission probability distributions from every state assign nonzero probability to each conformation in a dataset, while the parameterization procedure for the Markov transition probabilities assumes each conformation to be uniquely assigned to a single state.
-Relaxing this approximation is possible with the forward-backward algorithm,\cite{Rabiner1989Tutorial} but is beyond the scope of this work.
 
@@
 If we think of k-centers as a poor man's version of k-means, then we can apply this same likelihood.
-
 \begin{equation}
@@
 \begin{split}
-P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
-	\cdot \prod_{t=1}^N \frac{1}{V_{s(\mathbf{x}_{t})}} \cdot d\mathbf{x}^N
+%P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
+ P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{V_{s(\mathbf{x}_{t})}}
 \end{split}
@@
 
+A more tractable alternative is a Gaussian emission model where $P(\mathbf{x} | s(\mathbf{x}))$ is modeled as multivariate normal \cref{eq:like_mvn}.
 \begin{equation}
@@
 \begin{split}
-&P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
-	&\;\;\cdot \prod_{t=1}^N\Bigg[ \frac{1}{(2 \pi)^\frac{d}{2} |\Sigma_{s(\mathbf{x}_t)}|^\frac{1}{2}} \\
-    &\;\;\cdot \exp\left(-\frac{1}{2} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})^T \Sigma_{s(\mathbf{x}_t)}^{-1} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})\right)\Bigg] \cdot d\mathbf{x}^N
+P(\mathbf{x}_t | s(\mathbf{x}_t)) = \Bigg[ \frac{1}{(2 \pi)^\frac{d}{2} |\Sigma_{s(\mathbf{x}_t)}|^\frac{1}{2}} \\
+   \exp\left(-\frac{1}{2} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})^T \Sigma_{s(\mathbf{x}_t)}^{-1} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})\right)\Bigg] 
 \end{split}
-\end{equation}
-
-where $\bm{\mu}_{s(\mathbf{x}_t)}$ is the cluster center assigned to $\mathbf{x}_t$ and $\Sigma_{s(\mathbf{x}_t)}$ is the corresponding covariance matrix.
-For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which conditions eq.
-\ref{eq:like_mvn} reduces to
+\end{equation} where $\bm{\mu}_{s(\mathbf{x}_t)}$ is the cluster center assigned to $\mathbf{x}_t$ and $\Sigma_{s(\mathbf{x}_t)}$ is the corresponding covariance matrix.
+For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which it reduces to:
 \begin{equation}
 \label{eq:like_mvn2}
-\begin{split}
-P\left[\{\mathbf{x}_t\}_{t=1}^{N}\right] & d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
-	&\cdot \prod_{t=1}^N \left(2 \pi \hat{\sigma}^2\right)^{-\frac{d}{2}} \exp\left(-\frac{||\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)}||^2}{2 \hat{\sigma}^2}\right)
-\end{split}
+ P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{\left(2 \pi \hat{\sigma}^2\right)^{\frac{d}{2}}} \exp\left(-\frac{||\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)}||^2}{2 \hat{\sigma}^2} \right)
 \end{equation} where
@@
 \label{eq:mle_sigma}
-\hat{\sigma}^2 = \frac{1}{T - k} \sum_{i=0}^T || x_i - \mu_{s(x_i)} ||^2
-\end{equation} where $k$ is the number of states.
+\hat{\sigma}^2 = \frac{1}{N - k} \sum_{t=1}^N || \mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)} ||^2
+\end{equation} and $k$ is the number of states.
+
+Since the likelihood requires the emission distributions to have zero overlap -- such that each conformation only has nonzero emission probability from a single state, the nonzero overlap between the Gaussian emission distributions introduces a truncation error.
+One approach to ameliorate this issue is to truncate the density functions at the state boundaries, and renormalize.
+We assume that the overlaps are small, and use the density defined in \cref{eq:like_mvn2}.
+An alternative approach for relaxing this approximation is possible with the forward-backward algorithm,\cite{Rabiner1989Tutorial} but is beyond the scope of this work.
 
 \section{Cross Validation and the Bayes Information Criterion}
-
-%\begin{figure}
-%\centering
-%\includegraphics[width=3.1in]{figs/overfitting.png}
-%\caption{An example of overfitting.
-Left: in the presence of noise, a more complex model is able to fit the observed training data better, with lower residues and a higher empirical likelihood, %but fails to distinguish the signal from the noise.
-Right: more complex model classes exhibit lower training errors, but generalize poorly to unobserved data.}
-%\end{figure}
-
-Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to overfitting\cite{Liddle2007Information, Hastie01Elements}.
+Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to over-fitting\cite{Liddle2007Information, Hastie01Elements}.
 Statistical learning theory provides a number of alternative approaches for this problem.
@@
 
-Instead, we explore three alternative procedures: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC} for choosing the number of states in a Markov state model.
-In cross validation, we attempt to directly measure the Markov model's generalization error.
-We parameterize the model, building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, but evaluate the likelihood on the left-out portion.
+Instead, we explore three alternative procedures for choosing the number of states in an MSM: cross validation, Schwarz's Bayesian information criterion (BIC)\cite{Schwartz78Estimating} and the Akaike information criterion (AIC)\cite{Akaike1974AIC}.
+Cross-validation attempts to directly measure the Markov model's generalization error.
+First, the model is parameterized by building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, then the likelihood is evaluated on the left-out portion.
+This scheme can be repeated many times with different partitions of the data set.
 The AIC and BIC are augmented likelihood function approaches which do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.
-
 \begin{equation}
 \label{eq:bic}
-\mathrm{BIC} \equiv -2\cdot \ln L + k \cdot \ln N
+\mathrm{BIC} \equiv -2\cdot \ln L + \kappa \cdot \ln N
 \end{equation}
 \begin{equation}
-\label{eq:eic}
-\mathrm{AIC} \equiv -2\cdot \ln L + 2 \cdot k
-\end{equation}
-
-where $L$ is the likelihood, $k$ is the number of free parameters, and $N$ is the number of data points, assumed to be independent and identically distributed.
+\label{eq:aic}
+\mathrm{AIC} \equiv -2\cdot \ln L + 2 \cdot \kappa
+\end{equation} where $L$ is the likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points, assumed to be independent and identically distributed.
+Model selection is performed by a minimization of the criterion.
 
@@
 \caption{Systems studied in this work.
-(A) Langevin dynamics on the two dimensional M\"{u}ller potential.
+(A) Langevin dynamics on the two dimensional M\"{u}ller potential \cite{Muller1980Reaction}.
 (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E.
@@
 
-The uniform distribution emission model presents a computational challenge: its use requires the calculation of the (hyper)volume of the MSM's states, which, when defined by clustering are high-dimensional Voronoi cells.
+The uniform distribution emission model presents a computational challenge: its use requires the calculation of the (hyper)volume of the MSM's states, which, when defined by clustering, are high-dimensional Voronoi cells.
 While trivial in two or three dimensions, this computational geometry task becomes challenging in high-dimensional settings.
 The computation of such volumes has occupied significant attention in recent years in the computational geometry literature, especially via randomized algorithms\cite{Kannan97, Simonovits03, Lovasz03}.
-We opt to approximate the volumes using naive Monte Carlo rejection sampling, which we find tractable for large systems only when the molecular dynamics dataset is first projected into a suitable small vector space of up to perhaps ten dimensions.
+We opt to approximate the volumes using naive Monte Carlo rejection sampling, which we find tractable for large systems only when the molecular dynamics data set is first projected into a suitable small vector space of up to perhaps ten dimensions.
 
-A further challenge is the procedure by which to model the volume of states which are at the edge of the MSM -- whose Voronoi cells extend to infinity in some direction.
-In these cases the Voronoi cells are of unbounded volume.
-Because the convex hull of our simulation datasets are computationally inaccessible, we define the outer extent of our data sets to be the set of all trial points such that the nearest sampled configuration to the trial point is closer than a certain cutoff, $R$.
+A further challenge is the description of the states that are at the edge of the MSM -- whose Voronoi cells extend to infinity in at least one direction.
+In these cases, the Voronoi cells are of unbounded volume.
+Instead we wish to truncate these states by bounding them by the convex hull of the data set.
+Because the convex hull of our simulation data sets are computationally inaccessible, we define the outer extent of our data sets to be the set of all trial points such that the nearest sampled conformation to the trial point is closer than a certain cutoff, $R$.
 
 For cross validation, we find it essential to use a dense prior during transition matrix estimation.
-Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes with maximum likelihood methods.
-We used a dirichlet prior on each entry of the transition matrix, which
-
-For simplicity, we use the Jeffreys prior on each row of the transition matrix -- a Dirichlet prior with $\alpha=\frac{1}{2}$ -- which can be interpreted as a psuedocount of one half of an observed transition between each pair of states.
+Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes.
+We used a Dirichlet prior with a pseudo-count of $\frac{1}{k}$ for each possible transition, where $k$ is the number of states in the model.
 
@@
 \subsection{M\"{u}ller Potential}
-We simulated two trajectories with $10^6$ steps of Langevin dynamics on the M\"{u}ller potential\cite{}.
+We simulated two trajectories with $10^6$ steps of Langevin dynamics on the M\"{u}ller potential\cite{Muller1980Reaction}.
 Simulations were performed with a timestep $t=0.1$, friction coefficient of $\gamma=10^3$, and $kT=15$.
-The first trajectory was clustered using the $k$-centers clustering algorithm, and state volumes were computed for the uniform emission model using $10^5$ rounds of Monte Carlo rejection sampling with a padding distance of $R=0.2$ around the state centers from the 500 state model to define the dataset's envelope.
+The first trajectory was clustered using the $k$-centers clustering algorithm, and state volumes were computed for the uniform emission model using $10^5$ rounds of Monte Carlo rejection sampling with a cutoff of $R=0.28$ from the 50 state model's centroids.
 The second trajectory was used as a test set.
+All MSMs were built using a lag time of $30 steps$.
 
@@
 \includegraphics[width=3in]{figs_final/mull_likes.png}
-\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states.
-There are many methods that can be used to test the transferability of an MSM.
-We've plotted the AIC, BIC, and cross-validation scores for the M\"uller potential simulations.
-These results indicate that, at least for this dataset, the BIC and AIC may penalize the number of states too heavily.}
+\caption{We calculated the training and test set log-likelihoods and model selection criteria for models built with 50-1000 states using both the uniform ({\bf A}) and multivariate normal ({\bf B}) emission distributions.
+Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray).
+For both likelihoods, the BIC penalizes complexity more than the AIC and test set log-likelihood.
+The optimal window is consistent with the convergence of the relaxation timescales (right frames).}
 \label{fig:mullerlike}
@@
 As shown in \cref{fig:mullerlike}, models built with too few states achieve a drastically reduced likelihood, but above a threshold region the likelihood increases relatively slowly.
-Here, the AIC and BIC's penalty on the number of parameters, which scales with the square of the number of states, begins to dominate.
-The optimal models according to the three methods are between 200 and 1000 states for this system.
-This observation is consistent with the convergence of the relaxation timescales of the models.
+The penalty on the number of parameters in \cref{eq:bic} and \cref{eq:aic} begins to dominate.
+The optimal models, according to the three methods are between 200 and 600 states for this system, which is consistent with the convergence of the relaxation timescales of the models.
 
-Interestingly, the AIC and BIC penalize the larger state models much more than the cross-validation approach.
-This could be due to the simplicity of this system.
-Since the M\"uller potential is only two-dimensional, it is difficult to actually produce a drastically over-fit model.
-We suspect that the cross-validation score on a real protein system would be closer to the BIC and AIC.
-However, we note, that cross-validation requires estimating a model without some subset of the data, which is not feasible for large systems in the data-sparse regime.
+The AIC and BIC penalize the larger state models much more than the test set log-likelihood.
+On the two-dimensional potential, in our data-rich regime it is difficult to produce an over-fit model.
+Cross-validation requires fitting multiple models on subsets of the data, and so is not feasible for larger systems in the data-poor regime.
 
 We also tested the Gaussian likelihood on MSMs built using k-means clustering on the M\"uller potential.
-The likelihood (\cref{fig:kmeans_mull}) peaks at a five-state model and decreases with the addition of further states.
-This maximum can be interpreted by inspecting the k-means cluster centers on the M\"uller potential.
-The five state model is the first model to have a cluster center located in the least stable well and as more states are added, the most stable well ends up being heavily split.
-We hypothesize that this over-splitting in the most stable minima gives rise to the decrease in MSM likelihood, and indicates a deficiency in clustering-based procedures which determine the state space without directly maximizing equations \ref{eq:like_vol} or \ref{eq:like_mvn}.
+The training and test set log-likelihoods, BIC, and AIC are all optimized by a five state model (\cref{fig:kmeans_mull}).
+The decrease in the training set log-likelihood can be explained by a trade-off inherent in \cref{eq:like}, where adding more states leads to an improvement in the emission distribution term but causes the transition matrix term to suffer.
+This trade-off illustrates a deficiency in the step-wise approach in which neither the transition matrix nor the state space is parameterized to explicitly optimize the full \cref{eq:like}.
 
@@
 \centering
-\includegraphics[width=2.5in]{figs_final/kmeans_like2_lbl.png}
-\caption{MSMs were built using K-Means clustering on the M\"uller potential simulations.
-Since K-Means optimizes an objective function that is related to the emission probability, our likelihood picks only a few states as the most likely model.
-Interestingly, the likelihood decreases as you introduce additional states.
-This behavior is due to the state-to-state likelihood decreased rapidly when adding additional states, which may be a result of the large number of states in the most stable well.}
+\includegraphics[width=3in]{figs_final/kmeans_like2_lbl.png}
+\caption{MSMs were built using k-means clustering on the M\"uller potential simulations.
+Since k-means optimizes an objective function that is related to the emission probability, our likelihood picks only a few states as the most likely model.
+The likelihood decreases as new states are introduced, due to a competition between the transition matrix term and emission distribution term in \cref{eq:like}.}
 \label{fig:kmeans_mull}
@@
 
-To test the procedure on a larger protein system, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E.
+We reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E.
 Shaw Research \cite{Shaw2010Atomic}.
-Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification}, extracting only the the four slowest uncorrelated linear combinations of residue-residue distances.
-Using the Gaussian emission model, we compare both models built in the 3 dimensional tICA subspace and those built using k-centers clustering with the cartesian root-mean squared deviation (RMSD) over the Fip35 WW domain's 252 non-redundant heavy atoms, computed using the quaternion characteristic polynomial method\cite{Theobald2005Rapid}.
+Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification}, extracting only the four slowest uncorrelated linear combinations of residue-residue distances.
+All MSMs were built using a lag time of $50 ns$.
 
-How many states are required in a Markov state model?
-The two 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain simulated by Shaw et.
-al.\cite{Shaw2010Atomic} have been used to parameterize multiple Markov state models.
-\citet{Lane2011Markov} proposed a 26,000 state MSM after clustering the data by RMSD, whereas \citet{Kellogg2012Evaluation} clustering on contact maps, present a 1000 state model.
-We find the optimal number of states to be highly dependent on space in which the data is clustered.
-With RMSD, the Gaussian likelihood indicates that Markov models built with on the order of 5,000 states optimally balance model quality and generalization error.
-On the other hand, the tICA method's identification of slow order parameters in the simulations has the effect of reducing the number of states necessary when constructing Markov models: here we find high-quality models can be constructed using only on the order of 500 states.
-Even with 200-600 states, Markov models constructed using tICA are able to accurately identify the simulations folding process on the 1-10 $\mu s$ timescale, as well as the faster near-native dynamics identified in this dataset by \citet{McGibbon2013Learning}.
+The complexity of evaluating the Gaussian emission model likelihood does not increase with respect to dimension.
+Thus we used the Gaussian emission model with both the four-dimensional tICA models and those built using k-centers clustering with the minimal cartesian root-mean squared deviation (RMSD) over Fip35's 252 non-redundant heavy atoms, computed using the quaternion characteristic polynomial method\cite{Theobald2005Rapid}.
+For evaluating the Gaussian emission likelihood, we used a value of $3 \cdot 252 - 6 = 750$ for the dimension of the emission distribution.
 
@@
 \includegraphics[width=3in]{figs_final/ww_likes.png}
-\caption{For models built using tICA, we were able to compute volumes as the dimensionality was low (A).
-These likelihoods were qualitatively similar to the MVN likelihoods (B).
-Because the MVN likelihood is easy to compute, we can use it to calculate the likelihood of a model clustered using the RMSD distance metric (C).
-Our likelihood indicates we should construct a model with 5000 states in the RMSD metric, whereas previous studies used a model with 26,000 states \cite{Lane2011Markov}.}
-\label{fig:ww}
+\caption{On models whose states were defined after projecting into the four-dimensional tICA subspace, we computed likelihoods based on the uniform ({\bf A}) and Gaussian ({\bf B}) emission models.
+Models built using the RMSD metric were evaluated with the Gaussian emission model ({\bf C}).
+Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray).
+Test set log-likelihoods are omitted in {\bf C} as the dense matrix algebra is intractable for $k > 10^3$.\label{fig:ww}}
 \end{figure}
+
+For the models built in the tICA subspace, the AIC is comparable to the test set log-likelihood scores using either emission model, displaying a optimum at ~600 states (\cref{fig:ww}).
+The BIC penalizes complexity more strongly with an optimum at ~150 states.
+This window is consistent with the convergence of the relaxation timescales and the onset of Markovian behavior.
+Models built with RMSD require significantly more states than the tICA MSMs and show poor timescale convergence.
+These results indicate why model selection based on relaxation timescales is so difficult, as a less than optimal state decomposition can result in timescales that do not converge rapidly (if at all).
+Furthermore, selection based on only the relaxation timescales contains no guard against over-fitting, which is of particular concern in the models built using RMSD that require thousands of states.
+
+The two 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain simulated by \citet{Shaw2010Atomic} have been used to parameterize multiple Markov state models.
+\citet{Lane2011Markov} proposed a 26,000 state MSM after clustering the data by RMSD, whereas Kellogg et al.,\cite{Kellogg2012Evaluation} clustering on contact maps or secondary structure, proposed 100 and 175 state models, respectively.
+Consistent with this spread, we find the optimal number of states to be highly dependent on the space in which the data is clustered.
+With RMSD, the Gaussian likelihood indicates that Markov models built with on the order of 5,000 states optimally balance model quality and generalization error.
+On the other hand, the tICA method's identification of slow order parameters in the simulations has the effect of reducing the number of states necessary when constructing Markov models.
+Even with 200-600 states, Markov models constructed using tICA are able to accurately identify the simulations folding process on the 1-10 $\mu s$ timescale, as well as the faster near-native dynamics identified in this data set by McGibbon and Pande\cite{McGibbon2013Learning}.
 
@@
 
-The likelihood functions described herein permits the comparison of Markov state models with varying number of states, however it requires the definition of a vector space over which the emission model is defined.
-As such, models built on different projections of an underlying space cannot be directly compared.
-Furthermore, this approach is not amenable to optimizing the Markov lag time, the discretization interval for the Markov jump process.
-Nonetheless, we anticipate that automatic selection of the number of states in a Markov state model will remove much of the guess work in MSM construction, as the selection of the lag time is more amenable to existing Chapman-Kolmogorov tests.
+The likelihood functions described herein permit the comparison of Markov state models with varying number of states, however, they require that the compared models have the same support.
+As such, a direct comparison of the likelihoods between models built with tICA and models built with RMSD is not possible.
+Furthermore, the uniform emission model is intractable for all but the simplest systems.
+The Gaussian emission model does not have this limitation.
 
-In the future, we plan to extend this work to the consideration of models without discrete states, where the requirement that states strictly partition phase space into a set of discrete indicator functions is relaxed.
+In the future, we plan to extend this work to the consideration of models without discrete states, where the requirement that states strictly partition phase space into a set of discrete indicator functions is relaxed and the models are parameterized by a direct optimization of \cref{eq:like}.
+This strategy would complement approaches that generalize MSMs beyond discrete states\cite{Noe2013Variational}.
 
@@
 
-Markov State Models are powerful and popular tool for the analysis of molecular simulations, with a growing literature and two open source source software implementations\cite{Beauchamp2011Msmbuilder2, Senne2012EMMA}.
-There are, however a number of steps in the construction process that require human intervention, which limits the use of MSMs to experts and introduces significant biases into the model building process.
-Additionally, the ability to automatically construct Markov state models on the fly while simulations are in progress, an important point for so-called adaptive sampling procedures\cite{Bowman2010Enhanced}, is hampered when manual model selection is required.
-In the finite-data regime, additional complexity in the Markov model, in the form of increasing number of states, does come at a statistical cost.
-These results take a step toward fully automating the MSM construction process and controlling this statistical error by providing a quantitative method for by which the most suitable discretization procedure can be selected.
+Markov State Models are powerful and popular frameworks for the analysis of molecular simulations, with a growing literature and two open source software tools\cite{Beauchamp2011Msmbuilder2, Senne2012EMMA}.
+There are, however a number of steps in the construction process that require hand-tuning, which limits the use of MSMs to experts and introduces significant biases into the model building process.
+Additionally, the ability to automatically construct MSMs on the fly while simulations are in progress, an important point for so-called adaptive sampling procedures\cite{Bowman2010Enhanced}, is hampered when manual model selection is required.
+
+These results take a step toward fully automating the MSM construction process and controlling the complexity and generalization error by providing a quantitative method for selecting the most suitably complex model, given the data.
+
+\begin{acknowledgement}
+It is an honor and a pleasure to contribute this paper to the Swope Festschrift.
+I (V.S.P.) have been working with Bill since my early days at Stanford and have benefitted greatly from our numerous interactions with him.
+Bill's insight, clarity of thought, and careful approach has influenced my lab's work very broadly.
+I am very pleased that our collaborations continue through today and look forward to many more in the future.
+\end{acknowledgement}
 
@@
 
+In the finite-data regime, additional complexity in the Markov model, in the form of increasing number of states, does come at a statistical cost.
 \begin{itemize}
@@
 When you ask both a ``properly-fit'' model and an overfit model a question that basically falls within the training set -- and neither of them need to extrapolate or really infer out-of-sample information, they'll both do fine.
-Heck, the overfit model might do better at telling you exactly what was \emph{in} the dataset.
+Heck, the overfit model might do better at telling you exactly what was \emph{in} the data set.
 \item The folding timescale is relatively insensitive to differences in MSM construction methodology.
-This has been observed in many datasets (McGibbon 2013, others).
-We hypothesize that this behavior is due to the large structural deviations at play -- even a two state model, given the same dataset, could basically get the right timescale just from counting the number of transitions.
+This has been observed in many data sets (McGibbon 2013, others).
+We hypothesize that this behavior is due to the large structural deviations at play -- even a two state model, given the same data-set, could basically get the right timescale just from counting the number of transitions.
 But finding these other timescales that are structurally subtle is different challenge.
 Also, how robust are other properties like experimental projections or the MFPT distribution to the number of states?
-\item Single trajectory datasets are ``easier'' for MSMs.
+\item Single trajectory data-sets are ``easier'' for MSMs.
 One of the uses of MSMs is to ``stitch together'' multiple trajectories -- this function only happens when the states are large enough that they \emph{connect} trajectories, so you can't have too many states in this case.
@@
 This constant is set so that the distribution integrates to $1$, and is thus the reciprocal volume of the microstate.
+
+%\begin{figure}
+%\centering
+%\includegraphics[width=3.1in]{figs/overfitting.png}
+%\caption{An example of overfitting.
+Left: in the presence of noise, a more complex model is able to fit the observed training data better, with lower residues and a higher empirical likelihood, %but fails to distinguish the signal from the noise.
+Right: more complex model classes exhibit lower training errors, but generalize poorly to unobserved data.}
+%\end{figure}

### 2014-02-06 reviewer-comments/letter.tex
Added reviewer comments

\documentclass{letter}
\usepackage{hyperref}
\address{Clark Center S295 \\ Stanford University \\ Stanford, CA, 94305}
\signature{Robert T. McGibbon, Christian R. Schwantes, Vijay S. Pande}
\begin{document}

\begin{letter}{Prof. Sharon Hammes-Schiffer \\ Deputy Editor of JPC B \\ The Journal of Physical Chemistry}
\opening{Editor:}

Reviewer 1

\begin{quote}
In ``Optimal parameter selection in Markov state models for biomolecular conformational dynamics'', the authors elaborate on an earlier likelihood-based approach to the selection of the appropriate number of discrete states for crisp partitionings to attempt to obtain a Markov model that balances accuracy with the potential for overfitting that many-state models are capable of.  Three different critera are examined: a leave-some-out cross-validation likelihood estimate, as well as two criteria that penalize overfitting by using assumptions of asymptotic normality of the likelihood function (AIC and BIC).  Two different "emission probabilities" for the probability of observing a sample $x_t$ from a given discrete state s are considered that would produce a generative model---one that can be used to assign new data to a model: a uniform probability and Gaussian probability.  These are compared for consistency with each other, as well as compared with well-established metrics such as the stability of implied timescales of the model with the number of states.
\end{quote}

\begin{quote}
 This manuscript will be a useful contribution to the literature on Markov state model construction, but has a number of deficiencies that hinder publication in its current form.  Once these are addressed, it could be suitable for publication in JPC B.
\end{quote}

\begin{quote}
Most critically, the manuscript does not clearly explain the described work, fully justify the conclusions, or provide sufficient detail to reproduce the results.  A number of points below help address these deficiencies:
\end{quote}

\begin{quote}
The title ``Optimal parameter selection...'' does not reflect the work contained in the paper, which focuses only on the selection of the number of states.  Should the authors which to keep the general title, more work on selecting other kinds of parameters should be included; otherwise, the title should be amended to something like ``Optimal selection of the number of states... using a likelihood approach...''
\end{quote}

\begin{quote}
"As such, the eigenvalues, and correspondingly the relaxation timescales, should remain constant as the lag time increases."  The eigenvalues of which matrix should remain constant with what parameter?  The eigenvalues of T(t) actually change with t.  Perhaps you mean the eigenvalues of some implied rate matrix K where T(t) = exp[Kt]?
\end{quote}

\begin{quote}
``choosing models whose relaxation timescales have converged''  Give definition for ``relaxation timescales''.
\end{quote}

\begin{quote}
``With s(x) as the function mapping conformations into the indicator function basis set" Unclear: do you mean s(x) is the integer index of the basis function that is nonzero (has support) at x?  Please be precise.  The lack of precision in this manuscript substantially hampers readability and comprehension.
\end{quote}

\begin{quote}
Eq. 1: This neglects a term for the initial conditions $x_0$.  Eq. 1 actually refers to the conditional probability $P[{x_t}|x_0]$.  Is this the quantity you intend to work with, rather than $P[{x_t}]$?  If so, why?
\end{quote}

\begin{quote}
``By convention, we define $P(x | s(x))$ to be zero for all $x$ such that $s(x) \ne s_i$'' Be precise!  This statement is meaningless.  Do you mean ``we define $P(x | i) = 0$ whenever $s(x)\ne i$?''
\end{quote}

\begin{quote}
``We propose two possible emission distributions, $P(x | s(x))$''  Write this as $P(x | s)$ or $P(x | i)$, to make it clear that your generative probability model MUST be a probability over configurations $x$ given a discrete integer index $s$ or $i$.
\end{quote}

\begin{quote}
Eq. 3: Rewrite as $P(x|s)$ instead of $P(x_t|s(x_t))$, since these are supposed to be generative distributions for x given state index s.
\end{quote}

\begin{quote}
"A more tractable alternative is a Gaussian emission model..."  Why is this even anticipated to be a good model?  Won't it force your states to have gaussianlike clusters in them?  And isn't a gaussian infinite in support, while you SPECIFICALLY REQUIRE the state membership functions to have finite support because they form a crisp partitioning?  Why are you not truncating this model at the state boundaries, which would affect your normalization of the probability density in Eq 5.  This inconsistency seems like a major flaw in this work.
\end{quote}

\begin{quote}
"We assume that the overlaps are small..."  Is this enforced by some condition on sigma and mu for neighboring states?  Has this assumption been checked?  This is certainly, absolutely, positively not the case for even the simple model in Fig 3A.
\end{quote}

\begin{quote}
On choice of emission probabilities, an obvious question: Why not choose $P(x|i) \propto exp[-\beta U(x)] \delta_{i,s(x)}$, where $U(x)$ is the potential energy and $\beta$ the inverse temperature?  The normalization constants are ALSO difficult to compute, but have the advantage that they are proportional to the stationary probability elements computed from the transition matrix $T$.
\end{quote}

\begin{quote}
"Unfortunately such approaches are intractable for problems of the size encountered in biomolecular simulations because of the need to integrate over all possible Markov models with a given number of parameters."  Is this really intractable?  That's a very strong claim.  Perhaps it is just difficult or complex, since no evidence of intractability is presented?
\end{quote}

\begin{quote}
"N is the number of data points, assumed to be independent and identically distributed"  How reasonable is this assumption for molecular simulation data?  How do you even quantify the number of "iid normally distributed" data points in a molecular simulation?
\end{quote}

\begin{quote}
``For cross validation, we find it essential to use a prior...''  Give complete equation for log likelihood.  How do you initialize and sample posterior?  Or maximize likelihood?
\end{quote}

\begin{quote}
``We simulated two trajectories with ... Langevin dynamics...''  Which langevin integration scheme was used!  How were the initial conditions selected?
\end{quote}

\begin{quote}
* ``The first trajectory was clustered using the k-centers clustering algorithm...'' Cite.  What parameters were used for clustering?
\end{quote}

\begin{quote}
``The volumes did not change drastically...'' What does this mean, quantitatively?
\end{quote}

\begin{quote}
``...the likelihood remained essentially constant...'' What does this mean, quantitatively?
\end{quote}

\begin{quote}
according to the three methods" Which three methods exactly?
\end{quote}

\begin{quote}
Fig. 2: This figure is way too small.  What is "score"?  Is this a log likelihood?  Is lower or higher score representative of lower or higher likelihood?  The numbers/units of likelihood are actually meaningful, and should be shown.  What is the difference between "volume" and "Gaussian"?
\end{quote}

\begin{quote}
Fig. 2 caption: "form the boundaries of the optimal window"  Why is this region the optimal window?  It looks like the optimal BIC window is 100-200.  And why not just pick THE single optimal number of states?  Why do you need a window that contains highly nonoptimal models?
\end{quote}

\begin{quote}
"The AIC and BIC penalize the larger state models much more than the test set log-likelihood..."  Why is this discrepancy so large?  Doesn't this suggest AIC/BIC are unsuitable if they are not representative of the actual out-of-sample performance on the test set?
\end{quote}

\begin{quote}
"causes the transition matrix to suffer"  Clarify?
\end{quote}

\begin{quote}
For Mueller potential with Gaussian likelihood: Would be useful to show log-likelihood as a function of number of states so we can see how much the model prefers 5 states.
\end{quote}

\begin{quote}
Fig. 3: "Since k-means optimizes an objective function that is related to the emission probability" What is this objective function?  As I understand it, k-means simply minimizes the sum of intracluster variances, which has nothing to do with your choice of emission probability being uniform or Gaussian or otherwise. Or do you mean you optimize the log-likelihood form of Eq 1 using a voronoi decomposition based on generators that are optimized for each number of states?  Also: "The likelihood decreases as new states are added..."  Which likelihoods are printed in the figure panels?  Test set, uniform, or Gaussian emission probabilities?
\end{quote}

\begin{quote}
> * "...AIC is comparable to the test set log-likelihood...The BIC penalizes complexity more strongly..." Why the huge difference between the AIC and BIC?
\end{quote}

\begin{quote}
Fig. 4: Too small.  Missing figure title, like "Comparison of likelihood models for Fip35 WW domain trajectory data".  "Test set log-likelihoods are omitted in C as the dense matrix algebra is intratable for $k > 10^3$".  Intractable?  Really?  But this is just summing log $T_{ij}$ over N elements, isn't it?
\end{quote}

\begin{quote}
``For tICA, this window is consistent with the convergence of the relaxation timescales"  So....looking at convergence of the relaxation timescales is just as good as the new proposed likelihood method?
\end{quote}

\begin{quote}
``Lane et al. proposed...whereas Kellogg et al...Consistent with this spread, we find the optimal number of states to be highly dependent on the space in which the data is clustered."  Were these two examples using the same metric of model quality and criteria for selecting the number of states?
\end{quote}

\begin{quote}
``These results indicate why likelihood-based model selection is superior to approaches relying only on the timescale convergence.''  How?  I don't see clear evidence of this.  Which likelihood model is superior specifically?  How is this superior to choosing the smallest number of states for which the timescales become flat (assuming that such a thing exists)?  The only evidence presented here has judged success by examining concordance among AIC/BIC/leave-some-out cross validation and the timescale metric.  In Fig 4, BIC and AIC disagree, while AIC tracks (but does not quantitatively agree with) the test set method.  The agreement with the timescale convergence is shown as evidence of correctness.  But it seems like SOMETHING is missing to show that this approach (or, more properly, ONE of these likelihood approaches) is superior to simply checking when the timescales go flat.  A much more convincing test would be to show that these methods allow one to achieve smaller model error (eg in reproducing some quantitative features of the dynamics---maybe even the timescales?) for some smaller quantity of data when compared with a ``gold standard'' enormous dataset.  The behavior of the optimal number of states as a function of the quantity of data would also be of great interest.
\end{quote}

\begin{quote}
``These results take a step toward fully automating the MSM construction process...by providing a quantitative method for selecting the most suitable complex model.''  Which method in particular is recommended?  The Gaussian method?  AIC or BIC?  I'm still confused about what the actual, concrete demonstration of superiority is and which method the authors suggest should be used.
\end{quote}

\begin{quote}
Minor issues that the authors should address are below:
Abstract: ``in regimes in which these techniques encourage over-fitting''  It is unclear precisely what ``these techniques'' refers to.
\end{quote}

\begin{quote}
Two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space..."  There is a substantial quantity of literature on the use of basis functions that are not indicator functions (i.e. "fuzzy" rather than "crisp" partitionings) that should be mentioned.  See, for example, the work of Marcus Weber and Susanna Roeblitz (nee Kube).
\end{quote}

\begin{quote}
The model has the training data set as its support rather than phase space, which makes it unable to generalize and assign probability to new data."  It may also be worth noting that this is not a generative model, since the next paragraph states that a generative model is desired.
\end{quote}

\begin{quote}
``but is intractble in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation"  Be clear that *exact* volume computation is intractable in very high dimension due to the complexity of all known algorithms.  Monte Carlo schemes (such as the one used here, or methods such as [http://www.cs.elte.hu/~lovasz/vol4-focs-tr.pdf]) allow the approximation of this volume.  However, it's also worth noting that the Gaussian model you propose in Eq. 4 is ALSO INTRACTABLE for EXACTLY THE SAME REASON, since you do not compute the true normalizing constant for essentially identical reasons of computational tractability in high dimension.  As such, phrases such as "A more tractable alternative is a Gaussian emission model.." are entirely disingenuous and misleading.
\end{quote}

\begin{quote}
``All MSMs were built using a lag time of 50ns''  Why 50 ns?
\end{quote}

\begin{quote}
``identify the simulations folding process''  Something isn't quite right with this phrase.  What is being referred to?
\end{quote}

\begin{quote}
``the relaxation timescales can continually increase with the number of states used''  Even in the limit of an infinite quantity of data, the eigenvalue error bound theorems from Schuette et al. note that this behavior is precisely what we should expect, since the approximation error decreases as the eigenfunctions are better approximated via more states.  However, the statistical error grows, though this is not quantified or plotted here.
\end{quote}

Reviewer 2:

\begin{quote}
The authors present an alternative approach for choosing the Markov time for Markov state models that will be of value for automating the construction of these models by removing the need for subjective user input.  I have a few concerns, below.
\end{quote}

\begin{quote}
Line 34: Did you mean sacrificing ``accuracy'' instead of ``complexity''?
\end{quote}

\begin{quote}
Can the authors comment on whether k-means or k-centers is preferable based on their Muller potential results?
\end{quote}

\begin{quote}
The authors have access to many other data sets from their own lab and the Shaw group.  Would it be possible to show results for other systems (maybe one larger system and one system with more disorder)?  I think this would help to establish the generality of their results.
\end{quote}

Reviewer 3:

\begin{quote}
Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics by McGibbon, Schwantes and Pande. I enjoyed reading this paper. It is well suited for Bill Swope's birthday issue. It addresses an important problem in chemical physics, namely choosing amongst the various possible parameter settings when constructing a Markov State Model. However, the following general points should be addressed in a revision:
\end{quote}

\begin{quote}
I find the title “Optimal Parameter Selection in Markov State Models...” inappropriate: It suggests that the ``Parameter selection in MSMs'' is a well-defined problem, and that the solution approach here would be optimal, i.e. not further improvable. Neither is true. I suggest to pick a title that is specific to the content of the paper, e.g. ``Likelihood-approach for selecting the number of states / choosing between discretizations in MSMs''
\end{quote}

\begin{quote}
The paper starts with a discussion of approaches that choose the lag time based on implied timescales, and then proposes a way to choose the number of discrete states as an alternative. I can see the idea behind it, but to general reader it will probably be unclear what the connection between the two is. It should be spelled out that there is a change of viewpoint, where instead of fixing the discretization and varying the lagtime, one fixes the lagtime and varies the discretization. And the hope why both approaches are meaningful is that the implied timescales should converge to their true values for either (1) increasing lag time [see Djurdjevac, Sarich, Schütte, MMS 2011 and Prinz, Chodera, Noe, PRX 2014 (http://arxiv.org/abs/1207.0225)], or (2) better approximation of the dominant eigenfunctions via more discrete states [see (19) and Sarich, Noe, Sch\"{u}tte MMS 8, p1154 (2010)].
\end{quote}

\begin{quote}
My main concern is that the paper suggests that the best result is obtained by maximizing the present likelihood. But stationary and dynamical expectation values such as the relaxation timescales are well defined by the underlying dynamics, and the ultimate objective of any estimation procedure must be to correctly approximate them. Using the present likelihood could systematically give wrong estimates for various reasons. An inappropriate statistical model for the local output function (which is really hard to make) would lead to biased estimates. Then there is uncertainty in choosing the Bayesian criterion - AIC, BIC or something else? If the lag time chosen is too short, it will not lead to correct estimates for feasible numbers of states. The differences in Fig 4 (see Details below) indeed suggests that the present criterion fails to provide the correct estimate in at least some cases. If a new estimator is derived, it needs to be shown that it converges to the correct result in some limit. I suggest to build a fine discrete model, e.g. by direct gridding of the Müller potential, and constructing a simple MCMC dynamics on neighboring gridpoints, and then considering different clusterings on this data set. There, the exact timescales are known, and an embedding in a geometric space is given, such that the method can be applied such that the present method can be applied as is. It then needs to be shown that the present likelihood approach selects MSMs with timescales that coincide with the true values.In addition, it would be useful to compare the results here with implied timescale plots. Do the results agree in the cases where ITS do converge?
\end{quote}

\begin{quote}
Please check the timescales in Fig 2(A,B) and 4(A,B) - they look identical, but shouldn’t. Details / specific points:
\end{quote}

\begin{quote}
``ever-more states''. Ever-more states or ever-larger lag time? (when did number of states come in?)
\end{quote}

\begin{quote}
``Often the choice is made to use the model whose timescales better match an experimental observable:`` I hope that’s not true and I wouldn’t generally accuse the community of practising this. When timescales don’t converge, this is a numerical or statistical problem with the data or the estimators used. Ignoring this fact and deliberately choosing a model that coincidentially delivers the numbers one hopes to get would be simply cheating.
\end{quote}

\begin{quote}
Gaussians could be truncated and normalized - it seems that renormalization would require to solve an even harder problem than for uniform output probabilities, because you’d need to integrate the Gaussian density over a high-dimensional polytope.
\end{quote}

\begin{quote}
``We assume that the overlaps are small'' But are they small in your application? Ensuring this puts requirements on the sizes of states and sigma. How is it checked that this is then a good model?
\end{quote}

\begin{quote}
Are you sure that the timescales shown in Fig 2A(right) and 2B(right) are from different data? They look identical to me.
\end{quote}

\begin{quote}
I’m surprised that there are three slow relaxation processes for the Müller potential before the second gap. The potential looks like there should be only two (due to three metastable states). Were MSMs build with reversible transition matrix estimation, and are all eigenvalues thus real-valued?
\end{quote}

\begin{quote}
Muller potential: lag time was chosen to be 30 steps. How is this choice justified? Since the timescales in Fig. 2 are not independent of N, this must mean that at least for some state decompositions the ITS are not converged in the lag time. In these cases, it’s not even appropriate to describe the state-to-state dynamics by a Markov chain at all. Is the statistical model still meaningful in this case
\end{quote}

\begin{quote}
Müller potential, Gaussian likelihood: How were the Gaussian sigma-parameters chosen? The k-means states in Fig 3 have very different sizes and their centers are sometimes very close, so the assumption that output distributions shall not overlap seems to be hard to realize.
\end{quote}

\begin{quote}
WW domain, lag time was chosen to be 50 ns - same comment as for the Müller potential, above
\end{quote}

\begin{quote}
WW domain, Gaussian likelihood - same comment as for the Müller potential, above
\end{quote}

\begin{quote}
Are you sure that the timescales shown in Fig. 4A(right) and 4B(right) are from different data? They look identical to me.
\end{quote}

\begin{quote}
 Doesn’t Fig 4c suggest that there’s a problem here? For the likelihood-based choice made, the slowest timescale is estimated between 1500 and 3000 ns with RMSD clustering. In contrast, the TICA-based estimates suggest a timescale of around 5000 ns. So if the TICA results is indeed correct, the method should be going for larger state number in the RMSD case.
\end{quote}

\begin{quote}
In some timescale vs number of state plots the timescales descreases with increasing number of states (e.g. Fig 2A). That’s surprising. Can the authors explain that? Has a prior been used that may be responsible for this behavior?
\end{quote}

\closing{Sincerely,}

\end{letter}
\end{document}

### 2014-02-06 reviewer-comments/letter.tex
reviewer comments

@@
 \begin{quote}
-"As such, the eigenvalues, and correspondingly the relaxation timescales, should remain constant as the lag time increases."  The eigenvalues of which matrix should remain constant with what parameter?
-The eigenvalues of T(t) actually change with t.
-Perhaps you mean the eigenvalues of some implied rate matrix K where T(t) = exp[Kt]?
+``As such, the eigenvalues, and correspondingly the relaxation timescales, should remain constant as the lag time increases.''  The eigenvalues of which matrix should remain constant with what parameter?
+The eigenvalues of $T(t)$ actually change with t.
+Perhaps you mean the eigenvalues of some implied rate matrix $K$ where $T(t) = exp[Kt]$?
 \end{quote}
@@
 \begin{quote}
-"A more tractable alternative is a Gaussian emission model..."  Why is this even anticipated to be a good model?
+``A more tractable alternative is a Gaussian emission model...''  Why is this even anticipated to be a good model?
 Won't it force your states to have gaussianlike clusters in them?
@@
 \begin{quote}
-"We assume that the overlaps are small..."  Is this enforced by some condition on sigma and mu for neighboring states?
+``We assume that the overlaps are small...''  Is this enforced by some condition on sigma and mu for neighboring states?
 Has this assumption been checked?
@@
 \begin{quote}
-"Unfortunately such approaches are intractable for problems of the size encountered in biomolecular simulations because of the need to integrate over all possible Markov models with a given number of parameters."  Is this really intractable?
+``Unfortunately such approaches are intractable for problems of the size encountered in biomolecular simulations because of the need to integrate over all possible Markov models with a given number of parameters.''  Is this really intractable?
 That's a very strong claim.
@@
 \begin{quote}
-* ``The first trajectory was clustered using the k-centers clustering algorithm...'' Cite.  What parameters were used for clustering?
+``The first trajectory was clustered using the k-centers clustering algorithm...'' Cite.  What parameters were used for clustering?
 \end{quote}
@@
 2: This figure is way too small.
-What is "score"?
+What is ``score''?
 Is this a log likelihood?
@@
 Fig.
-2 caption: "form the boundaries of the optimal window"  Why is this region the optimal window?
+2 caption: ``form the boundaries of the optimal window''  Why is this region the optimal window?
 It looks like the optimal BIC window is 100-200.
@@
 \begin{quote}
-"The AIC and BIC penalize the larger state models much more than the test set log-likelihood..."  Why is this discrepancy so large?
+``The AIC and BIC penalize the larger state models much more than the test set log-likelihood...''  Why is this discrepancy so large?
 Doesn't this suggest AIC/BIC are unsuitable if they are not representative of the actual out-of-sample performance on the test set?
@@
 \begin{quote}
-"causes the transition matrix to suffer"  Clarify?
+``causes the transition matrix to suffer''  Clarify?
 \end{quote}
@@
 \begin{quote}
-``but is intractble in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation"  Be clear that *exact* volume computation is intractable in very high dimension due to the complexity of all known algorithms.
+``but is intractble in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation''  Be clear that *exact* volume computation is intractable in very high dimension due to the complexity of all known algorithms.
 Monte Carlo schemes (such as the one used here, or methods such as [http://www.cs.elte.hu/~lovasz/vol4-focs-tr.pdf]) allow the approximation of this volume.

### 2014-02-08 doublewell/doublewell.py
Bayes

@@ -1,0 +1,264 @@
+import os
+import IPython as ip
+import numpy as np
+from mdtraj import io
+import scipy.sparse
+import scipy.special
+from msmbuilder import MSMLib as msmlib
+from msmbuilder import  msm_analysis
+from matplotlib import mlab
+import matplotlib.pyplot as pp
+colors = ['#348ABD', 'g', '#7A68A6', '#A60628', '#467821', '#CF4457', '#188487', '#E24A33']
+
+DIFFUSION_CONST = 1
+DT = 1e-3
+SQRT_2D_DT = np.sqrt(2 * DIFFUSION_CONST * DT)
+grad_potential = lambda x : -2 * np.sin(2 * x)
+
+
+def reflect_pbc(x, min, max):
+    if x > max:
+        return 2*max - x
+    if x < min:
+        return 2*min - x
+    return x
+
+
+def propagate(n_steps):
+    import time; start = time.time()
+    n_steps = int(n_steps)
+
+    max = np.pi
+    min = -np.pi
+    rand = np.random.randn(n_steps)
+    x = np.zeros(n_steps+1)
+    for i in range(n_steps):
+        x_i_plus_1 = x[i] -DT * grad_potential(x[i]) + SQRT_2D_DT*rand[i]
+        x[i+1] = reflect_pbc(x_i_plus_1, -np.pi, np.pi)
+        # reflecting PBCs
+
+    print '%d steps/s' % (n_steps / (time.time() - start))
+    return x
+
+
+def exact_solution(n_grid, plot_eigfunctions=False):
+    ONE_OVER_SQRT_2PI = 1.0 / (np.sqrt(2*np.pi))
+    normalpdf = lambda x : ONE_OVER_SQRT_2PI * np.exp(-0.5 * (x*x))
+
+    grid = np.linspace(-np.pi, np.pi, n_grid)
+    width = grid[1]-grid[0]
+    transmat = np.zeros((n_grid, n_grid))
+    for i, x_i in enumerate(grid):
+        for offset in range(-(n_grid-1), n_grid):
+            x_j = x_i + (offset * width)
+            j = reflect_pbc(i+offset, 0, n_grid-1)
+
+            # What is the probability of going from x_i to x_j in one step?
+            diff = (x_j - x_i + DT * grad_potential(x_i)) / SQRT_2D_DT
+            transmat[i, j] += normalpdf(diff)
+        #print transmat[i, :]
[…]

### 2014-02-08 doublewell/doublewell.py
Stuff

@@ -111,6 +111,8 @@ class MarkovStateModel1D(object):
         self.grid_ = None
         self.starting_states_ = None
 
+        self._n_parameters = (n_states * (n_states + 1) / 2) - 1
+
     def fit(self, X):
         """
 
@@ -172,6 +174,40 @@ class MarkovStateModel1D(object):
             return emission_log_likelihood
         return transition_log_likelihood + emission_log_likelihood
 
+    def bic(self, X):
+        """
+        Bayesian information criterion for the current model fit
+        and the proposed data
+
+        Parameters
+        ----------
+        X : list of np.ndarrays
+            Each array is a 1D trajectory through phase space. This should
+            already be subsampled to your discretization interval / lag time
+            of choice.
+
+        Returns
+        -------
+        bic: float (the lower the better)
+        """
+        return -2 * self.loglikelihood(X) + self._n_parameters * np.log(sum(len(xx) for xx in X))
+
+    def aic(self, X):
+        """Akaike information criterion for the current model fit
+        and the proposed data
+
+        Parameters
+        ----------
+        X : list of np.ndarrays
+            Each array is a 1D trajectory through phase space. This should
+            already be subsampled to your discretization interval / lag time
+            of choice.
+
+        Returns
+        -------
+        aic: float (the lower the better)
+        """
+        return -2 * self.loglikelihood(X) + 2 * self._n_parameters
 
     def logevidence(self, terms='all'):
 
@@ -208,7 +244,7 @@ class MarkovStateModel1D(object):
         # this factors out when we integrate over model parameters
         # since the we're NOT integrating over the state-space parameters
         # as they're fixed given the number of states
-        if terms in ['emission' or 'all']:
+        if terms in ['emission', 'all']:
             emission_log_likelihood = - np.log(self.grid_[1]-self.grid_[0]) * \
                                       (np.sum(self.countsmats_) + len(self.countsmats_))
 
@@ -241,8 +277,10 @@ def main():
     model = MarkovStateModel1D(n_states=2)
     t = [trajectories[i][::100] for i in range(2)]
     model.fit(t)
-    print 'loglikelihood', model.loglikelihood(t, 'emission')
-    print 'evidence', model.logevidence('emission')
+    print 'loglikelihood', model.loglikelihood(t)
+    print 'evidence', model.logevidence()
+    print 'bic', -model.bic(t)/2
+    print 'aic', -model.aic(t)/2
 
 
 def test_logevidence():

### 2014-02-09 gridmsm/gridmsm.py
grid msm

@@ -1,0 +1,258 @@
+import numpy as np
+import scipy.special
+from msmbuilder import MSMLib as msmlib
+
+
+class GridMarkovStateModel(object):
+    """Markov State Model using a grid-based discretization
+
+    Parameters
+    ----------
+    n_bins : int
+        Number of bins along per degree of freedom. The total of states for
+        the Markov model will be :math:`n_{bins}^{n_{dofs}}`. Where
+        :math:`n_{dofs}` is the number of degrees of freedom.
+    min : float
+        The bins along each degree of freedom run from min to max.
+    max : float
+        The bins along each degree of freedom run from min to max.
+    prior : float
+        Strength of the symmetric edge reinforced random walk conjugate prior
+        on the (reversible) transition matrix. The interpretation of the prior
+        strength is very analogous to a symmetric Dirichlet prior -- the prior
+        acts like a "psuedocount" on each edge of the graph.
+
+    Attributes
+    ----------
+    """
+    def __init__(self, n_bins, min, max, transmat_prior=1.0):
+        self.n_bins = n_bins
+        self.min = min
+        self.max = max
+        self.transmat_prior = transmat_prior
+
+        self.n_states = None
+        self.n_features = None
+        self.transmat_ = None
+        self.grid = np.linspace(self.min, self.max, self.n_bins + 1)
+
+    def _n_parameters(self):
+        return (self.n_states * (self.n_states + 1) / 2) - 1
+
+    def _initialize_sequences(self, X):
+        if not isinstance(X, list):
+            raise ValueError('X must be a list')
+
+        if self.n_features is None:
+            if not isinstance(X, np.ndarray) or X[0].ndim != 2:
+                raise ValueError("Each element must be a 2d array")
+            self.n_features = X[0].shape[1]
+
+        self.n_states = self.n_bins**self.n_features
+        for xx in X:
+            if not isinstance(xx, np.ndarray) or xx.ndim != 2 or xx.shape[1] != self.n_features:
+                raise ValueError("Each element must be a 2D array "
+                                 "with of shape N by %d" % self.n_features)
+
+    def _discretize(self, datasequence):
+        """Discretize a real-valued data sequence, projecting each observation
+        into an integer-indexed state
+
+        Parameters
+        ----------
+        datasequence : np.ndarray, shape=(n_observations, n_features)
+            A single trajectory: a timeseries of n_dim dimensional observations
+            of the system moving through phase space
+
+        Returns
+        -------
+        labels : np.ndarray
[…]

### 2014-02-09 gridmsm/setup.py

@@ -1,0 +1,15 @@
+"""gridmsm: Markov state models with a grid discretization
+
+More description
+"""
+
+DOCLINES = __doc__.split("\n")
+from distutils.core import setup
+
+setup(name='gridmsm',
+    version='1.0',
+    author='Robert McGibbon',
+    description=DOCLINES[0],
+    long_description="\n".join(DOCLINES[2:]),
+    py_modules = ['gridmsm'])
+      

### 2014-02-09 gridmsm/gridmsm.py
add the tests

@@ -1,6 +1,59 @@
 import numpy as np
 import scipy.special
 from msmbuilder import MSMLib as msmlib
+__all__ = ['GridMarkovStateModel']
+
+
+def _log_transmat_evidence(countsmat, starting_state, prior):
+    """Compute the log evidence of a Markov model given directed transition counts
+    
+    This function implements Eq. 51 from [1], using a symmetric prior (all
+    :math:`\alpha_e`) are equal
+    
+    Parameters
+    ----------
+    countsmat : np.ndarray
+        Dense matrix of directed transition counts
+    starting_state : int
+        The initial state of the chain
+    prior : float
+        Strength of the symmetric prior. This is the starting weight on each
+        edge in the ERRW.
+    
+    References
+    ----------
+    .. [1] Diaconis, Persi, and Silke WW Rolles. "Bayesian analysis for
+           reversible Markov chains." The Annals of Statistics 34.3 (2006):
+           1270-1292.
+    """
+    
+    def logpochhammer(a, n):
+        "Natural logarithm of the Pochhammer symbol (a)_n"
+        return scipy.special.gammaln(a + n) - scipy.special.gammaln(a)
+
+    countsmat = np.asarray(countsmat)
+    n_states = countsmat.shape[0]
+    k_v = np.sum(countsmat, axis=1)
+    k_e = countsmat + countsmat.T
+
+    numerator1 = 0
+    for e in zip(*np.triu_indices(n_states, 1)):
+        # e in E\E_loop
+        numerator1 += logpochhammer(prior, k_e[e])
+
+    numerator2 = 0
+    for k_e in np.diag(k_e):
+        # e in E_loop
+        numerator2 += (k_e/2.0) * np.log(2.0) + \
+                      logpochhammer(prior / 2.0, k_e/2.0)
+
+    denominator = 0
+    for v in range(n_states):
+        offset = int(v != starting_state)
+        denominator += (k_v[v]) * np.log(2.0) + logpochhammer((
+            prior * n_states + offset) / 2.0, k_v[v])
+
+    return (numerator1 + numerator2 - denominator)
 
 
 class GridMarkovStateModel(object):
@@ -207,10 +260,6 @@ class GridMarkovStateModel(object):
             raise ValueError("terms must be one ['transmat', 'emission', 'all']")
         self._initialize_sequences(X)
 
-        def logpochhammer(a, n):
-            "Natural logarithm of the Pochhammer symbol (a)_n"
[…]

### 2014-02-09 gridmsm/test_gridmsm.py

@@ -1,0 +1,13 @@
+import numpy as np
+from gridmsm import _log_transmat_evidence
+
+def test_log_transmat_evidence():
+    "Example from Diaconis and Rolles (2006)"
+    countsmat = np.array([
+        [91, 160, 261, 108],
+        [213, 351, 161, 249],
+        [251, 224, 388, 201],
+        [66, 239, 254, 152]])
+    result = _log_transmat_evidence(countsmat, starting_state=3, prior=1)
+    expected = np.log(2.166939224648291) - 1961 * np.log(10)
+    np.testing.assert_almost_equal(expected, result)

### 2014-02-09 gridmsm/include/cycledeterminant.hpp
stuff

@@ -1,0 +1,22 @@
+#include <vector>
+
+namespace CycleDeterminant {
+class CycleMatrixBuilder {
+
+public:
+    CycleMatrixBuilder(const int n_states, std::string lapack_lite_lib);
+    double logSqrtDetCycleMatrix(const double* u);
+    ~CycleMatrixBuilder();
+
+private:
+    int indexInTriu(int i, int j);
+    double logdet(std::vector<double> const& A);
+
+    const int n_states_;
+    int n_cycles_;
+    void* lapack_lite;
+    std::vector<std::vector<int> > cycles_;
+    int (*dgetrf_)(const int*, const int *, const double *, const int *, int *, int *);
+};
+
+}; // closes namespace

### 2014-02-09 gridmsm/setup.py

@@ -4,12 +4,23 @@ More description
 """
 
 DOCLINES = __doc__.split("\n")
-from distutils.core import setup
+from distutils.core import setup, Extension
+import numpy as np
+from Cython.Distutils import build_ext
+
+map_transmat = Extension('gridmsm._map_transmat',
+    language='c++',
+    sources=['src/map_transmat.pyx', 'src/cycledeterminant.cpp'],
+    include_dirs=[np.get_include(), 'include']
+)
 
 setup(name='gridmsm',
     version='1.0',
     author='Robert McGibbon',
     description=DOCLINES[0],
     long_description="\n".join(DOCLINES[2:]),
-    py_modules = ['gridmsm'])
+    py_modules = ['gridmsm'],
+    ext_modules=[map_transmat],
+    cmdclass={'build_ext': build_ext})
+    
       

### 2014-02-09 gridmsm/src/cycledeterminant.cpp

@@ -1,0 +1,104 @@
+#include <cmath>
+#include <string>
+#include <stdexcept>
+#include <vector>
+#include <iostream>
+#include <dlfcn.h>
+#include "cycledeterminant.hpp"
+namespace CycleDeterminant {
+
+
+CycleMatrixBuilder::CycleMatrixBuilder(const int n_states, std::string lapack_lite_lib)
+    : n_states_(n_states),
+      n_cycles_(0)
+{
+    for (int i = 1; i < n_states_; i++) {
+        for (int j = i+1; j < n_states_; j++) {
+            std::vector<int> cycle;
+            cycle.push_back(indexInTriu(0, i));
+            cycle.push_back(indexInTriu(i, j));
+            cycle.push_back(indexInTriu(0, j));
+            cycles_.push_back(cycle);
+        }
+    }
+    n_cycles_ = cycles_.size();
+
+    //void* pyhandle = dlopen(python_lib.c_str(),  RTLD_LAZY | RTLD_LOCAL);
+    //if (!pyhandle)
+    //    throw std::runtime_error(dlerror());
+    lapack_lite = dlopen(lapack_lite_lib.c_str(), RTLD_LAZY | RTLD_LOCAL);
+    if (!lapack_lite)
+        throw std::runtime_error(dlerror());
+    //dlclose(pyhandle);
+    dgetrf_ = (int (*)(const int*, const int *, const double *, const int *, int *, int *)) dlsym(lapack_lite, "dgetrf_");
+    if (!dgetrf_)
+        throw std::runtime_error(dlerror());
+}
+
+CycleMatrixBuilder::~CycleMatrixBuilder() {
+    dlclose(lapack_lite);
+}
+
+double CycleMatrixBuilder::logSqrtDetCycleMatrix(const double* u) {
+    std::vector<double> A(n_cycles_ * n_cycles_);
+    for (int i = 0; i < n_cycles_; i++) {
+        for (int j = 0; j < n_cycles_; j++) {
+            if (i == j) {
+                A[i*n_cycles_ + i] = \
+                    exp(-u[cycles_[i][0]]) +  exp(-u[cycles_[i][1]]) +  exp(-u[cycles_[i][2]]);
+            } else {
+                for (int x = 0; x < 2; x++)
+                    for (int y = 0; y < 2; y++)
+                        if (cycles_[i][x] == cycles_[j][y])
[…]

### 2014-02-09 gridmsm/src/map_transmat.pyx

@@ -1,0 +1,115 @@
+from __future__ import division, print_function
+import os
+import sys
+import numpy as np
+import scipy.special
+from scipy.misc import logsumexp
+from mixtape._reversibility import reversible_transmat_likelihood, logsymsumexp
+
+cimport numpy as np
+from libcpp.string cimport string
+
+cdef extern from "cycledeterminant.hpp" namespace "CycleDeterminant":
+    cdef cppclass CycleMatrixBuilder:
+        CycleMatrixBuilder(const int n_states, string lapack_lite_lib) except +
+        double logSqrtDetCycleMatrix(const double* u)
+        
+    
+cdef class MAPTransmat:
+    cdef int n_states
+    cdef tuple triu_indices
+    cdef np.ndarray loop_indices
+    cdef np.ndarray edge_indices
+    cdef CycleMatrixBuilder * cycleMatrixBuilder
+
+    def __cinit__(self, n_states):
+        self.n_states = int(n_states)
+        cdef string python_lib = find_python_lib()
+        cdef string lapack_lite_lib = os.path.join(np.linalg.__path__[0], 'lapack_lite.so')
+        self.cycleMatrixBuilder = new CycleMatrixBuilder(n_states, lapack_lite_lib)
+
+        self.triu_indices = np.triu_indices(n_states)
+        # u[loop_indices] are entries of u corresponding to self transitions,
+        # u[edge_indices] are entries of u corresponding to non-self transitions
+        self.loop_indices = np.where(self.triu_indices[0] == self.triu_indices[1])[0]
+        self.edge_indices = np.where(self.triu_indices[0] != self.triu_indices[1])[0]
+        
+
+    def fit(self, countsmat, prior):
+        """Maximum a-posteriori reversible transition matrix given a set of
+        directed transition counts and a symmetric edge-reinforced random walk
+        prior.
+    
+        Parameters
+        ----------
+        countsmat : np.ndarray
+            Dense matrix of directed transition counts
+        prior : float
+            Strength of the symmetric prior. This is the starting weight on each
+            edge in the ERRW.
+        """
+        symcounts = (countsmat + countsmat.T)[self.triu_indices]
+        symcounts[self.loop_indices] -= np.diag(countsmat)
+        rowsums = np.sum(countsmat, axis=1)
+        logrowsums = np.log(rowsums)
+        u0 = np.log(symcounts)
+
[…]

### 2014-02-09 gridmsm/src/map_transmat.pyx
remove unused

@@ -24,7 +24,6 @@ cdef class MAPTransmat:
 
     def __cinit__(self, n_states):
         self.n_states = int(n_states)
-        cdef string python_lib = find_python_lib()
         cdef string lapack_lite_lib = os.path.join(np.linalg.__path__[0], 'lapack_lite.so')
         self.cycleMatrixBuilder = new CycleMatrixBuilder(n_states, lapack_lite_lib)
 
@@ -85,31 +84,3 @@ cdef class MAPTransmat:
         term4 = self.cycleMatrixBuilder.logSqrtDetCycleMatrix(&u[0]);
         
         return term4;
-
-
-def find_python_lib():
-    """Find the python library (libpython.so / libpython.dylib) for the
-    current process. This doesn't support windows.
-
-    Adapted from PySide's setup.py script
-    """
-    py_version = "%s.%s" % (sys.version_info[0], sys.version_info[1])
-    py_libdir = os.path.join(sys.prefix, "lib")
-
-    lib_exts = ['.so', '.so.1']
-    if sys.platform == 'darwin':
-        lib_exts.append('.dylib')
-    if sys.version_info[0] > 2:
-        lib_suff = getattr(sys, 'abiflags', None)
-    else: # Python 2
-        lib_suff = ''
-
-    libs_tried = []
-    for lib_ext in lib_exts:
-        lib_name = "libpython%s%s%s" % (py_version, lib_suff, lib_ext)
-        py_library = os.path.join(py_libdir, lib_name)
-        if os.path.exists(py_library):
-            return py_library
-        libs_tried.append(py_library)
-
-    raise ValueError('Could not find libpython. Tried looking at %s' % ','.join(libs_tried))

### 2014-03-19 paper/paper2.tex
Typo fix
minor

@@
 %\documentclass[twocolumn,floatfix,nofootinbib,acs]{revtex4-1}
-\documentclass[journal=jpcbfk, layout=twocolumn, manuscript=article]{achemso}
+\documentclass[journal=jpcbfk, layout=traditional, manuscript=article]{achemso}
 \setkeys{acs}{articletitle = true}
@@
 
-\title{Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics}
+\title{Statistical Model Selection for Markov Models of Biomolecular Dynamics}
 \author{Christian R. Schwantes}
@@
 \email{pande@stanford.edu}
+\keywords{MSM, molecular dynamics, AIC, BIC, Bayes factor}
 
@@
 However, to achieve satisfactory performance, existing methodologies often require expert intervention when defining the model's discrete state space.
-Here, we show that by consideration of the states' conditional distribution over conformations, the state space can be evaluated and optimized using the Markov state model's likelihood function and standard statistical model selection tools.
+While standard model selection methodologies focus on the minimization of systematic bias and disregard statistical error, we show that by consideration of the states' conditional distribution over conformations, both sources of error can be balanced evenhandedly.
 Application of these techniques on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those techniques encourage over-fitting.
@@
 \section{Introduction}
-Protein are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function.
+Proteins are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function.
 These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native sub-states, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}.
@@
 Computer simulations can complement experiments by providing atomic insight into conformational dynamics.
-With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation ensembles\cite{Gotz2012Routine, Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}.
-In fact, the minimally-biased kinetic analysis of such simulations is often a central bottleneck and presents a major challenge to the field.
+With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation ensembles with relative ease\cite{Gotz2012Routine, Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}.
+A central challenge in the field of molecular simulations is now often the kinetic analysis of simulation datasets.
 The analysis paradigms often entail the construction of lower resolution models parametrized from the high resolution simulation data set which capture the essential features in an interpretable framework\cite{Freddolino2010Challenges, Lane2013Milliseconds}.
@@
 
-Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models\cite{Chodera2007Automatic, Prinz2011Markov, Beauchamp2012Simple, Bowman2013Quantitative}.
-Such detailed models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity.
-In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states.
-The two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics.
-The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors to identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{}.
+Markov state models (MSMs) are one approach for analyzing MD data sets that are able to smoothly move between high and low-resolution models\cite{Chodera2007Automatic, Prinz2011Markov, Beauchamp2012Simple, Bowman2013Quantitative}.
+High-resolution  models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity.
+In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states which partition the phase space.
+Two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics.
+The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{}.
+
+Some explanation of notation...
+\begin{align}
+s(t) &= \sigma(x(t)), \\
+s(t+\tau) &\sim \mathbf{T}_{s(t), k} \sigma_k.
+\end{align}
 
 A significant challenge in the automated construction of Markov state models is the choice of the number of states\cite{McGibbon2013Learning}.
-Although classical Hamiltonian dynamics form a continuous-time Markov chain in $\mathbb{R}^{6N}$, the Markov property does not hold after the projecting the dynamics onto a basis of discrete indicator functions.
-In particular, when states contain within them free energy barriers of substantial magnitude, the validity of the Markov assumption begins to suffer considerably.
-While this source of modeling error can be addressed by increasing the number of microstates, the reduction in one error comes at the expense of the increase in another.
+Although classical Hamiltonian dynamics with $N$ particles form a continuous-time Markov chain in $\mathbb{R}^{6N}$\cite{vanKampen}, the projection of this high-dimensional process into a lower dimensional space does (such as the discrete indicator functions used by MSMs) generally experiences non-Markovian dynamics.
+In the Morri-Zwanzig approach, this corresponds to the generalized Langevin equation with a nontrivial memory kernel\cite{ZM}.
+Within the MSM context, on the other hand, this means that first order Markov chains incur a systematic error in their modeling of the high-dimensional system.
+In particular, when MSM states contain within them kinetic barriers of substantial magnitude, the bias due to the first-order Markov approximation is large.
+While this source of modeling error can be addressed by increasing the number of states, the reduction in one error comes at the expense of the increase in another.
 This second source of error is statistical in origin.
@@
 %After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.
-In the context of MSMs, model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time.
+MSM model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time.
 As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
@@
 That is, the likelihood of observing a conformation in phase space given that the conformation is assigned to state $i$ is $0$ if the conformation is outside of the bounding volume of the state and constant if the conformation is within the volume.
-This constant is set so that the distribution integrates to $1$, and is thus the reciprocal volume of the microstate.
+This constant is set so that the distribution integrates to $1$, and is thus the reciprocal volume of the state.
 

### 2014-03-24 paper/first_submission/paper2.tex
moved things around for revision

--- 2013-11-28 paper/paper2.tex
@@
 \begin{abstract}
-Markov state models provide a powerful framework for the analysis of biomolecular conformation dynamics in terms of their metastable states and transition rates.
-These models provide both a quantitative and comprehensible description of the long-timescale dynamics of large molecular dynamics with a Master equation, and have been successfully used to study protein folding, protein conformational change, and protein-ligand binding.
-However, to achieve satisfactory performance, existing methodologies often require expert intervention when defining the model's discrete state space.
-Here, we show that by consideration of the states' conditional distribution over conformations, the state space can be evaluated and optimized using the Markov state model's likelihood function and standard statistical model selection tools.
-Application of these techniques on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those techniques encourage over-fitting.
-By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markvov state model construction.
+Markov state models (MSMs) are a powerful tool for the analysis of molecular dynamics simulations, but have been hampered by the need for manual selection of the number of states.
+We report a new method for the optimal selection of the number of states in an MSM based on the Bayesian information criterion.
+We demonstrate the approach on three systems of increasing complexity...
 \end{abstract}
@@
 Protein are highly complex molecular machines, and their dynamics are an essential aspect of biomolecular function.
-These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native sub-states, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}.
+These dynamics span a wide range of length scales, timescales and complexity, including folding and aggregation, conformational change between functional native substates, ligand binding, and allostery \cite{Dobson2003Protein, Kim2008Real, Austin1975Dynamics, Bahar2007Intrinsic}.
 Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}.
@@
 Computer simulations can complement experiments by providing atomic insight into conformational dynamics.
-With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation ensembles\cite{Gotz2012Routine, Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}.
+With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation data sets\cite{Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}.
 In fact, the minimally-biased kinetic analysis of such simulations is often a central bottleneck and presents a major challenge to the field.
@@
 
-Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models\cite{Chodera2007Automatic, Prinz2011Markov, Beauchamp2012Simple, Bowman2013Quantitative}.
+Markov state models (MSMs) are one approach for analyzing MD data sets and driving further MD simulations that are able to smoothly move between high and low-resolution models\cite{Pande2010Everything, Beauchamp2012Simple, Prinz2011Markov, Bowman2013Quantitative}.
 Such detailed models maintain quantitative agreement with the underlying simulation data, while low-resolution models capture the salient features of the potential energy landscape, sacrificing some degree of model complexity.
@@
 
-In practice, while this test has been useful for many \cite{Noe2009Constructing}, it has some flaws.
-In particular, it is generally observed that most models' timescales do not converge, but steadily increase, making it is difficult to judge whether a particular model passes the test and encouraging the use of ever-more states.
+In practice, this test has been useful for many \cite{Noe2009Constructing}, however has some flaws.
+For instance, it is generally observed that most models' timescales do not converge, but steadily increase, so it is difficult to judge whether a particular model passes the test.
 Additionally, it is difficult to compare models built on two state decompositions, as neither may have convergent timescales.
@@
 Additionally, Bayesian methods that involve marginalizing over all possible MSMs have been developed \cite{Bacallado2009Bayesian, *Bacallado2011Bayesian}.
-However, these methods are often prohibitively expensive on larger data sets.
+However, these methods are often prohibitively expensive on larger datasets.
 In contrast, \citet{Kellogg2012Evaluation} proposed a tractable method for comparing MSMs built with different state decompositions, based on a likelihood function for an MD trajectory given an MSM, discussed below.
@@
 
-This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parametrizes a different statistical model.
+This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parameterizes a different statistical model.
 \citet{Kellogg2012Evaluation} used a discriminative model with:
@@
 P(\mathbf{x}_t | s(\mathbf{x}_t)) = \big| \{ \mathbf{x}_k : s(\mathbf{x}_k) = s(\mathbf{x}_t) \} \big|^{-1}
-\end{equation} This model has the training data set as its support rather than phase space, which makes it unable to generalize and assign probability to new data.
+\end{equation} This model has the training dataset as its support rather than phase space, which makes it unable to generalize and assign probability to new data.
 In certain circumstances, this is an undesirable property.
-For example, protocols that involve fitting and validating models on separate data sets (e.g.
+For example, protocols that involve fitting and validating models on separate datasets (e.g.
 cross-validation) are impossible when the statistical model lacks the capacity to describe new data.
@@
 \section{Cross Validation and the Bayes Information Criterion}
-Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to over-fitting\cite{Liddle2007Information, Hastie01Elements}.
+Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to overfitting\cite{Liddle2007Information, Hastie01Elements}.
 Statistical learning theory provides a number of alternative approaches for this problem.
@@
 First, the model is parameterized by building both the state space and transition matrix on a subset of the available molecular dynamics trajectories, then the likelihood is evaluated on the left-out portion.
-This scheme can be repeated many times with different partitions of the data set.
+This scheme can be repeated many times with different partitions of the dataset.
 The AIC and BIC are augmented likelihood function approaches which do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.
@@
 The computation of such volumes has occupied significant attention in recent years in the computational geometry literature, especially via randomized algorithms\cite{Kannan97, Simonovits03, Lovasz03}.
-We opt to approximate the volumes using naive Monte Carlo rejection sampling, which we find tractable for large systems only when the molecular dynamics data set is first projected into a suitable small vector space of up to perhaps ten dimensions.
+We opt to approximate the volumes using naive Monte Carlo rejection sampling, which we find tractable for large systems only when the molecular dynamics dataset is first projected into a suitable small vector space of up to perhaps ten dimensions.
 
@@
 In these cases, the Voronoi cells are of unbounded volume.
-Instead we wish to truncate these states by bounding them by the convex hull of the data set.
-Because the convex hull of our simulation data sets are computationally inaccessible, we define the outer extent of our data sets to be the set of all trial points such that the nearest sampled conformation to the trial point is closer than a certain cutoff, $R$.
+Instead we wish to truncate these states by bounding them by the convex hull of the dataset.
+Because the convex hull of our simulation datasets are computationally inaccessible, we define the outer extent of our data sets to be the set of all trial points such that the nearest sampled conformation to the trial point is closer than a certain cutoff, $R$.
 
@@
 For both likelihoods, the BIC penalizes complexity more than the AIC and test set log-likelihood.
-The optimal window is consistent with the convergence of the relaxation timescales (right frames).}
+The optimal window is consistent with the convergence of the implied timescales (right frames).}
 \label{fig:mullerlike}
@@
 \centering
-\includegraphics[width=3in]{figs_final/kmeans_like2_lbl.png}
+\includegraphics[width=2.5in]{figs_final/kmeans_like2_lbl.png}
 \caption{MSMs were built using k-means clustering on the M\"uller potential simulations.
@@
 The BIC penalizes complexity more strongly with an optimum at ~150 states.
-This window is consistent with the convergence of the relaxation timescales and the onset of Markovian behavior.
+This window is consistent with the convergence of the implied timescales and the onset of Markovian behavior.
 Models built with RMSD require significantly more states than the tICA MSMs and show poor timescale convergence.
-These results indicate why model selection based on relaxation timescales is so difficult, as a less than optimal state decomposition can result in timescales that do not converge rapidly (if at all).
-Furthermore, selection based on only the relaxation timescales contains no guard against over-fitting, which is of particular concern in the models built using RMSD that require thousands of states.
+These results indicate why model selection based on implied timescales is so difficult, as a less than optimal state decomposition can result in timescales that do not converge rapidly (if at all).
 
@@
 On the other hand, the tICA method's identification of slow order parameters in the simulations has the effect of reducing the number of states necessary when constructing Markov models.
-Even with 200-600 states, Markov models constructed using tICA are able to accurately identify the simulations folding process on the 1-10 $\mu s$ timescale, as well as the faster near-native dynamics identified in this data set by McGibbon and Pande\cite{McGibbon2013Learning}.
+Even with 200-600 states, Markov models constructed using tICA are able to accurately identify the simulations folding process on the 1-10 $\mu s$ timescale, as well as the faster near-native dynamics identified in this dataset by McGibbon and Pande\cite{McGibbon2013Learning}.
 
@@
 These results take a step toward fully automating the MSM construction process and controlling the complexity and generalization error by providing a quantitative method for selecting the most suitably complex model, given the data.
-
-\begin{acknowledgement}
-It is an honor and a pleasure to contribute this paper to the Swope Festschrift.
-I (V.S.P.) have been working with Bill since my early days at Stanford and have benefitted greatly from our numerous interactions with him.
-Bill's insight, clarity of thought, and careful approach has influenced my lab's work very broadly.
-I am very pleased that our collaborations continue through today and look forward to many more in the future.
-\end{acknowledgement}
-
 \bibliography{bibliography}
@@
 When you ask both a ``properly-fit'' model and an overfit model a question that basically falls within the training set -- and neither of them need to extrapolate or really infer out-of-sample information, they'll both do fine.
-Heck, the overfit model might do better at telling you exactly what was \emph{in} the data set.
+Heck, the overfit model might do better at telling you exactly what was \emph{in} the dataset.
 \item The folding timescale is relatively insensitive to differences in MSM construction methodology.
-This has been observed in many data sets (McGibbon 2013, others).
-We hypothesize that this behavior is due to the large structural deviations at play -- even a two state model, given the same data-set, could basically get the right timescale just from counting the number of transitions.
+This has been observed in many datasets (McGibbon 2013, others).
+We hypothesize that this behavior is due to the large structural deviations at play -- even a two state model, given the same dataset, could basically get the right timescale just from counting the number of transitions.
 But finding these other timescales that are structurally subtle is different challenge.
 Also, how robust are other properties like experimental projections or the MFPT distribution to the number of states?
-\item Single trajectory data-sets are ``easier'' for MSMs.
+\item Single trajectory datasets are ``easier'' for MSMs.
 One of the uses of MSMs is to ``stitch together'' multiple trajectories -- this function only happens when the states are large enough that they \emph{connect} trajectories, so you can't have too many states in this case.

### 2014-03-24 lib/likelihood.py

@@ -8,7 +8,7 @@ from msmbuilder import MSMLib
 import numpy as np
 
 def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False, 
-    kellogg=False, mapping=None):
+    kellogg=False, mapping=None, sliding_window=True):
     """
     Function to calculate the log likelihood according to:
 
@@ -44,8 +44,18 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False,
     if not mapping is None:
         MSMLib.apply_mapping_to_assignments(trajs, mapping)
 
-    counts = MSMLib.get_count_matrix_from_assignments(trajs, lag_time=lagtime,
-        n_states=tProb.shape[0])
+    if not sliding_window:
+        print "only using first trajectory"
+        counts = MSMLib.get_counts_from_traj(trajs[0], lag_time=lagtime, n_states=tProb.shape[0],
+            sliding_window=False)
+
+        counts = counts.tocsr()
+        counts.eliminate_zeros()
+        counts = counts.tolil()
+
+    else:
+        counts = MSMLib.get_count_matrix_from_assignments(trajs, lag_time=lagtime,
+            n_states=tProb.shape[0], sliding_window=True)
 
     counts = counts.tocsr()
     tProb = tProb.tocsr()
@@ -64,7 +74,10 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False,
     #log_like = np.sum(nonzero_counts * np.log(nonzero_tprobs)) - \
     #    np.sum(state_counts * np.log(volumes))
 
-    t_like = np.sum(nonzero_counts * np.log(nonzero_tprobs))
+    good_transitions = np.where(nonzero_tprobs != 0)
+    t_like = np.sum(nonzero_counts[good_transitions] * np.log(nonzero_tprobs[good_transitions]))
+    pct_lost = (nonzero_counts.sum() - nonzero_counts[good_transitions].sum()) / nonzero_counts.sum() * 100.
+    print "ignoring %.2f %% of counts that have zero probability according to the model"  % pct_lost
 
     if not kellogg:
         v_like = - np.sum(state_counts * np.log(volumes))

### 2014-04-05 paper/paper2.tex
edit

@@
 While standard model selection methodologies focus on the minimization of systematic bias and disregard statistical error, we show that by consideration of the states' conditional distribution over conformations, both sources of error can be balanced evenhandedly.
-Application of these techniques on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those techniques encourage over-fitting.
-By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markvov state model construction.
+Application of techniques that consider both systematic bias and statistical error on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those timescale-based techniques encourage over-fitting.
+By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markov state model construction.
 \end{abstract}
@@
 Two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics.
-The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{}.
+The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{who?}.
 
-Some explanation of notation...
-\begin{align}
-s(t) &= \sigma(x(t)), \\
-s(t+\tau) &\sim \mathbf{T}_{s(t), k} \sigma_k.
-\end{align}
+%The trajectory in phase space ($\mathbf{x}_t$) is assigned to one of $k$ states.
+We represent this assignment by the $s(\cdot)$ function.
+The dynamics in state space are described by the transition probability matrix.
+Since the model is Markovian, the probability that a given trajectory in state $s(\mathbf{x}_t)$ will be in some state $i$ is given by $\mathbf{T}_{s(\mathbf{x}_t), i}$:
+%\begin{equation}
+%\label{eq:markov_state}
+%s(\mathbf{x}_{t+\tau}) \sim_d \mathbf{T}_{s(\mathbf{x}_t), i}
+%\end{equation}
 
 A significant challenge in the automated construction of Markov state models is the choice of the number of states\cite{McGibbon2013Learning}.
-Although classical Hamiltonian dynamics with $N$ particles form a continuous-time Markov chain in $\mathbb{R}^{6N}$\cite{vanKampen}, the projection of this high-dimensional process into a lower dimensional space does (such as the discrete indicator functions used by MSMs) generally experiences non-Markovian dynamics.
-In the Morri-Zwanzig approach, this corresponds to the generalized Langevin equation with a nontrivial memory kernel\cite{ZM}.
+Although classical Hamiltonian dynamics with $N$ particles form a continuous-time Markov chain in $\mathbb{R}^{6N}$\cite{vanKampen}, the projection of this high-dimensional process into a lower dimensional space (such as the discrete indicator functions used by MSMs) is generally governed by non-Markovian dynamics.
+\hl{In the Morri-Zwanzig approach, this corresponds to the generalized Langevin equation with a nontrivial memory kernel}(What is this about?)\cite{ZM}.
 Within the MSM context, on the other hand, this means that first order Markov chains incur a systematic error in their modeling of the high-dimensional system.
@@
 
+\section{Background}
+\begin{itemize}
+\item For some system, we are interested in describing the arbitrarily long timescale dynamics. 
+\item define the phase space, and then an ensemble of systems propagating through phase space. 
+\item There exists an operator such that operator of p(0) is p(t). Formally this operator is dependent on the hamiltonian
+\item This operator has a set of characteristic timescales (and eigenvectors) such that p at t + ntau can be decomposed into a sum of exponential relaxation processes along each eigenvector (show the equation)
+\item A central focus for MD, is the estimation of the transfer operator 
+\item Calculating the hamiltonian integral is not possible, but estimators are available based on sampling the hamiltonian (MD) 
+\item One framework for estimating the transfer operator is the discrete-time Markov State Model (MSM), which is a projection of the operator onto an indicator function basis set. 
+\item The MSM is parameterized by a set of states corresponding to the indicator functions defined on phase space as well as the transition probabilities for transitioning between any state to any other state at a lag time (tau)
+\item It has been shown that using a finite number of discrete states in phase space, estimates of the transfer operator's timescales are biased to be too fast. 
+\item This result implies that one can always do better by varying the parameters  to increase the timescales (include the idea of a variational principle)
+\item However, these results are proved based on the assumption of infinite data, which is never the case
+\end{itemize}
+\section{another section}
+\begin{itemize}
+\item We always have finite data, and so this variational approach ignores another source of error due to the statistical error
+\item What we really want is a model that reproduces the dynamics due to the hamiltonian
+\item In practice, the transfer operator is inaccessible, and what you have is data and a model that attempts to describe the data.
+\item But the only way to access the true transfer operator is to sample from it and produce new data. Therefore, we should actually attempt to estimate a model that can best predict new data.
+\item This MSM has since been referred to as an object that takes in data, but in fact (as it is generative) we are parameterizing it to generate new data. 
+\item So if we build many MSMs, the ``best" model is the one that is most likely to have produced the newly observed data. 
+\item This means that we should construct a model such that it assigns a high probability to the new data. 
+\item This is precisely the idea of cross-validation or BIC/AIC (maybe we should define them here).
+\item The true data is a random variable (that we can sample), but in an MSM, the trajectory sampled is a random variable
+\end{itemize}
+
+\section{bayesian statistics}
+\begin{itemize}
+\item in the bayesian world, we would integrate over all possible models, but if we want a single model, then we should pick the most likely one
+\end{itemize}
+
+Given a Markov model parameterized at a lag time of $\tau$, we can propagate ensembles of trajectories via the transition probability matrix:
+$$\mathbf{p}_\tau^T = \mathbf{p}_0^T T(\tau)$$ where $\mathbf{p}_t$ is a probability distribution over states at time $t$ and $T(\tau)$ is a matrix with entries corresponding to the probability of transferring from one state to another at a time of $\tau$.
+As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a different model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
+$$\mathbf{p}_{n\tau}^T = \mathbf{p}_0^T T(n\tau) = \mathbf{p}_0^T \left[T(\tau)\right]^n $$ As such, the eigenvalues and eigenvectors of models built at many lag times are related.
+Specifically, when building multiple models for the same dataset, the eigenvectors should be invariant to changing lag time, while the eigenvalues change via the relation:
+$$\lambda(\tau)^n = \lambda(n \tau)$$ The largest, non-unit eigenvalues correspond to the slowest timescales in the data.
+It's often useful to compute the slowest implied timescales ($t_i$) via the relation:
+$$t_i = - \frac{\tau}{\log \lambda_i} $$ These implied timescales should be invariant with respect to the lag time and therefore provide a useful tool for testing the quality of an MSM.
+Specifically, many models are built at varying lag times, and one is selected by choosing the model whose timescales have converged.
+
 \section{Prior Work}
@@
 MSM model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time.
-As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
-$$ T(n\tau) = T(\tau)^n $$ As such, the eigenvalues, and correspondingly the relaxation timescales, should remain constant as the lag time increases.
-The resulting test of a model's quality consists of calculating the eigenspectra of models built at many lag times and choosing models whose eigenvalues have converged.
-
 In practice, while this test has been useful for many \cite{Noe2009Constructing}, it has some flaws.
-In particular, it is generally observed that most models' timescales do not converge, but steadily increase, making it is difficult to judge whether a particular model passes the test and encouraging the use of ever-more states.
+In particular, it is generally observed that most models' timescales do not converge, but steadily increase, making it difficult to judge whether a particular model passes the test and encouraging the use of ever-more states.
 Additionally, it is difficult to compare models built on two state decompositions, as neither may have convergent timescales.
@@
 The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis.
-With $s(\mathbf{x})$ as the function mapping conformations into the indicator function basis set, and $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory, the likelihood can be written as:
+With $s(\mathbf{x})$ as the function that maps a conformation in phase space ($\mathbf{x}$) to the index of a state in the discrete state space, and $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory, the likelihood can be written as:
 
@@
 
-With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state Markov transition probabilities, $T(s_i \rightarrow s_j) \equiv P(S_{t+1}=s_j | S_{t}=s_i)$, and so-called emission distributions of each state, $P(\mathbf{x} | s(\mathbf{x}))$, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}$, is within a certain state, $s_i$.
-By convention, we define $P(\mathbf{x}| s(\mathbf{x}))$ to be zero for all $\mathbf{x}$ such that $s(\mathbf{x}) \neq s_i$.
+With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state Markov transition probabilities, $T(s_i \rightarrow s_j) \equiv P(S_{t+1}=s_j | S_{t}=s_i)$, and so-called emission distributions of each state, $P(\mathbf{x} | s(\mathbf{x}))$, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}$, is within a certain state, $s(\mathbf{x})$.
+As we are using  non-overlapping states, the conditional probability of observing a conformation, $\mathbf{x}$ from state $i$ is zero unless $\mathbf{x}$ is assigned to $i$.
+Mathematically, this means:
+$$ P(\mathbf{x} | i) = 0 \;\;\;\; \forall \mathbf{x} \textnormal{ s.t. } s(\mathbf{x}) \neq i$$
 
-This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parametrizes a different statistical model.
+This emission distribution can be modeled in multiple ways, each of which, in combination with a transition matrix, parametrizes a different statistical model.
 \citet{Kellogg2012Evaluation} used a discriminative model with:
@@
 P(\mathbf{x}_t | s(\mathbf{x}_t)) = \big| \{ \mathbf{x}_k : s(\mathbf{x}_k) = s(\mathbf{x}_t) \} \big|^{-1}
-\end{equation} This model has the training data set as its support rather than phase space, which makes it unable to generalize and assign probability to new data.
+\end{equation} where $| \cdot |$ denotes the cardinality of a set.
+This is a discriminative model, whose support is the training data set rather than phase space, which makes it unable to generalize and assign probability to new data.
 In certain circumstances, this is an undesirable property.
@@
 standard distributions over $\mathbb{R}^{3N_{atoms}}$).
-We propose two possible emission distributions, $P(\mathbf{x} | s(\mathbf{x}))$: first, a uniform distribution over the (hyper)volume of each state, as shown in \cref{eq:like_vol}.
+We propose one possible emission distributions, $P(\mathbf{x} | s(\mathbf{x}))$: a uniform distribution over the (hyper)volume of each state, as shown in \cref{eq:like_vol}.
+\begin{equation}
+\label{eq:like_vol}
+\begin{split}
+%P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
+ P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{V_{s(\mathbf{x}_{t})}}
+\end{split}
+\end{equation} 
 This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation.
-
 %For example, consider two Markov state models sharing the same transition matrix, $T$.
@@
 If we think of k-centers as a poor man's version of k-means, then we can apply this same likelihood.
-\begin{equation}
-\label{eq:like_vol}
-\begin{split}
-%P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
- P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{V_{s(\mathbf{x}_{t})}}
-\end{split}
-\end{equation}
-
-A more tractable alternative is a Gaussian emission model where $P(\mathbf{x} | s(\mathbf{x}))$ is modeled as multivariate normal \cref{eq:like_mvn}.
-\begin{equation}
-\label{eq:like_mvn}
-\begin{split}
-P(\mathbf{x}_t | s(\mathbf{x}_t)) = \Bigg[ \frac{1}{(2 \pi)^\frac{d}{2} |\Sigma_{s(\mathbf{x}_t)}|^\frac{1}{2}} \\
-   \exp\left(-\frac{1}{2} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})^T \Sigma_{s(\mathbf{x}_t)}^{-1} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})\right)\Bigg] 
-\end{split}
-\end{equation} where $\bm{\mu}_{s(\mathbf{x}_t)}$ is the cluster center assigned to $\mathbf{x}_t$ and $\Sigma_{s(\mathbf{x}_t)}$ is the corresponding covariance matrix.
-For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which it reduces to:
-\begin{equation}
-\label{eq:like_mvn2}
- P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{\left(2 \pi \hat{\sigma}^2\right)^{\frac{d}{2}}} \exp\left(-\frac{||\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)}||^2}{2 \hat{\sigma}^2} \right)
-\end{equation} where
-\begin{equation}
-\label{eq:mle_sigma}
-\hat{\sigma}^2 = \frac{1}{N - k} \sum_{t=1}^N || \mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)} ||^2
-\end{equation} and $k$ is the number of states.
-
-Since the likelihood requires the emission distributions to have zero overlap -- such that each conformation only has nonzero emission probability from a single state, the nonzero overlap between the Gaussian emission distributions introduces a truncation error.
-One approach to ameliorate this issue is to truncate the density functions at the state boundaries, and renormalize.
-We assume that the overlaps are small, and use the density defined in \cref{eq:like_mvn2}.
-An alternative approach for relaxing this approximation is possible with the forward-backward algorithm,\cite{Rabiner1989Tutorial} but is beyond the scope of this work.
 
@@
 \label{eq:bic}
-\mathrm{BIC} \equiv -2\cdot \ln L + \kappa \cdot \ln N
+\mathrm{BIC} \equiv -2\cdot \ln L + \left(\ln N \right)\cdot \kappa
 \end{equation}
@@
 \label{eq:aic}
-\mathrm{AIC} \equiv -2\cdot \ln L + 2 \cdot \kappa
+\mathrm{AIC} \equiv -2\cdot \ln L + \left(2\right) \cdot \kappa
 \end{equation} where $L$ is the likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points, assumed to be independent and identically distributed.
@@
 Shaw research \cite{Shaw2010Atomic}.}
+\label{fig:pics}
 \end{figure}
@@
 
-For cross validation, we find it essential to use a dense prior during transition matrix estimation.
 Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes.
-We used a Dirichlet prior with a pseudo-count of $\frac{1}{k}$ for each possible transition, where $k$ is the number of states in the model.
+For this reason, we find it essential to use a dense prior during transition matrix estimation.
+Using a prior ensures that all transitions have a nonzero (albeit small) probability.
+We used a Dirichlet prior with a pseudo-count of $\frac{1}{k}$ for each possible transition, where $k$ is the number of states in the model\cite{msmb2?, nina?, wherever this happens to be}.
 
@@
 \subsection{M\"{u}ller Potential}
+The M\"{u}ller potential, is a simple two-dimensional surface with three energy minima.
 We simulated two trajectories with $10^6$ steps of Langevin dynamics on the M\"{u}ller potential\cite{Muller1980Reaction}.
 Simulations were performed with a timestep $t=0.1$, friction coefficient of $\gamma=10^3$, and $kT=15$.
+
 The first trajectory was clustered using the $k$-centers clustering algorithm, and state volumes were computed for the uniform emission model using $10^5$ rounds of Monte Carlo rejection sampling with a cutoff of $R=0.28$ from the 50 state model's centroids.
+(To give a sense of scale, the x-axis of the M\"{u}ller potential shown in \cref{fig:pics} is 4 units across).
 The second trajectory was used as a test set.
-All MSMs were built using a lag time of $30 steps$.
+All MSMs were built using a lag time of 30 steps.
 
-The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and the likelihood remained essentially constant, which indicates that the use of approximation volumes is sufficient to obtain robust estimates of the likelihood.
+The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and the likelihood remained essentially constant, which indicates that the use of approximate volumes is sufficient to obtain robust estimates of the likelihood.
 
@@
 \centering
-\includegraphics[width=3in]{figs_final/mull_likes.png}
-\caption{We calculated the training and test set log-likelihoods and model selection criteria for models built with 50-1000 states using both the uniform ({\bf A}) and multivariate normal ({\bf B}) emission distributions.
+\includegraphics[width=4in]{figs_final/mull_likes.png}
+\caption{\hl{Fix this figure so that only part (A) is depicted.} We calculated the training and test set log-likelihoods and model selection criteria for models built with 50-1000 states using the uniform emission distribution.
 Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray).
@@
 The penalty on the number of parameters in \cref{eq:bic} and \cref{eq:aic} begins to dominate.
-The optimal models, according to the three methods are between 200 and 600 states for this system, which is consistent with the convergence of the relaxation timescales of the models.
+The optimal models, according to the BIC, AIC, and test log-likelihood are between 200 and 600 states for this system, which is consistent with the convergence of the relaxation timescales of the models.
 
 The AIC and BIC penalize the larger state models much more than the test set log-likelihood.
-On the two-dimensional potential, in our data-rich regime it is difficult to produce an over-fit model.
-Cross-validation requires fitting multiple models on subsets of the data, and so is not feasible for larger systems in the data-poor regime.
-
-We also tested the Gaussian likelihood on MSMs built using k-means clustering on the M\"uller potential.
-The training and test set log-likelihoods, BIC, and AIC are all optimized by a five state model (\cref{fig:kmeans_mull}).
-The decrease in the training set log-likelihood can be explained by a trade-off inherent in \cref{eq:like}, where adding more states leads to an improvement in the emission distribution term but causes the transition matrix term to suffer.
-This trade-off illustrates a deficiency in the step-wise approach in which neither the transition matrix nor the state space is parameterized to explicitly optimize the full \cref{eq:like}.
-
-\begin{figure}
-\centering
-\includegraphics[width=3in]{figs_final/kmeans_like2_lbl.png}
-\caption{MSMs were built using k-means clustering on the M\"uller potential simulations.
-Since k-means optimizes an objective function that is related to the emission probability, our likelihood picks only a few states as the most likely model.
-The likelihood decreases as new states are introduced, due to a competition between the transition matrix term and emission distribution term in \cref{eq:like}.}
-\label{fig:kmeans_mull}
-\end{figure}
+It's not clear why the test log-likelihood is more lenient, however, it suggests that when possible, one should use a test log-likelihood approach as opposed to an approximate method like the AIC or BIC.
+This cross-validation requires fitting multiple models on subsets of the data, and so is not feasible for larger systems in the data-poor regime.
 
@@
 Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification}, extracting only the four slowest uncorrelated linear combinations of residue-residue distances.
-All MSMs were built using a lag time of $50 ns$.
-
-The complexity of evaluating the Gaussian emission model likelihood does not increase with respect to dimension.
-Thus we used the Gaussian emission model with both the four-dimensional tICA models and those built using k-centers clustering with the minimal cartesian root-mean squared deviation (RMSD) over Fip35's 252 non-redundant heavy atoms, computed using the quaternion characteristic polynomial method\cite{Theobald2005Rapid}.
-For evaluating the Gaussian emission likelihood, we used a value of $3 \cdot 252 - 6 = 750$ for the dimension of the emission distribution.
+All MSMs were built using a lag time of 50 ns.
 
@@
 \includegraphics[width=3in]{figs_final/ww_likes.png}
-\caption{On models whose states were defined after projecting into the four-dimensional tICA subspace, we computed likelihoods based on the uniform ({\bf A}) and Gaussian ({\bf B}) emission models.
-Models built using the RMSD metric were evaluated with the Gaussian emission model ({\bf C}).
+\caption{\hl{Fix this figure so that only part (A) is depicted.} On models whose states were defined after projecting into the four-dimensional tICA subspace, we computed likelihoods based on the uniform emission models.
 Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray).
-Test set log-likelihoods are omitted in {\bf C} as the dense matrix algebra is intractable for $k > 10^3$.\label{fig:ww}}
+\label{fig:ww}}
 \end{figure}
@@
 
-The two 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain simulated by \citet{Shaw2010Atomic} have been used to parameterize multiple Markov state models.
+\begin{figure}
+\centering
+\includegraphics[width=6in]{figs_final/timescales_suck2.pdf}
+\caption{For MSMs of the WW domain built on the top three PCA components,} 
+\end{figure}
+
+%The two 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain simulated by \citet{Shaw2010Atomic} have been used to parameterize multiple Markov state models.
 \citet{Lane2011Markov} proposed a 26,000 state MSM after clustering the data by RMSD, whereas Kellogg et al.,\cite{Kellogg2012Evaluation} clustering on contact maps or secondary structure, proposed 100 and 175 state models, respectively.
@@
 On the other hand, the tICA method's identification of slow order parameters in the simulations has the effect of reducing the number of states necessary when constructing Markov models.
-Even with 200-600 states, Markov models constructed using tICA are able to accurately identify the simulations folding process on the 1-10 $\mu s$ timescale, as well as the faster near-native dynamics identified in this data set by McGibbon and Pande\cite{McGibbon2013Learning}.
+While only using 200-600 states, Markov models constructed using tICA were still able to recapitulate the folding timescale as well as the faster processes identified by McGibbon and Pande\cite{McGibbon2013Learning}.
 
@@
 \bibliography{bibliography}
+
+\section{Appendix}
+\subsection{Other emission distributions}
+
+A more tractable alternative to the uniform distribution is a Gaussian emission model where $P(\mathbf{x} | s(\mathbf{x}))$ is modeled as multivariate normal \cref{eq:like_mvn}.
+\begin{equation}
+\label{eq:like_mvn}
+\begin{split}
+P(\mathbf{x}_t | s(\mathbf{x}_t)) = \Bigg[ \frac{1}{(2 \pi)^\frac{d}{2} |\Sigma_{s(\mathbf{x}_t)}|^\frac{1}{2}} \\
+   \exp\left(-\frac{1}{2} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})^T \Sigma_{s(\mathbf{x}_t)}^{-1} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})\right)\Bigg] 
+\end{split}
+\end{equation} where $\bm{\mu}_{s(\mathbf{x}_t)}$ is the cluster center assigned to $\mathbf{x}_t$ and $\Sigma_{s(\mathbf{x}_t)}$ is the corresponding covariance matrix.
+For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which it reduces to:
+\begin{equation}
+\label{eq:like_mvn2}
+ P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{\left(2 \pi \hat{\sigma}^2\right)^{\frac{d}{2}}} \exp\left(-\frac{||\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)}||^2}{2 \hat{\sigma}^2} \right)
+\end{equation} where
+\begin{equation}
+\label{eq:mle_sigma}
+\hat{\sigma}^2 = \frac{1}{N - k} \sum_{t=1}^N || \mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)} ||^2
+\end{equation} and $k$ is the number of states.
+
+Since the likelihood requires the emission distributions to have zero overlap -- such that each conformation only has nonzero emission probability from a single state, the nonzero overlap between the Gaussian emission distributions introduces a truncation error.
+One approach to ameliorate this issue is to truncate the density functions at the state boundaries, and renormalize.
+We assume that the overlaps are small, and use the density defined in \cref{eq:like_mvn2}.
+An alternative approach for relaxing this approximation is possible with the forward-backward algorithm,\cite{Rabiner1989Tutorial} but is beyond the scope of this work.
+
+We tested the Gaussian likelihood on MSMs built using k-means clustering on the M\"uller potential.
+The training and test set log-likelihoods, BIC, and AIC are all optimized by a five state model (\cref{fig:kmeans_mull}).
+The decrease in the training set log-likelihood can be explained by a trade-off inherent in \cref{eq:like}, where adding more states leads to an increase in the emission distribution term but causes the transition matrix term to decrease significantly more.
+This trade-off illustrates a deficiency in the step-wise approach in which neither the transition matrix nor the state space is parameterized to explicitly optimize the full \cref{eq:like}.
+
+\begin{figure}
+\centering
+\includegraphics[width=3in]{figs_final/kmeans_like2_lbl.png}
+\caption{MSMs were built using k-means clustering on the M\"uller potential simulations.
+Since k-means optimizes an objective function that is related to the emission probability, our likelihood picks only a few states as the most likely model.
+The likelihood decreases as new states are introduced, due to a competition between the transition matrix term and emission distribution term in \cref{eq:like}.\label{fig:kmeans_mull}}
+\end{figure}
+
+We also tested the Gaussian likelihood on the same WW simulations as discussed above.
+The complexity of evaluating the Gaussian emission model likelihood does not increase with respect to dimension.
+Thus we used the Gaussian emission model with both the four-dimensional tICA models and those built using k-centers clustering with the minimal cartesian root-mean squared deviation (RMSD) over Fip35's 252 non-redundant heavy atoms, computed using the quaternion characteristic polynomial method\cite{Theobald2005Rapid}.
+For evaluating the Gaussian emission likelihood, we used a value of $3 \cdot 252 - 6 = 750$ for the dimension of the emission distribution.
+
 \end{document}

### 2014-04-05 reviewer-comments/letter.tex
edit

@@
 \documentclass{letter}
+\usepackage{color,soul}
 \usepackage{hyperref}
+\usepackage{fullpage}
+\usepackage{cleveref}
+
 \address{Clark Center S295 \\ Stanford University \\ Stanford, CA, 94305}
 \signature{Robert T. McGibbon, Christian R. Schwantes, Vijay S. Pande}
+
+\newcommand{\separate}{\begin{center}--------\end{center}}
+
 \begin{document}
@@
 Reviewer 1
-
 \begin{quote}
@@
 The title ``Optimal parameter selection...'' does not reflect the work contained in the paper, which focuses only on the selection of the number of states.
-Should the authors which to keep the general title, more work on selecting other kinds of parameters should be included; otherwise, the title should be amended to something like ``Optimal selection of the number of states...
+Should the authors wish to keep the general title, more work on selecting other kinds of parameters should be included; otherwise, the title should be amended to something like ``Optimal selection of the number of states...
 using a likelihood approach...''
@@
 
+This is a fair point, as we focus only on picking the number of states.
+In fact, we've changed the title to reflect this issue as well as the fact that we're not attempting to argue that our selection is truly ``optimal," but merely suggesting to use a method that is ``{\it optimizable}." \hl{reference the title.}
+
+\separate
 \begin{quote}
@@
 
+The reviewer is exactly correct.
+We meant that the timescales will remain constant, but this means that the eigenvalues do change.
+We've corrected that sentence and also added a more detailed discussion of the transition probability matrix.
+The text can be found at the beginning of the section ``Prior Work."
+
+\hl{add the text here too}
+\separate
 \begin{quote}
@@
 
+We were too casual with our notation, but had meant that $s(\cdot)$ maps a conformation in phase space to an integer corresponding to the index of the state in the Markov model.
+This has been reworded to read:
+
+\begin{quote}
+With $s(\mathbf{x})$ as the function that maps a conformation in phase space ($\mathbf{x}$) to the index of a state in the discrete state space
+\end{quote}
+\separate
 \begin{quote}
@@
 
+This is most definitely true, but is not incredibly important.
+The reviewer is correct that technically we should include the probability of $x_0$ in the likelihood, however, when evaluating the relative likelihoods of models built on the same data, this term cancels out as it shows up in each model.
+Therefore, optimizing either function will arrive at the same result.
+\hl{We can add to the text if we think we need to, but this is really a nit-picky comment in my opinion.}
+
+\separate
 \begin{quote}
@@
 
+We've reworded this paragraph to be clearer. The text can be found right after Eq. (1) and is copied below:
+\begin{quote}
+With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state Markov transition probabilities, $T(s_i \rightarrow s_j) \equiv P(S_{t+1}=s_j | S_{t}=s_i)$, and so-called emission distributions of each state, $P(\mathbf{x} | s(\mathbf{x}))$, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}$, is within a certain state, $s(\mathbf{x})$.
+As we are using  non-overlapping states, the conditional probability of observing a conformation, $\mathbf{x}$ from state $i$ is zero unless $\mathbf{x}$ is assigned to $i$.
+Mathematically, this means:
+$$ P(\mathbf{x} | i) = 0 \;\;\;\; \forall \mathbf{x} \textnormal{ s.t. } s(\mathbf{x}) \neq i$$
+\end{quote}
+
+\separate
 \begin{quote}
@@
 \end{quote}
+\hl{I'm not sure that this change is really all that necessary.}
 
+\separate
 \begin{quote}
@@
 
+\hl{This is one of the two major criticisms in the review so we should word the response carefully}
+
+When we first added this emission model, our goal was to provide something that might be used in more realistic cases.
+As the volume calculation makes the volume emission model impractical for all but the simplest cases, we wanted a model whose probabilities could be evaluated analytically.
+One simple first guess would be to use a gaussian distribution.
+However, the reviewer brings up several issues with this model that make it too imprecise to be used in practice.
+As such, we've moved all of the Gaussian emission model results / discussion to an appendix section.
+It is still useful to think about as it is a stepping stone to using ``fuzzy" clustering or other methods whose state emission distributions may overlap.
+
 \begin{quote}
@@
 
+This is certainly one possibility, and is desirable because it is physically motivated. \hl{maybe add a discussion in the gaussian section?}
+\separate
 \begin{quote}
@@
 \end{quote}
+\hl{Is there some mathematical definition of intractability that I'm unaware of?}
+In order to compute the Bayes factor for selecting the number of states, we would need to sample all transition matrices that can be built in $k$ states.
+This is simply impractical as you need to integrate over all $k \times k$ transition matrices.
 
+\separate
 \begin{quote}
@@
 
+This is a subtle point, but is worth discussing.
+Since we are parameterizing a transition matrix, we choose the number of data points to be the number of transitions used to calculate the transition probabilities.
+However, as we are using the sliding window approach for counting transitions, we are in fact over-counting the number of identical transitions and so the BIC score is likely under-penalizing the models.
+This is a somewhat unanswered question in the field, however it has been discussed previously \cite{noe_sliding_window}.
+\hl{Do we want to add this to the main text?}.
+
+\separate
 \begin{quote}
@@
 
+We are being very unsophisticated in our use of the likelihood as we are not actually optimizing this likelihood to build our models.
+We are actually just using the likelihood to evaluate several models that are parameterized using the approach discussed in \cite{Beauchamp_msmb2}.
+\hl{Add this to the text, because the confusion is coming from the use of the word ``prior." We are not introducing a prior to the likelihood function, but merely using the prior when calculating the MAP for each model.}
+
+\separate
 \begin{quote}
@@
 \end{quote}
+\hl{Need to look this up. I just used robert's code though.}
 
+\separate
 \begin{quote}
@@
 \end{quote}
+\hl{Now we're getting really nit-picky, but fine.}
 
+\separate
 \begin{quote}
@@
 \end{quote}
+\hl{It's really just a qualitative change.
+I just meant that the changes in the likelihood are much less than the differences between models, and so are not important for evaluating the ``best" model.}
 
+\separate
 \begin{quote}
 according to the three methods" Which three methods exactly?
+\end{quote}
+We've clarified this to read:
+\begin{quote}
+according to the BIC, AIC, and test log-likelihood
 \end{quote}
@@
 \end{quote}
+\hl{I disagree with this comment. The absolute likelihoods are basically meaningless, while the relative differences are important. I think adding units would be distracting.}
 
+\separate
 \begin{quote}
@@
 
+We are not attempting to provide the ``best" model given the data, as there are many statistical tests for doing this.
+The window illustrated in these figures represents some reasonable range of models, and is a useful visualization of the differences in the three tests.
+In practice, however, one would choose one of the three and then pick the best model given that test.
+\hl{this was only mentioned in the captions of the figures, so I'll correct it when I change the figures.}
+
 \begin{quote}
@@
 
+It's quite possible that the AIC and BIC are over-penalizing the likelihood.
+Given this, we would always suggest that one uses the test log-likelihood approach when possible.
+However, the ``best" models according to the AIC and BIC were close to the best model from the test log-likelihood, so they may still be useful when it is impossible to leave out any data during the parameterization step.
+We are by no means advocating that either the AIC or BIC are perfect approximations to the out-of-sample performance.
+\hl{Add a sentence saying this in the main text.}
+
+\separate
 \begin{quote}
@@
 
+We should not anthropomorphize our mathematical expressions! We've corrected this to read more clearly:
+\begin{quote}
+The decrease in the training set log-likelihood can be explained by a trade-off inherent in Eq.
+(1), where adding more states leads to an increase in the emission distribution term but causes the transition matrix term to decrease significantly more.
+\end{quote}
+
+\separate
 \begin{quote}
@@
 \end{quote}
+\hl{I originally had this but you didn't like my graph. We could add it back or ignore this since it's moved to the appendix anyway}
 
+\separate
 \begin{quote}
@@
 
+The k-means clustering method minimizes the average mean squared distance from each data point to its cluster's mean.
+This is precisely the same as the $\hat{\sigma}^2$ term in Eq.
+(8).
+When you write out the entire log-likelihood of the training dataset, there ends up being a term proportional to $\log \hat{\sigma}$.
+This is what we meant by the k-means clustering optimizing the likelihood function (and is precisely why Pelleg and Moore selected the probability that they did).
+\hl{Add clarification in figure to indicate that these are training set likelihoods.}
+
+\separate
 \begin{quote}
-> * "...AIC is comparable to the test set log-likelihood...The BIC penalizes complexity more strongly..." Why the huge difference between the AIC and BIC?
+``...AIC is comparable to the test set log-likelihood...The BIC penalizes complexity more strongly..." Why the huge difference between the AIC and BIC?
 \end{quote}
 
+The difference just comes from the difference between the way both are defined.
+\hl{I want to brush this off as ``we're not advocating one or the other, but merely suggesting they may be useful in cases where cross-validation is not possible" but I'm not sure of the best wording.}
+
+\separate
 \begin{quote}
@@
 Missing figure title, like "Comparison of likelihood models for Fip35 WW domain trajectory data".
-"Test set log-likelihoods are omitted in C as the dense matrix algebra is intratable for $k > 10^3$".
+"Test set log-likelihoods are omitted in C as the dense matrix algebra is intractable for $k > 10^3$".
 Intractable?
@@
 \end{quote}
+We really shouldn't use this word here, as it's not intractable, but it is difficult, as you have to fit the transition matrix with $k^2$ (nonzero) terms.
+Our software leverages sparse matrices, so it is difficult for us to actually parameterize large transition matrices that are not sparse.
 
+\separate
 \begin{quote}
@@
 
+Yes it is.
+For well-sampled datasets built using the state of the art methods we can still use implied timecales, however, there are many times that the timescales are not as clean and easy to interpret.
+We've added a figure for models built with PCA instead of tICA on the WW dataset to illustrate how the implied timescales are not ``optimizable" in non-ideal cases.
+\hl{still need to make this figure.}
+
+\separate
 \begin{quote}
@@
 \end{quote}
+Lane et al.
+used an implied timescale approach with the RMSD metric, while Kellog et al.
+(discussed in the section ``Prior Work" \hl{right?}) used a likelihood model with a discriminative emission distribution.
+This line has been re-worked since we've moved the Gaussian emission model to the appendix.
 
+\separate
 \begin{quote}
@@
 
+\hl{This is another important note that we need to answer carefully.
+The idea is that in ideal cases, our methods don't disagree with the timescale test drastically, but they are optimizable, which is in itself a significant advantage.
+This is particularly useful when the timescales simply do not converge, and cannot be quantitatively used to select a model.
+Secondly, the model that the statistical test selects may not be the model that best represents the protein of interest, however it does represent our best estimate of the system given how much data we have.
+Cross-validation is a rigorous way to compare models based on newly observed data, and so our test log-likelihood serves as a way to see that adding new states may increase the timescales, but in fact decreases the ``quality" of the model.}
+
+\separate
 \begin{quote}
@@
 
+The advantage is precisely that this framework is {\it optimizable}.
+The statistical tools we've suggested match the Chapman-Kolmogorov when the dataset is ``well-behaved" (i.e.
+when we have a lot of data and are using tICA).
+However, there are many cases where the timescales do not converge at all and so the Chapman-Kolmogorov tests fail.
+We've illustrated this with a simple example (\hl{See section ``PCA on WW")}, but in fact there are many larger simulations that give rise to wildly miss-behaving implied timescales.
+
+\separate
 \begin{quote}
@@
 \end{quote}
+This is a good point, and we've rephrased to say:
+\begin{quote}
+Application of techniques that consider both systematic bias and statistical error on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those timescale-based techniques encourage over-fitting.
+\end{quote} \hl{this sentence feels a bit long...}
 
+\separate
 \begin{quote}
@@
 
+\hl{We should add a section discussing this as well as the future of directly optimizing the likelihood in the form of an HMM or other type of moel}
+
+\separate
 \begin{quote}
@@
 
+This is precisely what we are drawing attention to, and we've added to this section to explicitly say this is a discriminative model.
+
 \begin{quote}
-``but is intractble in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation''  Be clear that *exact* volume computation is intractable in very high dimension due to the complexity of all known algorithms.
+``but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation''  Be clear that *exact* volume computation is intractable in very high dimension due to the complexity of all known algorithms.
 Monte Carlo schemes (such as the one used here, or methods such as [http://www.cs.elte.hu/~lovasz/vol4-focs-tr.pdf]) allow the approximation of this volume.
@@
 
+The reviewer is exactly right. For these reasons (and others) we've chosen to discuss just the volume-based emission model in the main text. This is really what we'd like to focus on.
+
 \begin{quote}
@@
 
+This was the same lag time that was used in Beauchamp et al.\cite{Beauchamp_shaw_paper}.
+In this scheme, the lag time must be specified before selecting the models, and so we picked a number that had been successfully used in the past.
+It's worth noting, however, that our scheme does not provide a framework for the selection of the lag time.
+\separate
 \begin{quote}
 ``identify the simulations folding process''  Something isn't quite right with this phrase.  What is being referred to?
+\end{quote}
+
+This sentence was poorly worded, but we've reworked it and copied the new version below:
+\begin{quote}
+While only using 200-600 states, Markov models constructed using tICA were still able to recapitulate the folding timescale as well as the faster processes identified by McGibbon and Pande.
 \end{quote}
@@
 \end{quote}
+
+The reviewer is exactly right, and this is the point we're trying to drive home.
+It would be nice to compute the statistical error, but it is a difficult thing to calculate.
+\hl{This is in the minor complaints section so I don't know if we really need anything more than this}
 
@@
 
+We do mean complexity here, however the two words go hand in hand.
+In this case, we're talking about an MSM built either with thousands vs.
+tens of states.
+The simpler model therefore cannot describe transitions between two states from the complicated model that are lumped together in the simple model.
+In this sense, the simpler model is less accurate.
+However, this may not matter, if what you really care about is a large global transition between ensembles of small states.
+In this sense, the simpler model does not lose any accuracy.
+\hl{I hate how I worded this.}
+
+\separate
 \begin{quote}
@@
 
+\hl{I have this data, so we can add it as an appendix. It doesn't hurt.}
+
+\separate
 \begin{quote}
@@
 \end{quote}
+
+This is a very good idea and something we're interested in doing.
+However, due to the complexity of the volume calculation it's simply not feasible.
+We hope that our work serves as a stepping stone to new methods that optimize a likelihood supported in phase space, but admit it may not be the one discussed here.
 
@@
 Using the present likelihood could systematically give wrong estimates for various reasons.
+\end{quote}
+
+\begin{quote}
 An inappropriate statistical model for the local output function (which is really hard to make) would lead to biased estimates.
@@
 If the lag time chosen is too short, it will not lead to correct estimates for feasible numbers of states.
+\end{quote}
+
+\begin{quote}
 The differences in Fig 4 (see Details below) indeed suggests that the present criterion fails to provide the correct estimate in at least some cases.
 If a new estimator is derived, it needs to be shown that it converges to the correct result in some limit.
+\end{quote}
+
+\begin{quote}
 I suggest to build a fine discrete model, e.g.
-by direct gridding of the Müller potential, and constructing a simple MCMC dynamics on neighboring gridpoints, and then considering different clusterings on this data set.
+by direct gridding of the M\"uller potential, and constructing a simple MCMC dynamics on neighboring gridpoints, and then considering different clusterings on this data set.
 There, the exact timescales are known, and an embedding in a geometric space is given, such that the method can be applied such that the present method can be applied as is.
@@
 \begin{quote}
-Please check the timescales in Fig 2(A,B) and 4(A,B) - they look identical, but shouldn’t. Details / specific points:
+Please check the timescales in Fig 2(A,B) and 4(A,B) - they look identical, but shouldn't. Details / specific points:
 \end{quote}
+
+These timescales were correctly displayed.
+The timescales in A and B are calculated from the same set of models, but the evaluation of those models just provides a different ``score." Part B of each of these figures has been moved to the appendix.
 
@@
 \begin{quote}
-``Often the choice is made to use the model whose timescales better match an experimental observable:`` I hope that’s not true and I wouldn’t generally accuse the community of practising this.
-When timescales don’t converge, this is a numerical or statistical problem with the data or the estimators used.
-Ignoring this fact and deliberately choosing a model that coincidentially delivers the numbers one hopes to get would be simply cheating.
+``Often the choice is made to use the model whose timescales better match an experimental observable:`` I hope that's not true and I wouldn't generally accuse the community of practicing this.
+When timescales don't converge, this is a numerical or statistical problem with the data or the estimators used.
+Ignoring this fact and deliberately choosing a model that coincidentally delivers the numbers one hopes to get would be simply cheating.
 \end{quote}
 
+The reviewer is definitely correct here.
+We are not accusing anyone of ``cheating" but more drawing attention to the fact that with the absence of an optimizable objective function there is nothing preventing one from picking the model that gives a reasonable answer.
+\hl{we should rephrase this sentence, but I don't know what to do exactly}
+
+\separate
 \begin{quote}
@@
 \begin{quote}
-WW domain, lag time was chosen to be 50 ns - same comment as for the Müller potential, above
+WW domain, lag time was chosen to be 50 ns - same comment as for the Mueller potential, above
 \end{quote}
@@
 
+\bibliography{bibliography.bib}
 \end{letter}

### 2014-04-06 paper/paper2.tex
Wrote a new background section

@@
 Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}.
-But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Mertens2010Structural, Tzeng2011Protein}.
+But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{FRET_PAPER?, Mertens2010Structural, Tzeng2011Protein}.
 
@@
 With advances at the algorithmic, hardware, and software levels, modern molecular simulation paradigms, incorporating specialized or accelerated hardware, often in combination with highly parallel distributed computing frameworks, are capable of generating extensive simulation ensembles with relative ease\cite{Gotz2012Routine, Eastman2013OpenMM, Shirts2000Screen, Shaw2009Millisecond, Hess2008PLINCS, Buch2010High}.
-A central challenge in the field of molecular simulations is now often the kinetic analysis of simulation datasets.
-The analysis paradigms often entail the construction of lower resolution models parametrized from the high resolution simulation data set which capture the essential features in an interpretable framework\cite{Freddolino2010Challenges, Lane2013Milliseconds}.
+A central challenge in the field of molecular simulations is now often the kinetic analysis, which typically involves constructing a lower resolution model that captures the system's essential features in an interpretable framework\cite{Freddolino2010Challenges, Lane2013Milliseconds}.
 For example, by projecting the data down onto one or two degrees of freedom we create a simpler model for the system, such as one characterized by diffusion along a single reaction coordinate\cite{Best2010Coordinate}.
@@
 In an MSM, the dynamics are modeled as a memory-less jump process between a discrete set of conformational states which partition the phase space.
-Two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities or transition rates, which parameterize the kinetics.
-The matrix of transition probabilities can be used to locate the systems transition paths\cite{Weinan2006Towards}, and its dominant eigenvectors identify the metastable states\cite{Deuflhard2000Identification} and long-timescale dynamical modes\cite{who?}.
+Two key quantities that define the MSM are thus the state definitions, an indicator function basis over phase space, and the pairwise transition probabilities, which parameterize the kinetics.
+The matrix of transition probabilities can be used to locate the systems transition paths, and its dominant eigenvectors identify the metastable states and long-timescale dynamical modes\cite{Weinan2006Towards, Deuflhard2000Identification}.
 
@@
 
-A significant challenge in the automated construction of Markov state models is the choice of the number of states\cite{McGibbon2013Learning}.
-Although classical Hamiltonian dynamics with $N$ particles form a continuous-time Markov chain in $\mathbb{R}^{6N}$\cite{vanKampen}, the projection of this high-dimensional process into a lower dimensional space (such as the discrete indicator functions used by MSMs) is generally governed by non-Markovian dynamics.
+%A significant challenge in the automated construction of Markov state models is the choice of the number of states\cite{McGibbon2013Learning}.
+Although classical Hamiltonian dynamics with $N$ particles form a continuous-time Markov chain in $\mathbb{R}^{6N}$, the projection of this high-dimensional process into a lower dimensional space (such as the discrete indicator functions used by MSMs) is generally governed by non-Markovian dynamics\cite{van1992Stochastic}.
 \hl{In the Morri-Zwanzig approach, this corresponds to the generalized Langevin equation with a nontrivial memory kernel}(What is this about?)\cite{ZM}.
@@
 
+Two competing sources of error govern the accuracy of an MSM.
+The first source of error is systematic and is common to all dimensionality reduction methods.
+\hl{Any simplified representation of a complex system cannot generally describe arbitrary behavior of that system losslessly.} The second source of error is statistical in nature, in that the parameters of an MSM must be estimated from a finite number of stochastic MD trajectories.
+\hl{As we discuss later,?} These two sources of error compete in a bias-variance tradeoff\cite{bias_variance_paper}.
+Given a finite size data set, as the number of states increases, the systematic error (the bias) decreases, while the statistical error (the variance) increases.
+
 Here, we seek to build models that are \emph{suitably} complex, given the data, yielding complex descriptions of the system only to the extent that their additional parameters are implied by the observed dynamics.
-To that end, we introduce a procedure for scoring the likelihood of an MSM, which, together with standard statistical model selection techniques, enables the optimal selection of the state space, which we express both in terms of the number of states and the clustering algorithm employed to group sampled conformations into states.
-This approach complements validation procedures performed primarily based on human intuition, such as Chapman-Kolmogorov tests, and enables the treatment of model selection as an optimization problem amenable to automated methods.
+To that end, we introduce a procedure for scoring the likelihood of an MSM, which, together with standard statistical model selection techniques, enables a balanced selection of the model's parameters.
 
 \section{Background}
-\begin{itemize}
-\item For some system, we are interested in describing the arbitrarily long timescale dynamics. 
-\item define the phase space, and then an ensemble of systems propagating through phase space. 
-\item There exists an operator such that operator of p(0) is p(t). Formally this operator is dependent on the hamiltonian
-\item This operator has a set of characteristic timescales (and eigenvectors) such that p at t + ntau can be decomposed into a sum of exponential relaxation processes along each eigenvector (show the equation)
-\item A central focus for MD, is the estimation of the transfer operator 
-\item Calculating the hamiltonian integral is not possible, but estimators are available based on sampling the hamiltonian (MD) 
-\item One framework for estimating the transfer operator is the discrete-time Markov State Model (MSM), which is a projection of the operator onto an indicator function basis set. 
-\item The MSM is parameterized by a set of states corresponding to the indicator functions defined on phase space as well as the transition probabilities for transitioning between any state to any other state at a lag time (tau)
-\item It has been shown that using a finite number of discrete states in phase space, estimates of the transfer operator's timescales are biased to be too fast. 
-\item This result implies that one can always do better by varying the parameters  to increase the timescales (include the idea of a variational principle)
-\item However, these results are proved based on the assumption of infinite data, which is never the case
-\end{itemize}
-\section{another section}
-\begin{itemize}
-\item We always have finite data, and so this variational approach ignores another source of error due to the statistical error
-\item What we really want is a model that reproduces the dynamics due to the hamiltonian
-\item In practice, the transfer operator is inaccessible, and what you have is data and a model that attempts to describe the data.
-\item But the only way to access the true transfer operator is to sample from it and produce new data. Therefore, we should actually attempt to estimate a model that can best predict new data.
-\item This MSM has since been referred to as an object that takes in data, but in fact (as it is generative) we are parameterizing it to generate new data. 
-\item So if we build many MSMs, the ``best" model is the one that is most likely to have produced the newly observed data. 
-\item This means that we should construct a model such that it assigns a high probability to the new data. 
-\item This is precisely the idea of cross-validation or BIC/AIC (maybe we should define them here).
-\item The true data is a random variable (that we can sample), but in an MSM, the trajectory sampled is a random variable
-\end{itemize}
+Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$.
+Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$.
+There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
+$$
+p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
+$$
+
+The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
+$$
+p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
+$$
+where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
+$$
+\langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
+$$
+For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Generation}.
+
+For classical MD, $\mathcal{Q}^{(\tau)}$ is uniquely determined by the Hamiltonian and equations of motion\cite{}.
+However, it is not practical to explicitly construct the propagator from the Hamiltonian for complex systems.
+A Markov model provides an estimate for the propagator.
+A $k$-state Markov model is defined on a discrete set of states, $S = \{s_i\}_{i=1}^k$, where $s_i \subseteq \Omega$ and $\bigcup_{i=1}^k s_i = \Omega$.
+Furthermore, $s_i \cap s_j = \emptyset$ for all $i \neq j$.
+In words, every point in $\Omega$ is assigned to one (and only one) state in the MSM.
+Let $\sigma(\mathbf{x})$ be the function that maps a point $\mathbf{x} \in \Omega$ to the index $i$ of the state in $S$ such that $x \in s_i$.
+
+The MSM models the \hl{action?} of the propagator with a Markov jump process on $\{1, \ldots, k\}$.
+Let $\mathbf{p} \in (\mathbb{R} \cap [0,1])^k$ be a column vector whose entries sum to one.
+The MSM approximates $p(\mathbf{x})$ by
+$$
+p(\mathbf{x}) = \sum_{i=1}^{k} p_i \; \mathbf{1}_{s_i}(\mathbf{x}),
+$$
+where
+$$
+\mathbf{1}_{s}(\mathbf{x}) = \left\{\begin{array}{cc}
+1 & \mathbf{x} \in s \\
+0 & \text{otherwise}
+\end{array}\right.
+$$
+Consider a probability vector at time $t$ denoted $\mathbf{p}(t)$. The time-evolution of $\mathbf{p}(t)$ is described by
+$$
+\mathbf{p}(t+\tau)^T = \mathbf{p}(t)^T \mathbf{T},
+$$
+where $T_{ij}$ is the probability of transiting to state $j$ in time $\tau$ given that the system started in state $i$.
+With this construction, the eigenvalues and eigenvectors of $\mathbf{T}$ \hl{provide approximations (what about transfer operator?)} for the eigenvalues and eigenfunctions of $\mathcal{Q}$.
+
+Let $D$ be a set of molecular dynamics trajectories in $\Omega$.
+To construct an MSM, both $S$ and $\mathbf{T}$ must be determined from the data.
+${S}$ is defined by $\sigma$, which can be constructed by clustering the points in $D$.
+Given $S$, maximum-likelihood estimators for the the transition probabilities $\mathbf{T}$ exist\cite{kyle, frank}.
+
+Using this procedure, it has been shown that MSM estimates of the eigenvalues $\lambda_i$ are systematically underestimated in the limit of infinite data, and that the magnitude of this bias goes to zero as $k\rightarrow \infty$.
+This fact has inspired the development of variational methods for model selection, which optimize the MSM parameters in order to maximize the eigenvalues.
 
@@
 $$\mathbf{p}_{n\tau}^T = \mathbf{p}_0^T T(n\tau) = \mathbf{p}_0^T \left[T(\tau)\right]^n $$ As such, the eigenvalues and eigenvectors of models built at many lag times are related.
-Specifically, when building multiple models for the same dataset, the eigenvectors should be invariant to changing lag time, while the eigenvalues change via the relation:
+Specifically, when building multiple models for the same data set, the eigenvectors should be invariant to changing lag time, while the eigenvalues change via the relation:
 $$\lambda(\tau)^n = \lambda(n \tau)$$ The largest, non-unit eigenvalues correspond to the slowest timescales in the data.

### 2014-04-06 paper/paper2.tex
more edits

@@
 ${S}$ is defined by $\sigma$, which can be constructed by clustering the points in $D$.
-Given $S$, maximum-likelihood estimators for the the transition probabilities $\mathbf{T}$ exist\cite{kyle, frank}.
+Given $S$, maximum-likelihood estimators for the $O(k^2)$ transition probabilities $\mathbf{T}$ exist\cite{kyle, frank}.
 
 Using this procedure, it has been shown that MSM estimates of the eigenvalues $\lambda_i$ are systematically underestimated in the limit of infinite data, and that the magnitude of this bias goes to zero as $k\rightarrow \infty$.
-This fact has inspired the development of variational methods for model selection, which optimize the MSM parameters in order to maximize the eigenvalues.
+This fact has inspired the development of variational methods for model selection, which optimize the MSM parameters in order to maximize the eigenvalues \cite{feliks}.
+
+\section{Statistical Model Selection}
+For MSM methods to be valuable to practitioners, they must operate in limited data regimes.
+Thus, we seek a method for model selection that is sensitive to both the systematic error discussed above, as well as the statistical error associated with estimating the model's parameters.
+Whereas model selection based on the variational principle described above would tend to increase the number of states without bound, as the magnitude of the eigenvalue bias decreases with the number of states, practical application of this criteria would cause a catastrophic increase in statistical uncertainty in the model.
+This tradeoff between low-variance (small $k$) but biased estimators, and high-variance (large $k$) but unbiased estimators is a general feature of problems in statistical learning\cite{}.
+For example, as we show in appendix A, the uncertainty in the posterior distribution of $\mathbf{T}$ given $S$ and $D$ grows as $O(k)$.
+
+Introduction to MSMs via the propagator theory motivates the view of the parameterization problem as one of \emph{function estimation}.
+This perspective has a critical disadvantage: the natural error metric -- the difference between the estimated MSM and true propagator is intrinsically uncomputable.
+Instead, a compelling alternative is to view the problem as one of time series \emph{prediction}.
+We attempt to build models which predict new data.
+Here, the quality of the prediction (as measured for example by the likelihood that the model would generate an observed sample) is a natural measure of the quality of the model.
+
+%The basic framework for model selection intimated in the section involves minimization of an error metric characteristics of the model -- for example the norm of the between the model's eigenvalues and the propagator's true eigenvalues, or the L2 norm of the eigenvectors ( I think frank worked with this in one paper).
+Unfortunately, these quantities are not computable from a given model $M = \{S, T\}$ and data set $D$.
+The only access we have to the true propagator is via sampling new data.
+Therefore, the only way to know that we have constructed a model that is consistent with the propagator is to test how well the model would predict the new data.
+
+%The only available access to the information in the propgator is our sample $D$. For this reason, we instead emphasize the ability of the model to predict $D$.
+
+%On one hand, $M$ can be viewed as being fit _on_ $D$, but from another perspective $M$ is something that can _generate_ $D$. We thus can consider how predictive $M$ is.
+
+\begin{itemize}
+\item We always have finite data, and so this variational approach ignores another source of error due to the statistical error
+\item What we really want is a model that reproduces the dynamics due to the hamiltonian
+\item In practice, the transfer operator is inaccessible, and what you have is data and a model that attempts to describe the data.
+\item But the only way to access the true transfer operator is to sample from it and produce new data. Therefore, we should actually attempt to estimate a model that can best predict new data.
+\item This MSM has since been referred to as an object that takes in data, but in fact (as it is generative) we are parameterizing it to generate new data. 
+\item So if we build many MSMs, the ``best" model is the one that is most likely to have produced the newly observed data. 
+\item This means that we should construct a model such that it assigns a high probability to the new data. 
+\item This is precisely the idea of cross-validation or BIC/AIC (maybe we should define them here).
+\item The true data is a random variable (that we can sample), but in an MSM, the trajectory sampled is a random variable
+\end{itemize}
 

### 2014-04-06 paper/paper2.tex
got up to Results and Discussion.

--- 2014-04-06 paper/paper2.tex
@@
 \usepackage{achemso}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
 There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
-$$
+\begin{equation}
 p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
-$$
-
+\end{equation}
 The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
-$$
+\begin{equation}
 p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
-$$
+\end{equation}
 where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
-$$
+\begin{equation}
 \langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
-$$
+\end{equation}
 For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Generation}.
@@
 Let $\mathbf{p} \in (\mathbb{R} \cap [0,1])^k$ be a column vector whose entries sum to one.
-The MSM approximates $p(\mathbf{x})$ by
-$$
-p(\mathbf{x}) = \sum_{i=1}^{k} p_i \; \mathbf{1}_{s_i}(\mathbf{x}),
-$$
-where
-$$
-\mathbf{1}_{s}(\mathbf{x}) = \left\{\begin{array}{cc}
-1 & \mathbf{x} \in s \\
-0 & \text{otherwise}
-\end{array}\right.
-$$
+The elements in $\mathbf{p}$ are defined by a coarse-graining of $p(\mathbf{x})$ over $S$, as
+\begin{equation}
+\label{eq:coarse-grain-p}
+p_i = \int_{\mathbf{x} \in s_i} d\mathbf{x}\; p(\mathbf{x}).
+\end{equation} 
 Consider a probability vector at time $t$ denoted $\mathbf{p}(t)$. The time-evolution of $\mathbf{p}(t)$ is described by
-$$
+\begin{equation}
 \mathbf{p}(t+\tau)^T = \mathbf{p}(t)^T \mathbf{T},
-$$
+\end{equation}
 where $T_{ij}$ is the probability of transiting to state $j$ in time $\tau$ given that the system started in state $i$.
@@
 To construct an MSM, both $S$ and $\mathbf{T}$ must be determined from the data.
-${S}$ is defined by $\sigma$, which can be constructed by clustering the points in $D$.
-Given $S$, maximum-likelihood estimators for the the transition probabilities $\mathbf{T}$ exist\cite{kyle, frank}.
+The state-space, $S$, can be constructed by clustering the points in $D$ or via a grid-based discretization of $\Omega$.
+Given $S$, maximum-likelihood estimators for the $O(k^2)$ transition probabilities $\mathbf{T}$ exist\cite{kyle, frank}.
 
 Using this procedure, it has been shown that MSM estimates of the eigenvalues $\lambda_i$ are systematically underestimated in the limit of infinite data, and that the magnitude of this bias goes to zero as $k\rightarrow \infty$.
-This fact has inspired the development of variational methods for model selection, which optimize the MSM parameters in order to maximize the eigenvalues.
+This fact has inspired the development of variational methods for model selection, which optimize the MSM parameters in order to maximize the eigenvalues \cite{feliks}.
 
-\section{bayesian statistics}
-\begin{itemize}
-\item in the bayesian world, we would integrate over all possible models, but if we want a single model, then we should pick the most likely one
-\end{itemize}
+\section{The Bias-Variance Dilemma}
+For MSM methods to be valuable to practitioners, they must operate in limited data regimes.
+Thus, we seek a method for model selection that is sensitive to both the systematic error discussed above, as well as the statistical error associated with estimating the model's parameters.
+Whereas model selection based on the variational principle described above would tend to increase the number of states without bound, as the magnitude of the eigenvalue bias decreases with the number of states, practical application of this criteria would cause a catastrophic increase in statistical uncertainty in the model.
+This tradeoff between low-variance (small $k$) but biased estimators, and high-variance (large $k$) but unbiased estimators is a general feature of problems in statistical learning\cite{}.
+For example, as we show in appendix A, the uncertainty in the posterior distribution of $\mathbf{T}$ given $S$ and $D$ grows as $O(k)$.
 
-Given a Markov model parameterized at a lag time of $\tau$, we can propagate ensembles of trajectories via the transition probability matrix:
-$$\mathbf{p}_\tau^T = \mathbf{p}_0^T T(\tau)$$ where $\mathbf{p}_t$ is a probability distribution over states at time $t$ and $T(\tau)$ is a matrix with entries corresponding to the probability of transferring from one state to another at a time of $\tau$.
-As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a different model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
-$$\mathbf{p}_{n\tau}^T = \mathbf{p}_0^T T(n\tau) = \mathbf{p}_0^T \left[T(\tau)\right]^n $$ As such, the eigenvalues and eigenvectors of models built at many lag times are related.
-Specifically, when building multiple models for the same data set, the eigenvectors should be invariant to changing lag time, while the eigenvalues change via the relation:
-$$\lambda(\tau)^n = \lambda(n \tau)$$ The largest, non-unit eigenvalues correspond to the slowest timescales in the data.
-It's often useful to compute the slowest implied timescales ($t_i$) via the relation:
-$$t_i = - \frac{\tau}{\log \lambda_i} $$ These implied timescales should be invariant with respect to the lag time and therefore provide a useful tool for testing the quality of an MSM.
-Specifically, many models are built at varying lag times, and one is selected by choosing the model whose timescales have converged.
-
-\section{Prior Work}
-%After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.
-MSM model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time.
-In practice, while this test has been useful for many \cite{Noe2009Constructing}, it has some flaws.
-In particular, it is generally observed that most models' timescales do not converge, but steadily increase, making it difficult to judge whether a particular model passes the test and encouraging the use of ever-more states.
-Additionally, it is difficult to compare models built on two state decompositions, as neither may have convergent timescales.
-Often, the choice is made to use the model whose timescales better match an experimental observable.
-
-Other evaluation strategies include a test based on comparison to a second-order Markov model \cite{Park2006Validation}.
-Additionally, Bayesian methods that involve marginalizing over all possible MSMs have been developed \cite{Bacallado2009Bayesian, *Bacallado2011Bayesian}.
-However, these methods are often prohibitively expensive on larger data sets.
-In contrast, \citet{Kellogg2012Evaluation} proposed a tractable method for comparing MSMs built with different state decompositions, based on a likelihood function for an MD trajectory given an MSM, discussed below.
+Introduction to MSMs via the propagator theory motivates the view of the parameterization problem as one of \emph{function estimation}.
+This perspective has a critical disadvantage: the natural error metric -- the difference between the estimated MSM and true propagator is intrinsically uncomputable.
+The only access to the true propagator is via samples from the Hamiltonian with MD.
+Therefore, a compelling alternative is to view the problem as one of \emph{prediction}.
+Since the MSM is a generative model for trajectories in $\Omega$, it is possible to sample ``pseudo-trajectories" directly from the MSM.
+If the MSM were perfect, these ``pseudo-trajectories" would be indistinguishable from trajectories generated by the true propagator.
+Thus, the probability assigned by the MSM, to samples from the true propagator is a measure of the MSM's accuracy.
+This argument is formalized by Bayes' rule, which establishes a proportionality between the probability of a model given data to the probability of the data given the model
+\begin{equation}
+\label{eq:bayes}
+P\Big(\{S, \mathbf{T}\} \;\Big|\; D\Big) = P\Big(D \;\Big|\; \{S, \mathbf{T}\}\Big) \frac{P(\{S, \mathbf{T}\})}{P(D)},
+\end{equation}
+where $P\Big(\{S, \mathbf{T}\} \;\Big|\; D\Big)$ is the posterior probability of an MSM, $P\Big(D \;\Big|\; \{S, \mathbf{T}\}\Big)$ is the \emph{likelihood} of the data given a model, $P(\{S, \mathbf{T}\})$ is the prior probability of an MSM before observing any data, and $P(D)$ is the probability of the data, a constant.
 
@@
 Without a strong prior, model selection is reduced to the search for the model that maximizes the likelihood of the data.
-
-Thus we seek to formulate this likelihood function for Markov state models.
-With the kinetic model expressed as a set of pairwise state to state transition probabilities at a given lag time, the likelihood of an ensemble of trajectories after projection into the MSM state space is given simply by the product of the transition matrix elements along the observed trajectories.
-However, as we vary the number of states, it is not permissible to simply compare these likelihoods as part of an optimization of the state definitions.
-In doing so, the optimal model would always be the trivial one state model, whose computed likelihood is unity regardless of the data.
-
-The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis.
-With $s(\mathbf{x})$ as the function that maps a conformation in phase space ($\mathbf{x}$) to the index of a state in the discrete state space, and $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory, the likelihood can be written as:
-
+With $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory of length $N$, the likelihood can be written as:
 \begin{equation}
-\begin{split}
-P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
-\cdot \prod_{t=1}^{N} P(\mathbf{x}_t | s(\mathbf{x}_t)) \cdot d\mathbf{x}^N
+P\big(\{\mathbf{x}_t\}_{t=1}^N\;\big|\; \{S, T\}\big)d\mathbf{x}^N = \prod_{t=1}^{N-1} T_{\sigma(\mathbf{x}_t), \sigma(\mathbf{x}_{t+1})} \cdot \prod_{t=1}^{N} P(\mathbf{x}_t | \sigma(\mathbf{x}_t)) \cdot d\mathbf{x}^N,
 \label{eq:like}
-\end{split}
 \end{equation} 
-
-With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state Markov transition probabilities, $T(s_i \rightarrow s_j) \equiv P(S_{t+1}=s_j | S_{t}=s_i)$, and so-called emission distributions of each state, $P(\mathbf{x} | s(\mathbf{x}))$, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}$, is within a certain state, $s(\mathbf{x})$.
-As we are using  non-overlapping states, the conditional probability of observing a conformation, $\mathbf{x}$ from state $i$ is zero unless $\mathbf{x}$ is assigned to $i$.
-Mathematically, this means:
-$$ P(\mathbf{x} | i) = 0 \;\;\;\; \forall \mathbf{x} \textnormal{ s.t. } s(\mathbf{x}) \neq i$$
-
-This emission distribution can be modeled in multiple ways, each of which, in combination with a transition matrix, parametrizes a different statistical model.
-\citet{Kellogg2012Evaluation} used a discriminative model with:
+where $P(\mathbf{x}_t | \sigma(\mathbf{x}_t))$ is the probability of sampling a conformation $\mathbf{x}_t$ given that the process is in the state $\sigma(\mathbf{x}_t)$, referred to as the emission distribution of state $\sigma(\mathbf{x}_t)$.
+There is some flexibility in the choice of emission distribution, though \cref{eq:like} is valid only when the emission distributions satisfy
 \begin{equation}
-P(\mathbf{x}_t | s(\mathbf{x}_t)) = \big| \{ \mathbf{x}_k : s(\mathbf{x}_k) = s(\mathbf{x}_t) \} \big|^{-1}
-\end{equation} where $| \cdot |$ denotes the cardinality of a set.
-This is a discriminative model, whose support is the training data set rather than phase space, which makes it unable to generalize and assign probability to new data.
+P(\mathbf{x} | i) = 0 \;\;\;\; \forall \; \mathbf{x} \textnormal{ s.t. } \sigma(\mathbf{x}) \neq i.
+\end{equation}
+Given the coarse-graining of $p(\mathbf{x})$ in \cref{eq:coarse-grain-p}, the most natural choice emission distributions are normalized indicator functions
+\begin{equation}
+P(\mathbf{x} \;|\; i) = \frac{1}{V_i} \mathbf{1}_{s_i}(\mathbf{x}).
+\end{equation}
+An alternative emission distribution was proposed by \citet{Kellogg2012Evaluation} Instead of having support on $\Omega$, this distribution is supported on the training set $\{\mathbf{x}_t\}_{t=1}^N$.
+It is given by
+\begin{equation}
+P(\mathbf{x} \;|\; i) = \big| \{ \mathbf{x}_k : \sigma(\mathbf{x}_k) = \sigma(\mathbf{x}) \} \big|^{-1} \sum_{t=1}^N \delta(\mathbf{x}-\mathbf{x}_t),
+\end{equation} where $| \cdot |$ denotes the cardinality of a set and $\delta(\cdot)$ is the Dirac delta function.
+As this is a discriminative model, it is unable to generalize and assign probability to new data.
 In certain circumstances, this is an undesirable property.
@@
 
-Instead, we seek a generative model whose emission distributions are supported on phase space (i.e.
-standard distributions over $\mathbb{R}^{3N_{atoms}}$).
-We propose one possible emission distributions, $P(\mathbf{x} | s(\mathbf{x}))$: a uniform distribution over the (hyper)volume of each state, as shown in \cref{eq:like_vol}.
-\begin{equation}
-\label{eq:like_vol}
-\begin{split}
-%P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
- P(\mathbf{x}_t | s(\mathbf{x}_t)) = \frac{1}{V_{s(\mathbf{x}_{t})}}
-\end{split}
-\end{equation} 
-This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation.
-%For example, consider two Markov state models sharing the same transition matrix, $T$.
-In one model, the state emission distributions are highly peaked at specific locations in phase space, whereas in the other model the emission distributions are uniform over the volume of the states.
-If an observed trajectory does go through the first models' regions of high likelihood, it is appropriately termed a more likely model given the data.
-
-%The simplest definition would be a uniform distribution since we are constructing an MSM that is agnostic to intra-state dynamics.
-When properly normalized, the uniform distribution takes a value of the reciprocal of the state's volume.
-This makes for an easy expression, however, volumes in high-dimensional states can be difficult to compute \ref{eq:like_vol}.
-For this reason, we can also choose other emission distributions that may be more tractable.
-In fact, \citet{Pelleg2000Xmeans} have used a gaussian emission distribution to define a likelihood to compare two k-means clusterings.
-The result of this emission distribution, is that the likelihood is proportional to the sum of square distances, which is precisely what is optimized in k-means clustering.
-If we think of k-centers as a poor man's version of k-means, then we can apply this same likelihood.
-
-\section{Cross Validation and the Bayes Information Criterion}
-Likelihood maximization is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to over-fitting\cite{Liddle2007Information, Hastie01Elements}.
-Statistical learning theory provides a number of alternative approaches for this problem.
+\section{Statistical Model Selection}
+Likelihood maximization on the training data (i.e.
+the data used to fit the model) is insufficient for model selection when the number of parameters varies between proposed models, as more complex models generally exhibit higher empirical likelihoods, often at the cost of larger generalization errors due to over-fitting\cite{Liddle2007Information, Hastie01Elements}.
+Statistical learning theory provides a number of methods to overcome this problem.
 Conceptually, the most straightforward is a full Bayesian treatment in which all unknown model parameters are represented by probability distributions.
-The evidence for a model is computed by formally integrating over the model parameters and the evidence ratio, or Bayes factor\cite{Gelfand94Bayesian}, then provides a rigorous basis of model selection that appropriately punishes overly complex models as they become poorly constrained in parameter space.
-Unfortunately such approaches are intractable for problems of this size because of the need to integrate over all possible Markov models of a given size.
+The evidence for a $k$-state model is computed by formally integrating \cref{eq:bayes} over the model parameters and the evidence ratio, or Bayes factor\cite{Gelfand94Bayesian}, then provides a rigorous basis of model selection that appropriately punishes overly complex models as they become poorly constrained in parameter space.
+Unfortunately such approaches are impractical for problems of this size because of the need to integrate over all possible Markov models of a given size.
 
@@
 This scheme can be repeated many times with different partitions of the data set.
-The AIC and BIC are augmented likelihood function approaches which do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.
+The AIC and BIC are augmented likelihood function approaches that do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.
 \begin{equation}
@@
 
+%\section{Prior Work}
+
+%Given a Markov model parameterized at a lag time of $\tau$, we can propagate ensembles of trajectories via the transition probability matrix:
+%$$\mathbf{p}_\tau^T = \mathbf{p}_0^T T(\tau)$$ where $\mathbf{p}_t$ is a probability distribution over states at time $t$ and $T(\tau)$ is a matrix with entries corresponding to the probability of transferring from one state to another at a time of $\tau$.
+As suggested by \citet{Swope2004Describing}, given a Markov model at a lag time of $\tau$, $T(\tau)$, we expect that a different model with a lag time of $n\tau$ should merely be the $n^\textnormal{th}$ power of $T(\tau)$:
+%$$\mathbf{p}_{n\tau}^T = \mathbf{p}_0^T T(n\tau) = \mathbf{p}_0^T \left[T(\tau)\right]^n $$ As such, the eigenvalues and eigenvectors of models built at many lag times are related.
+Specifically, when building multiple models for the same data set, the eigenvectors should be invariant to changing lag time, while the eigenvalues change via the relation:
+%$$\lambda(\tau)^n = \lambda(n \tau)$$ The largest, non-unit eigenvalues correspond to the slowest timescales in the data.
+It's often useful to compute the slowest implied timescales ($t_i$) via the relation:
+%$$t_i = - \frac{\tau}{\log \lambda_i} $$ These implied timescales should be invariant with respect to the lag time and therefore provide a useful tool for testing the quality of an MSM.
+Specifically, many models are built at varying lag times, and one is selected by choosing the model whose timescales have converged.
+
+%After diagonalization of the transition matrix, the relaxation of any initial probability distribution can be described by a superposition of exponential relaxation processes, with a characteristic time given by $-\frac{1}{\ln \lambda_i}$.
+%MSM model selection has traditionally been performed by analyzing the self-consistency of the models with respect to changes in the Markov lag time.
+In practice, while this test has been useful for many \cite{Noe2009Constructing}, it has some flaws.
+In particular, it is generally observed that most models' timescales do not converge, but steadily increase, making it difficult to judge whether a particular model passes the test and encouraging the use of ever-more states.
+Additionally, it is difficult to compare models built on two state decompositions, as neither may have convergent timescales.
+Often, the choice is made to use the model whose timescales better match an experimental observable.
+
+%Other evaluation strategies include a test based on comparison to a second-order Markov model \cite{Park2006Validation}.
+Additionally, Bayesian methods that involve marginalizing over all possible MSMs have been developed \cite{Bacallado2009Bayesian, *Bacallado2011Bayesian}.
+However, these methods are often prohibitively expensive on larger data sets.
+In contrast, \citet{Kellogg2012Evaluation} proposed a tractable method for comparing MSMs built with different state decompositions, based on a likelihood function for an MD trajectory given an MSM, discussed below.
+
 \section{Computational Methods}
+
+\subsection{Volume Estimation}
+The uniform distribution emission model presents a computational challenge: its use requires the calculation of the (hyper)volume of the MSM's states, which, when defined by clustering, are high-dimensional Voronoi cells.
+While trivial in two or three dimensions, this computational geometry task becomes challenging in high-dimensional settings.
+The computation of such volumes has occupied significant attention in recent years in the computational geometry literature, especially via randomized algorithms\cite{Kannan97, Simonovits03, Lovasz03}.
+We opt to approximate the volumes using Monte Carlo integration, which we find tractable for large systems only when the molecular dynamics data set is first projected into a suitable small vector space of up to perhaps ten dimensions.
+
+A further challenge is the description of the states that are at the edge of the MSM -- whose Voronoi cells extend to infinity in at least one direction.
+In these cases, the Voronoi cells are of unbounded volume.
+Instead we wish to truncate these states by bounding them by the convex hull of the data set.
+Because the convex hull of our simulation data sets are computationally inaccessible, we defined the accessible volume of $\Omega$ denoted $A$, to be the set of all points within a distance $R$ within a set of test points $Y$.
+
+\begin{algorithm}
+\caption{Monte Carlo Estimation of the State Volumes}
+\begin{algorithmic}[1]
+\Procedure{MC Volume Estimation}{$\sigma, A, M$}
+	\State $\mathbf{c} \gets [0, 0, \ldots, 0]$\Comment{$\mathbf{c}$ is a vector of length $k$}
+	\While{$\sum_{i} c_i \leq M$}
+		\State $\mathbf{x} \gets $ sample from axis-aligned hyper cube containing $A$
+		\If{$\mathbf{x} \in A$}
+			\State $c_{\sigma(\mathbf{x})} \gets c_{\sigma(\mathbf{x})} + 1$
+		\EndIf
+	\EndWhile
+	\State $\mathbf{v} \gets \mathbf{c} \cdot(\sum_{i=1}^k c_i)^{-1}$ \Comment{$\mathbf{v}$ contains the relative volumes of each state}
+	\State \textbf{return} $\mathbf{v}$
+\EndProcedure
+\end{algorithmic}
+\end{algorithm} \noindent The result of this scheme produces the volumes of each state relative to the volume of $A$. It's important, therefore, to use the same set $A$ when comparing multiple models.
+\subsection{Simulation Protocol}
 
@@
 \caption{Systems studied in this work.
-(A) Langevin dynamics on the two dimensional M\"{u}ller potential \cite{Muller1980Reaction}.
+(A) Brownian dynamics on the two dimensional M\"{u}ller potential \cite{Muller1979Location}.
 (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E.
@@
 
-The uniform distribution emission model presents a computational challenge: its use requires the calculation of the (hyper)volume of the MSM's states, which, when defined by clustering, are high-dimensional Voronoi cells.
-While trivial in two or three dimensions, this computational geometry task becomes challenging in high-dimensional settings.
-The computation of such volumes has occupied significant attention in recent years in the computational geometry literature, especially via randomized algorithms\cite{Kannan97, Simonovits03, Lovasz03}.
-We opt to approximate the volumes using naive Monte Carlo rejection sampling, which we find tractable for large systems only when the molecular dynamics data set is first projected into a suitable small vector space of up to perhaps ten dimensions.
+Two systems were investigated using this likelihood scheme. The first was a simple two dimensional surface with three energy minima, called the M\"{u}ller potential. The dynamics are governed by
+$$
+\frac{d\mathbf{x}}{dt} = - \nabla V(\mathbf{x})\zeta + \sqrt{2kT\zeta} R(t),
+$$ where $\zeta = 10^{-3}$, $kT = 15$, $R(t)$ is a delta-correlated Gaussian process with zero mean, and $V(\mathbf{x})$ was defined as
+$$
+V(\mathbf{x}) =  \sum_{j=1}^4 A_j \cdot \exp\Big(a_j (x_1-X_j)^2+b_j(x_1-X_j)(x_2-Y_j)+c_j(x_2-Y_j)^2\Big),
+$$ where $a = (-1, -1, -6.5, 0.7)$; $b = (0, 0, 11, 0.6)$; $c = (-10, -10, -6.5, 0.7)$; $A = (-200, -100, -170, 15)$; $X = (1, 0, -0.5, -1)$; $Y = (0, 0.5, 1.5, 1)$ as suggested by \citet{Muller1979Location}.
+Using the Euler-Maruyama method and a time step of $0.1$, we produced two trajectories of length $10^6$ time steps.
 
-A further challenge is the description of the states that are at the edge of the MSM -- whose Voronoi cells extend to infinity in at least one direction.
-In these cases, the Voronoi cells are of unbounded volume.
-Instead we wish to truncate these states by bounding them by the convex hull of the data set.
-Because the convex hull of our simulation data sets are computationally inaccessible, we define the outer extent of our data sets to be the set of all trial points such that the nearest sampled conformation to the trial point is closer than a certain cutoff, $R$.
+The first trajectory was clustered using the $k$-centers clustering algorithm with the Euclidean distance.
+State volumes were computed for the uniform emission model using $M=10^5$ rounds of Monte Carlo integration defining the set $A$ using the cluster centers from the 50-state model as the test points $Y$ with a cutoff of $R=0.28$.
+The second trajectory was used as a test set.
+All MSMs were built using a lag time of 30 steps.
 
-Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes.
-For this reason, we find it essential to use a dense prior during transition matrix estimation.
-Using a prior ensures that all transitions have a nonzero (albeit small) probability.
-We used a Dirichlet prior with a pseudo-count of $\frac{1}{k}$ for each possible transition, where $k$ is the number of states in the model\cite{msmb2?, nina?, wherever this happens to be}.
+Next, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E.
+Shaw Research \cite{Shaw2010Atomic} (Amber ff99SB-ILDN force field,\cite{amber} TIP3P water model\cite{tip3p}).
+We projected the trajectories into a four dimensional space using time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification} on the first trajectory.
+To calculate the tICs, each conformation was represented as a vector of distances between all pairs of residues separated by at least three amino acids.
+The distance $d(A, B)$ between residues $A$ and $B$ was determined as follows.
+Let $\{a_i\}_{i=1}^{n_a}$ and $\{b_i\}_{i=1}^{n_b}$ be the cartesian coordinates of the heavy (non-hydrogen) atoms in $A$ and $B$.
+Then we define
+$$
+d(A,B) \equiv \min_{i, j} ||a_i - b_j||_2,
+$$ where $|| \cdot ||_2$ denotes the $\ell_2$ norm.
+The tICs were computed using a correlation lag time of 200 ns.
+We also performed Principal Components Analysis (PCA) on the same residue-residue distance representation and built MSMs using the top three PCs.
+In each projection, $k$-centers was used with the Euclidean distance metric.
+
+For the tICA MSMs, the cluster centers from the 100 state model were used as the test points $Y$ to define $A$ with a cutoff of $R=0.81$.
+In the PCA MSMs, the cluster centers from the 500 state model were used as the test points with a cutoff of $R=2.0$.
+In each case, $M=10^6$ rounds of Monte Carlo integration were performed to compute the state volumes.
+
+For all models, a maximum likelihood estimator for reversible transition matrices at a given lag time, with a pseudocount of $1 / k$ was used to compute $\mathbf{T}$.
+The pseudocount ensures that all transitions are assigned nonzero probability, which is especially important for evaluating data that was not used to train the model, as often new transitions will be observed.
 
@@
 \subsection{M\"{u}ller Potential}
-The M\"{u}ller potential, is a simple two-dimensional surface with three energy minima.
-We simulated two trajectories with $10^6$ steps of Langevin dynamics on the M\"{u}ller potential\cite{Muller1980Reaction}.
-Simulations were performed with a timestep $t=0.1$, friction coefficient of $\gamma=10^3$, and $kT=15$.
-
-The first trajectory was clustered using the $k$-centers clustering algorithm, and state volumes were computed for the uniform emission model using $10^5$ rounds of Monte Carlo rejection sampling with a cutoff of $R=0.28$ from the 50 state model's centroids.
-(To give a sense of scale, the x-axis of the M\"{u}ller potential shown in \cref{fig:pics} is 4 units across).
-The second trajectory was used as a test set.
-All MSMs were built using a lag time of 30 steps.
-
-The volumes did not change drastically upon doing ten times more Monte Carlo rejection sampling, and the likelihood remained essentially constant, which indicates that the use of approximate volumes is sufficient to obtain robust estimates of the likelihood.
 
@@
 
-We reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E.
-Shaw Research \cite{Shaw2010Atomic}.
-Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification}, extracting only the four slowest uncorrelated linear combinations of residue-residue distances.
  All MSMs were built using a lag time of 50 ns.
 
+Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with a dimensionality reduction technique.
 \begin{figure}

### 2014-04-06 paper/paper2.tex
added final figures and got through muller results

@@
 \begin{equation}
+\label{eq:prop-timescales}
 p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
@@
 With this construction, the eigenvalues and eigenvectors of $\mathbf{T}$ \hl{provide approximations (what about transfer operator?)} for the eigenvalues and eigenfunctions of $\mathcal{Q}$.
+In a direct analogy with \cref{eq:prop-timescales}, the eigenvalues $\lambda_i$ of $T$ correspond to relaxation timescales of the Markov model
+\begin{equation}
+\label{eq:msm-timescales}
+t_i = -\frac{\tau}{\ln \lambda_i}.
+\end{equation}
 
@@
 This fact has inspired the development of variational methods for model selection, which optimize the MSM parameters in order to maximize the eigenvalues \cite{feliks}.
+
+Selecting the size of $S$ is an important step in model construction, however there is not currently a widely-accepted method in use.
+Heuristic criteria, such as manually selecting $k = n_\textnormal{conformation} / 14$ or clustering $D$ such that each conformation is within a pre-specified distance to its cluster center (e.g.
+4.5\AA$\;$ or 1.2\AA $\;$ RMSD) have been previously employed\cite{Lane2012, Bowman2012, Beuchamp}.
+Selection of $\tau$ based convergence of the relaxation timescales (\cref{eq:msm-timescales}), as suggested by \citet{Swope2004Describing} and others is more well established.
+The rate of this convergence with respect to $\tau$ at fixed $S$ has also been used as a metric for evaluating a model's state decomposition.
+\hl{cite someone?}
 
@@
 
-\section{Computational Methods}
+\section{Methods}
+\label{sec:methods}
 
@@
 In each case, $M=10^6$ rounds of Monte Carlo integration were performed to compute the state volumes.
+All WW MSMs were built using a lag time of 50 ns, which is the same lag time used in previous analyses\cite{Beauchamp2012Simple}.
 
-For all models, a maximum likelihood estimator for reversible transition matrices at a given lag time, with a pseudocount of $1 / k$ was used to compute $\mathbf{T}$.
+For both the M\"{u}ller potential and WW models, a maximum likelihood estimator for reversible transition matrices, with a pseudocount of $1 / k$ was used to compute $\mathbf{T}$.
 The pseudocount ensures that all transitions are assigned nonzero probability, which is especially important for evaluating data that was not used to train the model, as often new transitions will be observed.
+All analysis and model construction was performed using MSMBuilder \cite{Beauchamp2011Msmbuilder2}.
 
@@
 
+How well can likelihood-based methods select parameters for an MSM in the data-rich regime?
+To answer this question, we simulated a simple two-dimensional potential and built a series of MSMs (see ``Methods").
+We compared four different model selection criteria: (1) test log-likelihood (2) BIC (3) AIC and (4) implied timescales convergence.
+The analysis \cref{fig:mullerlike} indicates that the log-likelihood on the training data set continually increases with the addition of further states.
+The increase is most dramatic at low $k$, and levels off at high $k$, indicating that once $k$ is large enough, the marginal increase in the likelihood with respect to further states is small.
+For this data set, the log-likelihoods computed on a separate test set exhibit a notably similar trend.
+Above $k=600$, the test log-likelihood stops increasing.
+These two trends are to be expected: the additional parameters in the model at higher $k$ never decreases the model's ability to fit the training data.
+On the other hand, this can lead to the model fitting the statistical noise in the training data set, rather than the system's true dynamics.
+As result, these "over fit" models at high $k$ are less predictive, as measured by the log-likelihood they assign to new test data.
+
+Leaving out a portion of a molecular dynamics data set from the training to serve as a test set is costly proposition for the applications of MSMs to the study of complex system.
+For large bimolecular simulations characterized by long intrinsic timescales, sampling the potential exhaustively is a significant challenge\cite{Lane2012}, making it difficult to afford discarding half of the dataset during the fitting of an MSM.
+For this reason, we also computed two penalized likelihood model selection criteria, the AIC and BIC, which augment the training set log-likelihood with an explicit penalty on model complexity to avoid over fitting.
+Applied to our simulations of the M\"{u}ller potential, both the AIC and BIC penalize model complexity more strongly than the test set log-likelihood.
+
+The likelihood-based methods are consistent with model selection based on maximization of the implied timescales.
+For this simple data set, the timescales are maximized by models with 200 states, which agrees with the results obtained with BIC.
+Model selection based on direct maximization of the implied timescales considers only the systematic error in the MSM, and neglects the statistical error.
+The consistency between the two approaches on this data set is an indication that the statistical uncertainty is low for these models, which is to be expected given the ease of sampling this two dimensional toy potential.
+
 \begin{figure}[h]
 \centering
-\includegraphics[width=4in]{figs_final/mull_likes.png}
+\includegraphics[width=6in]{figs/mull/mull.pdf}
 \caption{\hl{Fix this figure so that only part (A) is depicted.} We calculated the training and test set log-likelihoods and model selection criteria for models built with 50-1000 states using the uniform emission distribution.
@@
 
- All MSMs were built using a lag time of 50 ns.
-
 Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with a dimensionality reduction technique.
@@
 \centering
-\includegraphics[width=3in]{figs_final/ww_likes.png}
+\includegraphics[width=6in]{figs/ww_tica/ww_tica.pdf}
 \caption{\hl{Fix this figure so that only part (A) is depicted.} On models whose states were defined after projecting into the four-dimensional tICA subspace, we computed likelihoods based on the uniform emission models.
@@
 \centering
-\includegraphics[width=6in]{figs_final/timescales_suck2.pdf}
+\includegraphics[width=6in]{figs/ww_pca/ww_pca.pdf}
 \caption{For MSMs of the WW domain built on the top three PCA components,} 

### 2014-04-06 paper/paper2.tex
Finish the last section

@@
 \documentclass[journal=jpcbfk, layout=traditional, manuscript=article]{achemso}
-\setkeys{acs}{articletitle = true}
+% \setkeys{acs}{articletitle = true}
 \usepackage[utf8]{inputenc}
@@
 Two competing sources of error govern the accuracy of an MSM.
-The first source of error is systematic and is common to all dimensionality reduction methods.
-\hl{Any simplified representation of a complex system cannot generally describe arbitrary behavior of that system losslessly.} The second source of error is statistical in nature, in that the parameters of an MSM must be estimated from a finite number of stochastic MD trajectories.
-\hl{As we discuss later,?} These two sources of error compete in a bias-variance tradeoff\cite{bias_variance_paper}.
+The first source of error is systematic and is common to all dimensionality reduction methods, in that simplified representations of complex systems discard information.
+The second source of error is statistical in nature, in that the parameters of an MSM must be estimated from a finite number of stochastic MD trajectories.
+As we discuss below, these two sources of error compete in a bias-variance tradeoff\cite{bias_variance_paper}.
 Given a finite size data set, as the number of states increases, the systematic error (the bias) decreases, while the statistical error (the variance) increases.
@@
 
-The MSM models the \hl{action?} of the propagator with a Markov jump process on $\{1, \ldots, k\}$.
+The MSM models the dynamics of the propagator with a Markov jump process on $\{1, \ldots, k\}$.
 Let $\mathbf{p} \in (\mathbb{R} \cap [0,1])^k$ be a column vector whose entries sum to one.
@@
 where $T_{ij}$ is the probability of transiting to state $j$ in time $\tau$ given that the system started in state $i$.
-With this construction, the eigenvalues and eigenvectors of $\mathbf{T}$ \hl{provide approximations (what about transfer operator?)} for the eigenvalues and eigenfunctions of $\mathcal{Q}$.
+With this construction, the eigenvalues and eigenvectors of $\mathbf{T}$ provide approximations for the eigenvalues and eigenfunctions of $\mathcal{Q}$.
 In a direct analogy with \cref{eq:prop-timescales}, the eigenvalues $\lambda_i$ of $T$ correspond to relaxation timescales of the Markov model
@@
 This tradeoff between low-variance (small $k$) but biased estimators, and high-variance (large $k$) but unbiased estimators is a general feature of problems in statistical learning\cite{}.
-For example, as we show in appendix A, the uncertainty in the posterior distribution of $\mathbf{T}$ given $S$ and $D$ grows as $O(k)$.
+For example, it can be shown that the uncertainty in the posterior distribution of $\mathbf{T}$ given $S$ and $D$ approaches infinity as $k$ approaches infinity.
 
@@
 \caption{Monte Carlo Estimation of the State Volumes}
+\label{alg:volumes}
 \begin{algorithmic}[1]
@@
 How well can likelihood-based methods select parameters for an MSM in the data-rich regime?
-To answer this question, we simulated a simple two-dimensional potential and built a series of MSMs (see ``Methods").
+To answer this question, we simulated a simple two-dimensional potential and built a series of MSMs (see \emph{Methods}).
 We compared four different model selection criteria: (1) test log-likelihood (2) BIC (3) AIC and (4) implied timescales convergence.
@@
 \includegraphics[width=6in]{figs/mull/mull.pdf}
-\caption{\hl{Fix this figure so that only part (A) is depicted.} We calculated the training and test set log-likelihoods and model selection criteria for models built with 50-1000 states using the uniform emission distribution.
-Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray).
-For both likelihoods, the BIC penalizes complexity more than the AIC and test set log-likelihood.
-The optimal window is consistent with the convergence of the relaxation timescales (right frames).}
+\caption{Four model selection criteria were used (left panel) to select the state discretization for MSMs built on simulations of the M\"{u}ller potential.
+The log-likelihood was evaluated on the training set (black) and a separate test set (blue).
+These values are plotted as $-2\cdot \log L$ for ease of comparison with the BIC (red) and AIC (purple).
+Lower scores indicate better models.
+Dashed lines are used to indicate criterion-minimizing models.
+These criteria are consistent with the convergence of the implied timescales (right panel).}
 \label{fig:mullerlike}
@@
 \subsection{Fip35 WW Domain}
+How appropriate are likelihood-based model selection criteria for more realistic MD data sets?
+To answer this questions, we constructed Markov models based on simulations of the Fip35 WW domain\cite{Liu2008Experimental} performed on the Anton
+supercomputer\cite{Shaw2010Atomic}.
+As a preprocessing step, we used time-structure based independent components analysis (tICA) to project the data into a four dimensional space.
+This procedure was necessary in order to compute the volumes of the MSM states, as described in \cref{alg:volumes}.
+MSMs were constructed in this lower dimensional space with differing number of states using the k-centers clustering algorithm (see \emph{Methods}).
 
-Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with a dimensionality reduction technique.
-\begin{figure}
-\centering
-\includegraphics[width=6in]{figs/ww_tica/ww_tica.pdf}
-\caption{\hl{Fix this figure so that only part (A) is depicted.} On models whose states were defined after projecting into the four-dimensional tICA subspace, we computed likelihoods based on the uniform emission models.
-Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray).
-\label{fig:ww}}
-\end{figure}
-
-For the models built in the tICA subspace, the AIC is comparable to the test set log-likelihood scores using either emission model, displaying a optimum at ~600 states (\cref{fig:ww}).
-The BIC penalizes complexity more strongly with an optimum at ~150 states.
-This window is consistent with the convergence of the relaxation timescales and the onset of Markovian behavior.
-Models built with RMSD require significantly more states than the tICA MSMs and show poor timescale convergence.
-These results indicate why model selection based on relaxation timescales is so difficult, as a less than optimal state decomposition can result in timescales that do not converge rapidly (if at all).
-Furthermore, selection based on only the relaxation timescales contains no guard against over-fitting, which is of particular concern in the models built using RMSD that require thousands of states.
+As was the case with the M\"{u}ller MSMs, the log-likihood of the training data set increases with the addition of more states, while the test log-likelihood, AIC and BIC lack a monotonic trend (\cref{fig:ww-tica}).
+The test log likelihood has a maximum at 650 states, and decreases with the addition of further states.
+This trend is matched by the AIC, which is computed only on the training data, while the BIC on the other hand seems to over-penalize model complexity.
+The implied timescales of these MSMs converge rapidly, and selection based on their convergence is consistent with the statistical methods.
 
@@
 \centering
-\includegraphics[width=6in]{figs/ww_pca/ww_pca.pdf}
-\caption{For MSMs of the WW domain built on the top three PCA components,} 
+\label{fig:ww-tica}
+\includegraphics[width=6in]{figs/ww_tica/ww_tica.pdf}
+
+\caption{MSMs for the Fip35WW domain built with tICA were evaluated based on four model selection criteria.
+The log-likelihood was evaluated on the training set (black) and a separate test set (blue), and compared with the AIC (purple) and BIC (red) penalized log-likelihoods, computed on the training set.
+Dashed lines are used to indicate criterion-minimizing models.
+The models' ten slowest relaxation timescales (shown in the right panel) converge rapidly.
+\hl{right panel should be linear scale instead, for consistency with the PCA}}
 \end{figure}
 
-%The two 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain simulated by \citet{Shaw2010Atomic} have been used to parameterize multiple Markov state models.
-\citet{Lane2011Markov} proposed a 26,000 state MSM after clustering the data by RMSD, whereas Kellogg et al.,\cite{Kellogg2012Evaluation} clustering on contact maps or secondary structure, proposed 100 and 175 state models, respectively.
-Consistent with this spread, we find the optimal number of states to be highly dependent on the space in which the data is clustered.
-With RMSD, the Gaussian likelihood indicates that Markov models built with on the order of 5,000 states optimally balance model quality and generalization error.
-On the other hand, the tICA method's identification of slow order parameters in the simulations has the effect of reducing the number of states necessary when constructing Markov models.
-While only using 200-600 states, Markov models constructed using tICA were still able to recapitulate the folding timescale as well as the faster processes identified by McGibbon and Pande\cite{McGibbon2013Learning}.
+We also built a series of models on the same data set by replacing the tICA preprocessing with Principal Components Analysis (PCA).
+We projected the data set onto the top three principal components, which are the highest-variance uncorrelated linear combinations of the input degrees of freedom (see \emph{Methods}).
+Because PCA does not explicitly take into account the temporal structure of the data set, we expected that MSMs based on PCA would be less predictive than those based on tICA.
+Our results, shown in \cref{fig:ww-pca}, confirm this hypothesis.
+We built MSMs with as many as 25,000 states, and saw a linear trend in the longest relaxation timescale with respect to the number of states.
 
-\subsection{Limitations and Future Work}
+The statistical model selection methods indicate that we are only justified in using models with $k$ on the order of 100.
+The AIC, BIC and test set log-likelihood criteria each recommend a model built with 200 states, which is also the point just after the most dramatic increase in the longest relaxation timescale occurs with respect to $k$.
+On the other hand, the relaxation timescales continue to increase further, even as the test set log likelihood begins to decrease after 1000 states.
+
+\begin{figure}
+\centering
+\label{fig:ww-pca}
+\includegraphics[width=6in]{figs/ww_pca/ww_pca.pdf}
+\caption{The four selection criteria were used to compare MSMs for the Fip35WW domain built with PCA.
+The longest relaxation timescale (right panel) exhibits a linear increase up to 25,000 states.
+The AIC (blue), BIC (red) and test set log-likelihood each recommend a model with 200 states.}
+\end{figure}
 
 The likelihood functions described herein permit the comparison of Markov state models with varying number of states, however, they require that the compared models have the same support.
-As such, a direct comparison of the likelihoods between models built with tICA and models built with RMSD is not possible.
-Furthermore, the uniform emission model is intractable for all but the simplest systems.
-The Gaussian emission model does not have this limitation.
-
-In the future, we plan to extend this work to the consideration of models without discrete states, where the requirement that states strictly partition phase space into a set of discrete indicator functions is relaxed and the models are parameterized by a direct optimization of \cref{eq:like}.
-This strategy would complement approaches that generalize MSMs beyond discrete states\cite{Noe2013Variational}.
+As such, a direct comparison of the likelihoods between models built with tICA and models built with PCA is not justified.
 
@@
 
-These results take a step toward fully automating the MSM construction process and controlling the complexity and generalization error by providing a quantitative method for selecting the most suitably complex model, given the data.
+In the future, we plan to extend this work to the consideration of models without discrete states, where the requirement that states strictly partition phase space into a set of discrete indicator functions is relaxed and the models are parameterized by a direct optimization of \cref{eq:like}.
+This strategy would complement approaches that generalize MSMs beyond discrete states\cite{chiang2010markov, Noe2013Variational}.
 
@@
 
-\section{Appendix}
+\end{document}
+
 \subsection{Other emission distributions}
@@
 For evaluating the Gaussian emission likelihood, we used a value of $3 \cdot 252 - 6 = 750$ for the dimension of the emission distribution.
-
-\end{document}
 

### 2014-04-07 paper/paper2.tex
figure updates and fixed the citation issues

@@
 Whereas classical experimental probes have often been interpreted in two-state frameworks, ensemble measurements with increasingly high temporal resolution as well as sensitive single molecule probes have uncovered a vast array of complex multi-state kinetics \cite{Cosa2006Evidence, Zhang2011Direct}.
-But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{FRET_PAPER?, Mertens2010Structural, Tzeng2011Protein}.
+But the atomic-resolution characterization of these dynamics is often an enormous challenge -- as molecular probes like F\"{o}rster resonance energy transfer, small-angle x-ray scattering, and nuclear magnetic resonance techniques measure complex projections of the intrinsic structure, generally reporting simultaneously on many degrees of freedom\cite{Lipman2003Single, Mertens2010Structural, Tzeng2011Protein}.
 
@@
 The second source of error is statistical in nature, in that the parameters of an MSM must be estimated from a finite number of stochastic MD trajectories.
-As we discuss below, these two sources of error compete in a bias-variance tradeoff\cite{bias_variance_paper}.
+As we discuss below, these two sources of error compete in a bias-variance tradeoff\cite{Sammut2010Encyclopedia}.
 Given a finite size data set, as the number of states increases, the systematic error (the bias) decreases, while the statistical error (the variance) increases.
@@
 \end{equation}
-For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Generation}.
+For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Markov}.
 
-For classical MD, $\mathcal{Q}^{(\tau)}$ is uniquely determined by the Hamiltonian and equations of motion\cite{}.
+For classical MD, $\mathcal{Q}^{(\tau)}$ is determined by the Hamiltonian and equations of motion\cite{Schutte2001Transfer}.
 However, it is not practical to explicitly construct the propagator from the Hamiltonian for complex systems.
@@
 The state-space, $S$, can be constructed by clustering the points in $D$ or via a grid-based discretization of $\Omega$.
-Given $S$, maximum-likelihood estimators for the $O(k^2)$ transition probabilities $\mathbf{T}$ exist\cite{kyle, frank}.
+Given $S$, maximum-likelihood estimators for the $O(k^2)$ elements of a reversible transition probability matrix $\mathbf{T}$ exist\cite{Beauchamp2011Msmbuilder2, Prinz2011Markov}.
 
 Using this procedure, it has been shown that MSM estimates of the eigenvalues $\lambda_i$ are systematically underestimated in the limit of infinite data, and that the magnitude of this bias goes to zero as $k\rightarrow \infty$.
-This fact has inspired the development of variational methods for model selection, which optimize the MSM parameters in order to maximize the eigenvalues \cite{feliks}.
+This fact has inspired the development of variational methods for model selection, which optimize the MSM parameters in order to maximize the eigenvalues \cite{Noe2013Variational}.
 
@@
 Heuristic criteria, such as manually selecting $k = n_\textnormal{conformation} / 14$ or clustering $D$ such that each conformation is within a pre-specified distance to its cluster center (e.g.
-4.5\AA$\;$ or 1.2\AA $\;$ RMSD) have been previously employed\cite{Lane2012, Bowman2012, Beuchamp}.
+4.5\AA$\;$ or 1.2\AA $\;$ RMSD) have been previously employed\cite{Lane2011Markov, Bowman2012Equilibrium, Beauchamp2012Simple}.
 Selection of $\tau$ based convergence of the relaxation timescales (\cref{eq:msm-timescales}), as suggested by \citet{Swope2004Describing} and others is more well established.
-The rate of this convergence with respect to $\tau$ at fixed $S$ has also been used as a metric for evaluating a model's state decomposition.
-\hl{cite someone?}
+The rate of this convergence with respect to $\tau$ at fixed $S$ has also been used as a metric for evaluating a model's state decomposition \cite{someone}.
 
@@
 Whereas model selection based on the variational principle described above would tend to increase the number of states without bound, as the magnitude of the eigenvalue bias decreases with the number of states, practical application of this criteria would cause a catastrophic increase in statistical uncertainty in the model.
-This tradeoff between low-variance (small $k$) but biased estimators, and high-variance (large $k$) but unbiased estimators is a general feature of problems in statistical learning\cite{}.
+This tradeoff between low-variance (small $k$) but biased estimators, and high-variance (large $k$) but unbiased estimators is a general feature of problems in statistical learning\cite{Sammut2010Encyclopedia}.
 For example, it can be shown that the uncertainty in the posterior distribution of $\mathbf{T}$ given $S$ and $D$ approaches infinity as $k$ approaches infinity.
@@
 \begin{equation}
-P\big(\{\mathbf{x}_t\}_{t=1}^N\;\big|\; \{S, T\}\big)d\mathbf{x}^N = \prod_{t=1}^{N-1} T_{\sigma(\mathbf{x}_t), \sigma(\mathbf{x}_{t+1})} \cdot \prod_{t=1}^{N} P(\mathbf{x}_t | \sigma(\mathbf{x}_t)) \cdot d\mathbf{x}^N,
+P\big(\{\mathbf{x}_t\}_{t=1}^N\;\big|\; \{S, T\}\big)d\mathbf{x}^N = \pi_{\sigma(\mathbf{x}_1)}\prod_{t=1}^{N-1} T_{\sigma(\mathbf{x}_t), \sigma(\mathbf{x}_{t+1})} \cdot \prod_{t=1}^{N} P(\mathbf{x}_t | \sigma(\mathbf{x}_t)) \cdot d\mathbf{x}^N,
 \label{eq:like}
 \end{equation} 
-where $P(\mathbf{x}_t | \sigma(\mathbf{x}_t))$ is the probability of sampling a conformation $\mathbf{x}_t$ given that the process is in the state $\sigma(\mathbf{x}_t)$, referred to as the emission distribution of state $\sigma(\mathbf{x}_t)$.
+where $\pi_i$ is the stationary probability of state $i$ according to $\mathbf{T}$ and $P(\mathbf{x}_t | \sigma(\mathbf{x}_t))$ is the probability of sampling a conformation $\mathbf{x}_t$ given that the process is in the state $\sigma(\mathbf{x}_t)$, referred to as the emission distribution of state $\sigma(\mathbf{x}_t)$.
 There is some flexibility in the choice of emission distribution, though \cref{eq:like} is valid only when the emission distributions satisfy
@@
 Next, we reanalyzed two ultra-long 100 $\mu s$ molecular dynamics trajectories of the Fip35 WW domain\cite{Liu2008Experimental}, provided courtesy of D.E.
-Shaw Research \cite{Shaw2010Atomic} (Amber ff99SB-ILDN force field,\cite{amber} TIP3P water model\cite{tip3p}).
+Shaw Research \cite{Shaw2010Atomic} (Amber ff99SB-ILDN force field,\cite{Lindorff2010Improved} TIP3P water model\cite{Jorgensen1983Comparison}).
 We projected the trajectories into a four dimensional space using time-structure based independent components analysis (tICA) \cite{Schwantes2013Improvements, Perez2013Identification} on the first trajectory.
@@
 Leaving out a portion of a molecular dynamics data set from the training to serve as a test set is costly proposition for the applications of MSMs to the study of complex system.
-For large bimolecular simulations characterized by long intrinsic timescales, sampling the potential exhaustively is a significant challenge\cite{Lane2012}, making it difficult to afford discarding half of the dataset during the fitting of an MSM.
+For large bimolecular simulations characterized by long intrinsic timescales, sampling the potential exhaustively is a significant challenge\cite{Lane2013Milliseconds}, making it difficult to afford discarding half of the dataset during the fitting of an MSM.
 For this reason, we also computed two penalized likelihood model selection criteria, the AIC and BIC, which augment the training set log-likelihood with an explicit penalty on model complexity to avoid over fitting.
@@
 The models' ten slowest relaxation timescales (shown in the right panel) converge rapidly.
-\hl{right panel should be linear scale instead, for consistency with the PCA}}
+}
 \end{figure}
@@
 \centering
-\label{fig:ww-pca}
 \includegraphics[width=6in]{figs/ww_pca/ww_pca.pdf}
@@
 The longest relaxation timescale (right panel) exhibits a linear increase up to 25,000 states.
-The AIC (blue), BIC (red) and test set log-likelihood each recommend a model with 200 states.}
+The AIC (blue), BIC (red) and test set log-likelihood each recommend a model with 200 states.\label{fig:ww-pca}}
 \end{figure}

### 2014-04-07 lib/likelihood.py

@@ -60,6 +60,9 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False,
     counts = counts.tocsr()
     tProb = tProb.tocsr()
 
+    pi = msm_analysis.get_eigenvectors(tProb, k=1)
+    pi = pi.flatten() / pi.sum()
+
     ij = counts.nonzero()
 
     state_counts = np.array(counts.sum(axis=1)).flatten()
@@ -78,6 +81,7 @@ def get_log_likelihood(tProb, trajs, volumes, lagtime=1, separate=False,
     t_like = np.sum(nonzero_counts[good_transitions] * np.log(nonzero_tprobs[good_transitions]))
     pct_lost = (nonzero_counts.sum() - nonzero_counts[good_transitions].sum()) / nonzero_counts.sum() * 100.
     print "ignoring %.2f %% of counts that have zero probability according to the model"  % pct_lost
+    t_like += np.log(pi[trajs[0]])
 
     if not kellogg:
         v_like = - np.sum(state_counts * np.log(volumes))

### 2014-04-07 reviewer-comments/letter.tex

@@
 \usepackage{fullpage}
-\usepackage{cleveref}
+%\usepackage{cleveref}
 
@@
 
-This is a fair point, as we focus only on picking the number of states.
-In fact, we've changed the title to reflect this issue as well as the fact that we're not attempting to argue that our selection is truly ``optimal," but merely suggesting to use a method that is ``{\it optimizable}." \hl{reference the title.}
+We agree with the reviewer, as our contribution focuses only on picking the number of states.
+To address this comment, we've changed the title to reflect this fact.
+Furthermore, we don't wish to make the suggestion that our model selection criteria are themselves optimal.
+They are merely one set of statistical tools from among many.
+For this reason, we've changed the title to ``Statistical Model Selection for Markov Models of Biomolecular Dynamics''.
 
@@
 
-The reviewer is exactly correct.
+The reviewer is correct.
 We meant that the timescales will remain constant, but this means that the eigenvalues do change.
-We've corrected that sentence and also added a more detailed discussion of the transition probability matrix.
-The text can be found at the beginning of the section ``Prior Work."
+We've added a much more detailed discussion of the background on MSMs including a discussion of the propagator.
+This text can be found in the section titled ``Background''.
 
-\hl{add the text here too}
 \separate
@@
 \begin{quote}
-With $s(\mathbf{x})$ as the function that maps a conformation in phase space ($\mathbf{x}$) to the index of a state in the discrete state space
+A $k$-state Markov model is defined on a discrete set of states, $S = \{s_i\}_{i=1}^k$, where $s_i \subseteq \Omega$ and $\bigcup_{i=1}^k s_i = \Omega$.
+Furthermore, $s_i \cap s_j = \emptyset$ for all $i \neq j$.
+In words, every point in $\Omega$ is assigned to one (and only one) state in the MSM.
+Let $\sigma(\mathbf{x})$ be the function that maps a point $\mathbf{x} \in \Omega$ to the index $i$ of the state in $S$ such that $x \in s_i$.
 \end{quote}
+
 \separate
+
 \begin{quote}
@@
 
-This is most definitely true, but is not incredibly important.
-The reviewer is correct that technically we should include the probability of $x_0$ in the likelihood, however, when evaluating the relative likelihoods of models built on the same data, this term cancels out as it shows up in each model.
-Therefore, optimizing either function will arrive at the same result.
-\hl{We can add to the text if we think we need to, but this is really a nit-picky comment in my opinion.}
+We have changed this, such that we work directly $P\big(\{\mathbf{x}_t\}_{t=1}^N\;\big|\; \{S, \mathbf{T}\}\big)d\mathbf{x}^N$ as opposed to the conditional distribution in our work.
+We do not, however, fit a separate starting probability distribution.
+Because our goal is to build models for equilibrium dynamics, we choose $P\big(\mathbf{x}_0 | \{S, T\}\big) = \pi_{\sigma(\mathbf{x}_0)} \; P(\mathbf{x}_0 | \sigma(\mathbf{x}_0))$ where $\pi$ the stationary distribution of $\mathbf{T}$.
 
@@
 
-We've reworded this paragraph to be clearer. The text can be found right after Eq. (1) and is copied below:
-\begin{quote}
-With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state Markov transition probabilities, $T(s_i \rightarrow s_j) \equiv P(S_{t+1}=s_j | S_{t}=s_i)$, and so-called emission distributions of each state, $P(\mathbf{x} | s(\mathbf{x}))$, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}$, is within a certain state, $s(\mathbf{x})$.
-As we are using  non-overlapping states, the conditional probability of observing a conformation, $\mathbf{x}$ from state $i$ is zero unless $\mathbf{x}$ is assigned to $i$.
-Mathematically, this means:
-$$ P(\mathbf{x} | i) = 0 \;\;\;\; \forall \mathbf{x} \textnormal{ s.t. } s(\mathbf{x}) \neq i$$
-\end{quote}
+We've reworded this paragraph to be clearer. The text can be found after Eq. (8), where we emphasize that the likelihood function introduced is valid only when
+
+\begin{equation}
+P(\mathbf{x} | i) = 0 \;\;\;\; \forall \; \mathbf{x} \textnormal{ s.t. } \sigma(\mathbf{x}) \neq i.
+\end{equation}
 
@@
 \end{quote}
-\hl{I'm not sure that this change is really all that necessary.}
+
+We've rewritten the emission distribution as
+\begin{equation}
+P(\mathbf{x} \;|\; i) = \frac{1}{V_i} \mathbf{1}_{s_i}(\mathbf{x}).
+\end{equation}
 
@@
 
-When we first added this emission model, our goal was to provide something that might be used in more realistic cases.
-As the volume calculation makes the volume emission model impractical for all but the simplest cases, we wanted a model whose probabilities could be evaluated analytically.
+When we first added this emission model, our goal was to provide a protocol that would be practical for practitioners to apply in high-dimensional settings.
+As the scaling of the volume calculation makes the volume emission model challenging except in low dimensions, we wanted a model whose probabilities could be evaluated analytically.
 One simple first guess would be to use a gaussian distribution.
 However, the reviewer brings up several issues with this model that make it too imprecise to be used in practice.
-As such, we've moved all of the Gaussian emission model results / discussion to an appendix section.
-It is still useful to think about as it is a stepping stone to using ``fuzzy" clustering or other methods whose state emission distributions may overlap.
+As such, we've removed all of the Gaussian emission model from this work.
+Future work might be able to utilize this model as a tractable variational approximation to a hidden Markov model with Gaussian emission distributions.
 
 \begin{quote}
-On choice of emission probabilities, an obvious question: Why not choose $P(x|i) \propto exp[-\beta U(x)] \delta_{i,s(x)}$, where $U(x)$ is the potential energy and $\beta$ the inverse temperature?
+On choice of emission probabilities, an obvious question: Why not choose $P(x|i) \propto \exp[-\beta U(x)] \delta_{i,s(x)}$, where $U(x)$ is the potential energy and $\beta$ the inverse temperature?
 The normalization constants are ALSO difficult to compute, but have the advantage that they are proportional to the stationary probability elements computed from the transition matrix $T$.
@@
 
-This is certainly one possibility, and is desirable because it is physically motivated. \hl{maybe add a discussion in the gaussian section?}
+This is certainly one possibility, and is desirable because of its physical motivation.
+Our interest, however, is primarily in statistical models for molecular dynamics data that makes use of the potential function only by way of MD trajectories, and doesn't require further calculation based on the forcefield.
+
 \separate
@@
 \end{quote}
-\hl{Is there some mathematical definition of intractability that I'm unaware of?}
-In order to compute the Bayes factor for selecting the number of states, we would need to sample all transition matrices that can be built in $k$ states.
-This is simply impractical as you need to integrate over all $k \times k$ transition matrices.
+
+In order to compute the Bayes factor for selecting the number of states, we would need to integrate over the space of all crisp state decompositions $S$ with $|S|=k$ as well as the simplex of $k \times k$ reversible stochastic matrices.
+Although the second is possible analytically using the results of Diaconis and Rolles (2006), we're not aware of any practical methods for the former.
+To avoid confusion, we've replaced the word ``intractable'' with ``impractical''.
 

### 2014-04-07 paper/paper2.tex
small changes

@@
 
-%The trajectory in phase space ($\mathbf{x}_t$) is assigned to one of $k$ states.
-We represent this assignment by the $s(\cdot)$ function.
-The dynamics in state space are described by the transition probability matrix.
-Since the model is Markovian, the probability that a given trajectory in state $s(\mathbf{x}_t)$ will be in some state $i$ is given by $\mathbf{T}_{s(\mathbf{x}_t), i}$:
-%\begin{equation}
-%\label{eq:markov_state}
-%s(\mathbf{x}_{t+\tau}) \sim_d \mathbf{T}_{s(\mathbf{x}_t), i}
-%\end{equation}
-
-%A significant challenge in the automated construction of Markov state models is the choice of the number of states\cite{McGibbon2013Learning}.
-Although classical Hamiltonian dynamics with $N$ particles form a continuous-time Markov chain in $\mathbb{R}^{6N}$, the projection of this high-dimensional process into a lower dimensional space (such as the discrete indicator functions used by MSMs) is generally governed by non-Markovian dynamics\cite{van1992Stochastic}.
-\hl{In the Morri-Zwanzig approach, this corresponds to the generalized Langevin equation with a nontrivial memory kernel}(What is this about?)\cite{ZM}.
-Within the MSM context, on the other hand, this means that first order Markov chains incur a systematic error in their modeling of the high-dimensional system.
-In particular, when MSM states contain within them kinetic barriers of substantial magnitude, the bias due to the first-order Markov approximation is large.
-While this source of modeling error can be addressed by increasing the number of states, the reduction in one error comes at the expense of the increase in another.
-This second source of error is statistical in origin.
-As the number of states in the model grows, so does the number of parameters required to completely specify the kinetic model between all pairs of states.
-Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per model parameter, which makes the approach susceptible to over-fitting.
-
 Two competing sources of error govern the accuracy of an MSM.
@@
 \section{Background}
+\subsection{Markov State Models}
 Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$.
@@
 
-\section{The Bias-Variance Dilemma}
+\subsection{The Bias-Variance Dilemma}
 For MSM methods to be valuable to practitioners, they must operate in limited data regimes.
@@
 
-\section{Likelihood of a Markov State Model}
+\subsection{Likelihood of a Markov State Model}
 Bayes' rule provides a foundation for model selection, by establishing the proportionality of the probability of a model given the data to the probability of the data given the model.
@@
 
-\section{Statistical Model Selection}
+\subsection{Statistical Model Selection}
 Likelihood maximization on the training data (i.e.
@@
 
+For all models, we computed the BIC and AIC according to \cref{eq:bic} and \cref{eq:aid}.
+For the BIC, we need to define the ``number of data points" in the dataset.
+We used the total number of transitions observed in the data as the value for $N$.
+However, as transitions were counted using the ``sliding window" approach, these transitions are not necessarily independent.
+Therefore, the BIC may be over-penalizing the number of parameters due to this over-estimate of $N$.
+For a discussion of the sliding window approach, we refer the reader to \citet{Prinz2011Markov}.
+
 \section{Results and Discussion}
@@
 Bill's insight, clarity of thought, and careful approach has influenced my lab's work very broadly.
-I am very pleased that our collaborations continue through today and look forward to many more in the future.
+I am very happy that our collaborations continue through today and look forward to many more in the future.
 \end{acknowledgement}
@@
 %\end{figure}
+
+%The trajectory in phase space ($\mathbf{x}_t$) is assigned to one of $k$ states.
+We represent this assignment by the $s(\cdot)$ function.
+The dynamics in state space are described by the transition probability matrix.
+Since the model is Markovian, the probability that a given trajectory in state $s(\mathbf{x}_t)$ will be in some state $i$ is given by $\mathbf{T}_{s(\mathbf{x}_t), i}$:
+%\begin{equation}
+%\label{eq:markov_state}
+%s(\mathbf{x}_{t+\tau}) \sim_d \mathbf{T}_{s(\mathbf{x}_t), i}
+%\end{equation}
+
+%A significant challenge in the automated construction of Markov state models is the choice of the number of states\cite{McGibbon2013Learning}.
+Although classical Hamiltonian dynamics with $N$ particles form a continuous-time Markov chain in $\mathbb{R}^{6N}$, the projection of this high-dimensional process into a lower dimensional space (such as the discrete indicator functions used by MSMs) is generally governed by non-Markovian dynamics\cite{van1992Stochastic}.
+\hl{In the Morri-Zwanzig approach, this corresponds to the generalized Langevin equation with a nontrivial memory kernel}(What is this about?)\cite{ZM}.
+Within the MSM context, on the other hand, this means that first order Markov chains incur a systematic error in their modeling of the high-dimensional system.
+In particular, when MSM states contain within them kinetic barriers of substantial magnitude, the bias due to the first-order Markov approximation is large.
+While this source of modeling error can be addressed by increasing the number of states, the reduction in one error comes at the expense of the increase in another.
+This second source of error is statistical in origin.
+As the number of states in the model grows, so does the number of parameters required to completely specify the kinetic model between all pairs of states.
+Because the amount of data is constant, each additional parameter leads to a decrease in the amount of data available per model parameter, which makes the approach susceptible to over-fitting.

### 2014-04-07 paper/paper2.tex
more fixes to reviewer-comments
More citations and small tweaks

@@
 
-\section{Background}
+\section{Theory}
 \subsection{Markov State Models}
@@
 
-Introduction to MSMs via the propagator theory motivates the view of the parameterization problem as one of \emph{function estimation}.
+Introduction to MSMs via the propagator theory motivates the view of the parametrization problem as one of \emph{function estimation}.
 This perspective has a critical disadvantage: the natural error metric -- the difference between the estimated MSM and true propagator is intrinsically uncomputable.
@@
 \begin{equation}
+\label{eq:exclusive}
 P(\mathbf{x} | i) = 0 \;\;\;\; \forall \; \mathbf{x} \textnormal{ s.t. } \sigma(\mathbf{x}) \neq i.
 \end{equation}
-Given the coarse-graining of $p(\mathbf{x})$ in \cref{eq:coarse-grain-p}, the most natural choice emission distributions are normalized indicator functions
+An alternative approach involving fuzzy clustering, where conformations are not uniquely assigned to a single state, would not impose the constraint in \cref{eq:exclusive}, and would also involve a likelihood function that marginalizes over all possible paths\cite{gordon1992fuzzy}.
+In this work, we consider only models based on crisp partitioning.
+As such, given the coarse-graining of $p(\mathbf{x})$ in \cref{eq:coarse-grain-p}, the most natural choice emission distributions are normalized indicator functions
 \begin{equation}
@@
 This scheme can be repeated many times with different partitions of the data set.
-The AIC and BIC are augmented likelihood function approaches that do not require leaving out portions of the available data during fitting, and instead employ asymptotic approximations to the Bayesian evidence to directly penalize models with additional free parameters.
 \begin{equation}
@@
   \mathrm{AIC} \equiv -2\cdot \ln L + \left(2\right) \cdot \kappa
-\end{equation} where $L$ is the likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points, assumed to be independent and identically distributed.
+\end{equation} where $L$ is the maximum likelihood, $\kappa$ is the number of free parameters, and $N$ is the number of data points.
 Model selection is performed by a minimization of the criterion.
+The BIC is derived based on the dominant terms in the Laplace approximation to the logarithim of the Bayes factor with a vague prior, while the functional form of the AIC comes from an asymptotic approximation to the Kullback-Leibler divergence between the model and the true distribution\cite{kass1995bayes, liddle2004many}.
+The appearance of number of data points $N$ in \cref{eq:bic} assumes that the data is independent and identically distributed given the model, which is often poorly justified for time series.
+Nonetheless, the value in the AIC and BIC comes from their simple form, that they do not require leaving out portions of the available data during fitting, and acceptable performance in practice\cite{Kuha01112004}.
 
 %\section{Prior Work}
-
 %Given a Markov model parameterized at a lag time of $\tau$, we can propagate ensembles of trajectories via the transition probability matrix:
@@
 \caption{Systems studied in this work.
-(A) Brownian dynamics on the two dimensional M\"{u}ller potential \cite{Muller1979Location}.
+(A) Brownian dynamics on the two-dimensional M\"{u}ller potential \cite{Muller1979Location}.
 (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E.
@@
 
-For all models, we computed the BIC and AIC according to \cref{eq:bic} and \cref{eq:aid}.
-For the BIC, we need to define the ``number of data points" in the dataset.
-We used the total number of transitions observed in the data as the value for $N$.
-However, as transitions were counted using the ``sliding window" approach, these transitions are not necessarily independent.
+The BIC and AIC were calculated according to \cref{eq:bic} and \cref{eq:aic}, with $N$ equal to the total number of transitions observed in the data.
+However, as transitions were counted using the ``sliding window" approach, these transitions are not statistically independent\cite{Prinz2011Markov}.
 Therefore, the BIC may be over-penalizing the number of parameters due to this over-estimate of $N$.
-For a discussion of the sliding window approach, we refer the reader to \citet{Prinz2011Markov}.
 

### 2014-04-08 reviewer-comments/letter.tex
More letters

@@
-\documentclass{letter}
+\documentclass{article}
 \usepackage{color,soul}
@@
 \usepackage{fullpage}
-%\usepackage{cleveref}
-
-\address{Clark Center S295 \\ Stanford University \\ Stanford, CA, 94305}
-\signature{Robert T. McGibbon, Christian R. Schwantes, Vijay S. Pande}
+\usepackage{natbib}
+\usepackage{cleveref}
 
 \newcommand{\separate}{\begin{center}--------\end{center}}
-
 \begin{document}
 
-\begin{letter}{Prof. Sharon Hammes-Schiffer \\ Deputy Editor of JPC B \\ The Journal of Physical Chemistry}
-\opening{Editor:}
+\hfill \begin{minipage}{10em} Clark Center S295\\ Stanford University\\ Stanford, CA, 94305 \end{minipage}
 
-Reviewer 1
+\noindent Prof. Sharon Hammes-Schiffer \\ Deputy Editor of JPC B \\ The Journal of Physical Chemistry
+
+\vspace{2em}
+\noindent Dear Prof. Hammes-Schiffer,
+\vspace{1em}
+
+We are resubmitting this manuscript, whose title is now ``Statistical Model Selection for Markov Models of Biomolecular Dynamics.'' The reviews were very helpful.
+Our detailed responses are included below, inline with the comments that they respond to.
+
+\vspace{1em}
+\noindent Reviewer 1:
+
 \begin{quote}
@@
 Three different critera are examined: a leave-some-out cross-validation likelihood estimate, as well as two criteria that penalize overfitting by using assumptions of asymptotic normality of the likelihood function (AIC and BIC).
-Two different "emission probabilities" for the probability of observing a sample $x_t$ from a given discrete state s are considered that would produce a generative model---one that can be used to assign new data to a model: a uniform probability and Gaussian probability.
+Two different ``emission probabilities" for the probability of observing a sample $x_t$ from a given discrete state s are considered that would produce a generative model---one that can be used to assign new data to a model: a uniform probability and Gaussian probability.
 These are compared for consistency with each other, as well as compared with well-established metrics such as the stability of implied timescales of the model with the number of states.
@@
 They are merely one set of statistical tools from among many.
-For this reason, we've changed the title to ``Statistical Model Selection for Markov Models of Biomolecular Dynamics''.
+For this reason, we've changed the title to ``Statistical Model Selection for Markov Models of Biomolecular Dynamics.''
 
@@
 The eigenvalues of $T(t)$ actually change with t.
-Perhaps you mean the eigenvalues of some implied rate matrix $K$ where $T(t) = exp[Kt]$?
+Perhaps you mean the eigenvalues of some implied rate matrix $K$ where $T(t) = \exp[Kt]$?
 \end{quote}
@@
 We've added a much more detailed discussion of the background on MSMs including a discussion of the propagator.
-This text can be found in the section titled ``Background''.
+This text can be found in the subsection titled \emph{Theory: Markov State Models}
 
@@
 We were too casual with our notation, but had meant that $s(\cdot)$ maps a conformation in phase space to an integer corresponding to the index of the state in the Markov model.
-This has been reworded to read:
+We've changed this notation to $\sigma(\cdot)$, and reworded the relevant section to read:
 
@@
 Eq.
-1 actually refers to the conditional probability $P[{x_t}|x_0]$.
+1 actually refers to the conditional probability $P[{x_t}|x_1]$.
 Is this the quantity you intend to work with, rather than $P[{x_t}]$?
@@
 We do not, however, fit a separate starting probability distribution.
-Because our goal is to build models for equilibrium dynamics, we choose $P\big(\mathbf{x}_0 | \{S, T\}\big) = \pi_{\sigma(\mathbf{x}_0)} \; P(\mathbf{x}_0 | \sigma(\mathbf{x}_0))$ where $\pi$ the stationary distribution of $\mathbf{T}$.
+Because our goal is to build models for equilibrium dynamics, we choose $P\big(\mathbf{x}_1 | \{S, T\}\big) = \pi_{\sigma(\mathbf{x}_1)} \; P(\mathbf{x}_1| \sigma(\mathbf{x}_1))$ where $\pi$ the stationary distribution of $\mathbf{T}$.
 
@@
 
-We've rewritten the emission distribution as
+The reviewer's suggested notation is more clear. We've rewritten the emission distribution as
 \begin{equation}
@@
 
-\hl{This is one of the two major criticisms in the review so we should word the response carefully}
-
 When we first added this emission model, our goal was to provide a protocol that would be practical for practitioners to apply in high-dimensional settings.
-As the scaling of the volume calculation makes the volume emission model challenging except in low dimensions, we wanted a model whose probabilities could be evaluated analytically.
+As the scaling of the volume calculation makes the volume emission model challenging except in low dimensions, we looked for an approximation whose probabilities could be evaluated analytically.
 One simple first guess would be to use a gaussian distribution.
@@
 
+\separate
 \begin{quote}
@@
 In order to compute the Bayes factor for selecting the number of states, we would need to integrate over the space of all crisp state decompositions $S$ with $|S|=k$ as well as the simplex of $k \times k$ reversible stochastic matrices.
-Although the second is possible analytically using the results of Diaconis and Rolles (2006), we're not aware of any practical methods for the former.
-To avoid confusion, we've replaced the word ``intractable'' with ``impractical''.
+Although the second is possible analytically using the results of \cite{diaconis2006}, we're not aware of any practical methods for the former.
+To avoid confusion, we've replaced the word ``intractable'' with ``impractical.''
 
@@
 \begin{quote}
-"N is the number of data points, assumed to be independent and identically distributed"  How reasonable is this assumption for molecular simulation data?
+``N is the number of data points, assumed to be independent and identically distributed"  How reasonable is this assumption for molecular simulation data?
 How do you even quantify the number of "iid normally distributed" data points in a molecular simulation?
@@
 
-This is a subtle point, but is worth discussing.
+This is a subtle point, and we've added some clarification to the main text, both in the subsection titled \emph{Theory: Statistical Model Selection}, as well in the \emph{Methods} section.
 Since we are parameterizing a transition matrix, we choose the number of data points to be the number of transitions used to calculate the transition probabilities.
 However, as we are using the sliding window approach for counting transitions, we are in fact over-counting the number of identical transitions and so the BIC score is likely under-penalizing the models.
-This is a somewhat unanswered question in the field, however it has been discussed previously \cite{noe_sliding_window}.
-\hl{Do we want to add this to the main text?}.
+This is a weakness of the BIC for correlated timeseries applications, and is also a concern when estimating the transition matrix and its uncertainty from the observed transition counts\cite{Prinz2011Markov}.
 
@@
 
-We are being very unsophisticated in our use of the likelihood as we are not actually optimizing this likelihood to build our models.
-We are actually just using the likelihood to evaluate several models that are parameterized using the approach discussed in \cite{Beauchamp_msmb2}.
-\hl{Add this to the text, because the confusion is coming from the use of the word ``prior." We are not introducing a prior to the likelihood function, but merely using the prior when calculating the MAP for each model.}
+The prior refered to in the initial manuscript is in fact a psuedocount used during the estimation of $\mathbf{T}$. This is now clarified in the \emph{Methods} section.
 
@@
 \end{quote}
-\hl{Need to look this up. I just used robert's code though.}
 
-\separate
 \begin{quote}
@@
 \end{quote}
-\hl{Now we're getting really nit-picky, but fine.}
+
+We've expanded our discussion of the simulation methods as well as the parameters used in MSM construction to address these and other issues.
+The new subsection, entitled \emph{Simulation and MSM Construction} can be found in the \emph{Methods} section.
 
@@
 \end{quote}
-\hl{It's really just a qualitative change.
-I just meant that the changes in the likelihood are much less than the differences between models, and so are not important for evaluating the ``best" model.}
+
+We selected $M$, the number number rounds of Monte Carlo volume estimation, such that the log likelihood was converged with respect to $M$.
 
@@
 \begin{quote}
-according to the three methods" Which three methods exactly?
+``according to the three methods" Which three methods exactly?
 \end{quote}
@@
 The numbers/units of likelihood are actually meaningful, and should be shown.
-What is the difference between "volume" and "Gaussian"?
+What is the difference between ``volume" and ``Gaussian"?
 \end{quote}
-\hl{I disagree with this comment. The absolute likelihoods are basically meaningless, while the relative differences are important. I think adding units would be distracting.}
+All three figures have been reworked and are bigger.
+The y-axis label has been changed to ``Selection Criterion.'' As discussed by Liddle and others, ``the absolute value of the criterion is not of interest, only the relative value between different models.''\cite{liddle2004many}.
 
@@
 In practice, however, one would choose one of the three and then pick the best model given that test.
-\hl{this was only mentioned in the captions of the figures, so I'll correct it when I change the figures.}
 
@@
 We are by no means advocating that either the AIC or BIC are perfect approximations to the out-of-sample performance.
-\hl{Add a sentence saying this in the main text.}
+
+Additionally, it's worth noting that the AIC applied to the tICA MSMs for WW domain does qualitatively agree with the test-set log likelihood.
 
@@
 
-We should not anthropomorphize our mathematical expressions! We've corrected this to read more clearly:
-\begin{quote}
-The decrease in the training set log-likelihood can be explained by a trade-off inherent in Eq.
-(1), where adding more states leads to an increase in the emission distribution term but causes the transition matrix term to decrease significantly more.
-\end{quote}
+This original text has been removed, though the subsection entitled \emph{Theory: Bias-Variance Dilemma} discusses the tradeoff we were referring to in the original draft.
 
@@
 \end{quote}
-\hl{I originally had this but you didn't like my graph. We could add it back or ignore this since it's moved to the appendix anyway}
+
+As discussed above, we've removed the Gaussian likelihood results.
 
@@
 Fig.
-3: "Since k-means optimizes an objective function that is related to the emission probability" What is this objective function?
+3: ``Since k-means optimizes an objective function that is related to the emission probability'' What is this objective function?
 As I understand it, k-means simply minimizes the sum of intracluster variances, which has nothing to do with your choice of emission probability being uniform or Gaussian or otherwise.
 Or do you mean you optimize the log-likelihood form of Eq 1 using a voronoi decomposition based on generators that are optimized for each number of states?
-Also: "The likelihood decreases as new states are added..."  Which likelihoods are printed in the figure panels?
+Also: ``The likelihood decreases as new states are added...''  Which likelihoods are printed in the figure panels?
 Test set, uniform, or Gaussian emission probabilities?
@@
 The k-means clustering method minimizes the average mean squared distance from each data point to its cluster's mean.
-This is precisely the same as the $\hat{\sigma}^2$ term in Eq.
-(8).
-When you write out the entire log-likelihood of the training dataset, there ends up being a term proportional to $\log \hat{\sigma}$.
-This is what we meant by the k-means clustering optimizing the likelihood function (and is precisely why Pelleg and Moore selected the probability that they did).
-\hl{Add clarification in figure to indicate that these are training set likelihoods.}
+This is precisely the same as the $\hat{\sigma}^2$ term in the original equation.
+One can show that the resulting log-likelihood (evaluated on the data set) includes a term proportional to $\log \hat{\sigma}$.
+This is what we meant by the k-means clustering optimizing the likelihood function (and is precisely why \citet{Pelleg2000Xmeans} selected the probability that they did).
+
+This discussion, however, has been removed, since it only applies to the Gaussian likelihood.
 
@@
 \begin{quote}
-``...AIC is comparable to the test set log-likelihood...The BIC penalizes complexity more strongly..." Why the huge difference between the AIC and BIC?
+``...AIC is comparable to the test set log-likelihood...The BIC penalizes complexity more strongly...'' Why the huge difference between the AIC and BIC?
 \end{quote}
@@
 The difference just comes from the difference between the way both are defined.
-\hl{I want to brush this off as ``we're not advocating one or the other, but merely suggesting they may be useful in cases where cross-validation is not possible" but I'm not sure of the best wording.}
+The BIC penalizes the number of parameters more strongly.
+We've added a discussion of the origins of the BIC and AIC.
+It's important to note that we are not attempting to advocate the use of a specific criterion, but are more interested in convincing the reader to use a statistical model selection criterion.
+The gold standard, of course is a cross-validation approach.
+Since this is not practical for most applications, we also wanted to include other selection criteria that are easier to calculate.
+Given our results, we would suggest the use of the AIC, but it's always important to know where these criteria come from and what assumptions are implicitly made when using one.
 
@@
 4: Too small.
-Missing figure title, like "Comparison of likelihood models for Fip35 WW domain trajectory data".
-"Test set log-likelihoods are omitted in C as the dense matrix algebra is intractable for $k > 10^3$".
+Missing figure title, like ``Comparison of likelihood models for Fip35 WW domain trajectory data".
+``Test set log-likelihoods are omitted in C as the dense matrix algebra is intractable for $k > 10^3$".
 Intractable?
@@
 \end{quote}
-We really shouldn't use this word here, as it's not intractable, but it is difficult, as you have to fit the transition matrix with $k^2$ (nonzero) terms.
-Our software leverages sparse matrices, so it is difficult for us to actually parameterize large transition matrices that are not sparse.
+
+We did not intend to use the word ``intractable" in to indicate a particular computational complexity class. We've changed this word to ``impractical" to avoid confusion.
 
@@
 Yes it is.
-For well-sampled datasets built using the state of the art methods we can still use implied timecales, however, there are many times that the timescales are not as clean and easy to interpret.
-We've added a figure for models built with PCA instead of tICA on the WW dataset to illustrate how the implied timescales are not ``optimizable" in non-ideal cases.
-\hl{still need to make this figure.}
+For well-sampled datasets built using the state-of-the-art methods, the implied timescales, which consider only the systematic error in the MSM, yield consistent recommendations with the AIC, BIC and test log-likelihood.
+In other regimes, where the statistical error dominates, the results are different.
 
@@
 \end{quote}
-Lane et al.
-used an implied timescale approach with the RMSD metric, while Kellog et al.
-(discussed in the section ``Prior Work" \hl{right?}) used a likelihood model with a discriminative emission distribution.
-This line has been re-worked since we've moved the Gaussian emission model to the appendix.
+
+This discussion is no longer in this revision since we've removed the Gaussian emission model.
+However, in case the reviewer was curious, Lane et al.
+selected the number of states by insisting that each point was within 4.5 \AA$\,$ of the cluster center.
+Kellogg et al.
+used a likelihood (Eq.
+(11) in the revision) based on a discriminative probability model.
 
@@
 
-\hl{This is another important note that we need to answer carefully.
-The idea is that in ideal cases, our methods don't disagree with the timescale test drastically, but they are optimizable, which is in itself a significant advantage.
-This is particularly useful when the timescales simply do not converge, and cannot be quantitatively used to select a model.
-Secondly, the model that the statistical test selects may not be the model that best represents the protein of interest, however it does represent our best estimate of the system given how much data we have.
-Cross-validation is a rigorous way to compare models based on newly observed data, and so our test log-likelihood serves as a way to see that adding new states may increase the timescales, but in fact decreases the ``quality" of the model.}
-
-\separate
 \begin{quote}
-``These results take a step toward fully automating the MSM construction process...by providing a quantitative method for selecting the most suitable complex model.''  Which method in particular is recommended?
+``These results take a step toward fully automating the MSM construction process...by providing a quantitative method for selecting the most suitably complex model.''  Which method in particular is recommended?
 The Gaussian method?
@@
 
-The advantage is precisely that this framework is {\it optimizable}.
-The statistical tools we've suggested match the Chapman-Kolmogorov when the dataset is ``well-behaved" (i.e.
-when we have a lot of data and are using tICA).
-However, there are many cases where the timescales do not converge at all and so the Chapman-Kolmogorov tests fail.
-We've illustrated this with a simple example (\hl{See section ``PCA on WW")}, but in fact there are many larger simulations that give rise to wildly miss-behaving implied timescales.
+Our position in this manuscript is that in many cases the implied timescale tests, which are motivated by arguments made in the limit of infinite data, are insufficient for building MSMs on available data sets.
+We frequently observe that the timescales are not well-behaved for many large scale simulations.
+Therefore, model selection based on timescale convergence is not able to answer the question ``given the data that I have, what is the best MSM I can build?" In this regime, both the statistical and systematic sources of error must be taken into account simulataneously.
+The likelihood-based criteria are a few examples of such methods, but they may not be the gold standard for model selection.
+
+The statistical tools we've suggested match the timescale tests when the dataset is large and ``well-behaved.'' However, there are many cases where the timescales do not converge at all and so the timescale tests fail to pick a reasonable MSM, which we illustrate in Fig.
+4.
 
@@
 Application of techniques that consider both systematic bias and statistical error on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those timescale-based techniques encourage over-fitting.
-\end{quote} \hl{this sentence feels a bit long...}
-
+\end{quote} 
 \separate
 \begin{quote}
-Two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space..."  There is a substantial quantity of literature on the use of basis functions that are not indicator functions (i.e.
-"fuzzy" rather than "crisp" partitionings) that should be mentioned.
+``Two key quantities which define the MSM are thus the state definitions, an indicator function basis over phase space..."  There is a substantial quantity of literature on the use of basis functions that are not indicator functions (i.e.
+``fuzzy'' rather than ``crisp'' partitionings) that should be mentioned.
 See, for example, the work of Marcus Weber and Susanna Roeblitz (nee Kube).
@@
 
-\hl{We should add a section discussing this as well as the future of directly optimizing the likelihood in the form of an HMM or other type of moel}
+We've added a comment to the \emph{Likelihood of a Markov State Model} section, discussing ``fuzzy" clustering.
 
@@
 
-This is precisely what we are drawing attention to, and we've added to this section to explicitly say this is a discriminative model.
+This is precisely what we are drawing attention to, and we've added to this section (\emph{Likelihood of a Markov State Model}) to explicitly say this is a discriminative model.
 
@@
 4 is ALSO INTRACTABLE for EXACTLY THE SAME REASON, since you do not compute the true normalizing constant for essentially identical reasons of computational tractability in high dimension.
-As such, phrases such as "A more tractable alternative is a Gaussian emission model.." are entirely disingenuous and misleading.
+As such, phrases such as ``A more tractable alternative is a Gaussian emission model.." are entirely disingenuous and misleading.
 \end{quote}
 
-The reviewer is exactly right. For these reasons (and others) we've chosen to discuss just the volume-based emission model in the main text. This is really what we'd like to focus on.
+The reviewer is exactly right. For these reasons (and others) we've chosen to discuss only the volume-based emission model in this revision.
 
@@
 This was the same lag time that was used in Beauchamp et al.\cite{Beauchamp_shaw_paper}.
+This has also been specified in the methods section.
 In this scheme, the lag time must be specified before selecting the models, and so we picked a number that had been successfully used in the past.
@@
 
-This sentence was poorly worded, but we've reworked it and copied the new version below:
-\begin{quote}
-While only using 200-600 states, Markov models constructed using tICA were still able to recapitulate the folding timescale as well as the faster processes identified by McGibbon and Pande.
-\end{quote}
+This sentence was poorly worded, but we've reworked this section, and so this should be clearer.
 
+\separate
 \begin{quote}
@@
 
-The reviewer is exactly right, and this is the point we're trying to drive home.
-It would be nice to compute the statistical error, but it is a difficult thing to calculate.
-\hl{This is in the minor complaints section so I don't know if we really need anything more than this}
+The reviewer is exactly right, and this is precisely the point of our contribution.
+One advantage of the test-set likelihood approach is that it consolidates both the systematic and statistical contributions to the error into a single computable function.
 
-Reviewer 2:
+\vspace{3em}
+\noindent Reviewer 2:
 
@@
 We do mean complexity here, however the two words go hand in hand.
-In this case, we're talking about an MSM built either with thousands vs.
-tens of states.
-The simpler model therefore cannot describe transitions between two states from the complicated model that are lumped together in the simple model.
-In this sense, the simpler model is less accurate.
-However, this may not matter, if what you really care about is a large global transition between ensembles of small states.
-In this sense, the simpler model does not lose any accuracy.
-\hl{I hate how I worded this.}
+The complexity of a model can be generally thought of as how capable the model is at handling arbitrarily complicated data.
+As one increases the complexity of a model, it is more able to model the training set more accurately.
 
@@
 
-\hl{I have this data, so we can add it as an appendix. It doesn't hurt.}
+This is an interesting question, but we feel it is beyond the scope of this work. However, the approach based on likelihood functions should be able to be used to study this question.
 
@@
 This is a very good idea and something we're interested in doing.
-However, due to the complexity of the volume calculation it's simply not feasible.
+However, due to the complexity of the volume calculation it is a difficult task.
 We hope that our work serves as a stepping stone to new methods that optimize a likelihood supported in phase space, but admit it may not be the one discussed here.
 
-Reviewer 3:
+\vspace{3em}
+\noindent Reviewer 3:
+\vspace{1em}
+
+Many of the comments made by Reviewer 1 and Reviewer 3 were closely related, and our responses reflect the consolidated concerns of both reviewers.
 
@@
 \begin{quote}
-I find the title “Optimal Parameter Selection in Markov State Models...” inappropriate: It suggests that the ``Parameter selection in MSMs'' is a well-defined problem, and that the solution approach here would be optimal, i.e.
+I find the title ‚Optimal Parameter Selection in Markov State Models...‚ inappropriate: It suggests that the ``Parameter selection in MSMs'' is a well-defined problem, and that the solution approach here would be optimal, i.e.
 not further improvable.
@@
 
+As discussed above, we've changed the title to ``Statistical Model Selection for Markov Models of Biomolecular Dynamics.''
+
+\separate
 \begin{quote}
@@
 It should be spelled out that there is a change of viewpoint, where instead of fixing the discretization and varying the lagtime, one fixes the lagtime and varies the discretization.
-And the hope why both approaches are meaningful is that the implied timescales should converge to their true values for either (1) increasing lag time [see Djurdjevac, Sarich, Schütte, MMS 2011 and Prinz, Chodera, Noe, PRX 2014 (http://arxiv.org/abs/1207.0225)], or (2) better approximation of the dominant eigenfunctions via more discrete states [see (19) and Sarich, Noe, Sch\"{u}tte MMS 8, p1154 (2010)].
+And the hope why both approaches are meaningful is that the implied timescales should converge to their true values for either (1) increasing lag time [see Djurdjevac, Sarich, Sch√ºtte, MMS 2011 and Prinz, Chodera, Noe, PRX 2014 (http://arxiv.org/abs/1207.0225)], or (2) better approximation of the dominant eigenfunctions via more discrete states [see (19) and Sarich, Noe, Sch\"{u}tte MMS 8, p1154 (2010)].
 \end{quote}
 
+We've added significantly more exposition and background on the theory behind MSMs, and in particular the analysis of the error as a function of the number of discrete states.
+
+\separate
 \begin{quote}
@@
 
+These are valuable points, which we fully agree with.
+Our focus is on best fitting the available data, and ideally validating models on new data which was left out during training.
+The ultimate goal is to fit the underlying dynamics, but our only ``window'' into the underlying data is by way of MD trajectories.
+Clearly convergence proofs, that MSM methods converge to the exact transfer operator in the limit of infinite sampling and infinite model complexity (i.e.
+$k \rightarrow \infty$) are essential, but these results do not provide direct guidance to practitioners on balancing competing sources of error.
+
+\separate
 \begin{quote}
@@
 There, the exact timescales are known, and an embedding in a geometric space is given, such that the method can be applied such that the present method can be applied as is.
-It then needs to be shown that the present likelihood approach selects MSMs with timescales that coincide with the true values.In addition, it would be useful to compare the results here with implied timescale plots.
+It then needs to be shown that the present likelihood approach selects MSMs with timescales that coincide with the true values.
+In addition, it would be useful to compare the results here with implied timescale plots.
 Do the results agree in the cases where ITS do converge?
 \end{quote}
+
+In the finite-data regime, our methods will not select MSMs with timescales that correspond to the ``true values,'' precisely because there is a competing statistical error.
+Proofs of the consistency and asymptotic optimality of the AIC and BIC for linear regression do exist, which guarentee that as $n\rightarrow\infty$, the correct model will be selected with proability one\cite{nishii1984, shao1997asymptotic}.
+Our intuition is that similar results may apply for MSMs in the joint infinite-data / infinite number of states limit, but we have not shown this.
+Given that our approach is focused on practical tools in the limited data regime, we prefer to focus there.
 
@@
 These timescales were correctly displayed.
-The timescales in A and B are calculated from the same set of models, but the evaluation of those models just provides a different ``score." Part B of each of these figures has been moved to the appendix.
+The timescales in A and B are calculated from the same set of models, but the evaluation of those models just provides a different ``score." The results with the Gaussian emission distribution have been removed.
 
@@
 The reviewer is definitely correct here.
-We are not accusing anyone of ``cheating" but more drawing attention to the fact that with the absence of an optimizable objective function there is nothing preventing one from picking the model that gives a reasonable answer.
-\hl{we should rephrase this sentence, but I don't know what to do exactly}
+We are not accusing anyone of ``cheating" but more drawing attention to the fact that with the absence of an optimizable objective function, a variety of human factors may come into play.
+We've nonetheless removed this comment from the paper.
 
@@
 \begin{quote}
-Gaussians could be truncated and normalized - it seems that renormalization would require to solve an even harder problem than for uniform output probabilities, because you’d need to integrate the Gaussian density over a high-dimensional polytope.
+Gaussians could be truncated and normalized - it seems that renormalization would require to solve an even harder problem than for uniform output probabilities, because you‚Äôd need to integrate the Gaussian density over a high-dimensional polytope.
 \end{quote}
@@
 
+As discussed above, we've removed the Gaussian emission distribution.
+
+\separate
 \begin{quote}
-I’m surprised that there are three slow relaxation processes for the Müller potential before the second gap.
+I'm surprised that there are three slow relaxation processes for the M\"{u}ller potential before the second gap.
 The potential looks like there should be only two (due to three metastable states).
@@
 
+Yes, they are all real-valued.
+
+\separate
 \begin{quote}
-Muller potential: lag time was chosen to be 30 steps.
+M\"{u}ller potential: lag time was chosen to be 30 steps.
 How is this choice justified?
@@
 2 are not independent of N, this must mean that at least for some state decompositions the ITS are not converged in the lag time.
-In these cases, it’s not even appropriate to describe the state-to-state dynamics by a Markov chain at all.
-Is the statistical model still meaningful in this case
+In these cases, it's not even appropriate to describe the state-to-state dynamics by a Markov chain at all.
+Is the statistical model still meaningful in this case?
 \end{quote}
@@
 \begin{quote}
-Müller potential, Gaussian likelihood: How were the Gaussian sigma-parameters chosen?
+WW domain, lag time was chosen to be 50 ns - same comment as for the Mueller potential, above.
+\end{quote}
+
+Our goal is to build the best possible Markov model given the data and a prespectified lag time.
+From the perspective of model selection based on implied timescale convergence with respect to lag time, it's usually required that the modeller have some physical intuition about the correct number of states.
+Here, we work from the opposite direction, as likelihood based approach gives no guidance on the selection of $\tau$.
+
+\separate
+\begin{quote}
+M\"{u}ller potential, Gaussian likelihood: How were the Gaussian sigma-parameters chosen?
 The k-means states in Fig 3 have very different sizes and their centers are sometimes very close, so the assumption that output distributions shall not overlap seems to be hard to realize.
@@
 \begin{quote}
-WW domain, lag time was chosen to be 50 ns - same comment as for the Mueller potential, above
+WW domain, Gaussian likelihood - same comment as for the M\"{u}ller potential, above.
 \end{quote}
 
-\begin{quote}
-WW domain, Gaussian likelihood - same comment as for the Müller potential, above
-\end{quote}
+As discussed above, we've removed the Gaussian emission distribution.
+\separate
 
@@
 
+They were supposed to be the same timescales, but different likelihoods. Now that we have removed the Gaussian emission distribution, this point of confusion should not be present.
+
+\separate
 \begin{quote}
- Doesn’t Fig 4c suggest that there’s a problem here?
+ Doesn't Fig 4(c) suggest that there's a problem here?
 For the likelihood-based choice made, the slowest timescale is estimated between 1500 and 3000 ns with RMSD clustering.
-In contrast, the TICA-based estimates suggest a timescale of around 5000 ns.
-So if the TICA results is indeed correct, the method should be going for larger state number in the RMSD case.
+In contrast, the tICA-based estimates suggest a timescale of around 5000 ns.
+So if the tICA results is indeed correct, the method should be going for larger state number in the RMSD case.
 \end{quote}
 
+This is a question of perspective.
+Our view is that even if the models based on RMSD or PCA underestimate the timescales, it nonetheless introduces an unacceptable amount of statistical error in the finite-data context to add states without limit.
+
+\separate
 \begin{quote}
@@
 Fig 2A).
-That’s surprising.
+That's surprising.
 Can the authors explain that?
@@
 
-\closing{Sincerely,}
+We have no strong theortical justification for this observation, but we have seen slight decreases in the timescales with respect to $k$ on occation, regardless of the method by which $\mathbf{T}$ was estimated (i.e.
+with and without psuedocounts).
 
-\bibliography{bibliography.bib}
-\end{letter}
+\vspace{4em}
+\noindent Sincerely, \\
+
+\hspace{5em} Robert T. McGibbon
+
+\hspace{5em} Christian R. Schwantes
+
+\hspace{5em} Vijay S. Pande
+
+\bibliographystyle{plain}
+\bibliography{bibliography}
+
 \end{document}

### 2014-04-08 paper/paper2.tex
abbrevs.
Typo fixes and added a single new citation

@@
 \subsection{Markov State Models}
-Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$.
+Let $\mathbf{x}_t$ be a ergodic time-homogeneous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$.
 Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$.
@@
 Selection of $\tau$ based convergence of the relaxation timescales (\cref{eq:msm-timescales}), as suggested by \citet{Swope2004Describing} and others is more well established.
-The rate of this convergence with respect to $\tau$ at fixed $S$ has also been used as a metric for evaluating a model's state decomposition \cite{someone}.
+The rate of this convergence with respect to $\tau$ at fixed $S$ has also been used as a metric for evaluating a model's state decomposition \cite{bowman2009progress}.
 
@@
 Model selection is performed by a minimization of the criterion.
-The BIC is derived based on the dominant terms in the Laplace approximation to the logarithim of the Bayes factor with a vague prior, while the functional form of the AIC comes from an asymptotic approximation to the Kullback-Leibler divergence between the model and the true distribution\cite{kass1995bayes, liddle2004many}.
+The BIC is derived based on the dominant terms in the Laplace approximation to the logarithm of the Bayes factor with a vague prior, while the functional form of the AIC comes from an asymptotic approximation to the Kullback-Leibler divergence between the model and the true distribution\cite{kass1995bayes, liddle2004many}.
 The appearance of number of data points $N$ in \cref{eq:bic} assumes that the data is independent and identically distributed given the model, which is often poorly justified for time series.
@@
 \end{algorithm} \noindent The result of this scheme produces the volumes of each state relative to the volume of $A$. It's important, therefore, to use the same set $A$ when comparing multiple models.
-\subsection{Simulation Protocol}
+\subsection{Simulation Protocol and MSM Construction}
 
@@
 Using the Euler-Maruyama method and a time step of $0.1$, we produced two trajectories of length $10^6$ time steps.
+The initial positions were sampled via a uniform distribution over the box: $[-1.5, 2.0] \times [-0.2, 2.0]$.
 
@@
 The distance $d(A, B)$ between residues $A$ and $B$ was determined as follows.
-Let $\{a_i\}_{i=1}^{n_a}$ and $\{b_i\}_{i=1}^{n_b}$ be the cartesian coordinates of the heavy (non-hydrogen) atoms in $A$ and $B$.
+Let $\{a_i\}_{i=1}^{n_a}$ and $\{b_i\}_{i=1}^{n_b}$ be the Cartesian coordinates of the heavy (non-hydrogen) atoms in $A$ and $B$.
 Then we define
@@
 Leaving out a portion of a molecular dynamics data set from the training to serve as a test set is costly proposition for the applications of MSMs to the study of complex system.
-For large bimolecular simulations characterized by long intrinsic timescales, sampling the potential exhaustively is a significant challenge\cite{Lane2013Milliseconds}, making it difficult to afford discarding half of the dataset during the fitting of an MSM.
+For large biomolecular simulations characterized by long intrinsic timescales, sampling the potential exhaustively is a significant challenge\cite{Lane2013Milliseconds}, making it difficult to afford discarding half of the data set during the fitting of an MSM.
 For this reason, we also computed two penalized likelihood model selection criteria, the AIC and BIC, which augment the training set log-likelihood with an explicit penalty on model complexity to avoid over fitting.
@@
 
-As was the case with the M\"{u}ller MSMs, the log-likihood of the training data set increases with the addition of more states, while the test log-likelihood, AIC and BIC lack a monotonic trend (\cref{fig:ww-tica}).
+As was the case with the M\"{u}ller MSMs, the log-likelihood of the training data set increases with the addition of more states, while the test log-likelihood, AIC and BIC lack a monotonic trend (\cref{fig:ww-tica}).
 The test log likelihood has a maximum at 650 states, and decreases with the addition of further states.

### 2014-04-15 paper/manuscript/manuscript.tex
added manuscript dir and also fixed formatting things

--- 2014-04-08 paper/paper2.tex
@@
 
+\setkeys{acs}{usetitle=true}
+
 \title{Statistical Model Selection for Markov Models of Biomolecular Dynamics}
-\author{Christian R. Schwantes}
+\author{Robert T. McGibbon}
 \affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
 \altaffiliation{Contributed equally to this work}
-\author{Robert T. McGibbon}
+\author{Christian R. Schwantes}
 \affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
@@
 \email{pande@stanford.edu}
-\keywords{MSM, molecular dynamics, AIC, BIC, Bayes factor}
 
@@
 By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markov state model construction.
+
+\vspace{1em}
+\noindent KEYWORDS: MSM, molecular dynamics, AIC, BIC, Bayes factor, model selection
 \end{abstract}
@@
 \centering
-\includegraphics[width=3in]{figs_final/mull_ww.png}
+\includegraphics[width=3in]{figs/mull_ww.png}
 \caption{Systems studied in this work.
@@
 
-Two systems were investigated using this likelihood scheme. The first was a simple two dimensional surface with three energy minima, called the M\"{u}ller potential. The dynamics are governed by
+Two systems were investigated using this likelihood scheme (\cref{fig:pics}).
+The first was a simple two dimensional surface with three energy minima, called the M\"{u}ller potential.
+The dynamics are governed by
 $$
@@
 We compared four different model selection criteria: (1) test log-likelihood (2) BIC (3) AIC and (4) implied timescales convergence.
-The analysis \cref{fig:mullerlike} indicates that the log-likelihood on the training data set continually increases with the addition of further states.
+The analysis (\cref{fig:mullerlike}) indicates that the log-likelihood on the training data set continually increases with the addition of further states.
 The increase is most dramatic at low $k$, and levels off at high $k$, indicating that once $k$ is large enough, the marginal increase in the likelihood with respect to further states is small.
@@
 \centering
-\includegraphics[width=6in]{figs/mull/mull.pdf}
+\includegraphics[width=6in]{figs/mull.pdf}
 \caption{Four model selection criteria were used (left panel) to select the state discretization for MSMs built on simulations of the M\"{u}ller potential.
@@
 \label{fig:ww-tica}
-\includegraphics[width=6in]{figs/ww_tica/ww_tica.pdf}
+\includegraphics[width=6in]{figs/ww_tica.pdf}
 
@@
 \centering
-\includegraphics[width=6in]{figs/ww_pca/ww_pca.pdf}
+\includegraphics[width=6in]{figs/ww_pca.pdf}
 \caption{The four selection criteria were used to compare MSMs for the Fip35WW domain built with PCA.

### 2014-04-15 paper/manuscript/manuscript.tex
manuscript round 2

@@
 Model selection is performed by a minimization of the criterion.
-The BIC is derived based on the dominant terms in the Laplace approximation to the logarithm of the Bayes factor with a vague prior, while the functional form of the AIC comes from an asymptotic approximation to the Kullback-Leibler divergence between the model and the true distribution\cite{kass1995bayes, liddle2004many}.
+The BIC is derived based on the dominant terms in the Laplace approximation to the logarithm of the Bayes factor with a vague prior, while the functional form of the AIC comes from an asymptotic approximation to the Kullback-Leibler divergence between the model and the true distribution\cite{kass1995bayes,Liddle2007Information}.
 The appearance of number of data points $N$ in \cref{eq:bic} assumes that the data is independent and identically distributed given the model, which is often poorly justified for time series.
@@
 \centering
-\label{fig:ww-tica}
 \includegraphics[width=6in]{figs/ww_tica.pdf}
 
-\caption{MSMs for the Fip35WW domain built with tICA were evaluated based on four model selection criteria.
+\caption{\label{fig:ww-tica} MSMs for the Fip35WW domain built with tICA were evaluated based on four model selection criteria.
 The log-likelihood was evaluated on the training set (black) and a separate test set (blue), and compared with the AIC (purple) and BIC (red) penalized log-likelihoods, computed on the training set.
@@
 \bibliography{bibliography}
+
+\begin{figure}
+\includegraphics{figs/toc.png}
+\caption{TOC figure}
+\end{figure}
 
